### 2018-03-21 abstract.tex
Initial commit

The construction of phylogenies helps us answer evolutionary biological
questions.
Our current phylogenetic tools ignore the fact that speciation takes time,
which has an effect unknown in phylogeny reconstruction.
Here, we simulate true incipient species trees and their corresponding
DNA alignments, from which we measure the reconstruction of the phylogeny
using a standard birth-death model.
We measure the errors that the Bayesian phylogenetic software tool BEAST2
gives when recovering simulated phylogenies, for different times-to-speciate,
under a range of additional parameter settings.
It has been found that branch lengths are consistently and strongly 
underestimated for biologically relevant parameters.
This research shows that protractedness is a complexity of nature that
cannot always be ignored and should be incorporated in our phylogenetic
tools.

### 2018-03-21 introduction.tex
Initial commit

\section{Introduction}

% Biology: speciation takes time

It takes time for new species-to-be to achieve the status of a new species.
First, it takes time for the new species-to-be has to acquire sufficient different mutations to 
establish reproductive isolation \cite{schluter2001ecology}. Second, 
it takes time for us to recognize reproductive isolation
and reclassify that one single species has become two (or more) species. As an
example, \cite{fennessy2016multi} showed that Africa has four instead of only one
giraffe species, with a common ancestor estimated at around 2 million years ago.

% The use of speciation models

From DNA sequences obtained in the present, we can construct phylogenies
to infer the evolutionary relationship between species and the time 
from which ancestral species developed into separate.

As we only sample DNA of extant species, 
we can only infer a reconstructed tree, in which
the lineages that have gone extinct are absent.
Although most models assume extinctions (\cite{yule1925mathematical} being the
classic exception, as at its time computation was expensive), 
information on extinction times is lost due to our sampling.

The time when reproductive isolation started to take off, is
also lost in our sampling method, if the new species-to-be is not 
recognized yet as such. Without fossil date, one cannot observe the
possibly many times that reproductive isolation built up, but was lost again.

% First model: constant rate birth death model

Although we know that speciation takes time, we commonly ignore this when
constructing a phylogeny, by choosing a constant-rate birth-death model
as a speciation model. This can be justified, as for lineages 
that have speciated far back in the past, there is no error. 
Yet in the present, we do not recognize the future species-to-be
that are already genetically converging, resulting in an underestimation
of the number of species present today, which can only be 
concluded in retrospect.

The constant-rate birth-death model (as described in for example \cite{nee1994reconstructed}) 
is among the simplest speciation models, and assumes a constant speciation
 rate $\lambda$ and constant extinction rate $\mu$.

Additionally, it assumes speciation is instantaneous.
The constant-rate birth-death model is popular for its simplicity, yet
has also served as a starting point for more elaborate speciation models.

% Other non-protracted speciation models

Other speciation models may assume that speciation rate changes in 
time \cite{rabosky2008explosive}, is dependent on the amount of species 
present \cite{etienne2011diversity}, or is trait dependent \cite{fitzjohn2009estimating}.

In these examples, the original birth-death model has been extended by making 
speciation rate dependent on time, diversity or trait value respectively.
The assumption that all these extensions share is that speciation is instantaneous;
that after a new branching event, both two lineages are (or are recognized as being) different
species.

% Protracted speciation model

% Good and incipient states

The protracted speciation model \cite{etienne2012prolonging} allows for speciation taking time.

It adds an additional species state (see also figure \ref{fig:pbd_states}),
coined the 'incipient' stage, which a lineages has to complete before becoming
a 'good' species.

One view is to say that good species have achieved reproductive isolation,
where incipient species are in the process of achieving this \todo{Add reference}.

As genetic drift is unlikely to be a relevant speciation mechanism,
this is more likely to be ecological speciation [Sobel et al., 2009]. 
\todo{RSE: Ben je het daarmee eens? Zou drift niet een belangrijke ooraak van speciation op islands kunnen zijn bijvoorbeeld?}

Alternatively, an incipient species can be described as a good-species-to-be,
yet not recognized as such \todo{?[Purvis et al., 2009]?}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}
  \centering
  \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=4cm, semithick]   
  \tikzstyle{every state}=[]
  \node[state] (A)              {Good};   
  \node[state] (B) [right of=A] {Incipient};   
  \node[state] (C) [below of=B] {Extinct};   
  \path (A) edge [bend right] node {$b_g$} (B)
        (A) edge [bend right] node {$\mu_g$} (C)
        (B) edge [loop above] node {$b_i$} (B)
        (B) edge [bend right] node {$\lambda$} (A)
        (B) edge [bend right] node {$\mu_i$} (C); 
  \end{tikzpicture}

  \caption{
    The states and transitions of a species in the PBD model.
    $b_i$: speciation-initiation rate of incipient species. 
    $b_g$: speciation-initiation rate of good species. 
    $\lambda$: speciation completion rate. 
    $\mu_i$: extinction rate of incipient species. 
    $\mu_g$: extinction rate of good species. 
    Figure after Etienne et al, 2014, Evolution
  }
  \label{fig:pbd_states}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% Biological mechanism of reproductive isolation

The protracted speciation model makes no assumption about the exact biological
mechanism by which reproductive isolation is 
acquired \cite{etienne2012prolonging, etienne2014estimating,rosindell2010protracted}.

A 'good' state can be viewed as a state in which there is convergent selection
for that lineage.

The 'incipient' state can be seen as a state in which there is the potential
for speciation.

A first example can be from the BDM \todo{Add reference} model, the state in which there
are all three genotypes, before the intermediate genotype is lost.
A second example can be in a sexual selection model [van Doorn and Weissing, 2002], 
where there is the onset for convergent selection.
In both cases, these intermediate states may fall back to the original
state of one species, or complete to produce a new species.

% Effects on topology

A feature of the protracted birth-death model, is that it may result in
paraphylies when the rates between good and incipient species differ \todo{Add reference}.
Paraphylies may be more than an experimental artifact and are claimed to
be an inherent feature of nature \cite{funk2003species}.

\todo{Add something like: this research solves this by sampling from paraphyletic trees (at the subspecies level) to create non-paraphyletic trees at the species level}

% Good and incipient rates

Where the classic birth death model assumes constant speciation and extinction
 rates, the protracted birth-death model allows these rates to be state
 dependent.

The only additional parameter is the rate at which incipient species become
(recognized as) good species.
When setting that parameter to infinity, incipient species become good
species instantaneous and the model falls back to a constant-rate (non-protracted) birth-death model \todo{Add reference}.

% Relationship to constant-rate birth-death model

\todo{?Wat bedoelde ik met deze onzin?}
Although both standard and protracted birth-death models use a parameter 
$\lambda$, the $\lambda$ in the standard birth death model denotes
a speciation rates (initiation and completion), where in the PBD model
$\lambda$ denotes only the speciation initiation rate. 
There is no relationship known between the parameters of the constant-rate
birth-death model and the protracted birth-death model \cite{etienne2014estimating} 
\todo{refesh memory, then update paragraph}.

% Good and incipient rates do not need to be constant

The assumption that speciation takes time is independent of the dynamics
of the speciation and extinction rates.
The protracted speciation model in this research assumes these rates are
constant, but these rates can easily be made to depend on time, diversity
or trait.

% Use of a speciation model in inferring a phylogeny

Speciation models are widely used to make inferences from genetic data.
There are multiple computer programs to create phylogenies and/or parameter
estimates from aligned DNA sequences.
BEAST2 is a widely used tool that allows for a Bayesian approach to 
phylogenetics \cite{bouckaert2014beast}.

BEAST2 supplies the user with multiple speciation models, that all assume
instantaneous speciation.

% This study

This simulation investigates the consequence of BEAST2 using instantaneous
speciation, by simulating a 'true' tree that is protracted, simulate DNA
alignments following that tree, and seeing how well BEAST2 can recover
the original phylogeny.

% Novelty

This study is the first to measure the error made when acknowledging that
speciation takes time, yet using one of the many tools that ignores this
fact.

Part of the parameter space will create 'true' trees that are not protracted.
This will result in a test of BEAST2 against itself, which is, to the best
of our knowledge, not been done, probably due to the extensive computations
that are needed \todo{RSE: Heb je dit uitgebreid getest? Misschien zou je Remco kunnen vragen}.

% Analysis

The 'true' and inferred trees are compared with the nLTT statistic, which
is a novel and efficient summary statistic, that has been proven to have
better performance than the gamma statistics in Approximate
Bayesian Computation \cite{janzen2015approximate} (yet on par with the likelihood).

The nLTT statistics is the summed absolute difference between two 
Lineages-Through-Time plot lines that are normalized to 
let both time and number of lineages
range from zero to one.
Additionally, this statistic allows to pinpoint the time span that contributed
most to its value.

% Expectations

It is expected that for higher protractedness (thus deviation from BEAST2
its assumptions), the error will increase.
Trees are expected to be inferred better, when there is more information
available to construct them, thus we expect less error when there are more
taxa and when the simulation DNA alignments are longer.
The error is expected to be biggest close to the present, as there will
be the most incipient species present.

It is unknown, however, under which biological settings this error is relevant.

% Preview results

This study shows that ... \todo{put result here}.

### 2018-03-21 methods.tex
Initial commit

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Methods}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

This research measures the error our current BD tools make, when
the real process has been protracted. Figure \ref{fig:experiment} shows
the following overview of the experiment graphically. 
An in silico data set is created for a wide
range of parameters from which incipient species trees are simulated (A in figure \ref{fig:experiment}). 
These phylogenies serve as 'the truth' or 'nature'. From that incipient species tree,
species trees are created, by picking one representive per species.
This is done twice, as either the youngest (B1) or oldest (B2) subspecies is chosen
to represent a species. Per species tree, $n_a$ alignments are simulated (C1 and C2).
Per alignment, $n_b$ posteriors are inferred, 
each consisting of $n_s$ phylogenies and parameter estimates (D1 and D2). Each
species tree is then compared with the species tree its alignment is based on (for
example, B1 and D1).

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}
  \centering 
  \resizebox {0.8\textwidth} {!} {
    \CreateTikzFigureExperiment{} 
  }
  \caption{
    Overview of the experiment: 
    A: per parameter set, there are $n_i$ incipient species trees created. 
    B1: constructing a species trees by using the youngest subspecies $A1$ to represent good species $A$.
    B2: constructing a species trees by using the oldest subspecies $A2$ to represent good species $A$.
    C1: constructing one of the $n_a$ alignments of B1. 
    C2: constructing one of the $n_a$ alignments of B2. 
    D1: reconstructing a species trees from alignment B1.
    D2: reconstructing a species trees from alignment B2.
    Comparisons are done between a species tree and its reconstructed species tree, 
    for example, between B1 and D1.
    All phylogenies are drawn at the same scale.
  }
  \label{fig:experiment}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Parameter space}
\label{subsec:parameter_space}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{table}
  \centering 
  \begin{tabular}{l l}
    \hline
    Symbol & Description \\
    \hline
    \hline
    $b_g$ & Speciation initiation rate of a good species (probability per lineage, per million year) \\
    $b_i$ & Speciation initiation rate of an incipient species (probability per lineage, per million year) \\
    $\lambda$ & Speciation completion rate (probability per lineage, per million year) \\
    $\mu_g$ & Extinction rate of a good species (probability per lineage, per million year) \\
    $\mu_i$ & Extinction rate of an incipient species (probability per lineage, per million year) \\
    \hline
    $t_c$ & Crown age (million years) \\
    $r$ & Mutation rate (probability of mutation per nucleotide per million years) \\
    $l_a$ & DNA alignment length (base pairs) \\
    \hline
    $n_i$ & Number of incipient species trees per parameter set \\
    $n_a$ & Number of alignments per species tree \\
    $n_b$ & Number of BEAST2 runs per alignment \\
    \hline
    $n_s$ & Number of MCMC samples \\
    $i_s$ & MCMC sampling interval \\
    \hline
    $n_{ti}$ & Number of taxa in an incipient species tree \\
    $n_{t}$ & Number of taxa in a (good) species tree \\
    $t_{\mean{ds}}$ & Average duration of speciation \\
    $t_{sc}$ & Speciation completion time \\
    $\Delta_{nLTT}$ & nLTT statistic value \\
    $\mean{\Delta_{nLTT}}$ & Average nLTT statistic value of a posterior \\
    $\median{\Delta_{nLTT}}$ & Median nLTT statistic value of a posterior \\
    \hline
  \end{tabular}
  \caption{
    Parameter descriptions. The sections are (1) the biological parameters, (2) the
    biological tuning parameters, (3) parameters involved in replication, (4) MCMC
    tuning parameters, and (5) other symbols.
  }
  \label{table:parameter_descriptions}
\end{table}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\todo{@RSE: would you prefer $t_{\mean{ds}}$ or $\mean{t_{ds}}$ for average duration of speciation?}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{table}
  \centering 
  \begin{tabular}{l l}
    \hline
    Parameter             & Values \\
    \hline
    \hline
    $b = b_g = b_i$       & 0.1, 0.5, 1.0 \\
    $\lambda$             & 0.1, 0.3, 1.0, $10^6$ \\
    $\mu = \mu_g = \mu_i$ & 0.0, 0.1, 0.2, 0.4 \\
    \hline
    $t_c$                 & 15 \\
    $r$                   & 0.01 \\
    $l_a$                 & $10^3$, $10^4$ \\
    \hline
    $n_i$                 & 20 \\
    $n_a$                 & 2 \\
    $n_b$                 & 2 \\
    \hline
    $n_s$                 & $10^3$ \\
    $i_s$                 & $10^3$ \\
    \hline
  \end{tabular}
  \caption{
    Parameter values. The sections are (1) the biological parameters, (2) the
    biological tuning parameters, (3) parameters involved in replication, 
    and (4) MCMC tuning parameters.
  }
  \label{table:parameter_values}
\end{table}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% Biological parameters

This study primarily investigates combinations of the five biological 
parameters ($b_g$, $b_g$, $\lambda$, $\mu_g$ and $\mu_g$, 
see also figure \ref{fig:pbd_states} and table \ref{table:parameter_descriptions}), 
in a close to full factorial fashion, with two exceptions.
First, following \cite{etienne2014estimating}, birth and extinction rates were each 
set equal between good and incipient species ($b_g = b_i$, and $\mu_g = \mu_i$).
Second, only parameter combinations with speciation initiation rates greater
than extinction rates were used ($b_g > \mu_g$). The values 
used in the experiment (see table \ref{table:parameter_values}) 
were selected to have a reasonable expected number of incipient 
lineages, $E(n_{ti})$, where \todo{?too basic to have a reference to? I assume Kendall, 1948, \cite{kendall1948generalized}, but could not look it up here and now}:

\begin{equation}
  E(n_{ti}) = e^{ (b - \mu) \cdot t_c }
\end{equation}

One value of speciation completion rate $\lambda$ is set (close to) infinity, as
then the protracted speciation model falls back to a constant-rate 
(non-protracted) birth-death model [?Etienne et al., 2012 or already Rosindell and Etienne 2011?].

% Biological tuning parameters

Secondarily, there are three biological tuning parameters. The crown age of
the incipient species tree, $t_c$, is set to 15 million years for all simulations, 
following \cite{etienne2014estimating}. The per-nucleotide mutation rate, $r$, and the
DNA aligment length, $l_a$, together must be tuned to result in informative
DNA alignments (see section \ref{subsec:simulating_alignments}). $r$ is identical in all
simulations. Increasing $l_a$ results in more informative alignments, and this
effect is measured by using two values of $l_a$ 
per biological parameter combination.

% Replication parameters

All processes involving stochasticity have replicates. 
For each set of parameters, $n_i$ incipient species trees are simulated. 
From each species tree, $n_a$ alignments are simulated. From each alignment, 
$n_b$ posteriors are generated.

% MCMC tuning parameters

The posteriors are generated using BEAST2, which uses a Monte-Carlo Markov-Chain (MCMC)
algorithm. This algorithm is tuned to result in a high effective sample size (ESS) by
two constants. The first constant is the MCMC sampling interval, which is $i_s$ states.
$i_s$ which must be high enough to avoid correlation between subsequently sampled posterior states.  
The second constant is the number of MCMC state samples, $n_s$, which equals the ESS iff \todo{@RSE: Can I use iff?}
$i_s$ is high enough to have no correlation between subsequently sampled posterior states.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Simulating incipient species trees}
\label{subsec:simulating_incipient_species_trees}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

From each parameter set, $n_i$ incipient species trees with a crown age
of $t_c$ and at least two extant taxa, 
are created in R \cite{R} using the PBD package \cite{PBD}.
The values of the biological parameters encourage that the number of tips in
each phylogeny is kept within computational limits. However, due to
the stochastic nature of the Doob-Gillespie algorithm used in the creation
of these phylogenies, phylogenies can, by chance, become very big. 
There is no maximum number of taxa set, yet imposed instead by the R 
programming language itself, due to a limit on the amount of memory that can be 
allocated. The incipient species trees that could not be simulated are
counted and discarded from further analysis.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Sampling species trees}
\label{subsec:sampling_species_trees}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Per incipient species tree, two species trees are created. 
To create a species tree (a tree in which each species is represented once)
for an incipient species tree, one representative for each species is picked. 
If a species has initiated one or more incipient species, either the youngest
or the oldest representative is picked. See figure \ref{fig:sampling} for an example.

An alternative sampling method would be to pick 
random representatives of each species. This would decrease the variability
between the two species tree. Maximizing variability between the two
species trees allows for clearer measuments on the effect of sampling.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}
  \centering 
  \CreateTikzFigureSampling{}
  \caption{
    Sampling an incipient species tree. 
    A) At stem age $t_0$, $A1$ is present.
    At crown age $t_1$, its derived incipient species $A2$ is born. 
    At $t_2$, $A2$ itself speciates into a new (unnamed) species, 
    that becomes good species $B$ at $t_3$. At present time, $t_4$, there
    exist two possible representatives of species $A$. 
    B) Using the oldest representative, $A1$, to represent species $A$, the crown
    age of the resulting species tree lies at $t_1$. 
    C) Using the youngest representative, $A2$, to represent species $A$, the crown
    age of the resulting species tree lies at $t_2$. 
  }
  \label{fig:sampling}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Simulating alignments}
\label{subsec:simulating_alignments}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

From each species tree, $n_a$ DNA alignments were simulated that match the
phylogeny's structure, using the phangorn R package \cite{phangorn}. 
Each alignment contains one DNA sequence per species 
of $l_a$ nucleotides. The alignment follows the Jukes-Cantor nucleotide 
subsititution model \cite{cantor1969mammalian}, 
with per-nucleotide rate of mutation, $r$.

The Jukes-Cantor
nucleotide substitution model is regarded as an oversimplification of 
nature \cite{unknown}. In this in silico experiment, however, we simulate first
a DNA alignment from a phylogeny using nucleotide substitution model $M_N$, 
then infer a phylogeny from that DNA alignment, assuming nucleotide substitution model $M_N$.
Too reduce unnecessary complexity, $M_N$ is chosen to be the simplest possible.

The combination of $l_a$ and $r$ is tuned to result in informative alignments. 
Within a maximally informative alignment, each species has its own unique DNA sequence.
Additionally, the Jukes-Cantor distance \cite{unknown} between any 
two sequences must be calculable, which is when the proportion of matching nucleotides is below 75\%.
The tuning was done using an $l_a$ of $10^3$ nucleotides, 
as this appears to be the current median of the sequenced genome lengths, as currently published.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Simulation of posteriors}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

From each DNA alignment, $n_b$ posteriors of phylogenies and parameter combinations 
is constructed, using BEAST2, a Bayesian phylogenetic tool \cite{bouckaert2014beast}.
The nucleotide substitution model is set to 
be Jukes-Cantor (one nucleotide substition rate category, no invariant nucleotides) 
and a strict molecular clock, which matches the procedure the DNA alignments are generated, 
as described in section \ref{subsec:simulating_alignments}).
The tree prior is set to the (non-protracted) birth-death model, 
with an uniform prior for birth-rate of range $[0, 10^5]$ and 
a uniform death-per-birth rate prior of range $[0, 1]$. 
The MCMC creation parameters were set to have $n_s$ states, sampled at an interval of $i_s$ states.

An MCMC must have an effective sample size (ESS) of at least 200 to result in a 
posterior that is 'adequate' \cite{drummond2015bayesian}. Due to the multiple 
sources of stochasticity in the preceding steps, this ESS may not always be achieved.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Analysis}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

To quantify the error, each species tree is compared with the posterior derived
from its $n_a$ alignments (for example, B1 and D1 in figure \ref{fig:experiment}).
This is by calculating the nLTT statistic \cite{janzen2015approximate}
between the species tree and each of its posterior's trees (using 
the nLTT package \cite{nLTT}), resulting in an nLTT statistic distribution.

% BD trees

The nLTT statistic distribution for all simulations with a high speciation 
completion rate is visualized for the different parameter values. This allows
to see the baseline error. Instead of using a complete
nLTT statistic distribution, also the mean value of each posterior is 
sometimes used to reduce variance in this.

The effect of speciation completion time $\lambda$ is visualized, by showing
nLTT statistic distributions for the different parameter values, for all values
of $\lambda$.

Instead of using the parameter value of $\lambda$ as a proxy for protractedness,
the speciation completion time, $t_{sc}$, and average duration of speciation, $t_{\mean{ds}}$, 
are used. Speciation completion time, $t_{sc}$, is simply the inverse of the
speciation completion rate $\lambda$:

\begin{equation}
  t_{sc} = \frac{1}{\lambda}
  \label{eq:speciation_completion_time}
\end{equation}

The speciation completion time differs from the observed average speciation time, 
as incipient species may go extinct and may produce other incipient species 
that complete speciation before the original incipient one. The average duration 
of speciation, $t_{\mean{ds}}$, takes this into 
account \todo{check formula ?19 of?20 in Etienne et al., 2012, then cite}:

\begin{equation}
    t_{\mean{ds}} = \mean{\rho} = \frac{2}{ (D - \lambda + b_i - \mu_i) \cdot \ln{ \frac{2}{(1 + (\lambda - b_i + \mu_i)/D)}  } }
\end{equation}

where $D$ is defined as:

\begin{equation}
    D = \sqrt{ (\lambda + b_i)^2 + 2 \cdot (\lambda - b_i) \cdot \mu_i + \mu_i^2 }
\end{equation}

For each simulation, the expected $t_{\mean{ds}}$ is calculated from its parameter 
values (using the PBD package \cite{PBD}), after which the nLTT statistic 
distributed is plotted, to see the effect of protractedness on the nLTT statistic
distribution.

As not all simulations result in posteriors with an adequate ESS, the
effect that a below-median or above-median ESS has on the nLTT statistic
distribution is displayed.

To view the effect of alignment length, $l_a$, the nLTT statistic
distributions are plotted side-by-side for all other parameter values.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Peripherals}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Simulating incipient species trees, sampling species trees, simulating alignments
and creating posteriors, are done with the wiritttes package \cite{wiritttes}. The
analysis is done with the wiritttea package \cite{wiritttea}. Those packages rely on 
functionality in the APE \cite{APE}, geiger \cite{GEIGER}, ggplot2 \cite{ggplot2}, 
RBeast \cite{RBeast}. Quality of these packages is assured by 
the covr \cite{covr}, devtools \cite{devtools}, 
goodpractice \cite{goodpractice}, lintr \cite{lintr}, testit \cite{testit} 
and testthat \cite{testthat} packages, using the Travis CI \cite{travis} continuous integration service. 
Both packages are hosted at GitHub \cite{github}, a good practice for computational
scientists \cite{perez2016ten} and helps improve transparency \cite{gorgolewski2016practical}.
Code coverage, which correlates with code quality \cite{del1995correlation}, 
is above 90\% for both packages.

The simulated dataset can be downloaded from DataDyrad [but not yet]. The
summarized dataset can be downloaded at \cite{wirittte_data}. The
pre-publication of this article can be downloaded at \cite{wirittte_article}.

Simulations and analyis was mostly performed on the Peregrine 
high performance computing cluster of the University of Groningen.

### 2018-03-21 misc.tex
Initial commit

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Misc}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\todo{remove this section}

\subsection{The $\Delta r$ statistic from \cite{pigot2010shape}}

The $\Delta r$ statistic from \cite{pigot2010shape}:

\begin{equation}
  \Delta r=\frac{ln\left(\frac{L(T)}{L(\frac{1}{2}T)}\right)-ln\left(\frac{L(\frac{1}{2}T)}{L(0)}\right)}{ln\left(\frac{L(T)}{L(\frac{1}{2}T)}\right)+ln\left(\frac{L(\frac{1}{2}T)}{L(0)}\right)}
  \label{eq:delta_r}
\end{equation}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Gamma statistic}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Worse than nLTT \cite{janzen2015approximate}.

Worse than $\Delta r$, as gamma statistic depends on the size of the tree \cite{etienne2012prolonging}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{nLTT}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

By \cite{janzen2015approximate}, outperforms gamma statistic and phylogenetic diversity.

In the R package 'nLTT'

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Normalized rooted branch score}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\todo{Better distinguish NRBS and NRTS}

See \cite{heled2010bayesian}.

It is to estimate the distance between two trees:

First obtain the (non-normalized) rooted branch score:

\begin{equation}
  \left|T_{1},T_{2}\right|=\sqrt{\sum_{c\in T_{1}\cup T_{2}}\left(B\left(T_{1},c\right)-B\left(T_{2},c\right)\right)^{2}}
  \label{eq:root_branch_score}
\end{equation}

Where:

$T_{x}$ denotes tree $x$, 
$\left|T_{1},T_{2}\right|$ denotes the (absolute value of the) rooted branch score between tree T1 and T2
and $B(T,c)$ is 'the length of the branch connecting clade c to the tree if it is present in T, 0 otherwise'

Normalization:

\begin{equation}
  NRBS = \frac{1}{\left|T\right|}\frac{\sum_{i=0}^{N}\left|T,T_{i}\right|}{N}
  \label{eq:normalized_root_branch_score}
\end{equation}

Where:

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Normalized root tree score}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

See \cite{heled2010bayesian}.

It is to estimate the distance between two trees:

First obtain the (non-normalized) rooted branch score:

\begin{equation}
  \left|T_{1},T_{2}\right|=\sqrt{\sum_{c\in T_{1}\cup T_{2}}\left(D\left(T_{1},c\right)-D\left(T_{2},c\right)\right)^{2}}
  \label{eq:root_branch_score_huh_again}
\end{equation}

Where $T_{x}$: tree x

$\left|T_{1},T_{2}\right|$: (absolute value of the) rooted branch score between tree T1 and T2

$D(T,c)=\intop_{b_{0}}^{b_{1}}\frac{1}{N_{b}}dt$: b is the branch connecting clade c to the tree if present in T, 0 otherwise.

To cite \cite{heled2010bayesian}: 'This is not a true metric, as branch length and population size are confounded'

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Phylogenetic diversity}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Worse than nLTT \cite{janzen2015approximate}.

### 2018-03-21 results.tex
Initial commit

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Results}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Creation of data set}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\todo{First focus on the real results, not their quality}

% Simulation of incipient species trees

\todo{redo with new data}

\todo{create figure or table that shows this}

957 out of 960 incipient species trees have been simulated successfully. 
The 3 failed simulations created data files bigger than R can read, 
and have been excluded from further analysis.

The number of good species ranged from 2 to 14296, with a median of 87 and an average of 503.8. 
There were 6 phylogenies with only 2 good species.

% Sampling species trees}

\todo{redo with new data}

For each incipient species trees, two species trees were constructed. 
In 653 out of 960 cases, these trees had different branch lengths.
In 302 cases, the trees were identical. In the remaining 5 cases,
there were computational problems (see figure \ref{fig:exit_statuses}).

% Simulation of alignments

\todo{redo with new data}

For each species tree, $n_a$ alignments have been created, 
resulting in $960 \cdot 2 \cdot 2 = 3840$ alignments. 
$?$ of these were informative.

\todo{May result in an overrepresentation of the two same species trees derived from one incipient species tree}

% Simulation of posteriors}

A BEAST2 posterior contains multiple parameter estimates (see figure \ref{fig:ess_distribution} for all),
and each one has an ESS. An ESS of 200 is acceptable for any parameter estimate.
The ESS of the tree likelihood should have such a decent ESS. Figure \ref{fig:ess_distribution_likelihood}
shows the distribution of the likelihood ESS.

\todo{How tp quote https://www.beast2.org/what-is-ess/ ? It states that not all variables
need to have a high ESS, but 'The likelihoods (both of the tree and coalescent model) should have decent ESSs.}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\begin{figure}[!htbp]
%  \includesvg[width=\textwidth]{figure_ess_distribution_likelihood}
%  \caption{
%    Likelihood ESSes. 
%    The dotted line indicates an 'acceptable' ESS
%  }
%  \label{fig:ess_distribution_likelihood}
%\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Quantification of the error BEAST2 makes on all trees}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

To get a first global impression of the error BEAST2 makes, figure \ref{fig:error}
shows a histogram of the nLTT statistics, colored by the expected mean
duration of speciation. As BEAST2 assumes a mean duration of speciation of zero,
it is expected that the nLTT statistic (a measure of error) is lowest. Zooming
in on the left part of the distribution, there is no clean pattern.
The right part of the tail of distribution shows that the biggest nLTT
statistic values are found for high (but not highest) expected mean durations of
speciation.

\todo{figure \ref{fig:error_for_scr} shows the distribution of errors for all
combinations of PBD parameters}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\begin{figure}[!htbp]
%  %\def\svgwidth{0.8\textwidth}
%  %\input{figure_error_20171009.pdf_tex}
%  \includesvg[]{figure_error}
%  \caption{
%    Histograms of the nLTT statistic distribution for all simulations
%    The color indicates the expected mean duration of speciation from
%    red (zero) to purple (highest). Similar colors indicate similar
%    values of expected mean duration of speciation.
%  }
%  \label{fig:error}
%\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

As the expected correlation appears to be absent, figure 
\ref{fig:error_expected_mean_dur_spec}, subfigure $a$, 
shows the correlation between expected
mean duration of speciation and the nLTT statistic. A linear fit shows
that there is indeed an increase in error for an increases expected duration
of speciation. A trendline using locally weighted smoothing, however,
indicates a more complex relationship. Note that this figure only uses
a subset of all measurements, consisting of $n_s$ nLTT statistic values 
per posterior.

Taking the average nLTT value per posterior frees us from computational
constraints and allows us to use the full data set. The resulting
plot, figure \ref{fig:error_expected_mean_dur_spec}, subfigure $b$, shows the same pattern:
A linear fit shows
that there is indeed an increase in error for an increases expected duration
of speciation. But now, a trendline using locally weighted smoothing 
follows this linear relationship closely.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[!htbp]
  %\subfloat[]{
  %  \includesvg[width=0.33\textwidth]{figure_error_expected_mean_dur_spec}
  %}
  %\subfloat[]{
  %  \includesvg[width=0.33\textwidth]{figure_error_expected_mean_dur_spec_mean}
  %}
  %\subfloat[]{
  %  \includesvg[width=0.33\textwidth]{figure_error_expected_mean_dur_spec_mean}
  %}
  \caption{
    nLTT statistic distribution
    for different expected mean duration of speciation.
    Subfigure $a$ takes a sample of all $\Delta_{nLTT}$ values, where
    subfigure $b$ uses each posterior's $\mean{\Delta_{nLTT}}$.
    subfigure $c$ uses each posterior's $\median{\Delta_{nLTT}}$.
    The straight blue line shows a linear
    fit, the red line a fit using locally weighted smoothing.
    The dashed horizontal line indicates
    the mean nLTT statistic for the Birth-Death simulations ($SCR = 10^6$).
    Note that this figure uses only a subset of all individually measured
    nLTT statistics
  }
  \label{fig:error_expected_mean_dur_spec}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

The inference error made is expected to be less for a longer DNA
alignment length. 
Figure \ref{fig:error_expected_mean_dur_spec_alignment_length} 
show the correlation between expected mean duration of speciation 
and the nLTT statistic for the two different DNA alignment lengths. 
The first subfigure uses a sample of all $n_s$ individuals nLTT statistics,
where the second subfigure takes only the posterior's 
nLTT statistic mean value, for the complete dataset.

Both figures show again that there is an 
increase in error for an increases expected duration
of speciation, when assuming a linear relation, and
a more complex relationship when using locally weighted smoothing. 
The figures also show that, as expected, 
the error is less with more information. Additionally,
a longer DNA alignment length results in a slower accumulation of error
for increased expected mean duration of speciation.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[!htbp]

  %\subfloat[]{
  %  \includesvg[width=0.5\textwidth]{figure_error_expected_mean_dur_spec_alignment_length}
  %}
  %\subfloat[]{
  %  \includesvg[width=0.5\textwidth]{figure_error_expected_mean_dur_spec_mean_alignment_length}
  %}

  \caption{
    Relation between mean nLTT statistic distribution
    and  expected mean duration of speciation, for different DNA
    alignment lengths. Straigh lines show a linear
    fit, the curved lines use a locally weighted smoothing.
    The dashed horizontal lines indicates
    the mean nLTT statistic for the Birth-Death simulations ($SCR = 10^6$)
    for that DNA alignment length.
  }
  \label{fig:error_expected_mean_dur_spec_alignment_length}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

A higher ESS for longer DNA alignments may be an additional
explanation for the observed slower accumulation of error
for increased expected mean duration of speciation.
Figure \ref{fig:ess_expected_mean_dur_spec_alignment_length}
shows that longer DNA alignments have a higher likelihood ESS, 
but this trend is weaker than in figure 
\ref{fig:error_expected_mean_dur_spec_alignment_length} and
the 95\% confidence intervals of the linearily fitted trendlines only
stop overlapping for the highest expected mean duration of speciation.
The locally weighted smoothing trendlines show a different trend,
suggesting that alignment length has no effect on the
likelihood ESS.

\todo{Figure \ref{fig:error_alignment_length} shows the same pattern, seperated
for all combinations of PBD parameters.}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[!htbp]
  %\includesvg[width=\textwidth]{figure_ess_expected_mean_dur_spec_alignment_length}
  \caption{
    Likelihood ESS and expected mean duration of speciation, for different DNA
    alignment lengths. Straigh lines show a linear
    fit, the curved lines use a locally weighted smoothing.
  }
  \label{fig:ess_expected_mean_dur_spec_alignment_length}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

For each incipient species tree, two species trees were created, by taking
either the youngest or oldest subspecies to represent each species.
The effect of this is shown in figure \ref{fig:error_expected_mean_dur_spec_sampling}.
Both subfigures show the
error made for different $t_{\mean{ds}}$, where the first subfigure uses a sample
of all $\Delta_{nLTT}$ values, where the second subfigure uses the $\mean{\Delta_{nLTT}}$
of each posterior. In both subfigures, the linear and locally weighted smoothing
have overlapping 95\% confidence intervals, hinting that there is no strong effect
of the way subspecies are selected to represent their species.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[!htbp]
  %\subfloat[]{
  %  \includesvg[width=0.5\textwidth]{figure_error_expected_mean_dur_spec_sampling}
  %}
  %\subfloat[]{
  %  \includesvg[width=0.5\textwidth]{figure_error_expected_mean_dur_spec_mean_sampling}
  %}
  \caption{
    Likelihood ESS and expected mean duration of speciation, for the two
    different methods to sample species representatives from a subspecies. 
    Straigh lines show a linear
    fit, the curved lines use a locally weighted smoothing.
    See also figure \ref{fig:error_sampling_representative}
  }
  \label{fig:error_expected_mean_dur_spec_sampling}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\note{Figure \ref{fig:error_sampling_representative} shows the same data, ordered
by the different PBD parameters instead}

The effect of having a low or high tree likelihood ESS is shown in 
figure \ref{fig:figure_error_expected_mean_dur_spec_low_high_ess}. We
define a low ESS as being below median, where a high ESS is defined as 
above-median. Plotting (a sample of or the means of) the nLTT statistic for different mean
duration of speciation for a low an high ESS, shows that there is no clear
difference in error made for posteriors with either low or high tree likelihood ESSes.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[!htbp]
  %\subfloat[]{
  %  \includesvg[width=0.5\textwidth]{figure_error_expected_mean_dur_spec_low_high_ess}
  %}
  %\subfloat[]{
  %  \includesvg[width=0.5\textwidth]{figure_error_expected_mean_dur_spec_mean_low_high_ess}
  %}
  \caption{
    Likelihood ESS and expected mean duration of speciation, for the lower and upper half of the tree likelihood ESSes. 
    Straigh lines show a linear
    fit, the curved lines use a locally weighted smoothing.
    See also figure \ref{fig:error_sampling_representative}
  }
  \label{fig:figure_error_expected_mean_dur_spec_low_high_ess}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

It is expected that the bigger phylogeny is simulated, the less the error
will be. Figure \ref{fig:figure_error_tree_size} shows the relation between the
number of taxa and the mean nLTT statistic. Although most simulated
phylogenies have less than 250 taxa, there is a clear positive trend, 
which is the opposite of the
expactation: the bigger the phylogeny, the bigger the error made.
As the nLTT statistic normalized for the number of taxa, some other
mechanism must be involved.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[!htbp]
  %\subfloat[]{
  %  \includesvg[width=0.5\textwidth]{figure_error_tree_size}
  %}
  %\subfloat[]{
  %  \includesvg[width=0.5\textwidth]{figure_error_tree_size_mean}
  %}
  \caption{
    The effect of tree size on nLTT.
    Straigh lines show a linear
    fit, the curved lines use a locally weighted smoothing.
    The linear fit and $R^2$ valaues are shown
  }
  \label{fig:figure_error_tree_size}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

For each species tree, $n_a$ different alignments are simulated. For each of
these alignments, $n_b$ posteriors are simulated, using the same alignment. 
It is assumed that two posteriors from the same alignment
give rise to the same parameter estimates distributions.
distribution. Figure \ref{fig:figure_posterior_distributions_likelihood} 
shows the distribution of Mann-Whitney U test p-values for each pair of 
posteriors working on a same alignment, 
using their tree likelihood estimates.

The figure shows that the tree likelihood posterior distributions 
from a same alignment nearly always
are significantly different. Subfigure (b) and (c) show the distribution of
tree likelihood estimates for a posterior pair with the lowest (b) and 
highest (c) p-value.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[!htbp]
  %\subfloat[]{
  %  \includesvg[width=0.8\textwidth]{figure_posterior_distribution_likelihoods}
  %}
  %\subfloat[]{
  %  \includesvg[width=0.5\textwidth]{figure_posterior_distribution_likelihoods_low}
  %}
  %\subfloat[]{
  %  \includesvg[width=0.5\textwidth]{figure_posterior_distribution_likelihoods_high}
  %}
  \caption{
    (a) Mann-Whitney U test p-values between $n_b$ posteriors working on the same DNA alignment
    for tree likelihood estimates.
    Low p-values indicate the distributions are different.
    High p-values indicate the distributions can be rejected to be different.
    The vertical dotted line shows the p-value of 0.05.
    (b) and (c): distribution of tree likelihoods for the two posteriors that had the lowest
    (b) and highest (c) p-value.
  }
  \label{fig:figure_posterior_distributions_likelihood}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Posteriors based on the same alignment are also (next to estimated tree likelihood) 
assumed to have a similar nLTT statistic distribution. 
Figure \ref{fig:figure_posterior_distribution_nltt} 
shows the distribution of Mann-Whitney U test p-values for each pair of 
posteriors working on a same alignment, 
using their nLTT statistics.

The figure shows that the nLTT statistic distributions 
from a same alignment are commonly different and commonly the same. 
Subfigures $(b)$ and $(c)$ show the distribution of
nLTT statistics for a posterior pair with the lowest $(b)$ and 
highest $(c)$ p-value.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[!htbp]
  %\subfloat[]{
  %  \includesvg[width=0.8\textwidth]{figure_posterior_distribution_nltts}
  %}
  %
  %\subfloat[]{
  %  \includesvg[width=0.5\textwidth]{figure_posterior_distribution_nltts_low}
  %}
  %\subfloat[]{
  %  \includesvg[width=0.5\textwidth]{figure_posterior_distribution_nltts_high}
  %}
  \caption{
    $(a)$ Mann-Whitney U test p-values between $n_b$ posteriors working on the same DNA alignment
    for nLTT statistics.
    Low p-values indicate the distributions are different.
    High p-values indicate the distributions can be rejected to be different.
    The vertical dotted line shows the p-value of 0.05.
    $(b)$ and $(c)$: distribution of nLTT statistics for the two posteriors that had the lowest
    $(b)$ and highest $(c)$ p-value.
  }
  \label{fig:figure_posterior_distribution_nltt}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

The nLTT statistic normalizes a phylogeny to have a crown age of 1.
To prevent overlooking patterns in posterior crown ages, 
figure \ref{fig:figure_posterior_distribution_crown_ages}
shows the distribution of crown ages of all simulations.
Almost all simulations have a crown age close to $0.15$, but there
are rare cases with crown ages above $0.2$, ranging to $83.4$.

\todo{I think it is very suspicious that the range centers on 0.15, when
the crown age is 15. Check this!}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[!htbp]
  %\subfloat[]{
  %  \includesvg[width=0.8\textwidth]{figure_posterior_distribution_crown_ages}
  %}
  %
  %\subfloat[]{
  %  \includesvg[width=0.5\textwidth]{figure_posterior_distribution_crown_ages_low_count}
  %}
  %\subfloat[]{
  %  \includesvg[width=0.5\textwidth]{figure_posterior_distribution_crown_ages_high_count}
  %}
  \caption{
    $(a)$ Distribution of crown ages in all posteriors. 
    The horizontal line indicates the maximum count for subfigure $(b)$. 
    The vertical line shows the maximum crown age for subfigure $(c)$.
    $(b)$ Rare crown ages
    $(c)$ Common crown ages
  }
  \label{fig:figure_posterior_distribution_crown_ages}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Again, it is assumed that posteriors based on the exact same alignment
have a similar crown age distribution. Of all posterior pairs, the Mann-Whitney
U test p-values are shown in figure \ref{fig:figure_posterior_distribution_crown_ages_p_values}.
It shows that crown age distributions being different is common and these
being the same is uncommon.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[!htbp]
  %\subfloat[]{
  %  \includesvg[width=0.8\textwidth]{figure_posterior_distribution_crown_ages_p_values}
  %}
  %
  %\subfloat[]{
  %  \includesvg[width=0.5\textwidth]{figure_posterior_distribution_crown_ages_lowest_p_value}
  %}
  %\subfloat[]{
  %  \includesvg[width=0.5\textwidth]{figure_posterior_distribution_crown_ages_highest_p_value}
  %}
  \caption{
    $(a)$ Mann-Whitney U test p-values comparing the distribution of crown ages
    between each posterior pair.
    $(b)$ Crown age distributions of a posterior pair with the lowest p-value
    $(c)$ Crown age distributions of a posterior pair with the highest p-value
  }
  \label{fig:figure_posterior_distribution_crown_ages_p_values}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

It is expected that if a tree is created by a Birth-Death process, BEAST2 is able
to reasonably estimate those parameters.

%a) `figure_posterior_distribution_bd_bd` histogram of BirthDeath, vertical line at simulation value	
%b) `figure_posterior_distribution_bd_br2` histogram of birthRate2, vertical line at simulation value	
%c) `figure_posterior_distribution_bd_rdr2` histogram of relativeDeathRate2, vertical line at simulation value
%figure_posterior_distribution_bd

% Bad species tree:
% figure_error_posterior_nltt_bad_stree
% figure_error_posterior_nltt_bad_post_tree

### 2018-03-26 article.tex
Merged to two files

\documentclass{article}

% Bibliography
\usepackage{natbib}
\bibpunct{(}{)}{;}{a}{}{;}

% Use 'It was found that A is B (Name 1234)' style
\setcitestyle{authoryear,open={},close={}}

% Affiliations
\usepackage{authblk}
\title{babette: BEAUti 2, BEAST2 and Tracer for R}
\author[1]{Rich\`el J.C. Bilderbeek}
\author[1]{Rampal S. Etienne}
\affil[1]{Groningen Institute for Evolutionary Life Sciences, University of Groningen, Groningen, The Netherlands}

% Load my functions
\input{functions.tex}

% Use double spacing
\usepackage{setspace}
\doublespacing

\usepackage{listings}
\usepackage{hyperref}
\usepackage{todonotes}
\usepackage{verbatim}
\usepackage{pgf}
\usepackage{bm}

% TikZ and friends
\usepackage{tikz}
\usepackage{tkz-graph}
\usepackage{pgf}
\usetikzlibrary{arrows,automata}

% Style of listings
% From http://r.789695.n4.nabble.com/How-to-nicely-display-R-code-with-the-LaTeX-package-listings-tp4648110.html
\usepackage{fancyvrb} 
\definecolor{codegreen}{rgb}{0,0.6,0}
\definecolor{codegray}{rgb}{0.5,0.5,0.5}
\definecolor{codepurple}{rgb}{0.58,0,0.82}
\definecolor{backcolor}{rgb}{0.95,0.95,0.92}
\lstdefinestyle{mystyle}{
  language=R,% set programming language
  basicstyle=\ttfamily\small,% basic font style
  commentstyle=\color{gray},% comment style
  % numbers=left,% display line numbers on the left side
  numberstyle=\scriptsize,% use small line numbers
  numbersep=10pt,% space between line numbers and code
  tabsize=2,% sizes of tabs
  showstringspaces=false,% do not replace spaces in strings by a certain character
  captionpos=b,% positioning of the caption below
  breaklines=true,% automatic line breaking
  escapeinside={(*}{*)},% escaping to LaTeX
  fancyvrb=true,% verbatim code is typset by listings
  extendedchars=false,% prohibit extended chars (chars of codes 128--255)
  alsoletter={.<-},% becomes a letter
  alsoother={$},% becomes other
  otherkeywords={!=, ~, $, \&, \%/\%, \%*\%, \%\%, <-, <<-, /},% other keywords
  deletekeywords={c}% remove keywords 
}
\lstset{style=mystyle}

% Adds numbered lines
\usepackage{lineno}
\linenumbers

% Use TODO notes
\usepackage{todonotes}

% Rename 'Abstract' to 'Summary 
\usepackage[english]{babel}
\addto{\captionsenglish}{\renewcommand{\abstractname}{Summary}}

\begin{document}

\maketitle

\begin{abstract}

  The construction of phylogenies helps us answer evolutionary biological
  questions.
  Our current phylogenetic tools ignore the fact that speciation takes time,
  which has an effect unknown in phylogeny reconstruction.
  Here, we simulate true incipient species trees and their corresponding
  DNA alignments, from which we measure the reconstruction of the phylogeny
  using a standard birth-death model.
  We measure the errors that the Bayesian phylogenetic software tool BEAST2
  gives when recovering simulated phylogenies, for different times-to-speciate,
  under a range of additional parameter settings.
  It has been found that branch lengths are consistently and strongly 
  underestimated for biologically relevant parameters.
  This research shows that protractedness is a complexity of nature that
  cannot always be ignored and should be incorporated in our phylogenetic
  tools.

\end{abstract}

{\bf Keywords:} computational biology, evolution, phylogenetics, BEAST2, R

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Introduction}

% Biology: speciation takes time

It takes time for new species-to-be to achieve the status of a new species.
First, it takes time for the new species-to-be 
to acquire sufficient different mutations to 
establish reproductive isolation \cite{schluter2001ecology}. Second, 
it takes time for us to recognize reproductive isolation
and reclassify that one single species has become two 
(or more) species.

As an example, \cite{fennessy2016multi} 
showed that Africa has four instead of only one
giraffe species, with a common ancestor estimated 
at around 2 million years ago.

% The use of speciation models

From DNA sequences obtained in the present, we can construct phylogenies
to infer the evolutionary relationship between species and the time 
from which ancestral species developed into separate.

As we only sample DNA of extant species, 
we can only infer a reconstructed tree, in which
the lineages that have gone extinct are absent.
Although most models assume extinctions (\cite{yule1925mathematical} being the
classic exception, as at its time computation was expensive), 
information on extinction times is lost due to our sampling.

The time when reproductive isolation started to take off, is
also lost in our sampling method, if the new species-to-be is not 
recognized yet as such. Without fossil date, one cannot observe the
possibly many times that reproductive isolation built up, but was lost again.

% First model: constant rate birth death model

Although we know that speciation takes time, we commonly ignore this when
constructing a phylogeny, by choosing a constant-rate birth-death model
as a speciation model. This can be justified, as for lineages 
that have speciated far back in the past, there is no error. 
Yet in the present, we do not recognize the future species-to-be
that are already genetically converging, resulting in an underestimation
of the number of species present today, which can only be 
concluded in retrospect.

The constant-rate birth-death model (as described in for example \cite{nee1994reconstructed}) 
is among the simplest speciation models, and assumes a constant speciation
 rate $\lambda$ and constant extinction rate $\mu$.

Additionally, it assumes speciation is instantaneous.
The constant-rate birth-death model is popular for its simplicity, yet
has also served as a starting point for more elaborate speciation models.

% Other non-protracted speciation models

Other speciation models may assume that speciation rate changes in 
time \cite{rabosky2008explosive}, is dependent on the amount of species 
present \cite{etienne2011diversity}, or is trait 
dependent \cite{fitzjohn2009estimating}.

In these examples, the original birth-death model has been extended by making 
speciation rate dependent on time, diversity or trait value respectively.
The assumption that all these extensions share is that speciation is instantaneous;
that after a new branching event, both two lineages are (or are recognized as being) different
species.

% Protracted speciation model

% Good and incipient states

The protracted speciation model \cite{etienne2012prolonging} allows for speciation taking time.

It adds an additional species state (see also figure \ref{fig:pbd_states}),
coined the 'incipient' stage, which a lineages has to complete before becoming
a 'good' species.

One view is to say that good species have achieved reproductive isolation,
where incipient species are in the process of achieving this \todo{Add reference}.

As genetic drift is unlikely to be a relevant speciation mechanism,
this is more likely to be ecological speciation [Sobel et al., 2009]. 
\todo{RSE: Ben je het daarmee eens? Zou drift niet een belangrijke ooraak van speciation op islands kunnen zijn bijvoorbeeld?}

Alternatively, an incipient species can be described as a good-species-to-be,
yet not recognized as such \todo{?[Purvis et al., 2009]?}.

% Biological mechanism of reproductive isolation

The protracted speciation model makes no assumption about the exact biological
mechanism by which reproductive isolation is 
acquired \cite{etienne2012prolonging, etienne2014estimating,rosindell2010protracted}.

A 'good' state can be viewed as a state in which there is convergent selection
for that lineage.

The 'incipient' state can be seen as a state in which there is the potential
for speciation.

A first example can be from the BDM \todo{Add reference} model, the state in which there
are all three genotypes, before the intermediate genotype is lost.
A second example can be in a sexual selection model [van Doorn and Weissing, 2002], 
where there is the onset for convergent selection.
In both cases, these intermediate states may fall back to the original
state of one species, or complete to produce a new species.

% Effects on topology

A feature of the protracted birth-death model, is that it may result in
paraphylies when the rates between good and incipient species differ \todo{Add reference}.
Paraphylies may be more than an experimental artifact and are claimed to
be an inherent feature of nature \cite{funk2003species}.

\todo{Add something like: this research solves this by sampling from paraphyletic trees (at the subspecies level) to create non-paraphyletic trees at the species level}

% Good and incipient rates

Where the classic birth death model assumes constant speciation and extinction
 rates, the protracted birth-death model allows these rates to be state
 dependent.

The only additional parameter is the rate at which incipient species become
(recognized as) good species.
When setting that parameter to infinity, incipient species become good
species instantaneous and the model falls back to a constant-rate (non-protracted) 
birth-death model \todo{Add reference}.

% Relationship to constant-rate birth-death model

\todo{?Wat bedoelde ik met deze onzin?}
Although both standard and protracted birth-death models use a parameter 
$\lambda$, the $\lambda$ in the standard birth death model denotes
a speciation rates (initiation and completion), where in the PBD model
$\lambda$ denotes only the speciation initiation rate. 
There is no relationship known between the parameters of the constant-rate
birth-death model and the protracted birth-death model \cite{etienne2014estimating} 
\todo{refesh memory, then update paragraph}.

% Good and incipient rates do not need to be constant

The assumption that speciation takes time is independent of the dynamics
of the speciation and extinction rates.
The protracted speciation model in this research assumes these rates are
constant, but these rates can easily be made to depend on time, diversity
or trait.

% Use of a speciation model in inferring a phylogeny

Speciation models are widely used to make inferences from genetic data.
There are multiple computer programs to create phylogenies and/or parameter
estimates from aligned DNA sequences.
BEAST2 is a widely used tool that allows for a Bayesian approach to 
phylogenetics \cite{bouckaert2014beast}.

BEAST2 supplies the user with multiple speciation models, that all assume
instantaneous speciation.

% This study

This simulation investigates the consequence of BEAST2 using instantaneous
speciation, by simulating a 'true' tree that is protracted, simulate DNA
alignments following that tree, and seeing how well BEAST2 can recover
the original phylogeny.

% Novelty

This study is the first to measure the error made when acknowledging that
speciation takes time, yet using one of the many tools that ignores this
fact.

Part of the parameter space will create 'true' trees that are not protracted.
This will result in a test of BEAST2 against itself, which is, to the best
of our knowledge, not been done, probably due to the extensive computations
that are needed \todo{RSE: Heb je dit uitgebreid getest? Misschien zou je Remco kunnen vragen}.

% Analysis

The 'true' and inferred trees are compared with the nLTT statistic, which
is a novel and efficient summary statistic, that has been proven to have
better performance than the gamma statistics in Approximate
Bayesian Computation \cite{janzen2015approximate} (yet on par with the likelihood).

The nLTT statistics is the summed absolute difference between two 
Lineages-Through-Time plot lines that are normalized to 
let both time and number of lineages
range from zero to one.
Additionally, this statistic allows to pinpoint the time span that contributed
most to its value.

% Expectations

It is expected that for higher protractedness (thus deviation from BEAST2
its assumptions), the error will increase.
Trees are expected to be inferred better, when there is more information
available to construct them, thus we expect less error when there are more
taxa and when the simulation DNA alignments are longer.
The error is expected to be biggest close to the present, as there will
be the most incipient species present.

It is unknown, however, under which biological settings this error is relevant.

% Preview results

This study shows that ... \todo{put result here}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Methods}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

This research measures the error our current BD tools make, when
the real process has been protracted. Figure \ref{fig:experiment} shows
the following overview of the experiment graphically.
 An in silico data set is created for a wide
range of parameters from which incipient species trees are simulated (A in figure \ref{fig:experiment}). 
These phylogenies serve as 'the truth' or 'nature'. From that incipient species tree,
species trees are created, by picking one representive per species.
This is done twice, as either the youngest (B1) or oldest (B2) subspecies is chosen
to represent a species. Per species tree, $n_a$ alignments are simulated (C1 and C2).
Per alignment, $n_b$ posteriors are inferred, 
each consisting of $n_s$ phylogenies and parameter estimates (D1 and D2). Each
species tree is then compared with the species tree its alignment is based on (for
example, B1 and D1).

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Parameter space}
\label{subsec:parameter_space}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\todo{@RSE: would you prefer $t_{\mean{ds}}$ or $\mean{t_{ds}}$ for average duration of speciation?}

% Biological parameters

This study primarily investigates combinations of the five biological 
parameters ($b_g$, $b_g$, $\lambda$, $\mu_g$ and $\mu_g$, 
see also figure \ref{fig:pbd_states} and table \ref{table:parameter_descriptions}), 
in a close to full factorial fashion, with two exceptions.
First, following \cite{etienne2014estimating}, birth and extinction rates were each 
set equal between good and incipient species ($b_g = b_i$, and $\mu_g = \mu_i$).
Second, only parameter combinations with speciation initiation rates greater
than extinction rates were used ($b_g > \mu_g$). The values 
used in the experiment (see table \ref{table:parameter_values}) 
were selected to have a reasonable expected number of incipient 
lineages, $E(n_{ti})$, where \todo{?too basic to have a reference to? I assume Kendall, 1948, \cite{kendall1948generalized}, but could not look it up here and now}:

\begin{equation}
  E(n_{ti}) = e^{ (b - \mu) \cdot t_c }
\end{equation}

One value of speciation completion rate $\lambda$ is set (close to) infinity, as
then the protracted speciation model falls back to a constant-rate 
(non-protracted) birth-death model [?Etienne et al., 2012 or already Rosindell and Etienne 2011?].

% Biological tuning parameters

Secondarily, there are three biological tuning parameters. The crown age of
the incipient species tree, $t_c$, is set to 15 million years for all simulations, 
following \cite{etienne2014estimating}. The per-nucleotide mutation rate, $r$, and the
DNA aligment length, $l_a$, together must be tuned to result in informative
DNA alignments (see section \ref{subsec:simulating_alignments}). $r$ is identical in all
simulations. Increasing $l_a$ results in more informative alignments, and this
effect is measured by using two values of $l_a$ 
per biological parameter combination.

% Replication parameters

All processes involving stochasticity have replicates. 
For each set of parameters, $n_i$ incipient species trees are simulated. 
From each species tree, $n_a$ alignments are simulated. From each alignment, 
$n_b$ posteriors are generated.

% MCMC tuning parameters

The posteriors are generated using BEAST2, which uses a Monte-Carlo Markov-Chain (MCMC)
algorithm. This algorithm is tuned to result in a high effective sample size (ESS) by
two constants. The first constant is the MCMC sampling interval, which is $i_s$ states.
$i_s$ which must be high enough to avoid correlation between subsequently sampled posterior states.  
The second constant is the number of MCMC state samples, $n_s$, which equals the ESS iff \todo{@RSE: Can I use iff?}
$i_s$ is high enough to have no correlation between subsequently sampled posterior states.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Simulating incipient species trees}
\label{subsec:simulating_incipient_species_trees}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

From each parameter set, $n_i$ incipient species trees with a crown age
of $t_c$ and at least two extant taxa, 
are created in R \cite{R} using the PBD package \cite{PBD}.
The values of the biological parameters encourage that the number of tips in
each phylogeny is kept within computational limits. However, due to
the stochastic nature of the Doob-Gillespie algorithm used in the creation
of these phylogenies, phylogenies can, by chance, become very big. 
There is no maximum number of taxa set, yet imposed instead by the R 
programming language itself, due to a limit on the amount of memory that can be 
allocated. The incipient species trees that could not be simulated are
counted and discarded from further analysis.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Sampling species trees}
\label{subsec:sampling_species_trees}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Per incipient species tree, two species trees are created. 
To create a species tree (a tree in which each species is represented once)
for an incipient species tree, one representative for each species is picked. 
If a species has initiated one or more incipient species, either the youngest
or the oldest representative is picked. See figure \ref{fig:sampling} for an example.

An alternative sampling method would be to pick 
random representatives of each species. This would decrease the variability
between the two species tree. Maximizing variability between the two
species trees allows for clearer measuments on the effect of sampling.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Simulating alignments}
\label{subsec:simulating_alignments}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

From each species tree, $n_a$ DNA alignments were simulated that match the
phylogeny's structure, using the phangorn R package \cite{phangorn}. 
Each alignment contains one DNA sequence per species 
of $l_a$ nucleotides. The alignment follows the Jukes-Cantor nucleotide 
subsititution model \cite{cantor1969mammalian}, 
with per-nucleotide rate of mutation, $r$.

The Jukes-Cantor
nucleotide substitution model is regarded as an oversimplification of 
nature \cite{unknown}. In this in silico experiment, however, we simulate first
a DNA alignment from a phylogeny using nucleotide substitution model $M_N$, 
then infer a phylogeny from that DNA alignment, assuming nucleotide substitution model $M_N$.
Too reduce unnecessary complexity, $M_N$ is chosen to be the simplest possible.

The combination of $l_a$ and $r$ is tuned to result in informative alignments. 
Within a maximally informative alignment, each species has its own unique DNA sequence.
Additionally, the Jukes-Cantor distance \cite{unknown} between any 
two sequences must be calculable, which is when the proportion of matching nucleotides is below 75\%.
The tuning was done using an $l_a$ of $10^3$ nucleotides, 
as this appears to be the current median of the sequenced genome lengths, as currently published.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Simulation of posteriors}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

From each DNA alignment, $n_b$ posteriors of phylogenies and parameter combinations 
is constructed, using BEAST2, a Bayesian phylogenetic tool \cite{bouckaert2014beast}.
The nucleotide substitution model is set to 
be Jukes-Cantor (one nucleotide substition rate category, no invariant nucleotides) 
and a strict molecular clock, which matches the procedure the DNA alignments are generated, 
as described in section \ref{subsec:simulating_alignments}).
The tree prior is set to the (non-protracted) birth-death model, 
with an uniform prior for birth-rate of range $[0, 10^5]$ and 
a uniform death-per-birth rate prior of range $[0, 1]$. 
The MCMC creation parameters were set to have $n_s$ states, sampled at an interval of $i_s$ states.

An MCMC must have an effective sample size (ESS) of at least 200 to result in a 
posterior that is 'adequate' \cite{drummond2015bayesian}. Due to the multiple 
sources of stochasticity in the preceding steps, this ESS may not always be achieved.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Analysis}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

To quantify the error, each species tree is compared with the posterior derived
from its $n_a$ alignments (for example, B1 and D1 in figure \ref{fig:experiment}).
This is by calculating the nLTT statistic \cite{janzen2015approximate}
between the species tree and each of its posterior's trees (using 
the nLTT package \cite{nLTT}), resulting in an nLTT statistic distribution.

% BD trees

The nLTT statistic distribution for all simulations with a high speciation 
completion rate is visualized for the different parameter values. This allows
to see the baseline error. Instead of using a complete
nLTT statistic distribution, also the mean value of each posterior is 
sometimes used to reduce variance in this.

The effect of speciation completion time $\lambda$ is visualized, by showing
nLTT statistic distributions for the different parameter values, for all values
of $\lambda$.

Instead of using the parameter value of $\lambda$ as a proxy for protractedness,
the speciation completion time, $t_{sc}$, and average duration of speciation, $t_{\mean{ds}}$, 
are used. Speciation completion time, $t_{sc}$, is simply the inverse of the
speciation completion rate $\lambda$:

\begin{equation}
  t_{sc} = \frac{1}{\lambda}
  \label{eq:speciation_completion_time}
\end{equation}

The speciation completion time differs from the observed average speciation time, 
as incipient species may go extinct and may produce other incipient species 
that complete speciation before the original incipient one. The average duration 
of speciation, $t_{\mean{ds}}$, takes this into 
account \todo{check formula ?19 of?20 in Etienne et al., 2012, then cite}:

\begin{equation}
    t_{\mean{ds}} = \mean{\rho} = \frac{2}{ (D - \lambda + b_i - \mu_i) \cdot \ln{ \frac{2}{(1 + (\lambda - b_i + \mu_i)/D)}  } }
\end{equation}

where $D$ is defined as:

\begin{equation}
    D = \sqrt{ (\lambda + b_i)^2 + 2 \cdot (\lambda - b_i) \cdot \mu_i + \mu_i^2 }
\end{equation}

For each simulation, the expected $t_{\mean{ds}}$ is calculated from its parameter 
values (using the PBD package \cite{PBD}), after which the nLTT statistic 
distributed is plotted, to see the effect of protractedness on the nLTT statistic
distribution.

As not all simulations result in posteriors with an adequate ESS, the
effect that a below-median or above-median ESS has on the nLTT statistic
distribution is displayed.

To view the effect of alignment length, $l_a$, the nLTT statistic
distributions are plotted side-by-side for all other parameter values.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Peripherals}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Simulating incipient species trees, sampling species trees, simulating alignments
and creating posteriors, are done with the wiritttes package \cite{wiritttes}. The
analysis is done with the wiritttea package \cite{wiritttea}. Those packages rely on 
functionality in the APE \cite{APE}, geiger \cite{GEIGER}, ggplot2 \cite{ggplot2}, 
RBeast \cite{RBeast}. Quality of these packages is assured by 
the covr \cite{covr}, devtools \cite{devtools}, 
goodpractice \cite{goodpractice}, lintr \cite{lintr}, testit \cite{testit} 
and testthat \cite{testthat} packages, using the Travis CI \cite{travis} continuous integration service. 
Both packages are hosted at GitHub \cite{github}, a good practice for computational
scientists \cite{perez2016ten} and helps improve transparency \cite{gorgolewski2016practical}.
Code coverage, which correlates with code quality \cite{del1995correlation}, 
is above 90\% for both packages.

The simulated dataset can be downloaded from DataDyrad [but not yet]. The
summarized dataset can be downloaded at \cite{wirittte_data}. The
pre-publication of this article can be downloaded at \cite{wirittte_article}.

Simulations and analyis was mostly performed on the Peregrine 
high performance computing cluster of the University of Groningen.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Results}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Creation of data set}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\todo{First focus on the real results, not their quality}

% Simulation of incipient species trees

\todo{redo with new data}

\todo{create figure or table that shows this}

957 out of 960 incipient species trees have been simulated successfully. 
The 3 failed simulations created data files bigger than R can read, 
and have been excluded from further analysis.

The number of good species ranged from 2 to 14296, with a median of 87 and an average of 503.8. 
There were 6 phylogenies with only 2 good species.

% Sampling species trees}

\todo{redo with new data}

For each incipient species trees, two species trees were constructed. 
In 653 out of 960 cases, these trees had different branch lengths.
In 302 cases, the trees were identical. In the remaining 5 cases,
there were computational problems (see figure \ref{fig:exit_statuses}).

% Simulation of alignments

\todo{redo with new data}

For each species tree, $n_a$ alignments have been created, 
resulting in $960 \cdot 2 \cdot 2 = 3840$ alignments. 
$?$ of these were informative.

\todo{May result in an overrepresentation of the two same species trees derived from one incipient species tree}

% Simulation of posteriors}

A BEAST2 posterior contains multiple parameter estimates (see figure \ref{fig:ess_distribution} for all),
and each one has an ESS. An ESS of 200 is acceptable for any parameter estimate.
The ESS of the tree likelihood should have such a decent ESS. Figure \ref{fig:ess_distribution_likelihood}
shows the distribution of the likelihood ESS.

\todo{How tp quote https://www.beast2.org/what-is-ess/ ? It states that not all variables
need to have a high ESS, but 'The likelihoods (both of the tree and coalescent model) should have decent ESSs.}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Quantification of the error BEAST2 makes on all trees}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

To get a first global impression of the error BEAST2 makes, figure \ref{fig:error}
shows a histogram of the nLTT statistics, colored by the expected mean
duration of speciation. As BEAST2 assumes a mean duration of speciation of zero,
it is expected that the nLTT statistic (a measure of error) is lowest. Zooming
in on the left part of the distribution, there is no clean pattern.
The right part of the tail of distribution shows that the biggest nLTT
statistic values are found for high (but not highest) expected mean durations of
speciation.

\todo{figure \ref{fig:error_for_scr} shows the distribution of errors for all
combinations of PBD parameters}

As the expected correlation appears to be absent, figure 
\ref{fig:error_expected_mean_dur_spec}, subfigure $a$, 
shows the correlation between expected
mean duration of speciation and the nLTT statistic. A linear fit shows
that there is indeed an increase in error for an increases expected duration
of speciation. A trendline using locally weighted smoothing, however,
indicates a more complex relationship. Note that this figure only uses
a subset of all measurements, consisting of $n_s$ nLTT statistic values 
per posterior.

Taking the average nLTT value per posterior frees us from computational
constraints and allows us to use the full data set. The resulting
plot, figure \ref{fig:error_expected_mean_dur_spec}, subfigure $b$, shows the same pattern:
A linear fit shows
that there is indeed an increase in error for an increases expected duration
of speciation. But now, a trendline using locally weighted smoothing 
follows this linear relationship closely.

The inference error made is expected to be less for a longer DNA
alignment length. 
Figure \ref{fig:error_expected_mean_dur_spec_alignment_length} 
show the correlation between expected mean duration of speciation 
and the nLTT statistic for the two different DNA alignment lengths. 
The first subfigure uses a sample of all $n_s$ individuals nLTT statistics,
where the second subfigure takes only the posterior's 
nLTT statistic mean value, for the complete dataset.

Both figures show again that there is an 
increase in error for an increases expected duration
of speciation, when assuming a linear relation, and
a more complex relationship when using locally weighted smoothing. 
The figures also show that, as expected, 
the error is less with more information. Additionally,
a longer DNA alignment length results in a slower accumulation of error
for increased expected mean duration of speciation.

A higher ESS for longer DNA alignments may be an additional
explanation for the observed slower accumulation of error
for increased expected mean duration of speciation.
Figure \ref{fig:ess_expected_mean_dur_spec_alignment_length}
shows that longer DNA alignments have a higher likelihood ESS, 
but this trend is weaker than in figure 
\ref{fig:error_expected_mean_dur_spec_alignment_length} and
the 95\% confidence intervals of the linearily fitted trendlines only
stop overlapping for the highest expected mean duration of speciation.
The locally weighted smoothing trendlines show a different trend,
suggesting that alignment length has no effect on the
likelihood ESS.

\todo{Figure \ref{fig:error_alignment_length} shows the same pattern, seperated
for all combinations of PBD parameters.}

For each incipient species tree, two species trees were created, by taking
either the youngest or oldest subspecies to represent each species.
The effect of this is shown in figure \ref{fig:error_expected_mean_dur_spec_sampling}.
Both subfigures show the
error made for different $t_{\mean{ds}}$, where the first subfigure uses a sample
of all $\Delta_{nLTT}$ values, where the second subfigure uses the $\mean{\Delta_{nLTT}}$
of each posterior. In both subfigures, the linear and locally weighted smoothing
have overlapping 95\% confidence intervals, hinting that there is no strong effect
of the way subspecies are selected to represent their species.

\note{Figure \ref{fig:error_sampling_representative} shows the same data, ordered
by the different PBD parameters instead}

The effect of having a low or high tree likelihood ESS is shown in 
figure \ref{fig:figure_error_expected_mean_dur_spec_low_high_ess}. We
define a low ESS as being below median, where a high ESS is defined as 
above-median. Plotting (a sample of or the means of) the nLTT statistic for different mean
duration of speciation for a low an high ESS, shows that there is no clear
difference in error made for posteriors with either low or high tree likelihood ESSes.

It is expected that the bigger phylogeny is simulated, the less the error
will be. Figure \ref{fig:figure_error_tree_size} shows the relation between the
number of taxa and the mean nLTT statistic. Although most simulated
phylogenies have less than 250 taxa, there is a clear positive trend, 
which is the opposite of the
expactation: the bigger the phylogeny, the bigger the error made.
As the nLTT statistic normalized for the number of taxa, some other
mechanism must be involved.

For each species tree, $n_a$ different alignments are simulated. For each of
these alignments, $n_b$ posteriors are simulated, using the same alignment. 
It is assumed that two posteriors from the same alignment
give rise to the same parameter estimates distributions.
distribution. Figure \ref{fig:figure_posterior_distributions_likelihood} 
shows the distribution of Mann-Whitney U test p-values for each pair of 
posteriors working on a same alignment, 
using their tree likelihood estimates.

The figure shows that the tree likelihood posterior distributions 
from a same alignment nearly always
are significantly different. Subfigure (b) and (c) show the distribution of
tree likelihood estimates for a posterior pair with the lowest (b) and 
highest (c) p-value.

Posteriors based on the same alignment are also (next to estimated tree likelihood) 
assumed to have a similar nLTT statistic distribution. 
Figure \ref{fig:figure_posterior_distribution_nltt} 
shows the distribution of Mann-Whitney U test p-values for each pair of 
posteriors working on a same alignment, 
using their nLTT statistics.

The figure shows that the nLTT statistic distributions 
from a same alignment are commonly different and commonly the same. 
Subfigures $(b)$ and $(c)$ show the distribution of
nLTT statistics for a posterior pair with the lowest $(b)$ and 
highest $(c)$ p-value.

The nLTT statistic normalizes a phylogeny to have a crown age of 1.
To prevent overlooking patterns in posterior crown ages, 
figure \ref{fig:figure_posterior_distribution_crown_ages}
shows the distribution of crown ages of all simulations.
Almost all simulations have a crown age close to $0.15$, but there
are rare cases with crown ages above $0.2$, ranging to $83.4$.

\todo{I think it is very suspicious that the range centers on 0.15, when
the crown age is 15. Check this!}

Again, it is assumed that posteriors based on the exact same alignment
have a similar crown age distribution. Of all posterior pairs, the Mann-Whitney
U test p-values are shown in figure \ref{fig:figure_posterior_distribution_crown_ages_p_values}.
It shows that crown age distributions being different is common and these
being the same is uncommon.

It is expected that if a tree is created by a Birth-Death process, BEAST2 is able
to reasonably estimate those parameters.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Discussion}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

This article has quantified the error made in inference when (1) nature would follow the
protracted speciation model, and (2) her DNA sequences would be analysed with a birth-death
model.

% The PBD model is just one implementation of speciation taking time

The key feature of the PBD model is that it encompasses speciation taking time in the
simplest way possible (that allows for mathematical analysis). The PBD model allows for
multiple incipient species stages (where its earliest version had infinite
stages \cite{rosindell2010protracted}). This research assumes that there is only one
incipient species stage. One could argue that speciation takes place in more
biological stages \todo{Add reference}, but we tried to use the simplest incarnation of the PBD possible.
Also, the PBD model can also be viewed as a model of species recognition \todo{?Purvis et al., 2009},
in which (human) recognition needs to build up, which we consider to be sufficiently
encapsulated in one single stage. Future research may investigate if adding multiple incipient
species stages for simulating nature changes the error made by (non-protracted) birth-death models.

% Equal and constant rates

The PBD model allows for multiple incipient species stages. 
Each stage has its own extinction and speciation-initiation and 
speciation-completion rate (except for good species, which lacks 
a speciation-completion rate). These rates are assumed to be constant
in time and equal for all stages. We chose to use the constant-rate
birth-death model to keep this investigation as simple as possible.
Future research may investigate whether a biologically more realistic
rate model, for example, by adding diversity-dependence, results in
the same inference error.

We chose to set the extinction and speciation-initiation rates equal for all stages,
again for choosing the simplest model. We expect these rates to differ in reality:
incipient species have low initial abundances, which has an effect on extinction
and speciation-initiation rates.

In this research, we nevertheless set these rates equal. 
Not only because it helps constrain the parameter space investigated,
also paraphylies would pop up in the simulated trees. This would violate more assumptions
of the BEAST2 model used as a prior and increase the complexity of the analysis.

% Sampling species trees from incipient species trees

An incipient species tree may have multiple incipient lineages from the
same ancestor. We constructed two species trees from each incipient species
tree, by selecting the youngest and oldest subspecies to represent a species.
This selection maximizes the chance the species trees are different, where
random sampling of representatives would also be a valid approach, closer
to what would happen when scientists take measurements in the field. Our
approach increases the error made.

% Nucleotide substitution model

For the simulation of DNA alignments, the Jukes-Cantor nucleotide substitution model
is used. This substitution model is a simplification of the DNA mutation processes
happening in reality \todo{Add reference}. In this simulation research, however, we could also
state in BEAST2 that this nucleotide substitution model is used. Thus, we kept
the simulation of the nucleotides in time as simple as possible, and used that
knowledge in later steps. We expect that any nucleotide substitution model, when
used in both alignment generation and in phylogeny inference, will do equally well,
except that the full computational pipeline would take needlessly longer.

% Allopatric speciation only

When a speciation event is initiated, this investigation assumes no gene flow
between the two incipient lineages. This follows the allopatric mode of speciation, 
which is assumed to be most prominent of nature \todo{Add reference}. It would be interesting to
allow gene flow between related incipient species, 
so also peripatric and sympatric modes of speciation would be allowed for.
We predict that this would increase the error in inference.

% Why not put the tree prior into an MCMC tool directly?
One could argue we should have written a PBD tree prior for an MCMC phylogenetic tool.
We did not do that for two reasons. First, we think it is rational to first
measure if there is need for a more complex tree prior. Second, 
we could not have done so, as a phylogenetic tool like BEAST2 assumes species
trees, where our input consists of incipient species trees. Before this
research, we were unsure how to sample species representatives. This research
has shown that any sampling is just as good, opening up avenues for
implementing a PBD species tree prior.

% BEAST2 is just one of many tools

BEAST2 is one of many MCMC phylogenetic tools with its
own internal technicalities. Similar tools are the R phytools
package \cite{phytools}, BAli-Phy \cite{suchard2006bali}
and BayesPhylogenies \cite{pagel2007bayesphylogenies}, to name a few.
BEAST2 was chosen, because (1) it's FOSS, (2) its software
architecture is modular, encouraging extension, (3) its development
team follows good practices in software development.

% BEAST2 parameter settings: strict clock

\todo{elaborate}

Use of strict clock is clean, but unrealistic.

\todo{elaborate}

% Alignment generation

Use of JC69 is rational, but unrealistic.

% Use for empiricists

\todo{elaborate}

This reseach shows that a DNA sequence length of 10k [is good].
[must be effective DNA sequence length]
[no indels generated]

% Low number of incipient species trees

\todo{elaborate}

For each point in parameter space, twenty incipient species tree are created.
Was fitted to 1000 jobs.

% Failed simulations

Not all parameter combinations resulted in posteriors with a high ESS,
due to computational constraints. We think this is a valid approach:
would a phylogeny get too big, we as scientists would cut it to
pieces that are analytically tractable. Because these simulations
are only a minority (as shown in the Results section), we do not
expect this to influence our conclusions.

% nLTT statistic is just one of many summary statistics

We used the nLTT statistic for quantifying the error between the
true and inferred phylogeny. There are simpler statistics that
can do the same, like the gamma \cite{pagel1999inferring}, 
delta-r \cite{pigot2010shape} \todo{check citation}, 
tree balance \todo{Add reference} or
Normalized Rooted Branch Score [Drummond] statistic, 
or more complex, like the spectral analysis [Hendy and Penny, 1993].
We consider those other summary statistics equally good, we just picked the
one we thought of the most informative per unit of added computational time.

% Add:
%
% Computational Performance and Statistical Accuracy of *BEAST and Comparisons with Other Methods (by Huw A. Ogilvie Joseph Heled Dong Xie Alexei J. Drummond)
% cites rooted branch score, RBS, as being from Heled and Bouckaert 2013, `Looking for trees in the forest: Summary tree from posterior samples`

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Acknowledgements}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

We would like to thank the Center for Information Technology of the University of Groningen for their support
and for providing access to the Peregrine high performance computing cluster.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Authors' contributions}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

RJCB and RSE conceived the idea for the package. RJCB created and tested the package, and wrote the first draft of the manuscript. RSE contributed substantially to revisions.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Bibliography
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% MEE style
\bibliographystyle{mee}
\bibliography{article}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Tables}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{table}
  \centering 
  \begin{tabular}{l l}
    \hline
    Symbol & Description \\
    \hline
    \hline
    $b_g$ & Speciation initiation rate of a good species (probability per lineage, per million year) \\
    $b_i$ & Speciation initiation rate of an incipient species (probability per lineage, per million year) \\
    $\lambda$ & Speciation completion rate (probability per lineage, per million year) \\
    $\mu_g$ & Extinction rate of a good species (probability per lineage, per million year) \\
    $\mu_i$ & Extinction rate of an incipient species (probability per lineage, per million year) \\
    \hline
    $t_c$ & Crown age (million years) \\
    $r$ & Mutation rate (probability of mutation per nucleotide per million years) \\
    $l_a$ & DNA alignment length (base pairs) \\
    \hline
    $n_i$ & Number of incipient species trees per parameter set \\
    $n_a$ & Number of alignments per species tree \\
    $n_b$ & Number of BEAST2 runs per alignment \\
    \hline
    $n_s$ & Number of MCMC samples \\
    $i_s$ & MCMC sampling interval \\
    \hline
    $n_{ti}$ & Number of taxa in an incipient species tree \\
    $n_{t}$ & Number of taxa in a (good) species tree \\
    $t_{\mean{ds}}$ & Average duration of speciation \\
    $t_{sc}$ & Speciation completion time \\
    $\Delta_{nLTT}$ & nLTT statistic value \\
    $\mean{\Delta_{nLTT}}$ & Average nLTT statistic value of a posterior \\
    $\median{\Delta_{nLTT}}$ & Median nLTT statistic value of a posterior \\
    \hline
  \end{tabular}
  \caption{
    Parameter descriptions. The sections are (1) the biological parameters, (2) the
    biological tuning parameters, (3) parameters involved in replication, (4) MCMC
    tuning parameters, and (5) other symbols.
  }
  \label{table:parameter_descriptions}
\end{table}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{table}
  \centering 
  \begin{tabular}{l l}
    \hline
    Parameter             & Values \\
    \hline
    \hline
    $b = b_g = b_i$       & 0.1, 0.5, 1.0 \\
    $\lambda$             & 0.1, 0.3, 1.0, $10^6$ \\
    $\mu = \mu_g = \mu_i$ & 0.0, 0.1, 0.2, 0.4 \\
    \hline
    $t_c$                 & 15 \\
    $r$                   & 0.01 \\
    $l_a$                 & $10^3$, $10^4$ \\
    \hline
    $n_i$                 & 20 \\
    $n_a$                 & 2 \\
    $n_b$                 & 2 \\
    \hline
    $n_s$                 & $10^3$ \\
    $i_s$                 & $10^3$ \\
    \hline
  \end{tabular}
  \caption{
    Parameter values. The sections are (1) the biological parameters, (2) the
    biological tuning parameters, (3) parameters involved in replication, 
    and (4) MCMC tuning parameters.
  }
  \label{table:parameter_values}
\end{table}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Figures}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}
  \centering
  \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=4cm, semithick]   
  \tikzstyle{every state}=[]
  \node[state] (A)              {Good};   
  \node[state] (B) [right of=A] {Incipient};   
  \node[state] (C) [below of=B] {Extinct};   
  \path (A) edge [bend right] node {$b_g$} (B)
        (A) edge [bend right] node {$\mu_g$} (C)
        (B) edge [loop above] node {$b_i$} (B)
        (B) edge [bend right] node {$\lambda$} (A)
        (B) edge [bend right] node {$\mu_i$} (C); 
  \end{tikzpicture}

  \caption{
    The states and transitions of a species in the PBD model.
    $b_i$: speciation-initiation rate of incipient species. 
    $b_g$: speciation-initiation rate of good species. 
    $\lambda$: speciation completion rate. 
    $\mu_i$: extinction rate of incipient species. 
    $\mu_g$: extinction rate of good species. 
    Figure after Etienne et al, 2014, Evolution
  }
  \label{fig:pbd_states}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}
  \centering 
  \resizebox {0.8\textwidth} {!} {
    \CreateTikzFigureExperiment{} 
  }
  \caption{
    Overview of the experiment: 
    A: per parameter set, there are $n_i$ incipient species trees created. 
    B1: constructing a species trees by using the youngest subspecies $A1$ to represent good species $A$.
    B2: constructing a species trees by using the oldest subspecies $A2$ to represent good species $A$.
    C1: constructing one of the $n_a$ alignments of B1. 
    C2: constructing one of the $n_a$ alignments of B2. 
    D1: reconstructing a species trees from alignment B1.
    D2: reconstructing a species trees from alignment B2.
    Comparisons are done between a species tree and its reconstructed species tree, 
    for example, between B1 and D1.
    All phylogenies are drawn at the same scale.
  }
  \label{fig:experiment}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}
  \centering 
  \CreateTikzFigureSampling{}
  \caption{
    Sampling an incipient species tree. 
    A) At stem age $t_0$, $A1$ is present.
    At crown age $t_1$, its derived incipient species $A2$ is born. 
    At $t_2$, $A2$ itself speciates into a new (unnamed) species, 
    that becomes good species $B$ at $t_3$. At present time, $t_4$, there
    exist two possible representatives of species $A$. 
    B) Using the oldest representative, $A1$, to represent species $A$, the crown
    age of the resulting species tree lies at $t_1$. 
    C) Using the youngest representative, $A2$, to represent species $A$, the crown
    age of the resulting species tree lies at $t_2$. 
  }
  \label{fig:sampling}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\listoftodos

\end{document}

### 2018-03-26 obsolete/discussion.tex
Merged to two files

--- 2018-03-26 article.tex
@@
-\documentclass{article}
-
-% Bibliography
-\usepackage{natbib}
-\bibpunct{(}{)}{;}{a}{}{;}
-
-% Use 'It was found that A is B (Name 1234)' style
-\setcitestyle{authoryear,open={},close={}}
-
-% Affiliations
-\usepackage{authblk}
-\title{babette: BEAUti 2, BEAST2 and Tracer for R}
-\author[1]{Rich\`el J.C. Bilderbeek}
-\author[1]{Rampal S. Etienne}
-\affil[1]{Groningen Institute for Evolutionary Life Sciences, University of Groningen, Groningen, The Netherlands}
-
-% Load my functions
-\input{functions.tex}
-
-% Use double spacing
-\usepackage{setspace}
-\doublespacing
-
-\usepackage{listings}
-\usepackage{hyperref}
-\usepackage{todonotes}
-\usepackage{verbatim}
-\usepackage{pgf}
-\usepackage{bm}
-
-% TikZ and friends
-\usepackage{tikz}
-\usepackage{tkz-graph}
-\usepackage{pgf}
-\usetikzlibrary{arrows,automata}
-
-% Style of listings
-% From http://r.789695.n4.nabble.com/How-to-nicely-display-R-code-with-the-LaTeX-package-listings-tp4648110.html
-\usepackage{fancyvrb} 
-\definecolor{codegreen}{rgb}{0,0.6,0}
-\definecolor{codegray}{rgb}{0.5,0.5,0.5}
-\definecolor{codepurple}{rgb}{0.58,0,0.82}
-\definecolor{backcolor}{rgb}{0.95,0.95,0.92}
-\lstdefinestyle{mystyle}{
-  language=R,% set programming language
-  basicstyle=\ttfamily\small,% basic font style
-  commentstyle=\color{gray},% comment style
-  % numbers=left,% display line numbers on the left side
-  numberstyle=\scriptsize,% use small line numbers
-  numbersep=10pt,% space between line numbers and code
-  tabsize=2,% sizes of tabs
-  showstringspaces=false,% do not replace spaces in strings by a certain character
-  captionpos=b,% positioning of the caption below
-  breaklines=true,% automatic line breaking
-  escapeinside={(*}{*)},% escaping to LaTeX
-  fancyvrb=true,% verbatim code is typset by listings
-  extendedchars=false,% prohibit extended chars (chars of codes 128--255)
-  alsoletter={.<-},% becomes a letter
-  alsoother={$},% becomes other
-  otherkeywords={!=, ~, $, \&, \%/\%, \%*\%, \%\%, <-, <<-, /},% other keywords
-  deletekeywords={c}% remove keywords 
-}
-\lstset{style=mystyle}
-
-% Adds numbered lines
-\usepackage{lineno}
-\linenumbers
-
-% Use TODO notes
-\usepackage{todonotes}
-
-% Rename 'Abstract' to 'Summary 
-\usepackage[english]{babel}
-\addto{\captionsenglish}{\renewcommand{\abstractname}{Summary}}
-
-\begin{document}
-
-\maketitle
-
-\begin{abstract}
-
-  The construction of phylogenies helps us answer evolutionary biological
-  questions.
-  Our current phylogenetic tools ignore the fact that speciation takes time,
-  which has an effect unknown in phylogeny reconstruction.
-  Here, we simulate true incipient species trees and their corresponding
-  DNA alignments, from which we measure the reconstruction of the phylogeny
-  using a standard birth-death model.
-  We measure the errors that the Bayesian phylogenetic software tool BEAST2
-  gives when recovering simulated phylogenies, for different times-to-speciate,
-  under a range of additional parameter settings.
-  It has been found that branch lengths are consistently and strongly 
-  underestimated for biologically relevant parameters.
-  This research shows that protractedness is a complexity of nature that
-  cannot always be ignored and should be incorporated in our phylogenetic
-  tools.
-
-\end{abstract}
-
-{\bf Keywords:} computational biology, evolution, phylogenetics, BEAST2, R
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Introduction}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{Introduction}
-
-% Biology: speciation takes time
-
-It takes time for new species-to-be to achieve the status of a new species.
-First, it takes time for the new species-to-be 
-to acquire sufficient different mutations to 
-establish reproductive isolation \cite{schluter2001ecology}. Second, 
-it takes time for us to recognize reproductive isolation
-and reclassify that one single species has become two 
-(or more) species.
-
-As an example, \cite{fennessy2016multi} 
-showed that Africa has four instead of only one
-giraffe species, with a common ancestor estimated 
-at around 2 million years ago.
-
-% The use of speciation models
-
-From DNA sequences obtained in the present, we can construct phylogenies
-to infer the evolutionary relationship between species and the time 
-from which ancestral species developed into separate.
-
-As we only sample DNA of extant species, 
-we can only infer a reconstructed tree, in which
-the lineages that have gone extinct are absent.
-Although most models assume extinctions (\cite{yule1925mathematical} being the
-classic exception, as at its time computation was expensive), 
-information on extinction times is lost due to our sampling.
-
-The time when reproductive isolation started to take off, is
-also lost in our sampling method, if the new species-to-be is not 
-recognized yet as such. Without fossil date, one cannot observe the
-possibly many times that reproductive isolation built up, but was lost again.
-
-% First model: constant rate birth death model
-
-Although we know that speciation takes time, we commonly ignore this when
-constructing a phylogeny, by choosing a constant-rate birth-death model
-as a speciation model. This can be justified, as for lineages 
-that have speciated far back in the past, there is no error. 
-Yet in the present, we do not recognize the future species-to-be
-that are already genetically converging, resulting in an underestimation
-of the number of species present today, which can only be 
-concluded in retrospect.
-
-The constant-rate birth-death model (as described in for example \cite{nee1994reconstructed}) 
-is among the simplest speciation models, and assumes a constant speciation
- rate $\lambda$ and constant extinction rate $\mu$.
-
-Additionally, it assumes speciation is instantaneous.
-The constant-rate birth-death model is popular for its simplicity, yet
-has also served as a starting point for more elaborate speciation models.
-
-% Other non-protracted speciation models
-
-Other speciation models may assume that speciation rate changes in 
-time \cite{rabosky2008explosive}, is dependent on the amount of species 
-present \cite{etienne2011diversity}, or is trait 
-dependent \cite{fitzjohn2009estimating}.
-
-In these examples, the original birth-death model has been extended by making 
-speciation rate dependent on time, diversity or trait value respectively.
-The assumption that all these extensions share is that speciation is instantaneous;
-that after a new branching event, both two lineages are (or are recognized as being) different
-species.
-
-% Protracted speciation model
-
-% Good and incipient states
-
-The protracted speciation model \cite{etienne2012prolonging} allows for speciation taking time.
-
-It adds an additional species state (see also figure \ref{fig:pbd_states}),
-coined the 'incipient' stage, which a lineages has to complete before becoming
-a 'good' species.
-
-One view is to say that good species have achieved reproductive isolation,
-where incipient species are in the process of achieving this \todo{Add reference}.
-
-As genetic drift is unlikely to be a relevant speciation mechanism,
-this is more likely to be ecological speciation [Sobel et al., 2009]. 
-\todo{RSE: Ben je het daarmee eens? Zou drift niet een belangrijke ooraak van speciation op islands kunnen zijn bijvoorbeeld?}
-
-Alternatively, an incipient species can be described as a good-species-to-be,
-yet not recognized as such \todo{?[Purvis et al., 2009]?}.
-
-% Biological mechanism of reproductive isolation
-
-The protracted speciation model makes no assumption about the exact biological
-mechanism by which reproductive isolation is 
-acquired \cite{etienne2012prolonging, etienne2014estimating,rosindell2010protracted}.
-
-A 'good' state can be viewed as a state in which there is convergent selection
-for that lineage.
-
-The 'incipient' state can be seen as a state in which there is the potential
-for speciation.
-
-A first example can be from the BDM \todo{Add reference} model, the state in which there
-are all three genotypes, before the intermediate genotype is lost.
-A second example can be in a sexual selection model [van Doorn and Weissing, 2002], 
-where there is the onset for convergent selection.
-In both cases, these intermediate states may fall back to the original
-state of one species, or complete to produce a new species.
-
-% Effects on topology
-
-A feature of the protracted birth-death model, is that it may result in
-paraphylies when the rates between good and incipient species differ \todo{Add reference}.
-Paraphylies may be more than an experimental artifact and are claimed to
-be an inherent feature of nature \cite{funk2003species}.
-
-\todo{Add something like: this research solves this by sampling from paraphyletic trees (at the subspecies level) to create non-paraphyletic trees at the species level}
-
-% Good and incipient rates
-
-Where the classic birth death model assumes constant speciation and extinction
- rates, the protracted birth-death model allows these rates to be state
- dependent.
-
-The only additional parameter is the rate at which incipient species become
-(recognized as) good species.
-When setting that parameter to infinity, incipient species become good
-species instantaneous and the model falls back to a constant-rate (non-protracted) 
-birth-death model \todo{Add reference}.
-
-% Relationship to constant-rate birth-death model
-
-\todo{?Wat bedoelde ik met deze onzin?}
-Although both standard and protracted birth-death models use a parameter 
-$\lambda$, the $\lambda$ in the standard birth death model denotes
-a speciation rates (initiation and completion), where in the PBD model
-$\lambda$ denotes only the speciation initiation rate. 
-There is no relationship known between the parameters of the constant-rate
-birth-death model and the protracted birth-death model \cite{etienne2014estimating} 
-\todo{refesh memory, then update paragraph}.
-
-% Good and incipient rates do not need to be constant
-
-The assumption that speciation takes time is independent of the dynamics
-of the speciation and extinction rates.
-The protracted speciation model in this research assumes these rates are
-constant, but these rates can easily be made to depend on time, diversity
-or trait.
-
-% Use of a speciation model in inferring a phylogeny
-
-Speciation models are widely used to make inferences from genetic data.
-There are multiple computer programs to create phylogenies and/or parameter
-estimates from aligned DNA sequences.
-BEAST2 is a widely used tool that allows for a Bayesian approach to 
-phylogenetics \cite{bouckaert2014beast}.
-
-BEAST2 supplies the user with multiple speciation models, that all assume
-instantaneous speciation.
-
-% This study
-
-This simulation investigates the consequence of BEAST2 using instantaneous
-speciation, by simulating a 'true' tree that is protracted, simulate DNA
-alignments following that tree, and seeing how well BEAST2 can recover
-the original phylogeny.
-
-% Novelty
-
-This study is the first to measure the error made when acknowledging that
-speciation takes time, yet using one of the many tools that ignores this
-fact.
-
-Part of the parameter space will create 'true' trees that are not protracted.
-This will result in a test of BEAST2 against itself, which is, to the best
-of our knowledge, not been done, probably due to the extensive computations
-that are needed \todo{RSE: Heb je dit uitgebreid getest? Misschien zou je Remco kunnen vragen}.
-
-% Analysis
-
-The 'true' and inferred trees are compared with the nLTT statistic, which
-is a novel and efficient summary statistic, that has been proven to have
-better performance than the gamma statistics in Approximate
-Bayesian Computation \cite{janzen2015approximate} (yet on par with the likelihood).
-
-The nLTT statistics is the summed absolute difference between two 
-Lineages-Through-Time plot lines that are normalized to 
-let both time and number of lineages
-range from zero to one.
-Additionally, this statistic allows to pinpoint the time span that contributed
-most to its value.
-
-% Expectations
-
-It is expected that for higher protractedness (thus deviation from BEAST2
-its assumptions), the error will increase.
-Trees are expected to be inferred better, when there is more information
-available to construct them, thus we expect less error when there are more
-taxa and when the simulation DNA alignments are longer.
-The error is expected to be biggest close to the present, as there will
-be the most incipient species present.
-
-It is unknown, however, under which biological settings this error is relevant.
-
-% Preview results
-
-This study shows that ... \todo{put result here}.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Methods}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-This research measures the error our current BD tools make, when
-the real process has been protracted. Figure \ref{fig:experiment} shows
-the following overview of the experiment graphically.
- An in silico data set is created for a wide
-range of parameters from which incipient species trees are simulated (A in figure \ref{fig:experiment}). 
-These phylogenies serve as 'the truth' or 'nature'. From that incipient species tree,
-species trees are created, by picking one representive per species.
-This is done twice, as either the youngest (B1) or oldest (B2) subspecies is chosen
-to represent a species. Per species tree, $n_a$ alignments are simulated (C1 and C2).
-Per alignment, $n_b$ posteriors are inferred, 
-each consisting of $n_s$ phylogenies and parameter estimates (D1 and D2). Each
-species tree is then compared with the species tree its alignment is based on (for
-example, B1 and D1).
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Parameter space}
-\label{subsec:parameter_space}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\todo{@RSE: would you prefer $t_{\mean{ds}}$ or $\mean{t_{ds}}$ for average duration of speciation?}
-
-% Biological parameters
-
-This study primarily investigates combinations of the five biological 
-parameters ($b_g$, $b_g$, $\lambda$, $\mu_g$ and $\mu_g$, 
-see also figure \ref{fig:pbd_states} and table \ref{table:parameter_descriptions}), 
-in a close to full factorial fashion, with two exceptions.
-First, following \cite{etienne2014estimating}, birth and extinction rates were each 
-set equal between good and incipient species ($b_g = b_i$, and $\mu_g = \mu_i$).
-Second, only parameter combinations with speciation initiation rates greater
-than extinction rates were used ($b_g > \mu_g$). The values 
-used in the experiment (see table \ref{table:parameter_values}) 
-were selected to have a reasonable expected number of incipient 
-lineages, $E(n_{ti})$, where \todo{?too basic to have a reference to? I assume Kendall, 1948, \cite{kendall1948generalized}, but could not look it up here and now}:
-
-\begin{equation}
-  E(n_{ti}) = e^{ (b - \mu) \cdot t_c }
-\end{equation}
-
-One value of speciation completion rate $\lambda$ is set (close to) infinity, as
-then the protracted speciation model falls back to a constant-rate 
-(non-protracted) birth-death model [?Etienne et al., 2012 or already Rosindell and Etienne 2011?].
-
-% Biological tuning parameters
-
-Secondarily, there are three biological tuning parameters. The crown age of
-the incipient species tree, $t_c$, is set to 15 million years for all simulations, 
-following \cite{etienne2014estimating}. The per-nucleotide mutation rate, $r$, and the
-DNA aligment length, $l_a$, together must be tuned to result in informative
-DNA alignments (see section \ref{subsec:simulating_alignments}). $r$ is identical in all
-simulations. Increasing $l_a$ results in more informative alignments, and this
-effect is measured by using two values of $l_a$ 
-per biological parameter combination.
-
-% Replication parameters
-
-All processes involving stochasticity have replicates. 
-For each set of parameters, $n_i$ incipient species trees are simulated. 
-From each species tree, $n_a$ alignments are simulated. From each alignment, 
-$n_b$ posteriors are generated.
-
-% MCMC tuning parameters
-
-The posteriors are generated using BEAST2, which uses a Monte-Carlo Markov-Chain (MCMC)
-algorithm. This algorithm is tuned to result in a high effective sample size (ESS) by
-two constants. The first constant is the MCMC sampling interval, which is $i_s$ states.
-$i_s$ which must be high enough to avoid correlation between subsequently sampled posterior states.  
-The second constant is the number of MCMC state samples, $n_s$, which equals the ESS iff \todo{@RSE: Can I use iff?}
-$i_s$ is high enough to have no correlation between subsequently sampled posterior states.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Simulating incipient species trees}
-\label{subsec:simulating_incipient_species_trees}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-From each parameter set, $n_i$ incipient species trees with a crown age
-of $t_c$ and at least two extant taxa, 
-are created in R \cite{R} using the PBD package \cite{PBD}.
-The values of the biological parameters encourage that the number of tips in
-each phylogeny is kept within computational limits. However, due to
-the stochastic nature of the Doob-Gillespie algorithm used in the creation
-of these phylogenies, phylogenies can, by chance, become very big. 
-There is no maximum number of taxa set, yet imposed instead by the R 
-programming language itself, due to a limit on the amount of memory that can be 
-allocated. The incipient species trees that could not be simulated are
-counted and discarded from further analysis.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Sampling species trees}
-\label{subsec:sampling_species_trees}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-Per incipient species tree, two species trees are created. 
-To create a species tree (a tree in which each species is represented once)
-for an incipient species tree, one representative for each species is picked. 
-If a species has initiated one or more incipient species, either the youngest
-or the oldest representative is picked. See figure \ref{fig:sampling} for an example.
-
-An alternative sampling method would be to pick 
-random representatives of each species. This would decrease the variability
-between the two species tree. Maximizing variability between the two
-species trees allows for clearer measuments on the effect of sampling.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Simulating alignments}
-\label{subsec:simulating_alignments}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-From each species tree, $n_a$ DNA alignments were simulated that match the
-phylogeny's structure, using the phangorn R package \cite{phangorn}. 
-Each alignment contains one DNA sequence per species 
-of $l_a$ nucleotides. The alignment follows the Jukes-Cantor nucleotide 
-subsititution model \cite{cantor1969mammalian}, 
-with per-nucleotide rate of mutation, $r$.
-
-The Jukes-Cantor
-nucleotide substitution model is regarded as an oversimplification of 
-nature \cite{unknown}. In this in silico experiment, however, we simulate first
-a DNA alignment from a phylogeny using nucleotide substitution model $M_N$, 
-then infer a phylogeny from that DNA alignment, assuming nucleotide substitution model $M_N$.
-Too reduce unnecessary complexity, $M_N$ is chosen to be the simplest possible.
-
-The combination of $l_a$ and $r$ is tuned to result in informative alignments. 
-Within a maximally informative alignment, each species has its own unique DNA sequence.
-Additionally, the Jukes-Cantor distance \cite{unknown} between any 
-two sequences must be calculable, which is when the proportion of matching nucleotides is below 75\%.
-The tuning was done using an $l_a$ of $10^3$ nucleotides, 
-as this appears to be the current median of the sequenced genome lengths, as currently published.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Simulation of posteriors}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-From each DNA alignment, $n_b$ posteriors of phylogenies and parameter combinations 
-is constructed, using BEAST2, a Bayesian phylogenetic tool \cite{bouckaert2014beast}.
-The nucleotide substitution model is set to 
-be Jukes-Cantor (one nucleotide substition rate category, no invariant nucleotides) 
-and a strict molecular clock, which matches the procedure the DNA alignments are generated, 
-as described in section \ref{subsec:simulating_alignments}).
-The tree prior is set to the (non-protracted) birth-death model, 
-with an uniform prior for birth-rate of range $[0, 10^5]$ and 
-a uniform death-per-birth rate prior of range $[0, 1]$. 
-The MCMC creation parameters were set to have $n_s$ states, sampled at an interval of $i_s$ states.
-
-An MCMC must have an effective sample size (ESS) of at least 200 to result in a 
-posterior that is 'adequate' \cite{drummond2015bayesian}. Due to the multiple 
-sources of stochasticity in the preceding steps, this ESS may not always be achieved.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Analysis}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-To quantify the error, each species tree is compared with the posterior derived
-from its $n_a$ alignments (for example, B1 and D1 in figure \ref{fig:experiment}).
-This is by calculating the nLTT statistic \cite{janzen2015approximate}
-between the species tree and each of its posterior's trees (using 
-the nLTT package \cite{nLTT}), resulting in an nLTT statistic distribution.
-
-% BD trees
-
-The nLTT statistic distribution for all simulations with a high speciation 
-completion rate is visualized for the different parameter values. This allows
-to see the baseline error. Instead of using a complete
-nLTT statistic distribution, also the mean value of each posterior is 
-sometimes used to reduce variance in this.
-
-The effect of speciation completion time $\lambda$ is visualized, by showing
-nLTT statistic distributions for the different parameter values, for all values
-of $\lambda$.
-
-Instead of using the parameter value of $\lambda$ as a proxy for protractedness,
-the speciation completion time, $t_{sc}$, and average duration of speciation, $t_{\mean{ds}}$, 
-are used. Speciation completion time, $t_{sc}$, is simply the inverse of the
-speciation completion rate $\lambda$:
-
-\begin{equation}
-  t_{sc} = \frac{1}{\lambda}
-  \label{eq:speciation_completion_time}
-\end{equation}
-
-The speciation completion time differs from the observed average speciation time, 
-as incipient species may go extinct and may produce other incipient species 
-that complete speciation before the original incipient one. The average duration 
-of speciation, $t_{\mean{ds}}$, takes this into 
-account \todo{check formula ?19 of?20 in Etienne et al., 2012, then cite}:
-
-\begin{equation}
-    t_{\mean{ds}} = \mean{\rho} = \frac{2}{ (D - \lambda + b_i - \mu_i) \cdot \ln{ \frac{2}{(1 + (\lambda - b_i + \mu_i)/D)}  } }
-\end{equation}
-
-where $D$ is defined as:
-
-\begin{equation}
-    D = \sqrt{ (\lambda + b_i)^2 + 2 \cdot (\lambda - b_i) \cdot \mu_i + \mu_i^2 }
-\end{equation}
-
-For each simulation, the expected $t_{\mean{ds}}$ is calculated from its parameter 
-values (using the PBD package \cite{PBD}), after which the nLTT statistic 
-distributed is plotted, to see the effect of protractedness on the nLTT statistic
-distribution.
-
-As not all simulations result in posteriors with an adequate ESS, the
-effect that a below-median or above-median ESS has on the nLTT statistic
-distribution is displayed.
-
-To view the effect of alignment length, $l_a$, the nLTT statistic
-distributions are plotted side-by-side for all other parameter values.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Peripherals}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-Simulating incipient species trees, sampling species trees, simulating alignments
-and creating posteriors, are done with the wiritttes package \cite{wiritttes}. The
-analysis is done with the wiritttea package \cite{wiritttea}. Those packages rely on 
-functionality in the APE \cite{APE}, geiger \cite{GEIGER}, ggplot2 \cite{ggplot2}, 
-RBeast \cite{RBeast}. Quality of these packages is assured by 
-the covr \cite{covr}, devtools \cite{devtools}, 
-goodpractice \cite{goodpractice}, lintr \cite{lintr}, testit \cite{testit} 
-and testthat \cite{testthat} packages, using the Travis CI \cite{travis} continuous integration service. 
-Both packages are hosted at GitHub \cite{github}, a good practice for computational
-scientists \cite{perez2016ten} and helps improve transparency \cite{gorgolewski2016practical}.
-Code coverage, which correlates with code quality \cite{del1995correlation}, 
-is above 90\% for both packages.
-
-The simulated dataset can be downloaded from DataDyrad [but not yet]. The
-summarized dataset can be downloaded at \cite{wirittte_data}. The
-pre-publication of this article can be downloaded at \cite{wirittte_article}.
-
-Simulations and analyis was mostly performed on the Peregrine 
-high performance computing cluster of the University of Groningen.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Results}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Creation of data set}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\todo{First focus on the real results, not their quality}
-
-% Simulation of incipient species trees
-
-\todo{redo with new data}
-
-\todo{create figure or table that shows this}
-
-957 out of 960 incipient species trees have been simulated successfully. 
-The 3 failed simulations created data files bigger than R can read, 
-and have been excluded from further analysis.
-
-The number of good species ranged from 2 to 14296, with a median of 87 and an average of 503.8. 
-There were 6 phylogenies with only 2 good species.
-
-% Sampling species trees}
-
-\todo{redo with new data}
-
-For each incipient species trees, two species trees were constructed. 
-In 653 out of 960 cases, these trees had different branch lengths.
-In 302 cases, the trees were identical. In the remaining 5 cases,
-there were computational problems (see figure \ref{fig:exit_statuses}).
-
-% Simulation of alignments
-
-\todo{redo with new data}
-
-For each species tree, $n_a$ alignments have been created, 
-resulting in $960 \cdot 2 \cdot 2 = 3840$ alignments. 
-$?$ of these were informative.
-
-\todo{May result in an overrepresentation of the two same species trees derived from one incipient species tree}
-
-% Simulation of posteriors}
-
-A BEAST2 posterior contains multiple parameter estimates (see figure \ref{fig:ess_distribution} for all),
-and each one has an ESS. An ESS of 200 is acceptable for any parameter estimate.
-The ESS of the tree likelihood should have such a decent ESS. Figure \ref{fig:ess_distribution_likelihood}
-shows the distribution of the likelihood ESS.
-
-\todo{How tp quote https://www.beast2.org/what-is-ess/ ? It states that not all variables
-need to have a high ESS, but 'The likelihoods (both of the tree and coalescent model) should have decent ESSs.}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Quantification of the error BEAST2 makes on all trees}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-To get a first global impression of the error BEAST2 makes, figure \ref{fig:error}
-shows a histogram of the nLTT statistics, colored by the expected mean
-duration of speciation. As BEAST2 assumes a mean duration of speciation of zero,
-it is expected that the nLTT statistic (a measure of error) is lowest. Zooming
-in on the left part of the distribution, there is no clean pattern.
-The right part of the tail of distribution shows that the biggest nLTT
-statistic values are found for high (but not highest) expected mean durations of
-speciation.
-
-\todo{figure \ref{fig:error_for_scr} shows the distribution of errors for all
-combinations of PBD parameters}
-
-As the expected correlation appears to be absent, figure 
-\ref{fig:error_expected_mean_dur_spec}, subfigure $a$, 
-shows the correlation between expected
-mean duration of speciation and the nLTT statistic. A linear fit shows
-that there is indeed an increase in error for an increases expected duration
-of speciation. A trendline using locally weighted smoothing, however,
-indicates a more complex relationship. Note that this figure only uses
-a subset of all measurements, consisting of $n_s$ nLTT statistic values 
-per posterior.
-
-Taking the average nLTT value per posterior frees us from computational
-constraints and allows us to use the full data set. The resulting
-plot, figure \ref{fig:error_expected_mean_dur_spec}, subfigure $b$, shows the same pattern:
-A linear fit shows
-that there is indeed an increase in error for an increases expected duration
-of speciation. But now, a trendline using locally weighted smoothing 
-follows this linear relationship closely.
-
-The inference error made is expected to be less for a longer DNA
-alignment length. 
-Figure \ref{fig:error_expected_mean_dur_spec_alignment_length} 
-show the correlation between expected mean duration of speciation 
-and the nLTT statistic for the two different DNA alignment lengths. 
-The first subfigure uses a sample of all $n_s$ individuals nLTT statistics,
-where the second subfigure takes only the posterior's 
-nLTT statistic mean value, for the complete dataset.
-
-Both figures show again that there is an 
-increase in error for an increases expected duration
-of speciation, when assuming a linear relation, and
-a more complex relationship when using locally weighted smoothing. 
-The figures also show that, as expected, 
-the error is less with more information. Additionally,
-a longer DNA alignment length results in a slower accumulation of error
-for increased expected mean duration of speciation.
-
-A higher ESS for longer DNA alignments may be an additional
-explanation for the observed slower accumulation of error
-for increased expected mean duration of speciation.
-Figure \ref{fig:ess_expected_mean_dur_spec_alignment_length}
-shows that longer DNA alignments have a higher likelihood ESS, 
-but this trend is weaker than in figure 
-\ref{fig:error_expected_mean_dur_spec_alignment_length} and
-the 95\% confidence intervals of the linearily fitted trendlines only
-stop overlapping for the highest expected mean duration of speciation.
-The locally weighted smoothing trendlines show a different trend,
-suggesting that alignment length has no effect on the
-likelihood ESS.
-
-\todo{Figure \ref{fig:error_alignment_length} shows the same pattern, seperated
-for all combinations of PBD parameters.}
-
-For each incipient species tree, two species trees were created, by taking
-either the youngest or oldest subspecies to represent each species.
-The effect of this is shown in figure \ref{fig:error_expected_mean_dur_spec_sampling}.
-Both subfigures show the
-error made for different $t_{\mean{ds}}$, where the first subfigure uses a sample
-of all $\Delta_{nLTT}$ values, where the second subfigure uses the $\mean{\Delta_{nLTT}}$
-of each posterior. In both subfigures, the linear and locally weighted smoothing
-have overlapping 95\% confidence intervals, hinting that there is no strong effect
-of the way subspecies are selected to represent their species.
-
-\note{Figure \ref{fig:error_sampling_representative} shows the same data, ordered
-by the different PBD parameters instead}
-
-The effect of having a low or high tree likelihood ESS is shown in 
-figure \ref{fig:figure_error_expected_mean_dur_spec_low_high_ess}. We
-define a low ESS as being below median, where a high ESS is defined as 
-above-median. Plotting (a sample of or the means of) the nLTT statistic for different mean
-duration of speciation for a low an high ESS, shows that there is no clear
-difference in error made for posteriors with either low or high tree likelihood ESSes.
-
-It is expected that the bigger phylogeny is simulated, the less the error
-will be. Figure \ref{fig:figure_error_tree_size} shows the relation between the
-number of taxa and the mean nLTT statistic. Although most simulated
-phylogenies have less than 250 taxa, there is a clear positive trend, 
-which is the opposite of the
-expactation: the bigger the phylogeny, the bigger the error made.
-As the nLTT statistic normalized for the number of taxa, some other
-mechanism must be involved.
-
-For each species tree, $n_a$ different alignments are simulated. For each of
-these alignments, $n_b$ posteriors are simulated, using the same alignment. 
-It is assumed that two posteriors from the same alignment
-give rise to the same parameter estimates distributions.
-distribution. Figure \ref{fig:figure_posterior_distributions_likelihood} 
-shows the distribution of Mann-Whitney U test p-values for each pair of 
-posteriors working on a same alignment, 
-using their tree likelihood estimates.
-
-The figure shows that the tree likelihood posterior distributions 
-from a same alignment nearly always
-are significantly different. Subfigure (b) and (c) show the distribution of
-tree likelihood estimates for a posterior pair with the lowest (b) and 
-highest (c) p-value.
-
-Posteriors based on the same alignment are also (next to estimated tree likelihood) 
-assumed to have a similar nLTT statistic distribution. 
-Figure \ref{fig:figure_posterior_distribution_nltt} 
-shows the distribution of Mann-Whitney U test p-values for each pair of 
-posteriors working on a same alignment, 
-using their nLTT statistics.
-
-The figure shows that the nLTT statistic distributions 
-from a same alignment are commonly different and commonly the same. 
-Subfigures $(b)$ and $(c)$ show the distribution of
-nLTT statistics for a posterior pair with the lowest $(b)$ and 
-highest $(c)$ p-value.
-
-The nLTT statistic normalizes a phylogeny to have a crown age of 1.
-To prevent overlooking patterns in posterior crown ages, 
-figure \ref{fig:figure_posterior_distribution_crown_ages}
-shows the distribution of crown ages of all simulations.
-Almost all simulations have a crown age close to $0.15$, but there
-are rare cases with crown ages above $0.2$, ranging to $83.4$.
-
-\todo{I think it is very suspicious that the range centers on 0.15, when
-the crown age is 15. Check this!}
-
-Again, it is assumed that posteriors based on the exact same alignment
-have a similar crown age distribution. Of all posterior pairs, the Mann-Whitney
-U test p-values are shown in figure \ref{fig:figure_posterior_distribution_crown_ages_p_values}.
-It shows that crown age distributions being different is common and these
-being the same is uncommon.
-
-It is expected that if a tree is created by a Birth-Death process, BEAST2 is able
-to reasonably estimate those parameters.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Discussion}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 % cites rooted branch score, RBS, as being from Heled and Bouckaert 2013, `Looking for trees in the forest: Summary tree from posterior samples`
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Acknowledgements}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-We would like to thank the Center for Information Technology of the University of Groningen for their support
-and for providing access to the Peregrine high performance computing cluster.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Authors' contributions}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-RJCB and RSE conceived the idea for the package. RJCB created and tested the package, and wrote the first draft of the manuscript. RSE contributed substantially to revisions.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Bibliography
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% MEE style
-\bibliographystyle{mee}
-\bibliography{article}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Tables}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{table}
-  \centering 
-  \begin{tabular}{l l}
-    \hline
-    Symbol & Description \\
-    \hline
-    \hline
-    $b_g$ & Speciation initiation rate of a good species (probability per lineage, per million year) \\
-    $b_i$ & Speciation initiation rate of an incipient species (probability per lineage, per million year) \\
-    $\lambda$ & Speciation completion rate (probability per lineage, per million year) \\
-    $\mu_g$ & Extinction rate of a good species (probability per lineage, per million year) \\
-    $\mu_i$ & Extinction rate of an incipient species (probability per lineage, per million year) \\
-    \hline
-    $t_c$ & Crown age (million years) \\
-    $r$ & Mutation rate (probability of mutation per nucleotide per million years) \\
-    $l_a$ & DNA alignment length (base pairs) \\
-    \hline
-    $n_i$ & Number of incipient species trees per parameter set \\
-    $n_a$ & Number of alignments per species tree \\
-    $n_b$ & Number of BEAST2 runs per alignment \\
-    \hline
-    $n_s$ & Number of MCMC samples \\
-    $i_s$ & MCMC sampling interval \\
-    \hline
-    $n_{ti}$ & Number of taxa in an incipient species tree \\
-    $n_{t}$ & Number of taxa in a (good) species tree \\
-    $t_{\mean{ds}}$ & Average duration of speciation \\
-    $t_{sc}$ & Speciation completion time \\
-    $\Delta_{nLTT}$ & nLTT statistic value \\
-    $\mean{\Delta_{nLTT}}$ & Average nLTT statistic value of a posterior \\
-    $\median{\Delta_{nLTT}}$ & Median nLTT statistic value of a posterior \\
-    \hline
-  \end{tabular}
-  \caption{
-    Parameter descriptions. The sections are (1) the biological parameters, (2) the
-    biological tuning parameters, (3) parameters involved in replication, (4) MCMC
-    tuning parameters, and (5) other symbols.
-  }
-  \label{table:parameter_descriptions}
-\end{table}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{table}
-  \centering 
-  \begin{tabular}{l l}
-    \hline
-    Parameter             & Values \\
-    \hline
-    \hline
-    $b = b_g = b_i$       & 0.1, 0.5, 1.0 \\
-    $\lambda$             & 0.1, 0.3, 1.0, $10^6$ \\
-    $\mu = \mu_g = \mu_i$ & 0.0, 0.1, 0.2, 0.4 \\
-    \hline
-    $t_c$                 & 15 \\
-    $r$                   & 0.01 \\
-    $l_a$                 & $10^3$, $10^4$ \\
-    \hline
-    $n_i$                 & 20 \\
-    $n_a$                 & 2 \\
-    $n_b$                 & 2 \\
-    \hline
-    $n_s$                 & $10^3$ \\
-    $i_s$                 & $10^3$ \\
-    \hline
-  \end{tabular}
-  \caption{
-    Parameter values. The sections are (1) the biological parameters, (2) the
-    biological tuning parameters, (3) parameters involved in replication, 
-    and (4) MCMC tuning parameters.
-  }
-  \label{table:parameter_values}
-\end{table}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Figures}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{figure}
-  \centering
-  \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=4cm, semithick]   
-  \tikzstyle{every state}=[]
-  \node[state] (A)              {Good};   
-  \node[state] (B) [right of=A] {Incipient};   
-  \node[state] (C) [below of=B] {Extinct};   
-  \path (A) edge [bend right] node {$b_g$} (B)
-        (A) edge [bend right] node {$\mu_g$} (C)
-        (B) edge [loop above] node {$b_i$} (B)
-        (B) edge [bend right] node {$\lambda$} (A)
-        (B) edge [bend right] node {$\mu_i$} (C); 
-  \end{tikzpicture}
-
-  \caption{
-    The states and transitions of a species in the PBD model.
-    $b_i$: speciation-initiation rate of incipient species. 
-    $b_g$: speciation-initiation rate of good species. 
-    $\lambda$: speciation completion rate. 
-    $\mu_i$: extinction rate of incipient species. 
-    $\mu_g$: extinction rate of good species. 
-    Figure after Etienne et al, 2014, Evolution
-  }
-  \label{fig:pbd_states}
-\end{figure}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{figure}
-  \centering 
-  \resizebox {0.8\textwidth} {!} {
-    \CreateTikzFigureExperiment{} 
-  }
-  \caption{
-    Overview of the experiment: 
-    A: per parameter set, there are $n_i$ incipient species trees created. 
-    B1: constructing a species trees by using the youngest subspecies $A1$ to represent good species $A$.
-    B2: constructing a species trees by using the oldest subspecies $A2$ to represent good species $A$.
-    C1: constructing one of the $n_a$ alignments of B1. 
-    C2: constructing one of the $n_a$ alignments of B2. 
-    D1: reconstructing a species trees from alignment B1.
-    D2: reconstructing a species trees from alignment B2.
-    Comparisons are done between a species tree and its reconstructed species tree, 
-    for example, between B1 and D1.
-    All phylogenies are drawn at the same scale.
-  }
-  \label{fig:experiment}
-\end{figure}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{figure}
-  \centering 
-  \CreateTikzFigureSampling{}
-  \caption{
-    Sampling an incipient species tree. 
-    A) At stem age $t_0$, $A1$ is present.
-    At crown age $t_1$, its derived incipient species $A2$ is born. 
-    At $t_2$, $A2$ itself speciates into a new (unnamed) species, 
-    that becomes good species $B$ at $t_3$. At present time, $t_4$, there
-    exist two possible representatives of species $A$. 
-    B) Using the oldest representative, $A1$, to represent species $A$, the crown
-    age of the resulting species tree lies at $t_1$. 
-    C) Using the youngest representative, $A2$, to represent species $A$, the crown
-    age of the resulting species tree lies at $t_2$. 
-  }
-  \label{fig:sampling}
-\end{figure}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\listoftodos
-
-\end{document}

### 2018-03-26 obsolete/introduction.tex
Merged to two files

--- 2018-03-26 article.tex
@@
-\documentclass{article}
-
-% Bibliography
-\usepackage{natbib}
-\bibpunct{(}{)}{;}{a}{}{;}
-
-% Use 'It was found that A is B (Name 1234)' style
-\setcitestyle{authoryear,open={},close={}}
-
-% Affiliations
-\usepackage{authblk}
-\title{babette: BEAUti 2, BEAST2 and Tracer for R}
-\author[1]{Rich\`el J.C. Bilderbeek}
-\author[1]{Rampal S. Etienne}
-\affil[1]{Groningen Institute for Evolutionary Life Sciences, University of Groningen, Groningen, The Netherlands}
-
-% Load my functions
-\input{functions.tex}
-
-% Use double spacing
-\usepackage{setspace}
-\doublespacing
-
-\usepackage{listings}
-\usepackage{hyperref}
-\usepackage{todonotes}
-\usepackage{verbatim}
-\usepackage{pgf}
-\usepackage{bm}
-
-% TikZ and friends
-\usepackage{tikz}
-\usepackage{tkz-graph}
-\usepackage{pgf}
-\usetikzlibrary{arrows,automata}
-
-% Style of listings
-% From http://r.789695.n4.nabble.com/How-to-nicely-display-R-code-with-the-LaTeX-package-listings-tp4648110.html
-\usepackage{fancyvrb} 
-\definecolor{codegreen}{rgb}{0,0.6,0}
-\definecolor{codegray}{rgb}{0.5,0.5,0.5}
-\definecolor{codepurple}{rgb}{0.58,0,0.82}
-\definecolor{backcolor}{rgb}{0.95,0.95,0.92}
-\lstdefinestyle{mystyle}{
-  language=R,% set programming language
-  basicstyle=\ttfamily\small,% basic font style
-  commentstyle=\color{gray},% comment style
-  % numbers=left,% display line numbers on the left side
-  numberstyle=\scriptsize,% use small line numbers
-  numbersep=10pt,% space between line numbers and code
-  tabsize=2,% sizes of tabs
-  showstringspaces=false,% do not replace spaces in strings by a certain character
-  captionpos=b,% positioning of the caption below
-  breaklines=true,% automatic line breaking
-  escapeinside={(*}{*)},% escaping to LaTeX
-  fancyvrb=true,% verbatim code is typset by listings
-  extendedchars=false,% prohibit extended chars (chars of codes 128--255)
-  alsoletter={.<-},% becomes a letter
-  alsoother={$},% becomes other
-  otherkeywords={!=, ~, $, \&, \%/\%, \%*\%, \%\%, <-, <<-, /},% other keywords
-  deletekeywords={c}% remove keywords 
-}
-\lstset{style=mystyle}
-
-% Adds numbered lines
-\usepackage{lineno}
-\linenumbers
-
-% Use TODO notes
-\usepackage{todonotes}
-
-% Rename 'Abstract' to 'Summary 
-\usepackage[english]{babel}
-\addto{\captionsenglish}{\renewcommand{\abstractname}{Summary}}
-
-\begin{document}
-
-\maketitle
-
-\begin{abstract}
-
-  The construction of phylogenies helps us answer evolutionary biological
-  questions.
-  Our current phylogenetic tools ignore the fact that speciation takes time,
-  which has an effect unknown in phylogeny reconstruction.
-  Here, we simulate true incipient species trees and their corresponding
-  DNA alignments, from which we measure the reconstruction of the phylogeny
-  using a standard birth-death model.
-  We measure the errors that the Bayesian phylogenetic software tool BEAST2
-  gives when recovering simulated phylogenies, for different times-to-speciate,
-  under a range of additional parameter settings.
-  It has been found that branch lengths are consistently and strongly 
-  underestimated for biologically relevant parameters.
-  This research shows that protractedness is a complexity of nature that
-  cannot always be ignored and should be incorporated in our phylogenetic
-  tools.
-
-\end{abstract}
-
-{\bf Keywords:} computational biology, evolution, phylogenetics, BEAST2, R
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Introduction}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
 \section{Introduction}
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\begin{figure}
+  \centering
+  \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=4cm, semithick]   
+  \tikzstyle{every state}=[]
+  \node[state] (A)              {Good};   
+  \node[state] (B) [right of=A] {Incipient};   
+  \node[state] (C) [below of=B] {Extinct};   
+  \path (A) edge [bend right] node {$b_g$} (B)
+        (A) edge [bend right] node {$\mu_g$} (C)
+        (B) edge [loop above] node {$b_i$} (B)
+        (B) edge [bend right] node {$\lambda$} (A)
+        (B) edge [bend right] node {$\mu_i$} (C); 
+  \end{tikzpicture}
+
+  \caption{
+    The states and transitions of a species in the PBD model.
+    $b_i$: speciation-initiation rate of incipient species. 
+    $b_g$: speciation-initiation rate of good species. 
+    $\lambda$: speciation completion rate. 
+    $\mu_i$: extinction rate of incipient species. 
+    $\mu_g$: extinction rate of good species. 
+    Figure after Etienne et al, 2014, Evolution
+  }
+  \label{fig:pbd_states}
+\end{figure}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 % Biological mechanism of reproductive isolation
@@
 When setting that parameter to infinity, incipient species become good
-species instantaneous and the model falls back to a constant-rate (non-protracted) 
-birth-death model \todo{Add reference}.
+species instantaneous and the model falls back to a constant-rate (non-protracted) birth-death model \todo{Add reference}.
 
@@
 This study shows that ... \todo{put result here}.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Methods}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-This research measures the error our current BD tools make, when
-the real process has been protracted. Figure \ref{fig:experiment} shows
-the following overview of the experiment graphically.
- An in silico data set is created for a wide
-range of parameters from which incipient species trees are simulated (A in figure \ref{fig:experiment}). 
-These phylogenies serve as 'the truth' or 'nature'. From that incipient species tree,
-species trees are created, by picking one representive per species.
-This is done twice, as either the youngest (B1) or oldest (B2) subspecies is chosen
-to represent a species. Per species tree, $n_a$ alignments are simulated (C1 and C2).
-Per alignment, $n_b$ posteriors are inferred, 
-each consisting of $n_s$ phylogenies and parameter estimates (D1 and D2). Each
-species tree is then compared with the species tree its alignment is based on (for
-example, B1 and D1).
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Parameter space}
-\label{subsec:parameter_space}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\todo{@RSE: would you prefer $t_{\mean{ds}}$ or $\mean{t_{ds}}$ for average duration of speciation?}
-
-% Biological parameters
-
-This study primarily investigates combinations of the five biological 
-parameters ($b_g$, $b_g$, $\lambda$, $\mu_g$ and $\mu_g$, 
-see also figure \ref{fig:pbd_states} and table \ref{table:parameter_descriptions}), 
-in a close to full factorial fashion, with two exceptions.
-First, following \cite{etienne2014estimating}, birth and extinction rates were each 
-set equal between good and incipient species ($b_g = b_i$, and $\mu_g = \mu_i$).
-Second, only parameter combinations with speciation initiation rates greater
-than extinction rates were used ($b_g > \mu_g$). The values 
-used in the experiment (see table \ref{table:parameter_values}) 
-were selected to have a reasonable expected number of incipient 
-lineages, $E(n_{ti})$, where \todo{?too basic to have a reference to? I assume Kendall, 1948, \cite{kendall1948generalized}, but could not look it up here and now}:
-
-\begin{equation}
-  E(n_{ti}) = e^{ (b - \mu) \cdot t_c }
-\end{equation}
-
-One value of speciation completion rate $\lambda$ is set (close to) infinity, as
-then the protracted speciation model falls back to a constant-rate 
-(non-protracted) birth-death model [?Etienne et al., 2012 or already Rosindell and Etienne 2011?].
-
-% Biological tuning parameters
-
-Secondarily, there are three biological tuning parameters. The crown age of
-the incipient species tree, $t_c$, is set to 15 million years for all simulations, 
-following \cite{etienne2014estimating}. The per-nucleotide mutation rate, $r$, and the
-DNA aligment length, $l_a$, together must be tuned to result in informative
-DNA alignments (see section \ref{subsec:simulating_alignments}). $r$ is identical in all
-simulations. Increasing $l_a$ results in more informative alignments, and this
-effect is measured by using two values of $l_a$ 
-per biological parameter combination.
-
-% Replication parameters
-
-All processes involving stochasticity have replicates. 
-For each set of parameters, $n_i$ incipient species trees are simulated. 
-From each species tree, $n_a$ alignments are simulated. From each alignment, 
-$n_b$ posteriors are generated.
-
-% MCMC tuning parameters
-
-The posteriors are generated using BEAST2, which uses a Monte-Carlo Markov-Chain (MCMC)
-algorithm. This algorithm is tuned to result in a high effective sample size (ESS) by
-two constants. The first constant is the MCMC sampling interval, which is $i_s$ states.
-$i_s$ which must be high enough to avoid correlation between subsequently sampled posterior states.  
-The second constant is the number of MCMC state samples, $n_s$, which equals the ESS iff \todo{@RSE: Can I use iff?}
-$i_s$ is high enough to have no correlation between subsequently sampled posterior states.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Simulating incipient species trees}
-\label{subsec:simulating_incipient_species_trees}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-From each parameter set, $n_i$ incipient species trees with a crown age
-of $t_c$ and at least two extant taxa, 
-are created in R \cite{R} using the PBD package \cite{PBD}.
-The values of the biological parameters encourage that the number of tips in
-each phylogeny is kept within computational limits. However, due to
-the stochastic nature of the Doob-Gillespie algorithm used in the creation
-of these phylogenies, phylogenies can, by chance, become very big. 
-There is no maximum number of taxa set, yet imposed instead by the R 
-programming language itself, due to a limit on the amount of memory that can be 
-allocated. The incipient species trees that could not be simulated are
-counted and discarded from further analysis.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Sampling species trees}
-\label{subsec:sampling_species_trees}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-Per incipient species tree, two species trees are created. 
-To create a species tree (a tree in which each species is represented once)
-for an incipient species tree, one representative for each species is picked. 
-If a species has initiated one or more incipient species, either the youngest
-or the oldest representative is picked. See figure \ref{fig:sampling} for an example.
-
-An alternative sampling method would be to pick 
-random representatives of each species. This would decrease the variability
-between the two species tree. Maximizing variability between the two
-species trees allows for clearer measuments on the effect of sampling.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Simulating alignments}
-\label{subsec:simulating_alignments}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-From each species tree, $n_a$ DNA alignments were simulated that match the
-phylogeny's structure, using the phangorn R package \cite{phangorn}. 
-Each alignment contains one DNA sequence per species 
-of $l_a$ nucleotides. The alignment follows the Jukes-Cantor nucleotide 
-subsititution model \cite{cantor1969mammalian}, 
-with per-nucleotide rate of mutation, $r$.
-
-The Jukes-Cantor
-nucleotide substitution model is regarded as an oversimplification of 
-nature \cite{unknown}. In this in silico experiment, however, we simulate first
-a DNA alignment from a phylogeny using nucleotide substitution model $M_N$, 
-then infer a phylogeny from that DNA alignment, assuming nucleotide substitution model $M_N$.
-Too reduce unnecessary complexity, $M_N$ is chosen to be the simplest possible.
-
-The combination of $l_a$ and $r$ is tuned to result in informative alignments. 
-Within a maximally informative alignment, each species has its own unique DNA sequence.
-Additionally, the Jukes-Cantor distance \cite{unknown} between any 
-two sequences must be calculable, which is when the proportion of matching nucleotides is below 75\%.
-The tuning was done using an $l_a$ of $10^3$ nucleotides, 
-as this appears to be the current median of the sequenced genome lengths, as currently published.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Simulation of posteriors}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-From each DNA alignment, $n_b$ posteriors of phylogenies and parameter combinations 
-is constructed, using BEAST2, a Bayesian phylogenetic tool \cite{bouckaert2014beast}.
-The nucleotide substitution model is set to 
-be Jukes-Cantor (one nucleotide substition rate category, no invariant nucleotides) 
-and a strict molecular clock, which matches the procedure the DNA alignments are generated, 
-as described in section \ref{subsec:simulating_alignments}).
-The tree prior is set to the (non-protracted) birth-death model, 
-with an uniform prior for birth-rate of range $[0, 10^5]$ and 
-a uniform death-per-birth rate prior of range $[0, 1]$. 
-The MCMC creation parameters were set to have $n_s$ states, sampled at an interval of $i_s$ states.
-
-An MCMC must have an effective sample size (ESS) of at least 200 to result in a 
-posterior that is 'adequate' \cite{drummond2015bayesian}. Due to the multiple 
-sources of stochasticity in the preceding steps, this ESS may not always be achieved.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Analysis}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-To quantify the error, each species tree is compared with the posterior derived
-from its $n_a$ alignments (for example, B1 and D1 in figure \ref{fig:experiment}).
-This is by calculating the nLTT statistic \cite{janzen2015approximate}
-between the species tree and each of its posterior's trees (using 
-the nLTT package \cite{nLTT}), resulting in an nLTT statistic distribution.
-
-% BD trees
-
-The nLTT statistic distribution for all simulations with a high speciation 
-completion rate is visualized for the different parameter values. This allows
-to see the baseline error. Instead of using a complete
-nLTT statistic distribution, also the mean value of each posterior is 
-sometimes used to reduce variance in this.
-
-The effect of speciation completion time $\lambda$ is visualized, by showing
-nLTT statistic distributions for the different parameter values, for all values
-of $\lambda$.
-
-Instead of using the parameter value of $\lambda$ as a proxy for protractedness,
-the speciation completion time, $t_{sc}$, and average duration of speciation, $t_{\mean{ds}}$, 
-are used. Speciation completion time, $t_{sc}$, is simply the inverse of the
-speciation completion rate $\lambda$:
-
-\begin{equation}
-  t_{sc} = \frac{1}{\lambda}
-  \label{eq:speciation_completion_time}
-\end{equation}
-
-The speciation completion time differs from the observed average speciation time, 
-as incipient species may go extinct and may produce other incipient species 
-that complete speciation before the original incipient one. The average duration 
-of speciation, $t_{\mean{ds}}$, takes this into 
-account \todo{check formula ?19 of?20 in Etienne et al., 2012, then cite}:
-
-\begin{equation}
-    t_{\mean{ds}} = \mean{\rho} = \frac{2}{ (D - \lambda + b_i - \mu_i) \cdot \ln{ \frac{2}{(1 + (\lambda - b_i + \mu_i)/D)}  } }
-\end{equation}
-
-where $D$ is defined as:
-
-\begin{equation}
-    D = \sqrt{ (\lambda + b_i)^2 + 2 \cdot (\lambda - b_i) \cdot \mu_i + \mu_i^2 }
-\end{equation}
-
-For each simulation, the expected $t_{\mean{ds}}$ is calculated from its parameter 
-values (using the PBD package \cite{PBD}), after which the nLTT statistic 
-distributed is plotted, to see the effect of protractedness on the nLTT statistic
-distribution.
-
-As not all simulations result in posteriors with an adequate ESS, the
-effect that a below-median or above-median ESS has on the nLTT statistic
-distribution is displayed.
-
-To view the effect of alignment length, $l_a$, the nLTT statistic
-distributions are plotted side-by-side for all other parameter values.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Peripherals}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-Simulating incipient species trees, sampling species trees, simulating alignments
-and creating posteriors, are done with the wiritttes package \cite{wiritttes}. The
-analysis is done with the wiritttea package \cite{wiritttea}. Those packages rely on 
-functionality in the APE \cite{APE}, geiger \cite{GEIGER}, ggplot2 \cite{ggplot2}, 
-RBeast \cite{RBeast}. Quality of these packages is assured by 
-the covr \cite{covr}, devtools \cite{devtools}, 
-goodpractice \cite{goodpractice}, lintr \cite{lintr}, testit \cite{testit} 
-and testthat \cite{testthat} packages, using the Travis CI \cite{travis} continuous integration service. 
-Both packages are hosted at GitHub \cite{github}, a good practice for computational
-scientists \cite{perez2016ten} and helps improve transparency \cite{gorgolewski2016practical}.
-Code coverage, which correlates with code quality \cite{del1995correlation}, 
-is above 90\% for both packages.
-
-The simulated dataset can be downloaded from DataDyrad [but not yet]. The
-summarized dataset can be downloaded at \cite{wirittte_data}. The
-pre-publication of this article can be downloaded at \cite{wirittte_article}.
-
-Simulations and analyis was mostly performed on the Peregrine 
-high performance computing cluster of the University of Groningen.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Results}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Creation of data set}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\todo{First focus on the real results, not their quality}
-
-% Simulation of incipient species trees
-
-\todo{redo with new data}
-
-\todo{create figure or table that shows this}
-
-957 out of 960 incipient species trees have been simulated successfully. 
-The 3 failed simulations created data files bigger than R can read, 
-and have been excluded from further analysis.
-
-The number of good species ranged from 2 to 14296, with a median of 87 and an average of 503.8. 
-There were 6 phylogenies with only 2 good species.
-
-% Sampling species trees}
-
-\todo{redo with new data}
-
-For each incipient species trees, two species trees were constructed. 
-In 653 out of 960 cases, these trees had different branch lengths.
-In 302 cases, the trees were identical. In the remaining 5 cases,
-there were computational problems (see figure \ref{fig:exit_statuses}).
-
-% Simulation of alignments
-
-\todo{redo with new data}
-
-For each species tree, $n_a$ alignments have been created, 
-resulting in $960 \cdot 2 \cdot 2 = 3840$ alignments. 
-$?$ of these were informative.
-
-\todo{May result in an overrepresentation of the two same species trees derived from one incipient species tree}
-
-% Simulation of posteriors}
-
-A BEAST2 posterior contains multiple parameter estimates (see figure \ref{fig:ess_distribution} for all),
-and each one has an ESS. An ESS of 200 is acceptable for any parameter estimate.
-The ESS of the tree likelihood should have such a decent ESS. Figure \ref{fig:ess_distribution_likelihood}
-shows the distribution of the likelihood ESS.
-
-\todo{How tp quote https://www.beast2.org/what-is-ess/ ? It states that not all variables
-need to have a high ESS, but 'The likelihoods (both of the tree and coalescent model) should have decent ESSs.}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Quantification of the error BEAST2 makes on all trees}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-To get a first global impression of the error BEAST2 makes, figure \ref{fig:error}
-shows a histogram of the nLTT statistics, colored by the expected mean
-duration of speciation. As BEAST2 assumes a mean duration of speciation of zero,
-it is expected that the nLTT statistic (a measure of error) is lowest. Zooming
-in on the left part of the distribution, there is no clean pattern.
-The right part of the tail of distribution shows that the biggest nLTT
-statistic values are found for high (but not highest) expected mean durations of
-speciation.
-
-\todo{figure \ref{fig:error_for_scr} shows the distribution of errors for all
-combinations of PBD parameters}
-
-As the expected correlation appears to be absent, figure 
-\ref{fig:error_expected_mean_dur_spec}, subfigure $a$, 
-shows the correlation between expected
-mean duration of speciation and the nLTT statistic. A linear fit shows
-that there is indeed an increase in error for an increases expected duration
-of speciation. A trendline using locally weighted smoothing, however,
-indicates a more complex relationship. Note that this figure only uses
-a subset of all measurements, consisting of $n_s$ nLTT statistic values 
-per posterior.
-
-Taking the average nLTT value per posterior frees us from computational
-constraints and allows us to use the full data set. The resulting
-plot, figure \ref{fig:error_expected_mean_dur_spec}, subfigure $b$, shows the same pattern:
-A linear fit shows
-that there is indeed an increase in error for an increases expected duration
-of speciation. But now, a trendline using locally weighted smoothing 
-follows this linear relationship closely.
-
-The inference error made is expected to be less for a longer DNA
-alignment length. 
-Figure \ref{fig:error_expected_mean_dur_spec_alignment_length} 
-show the correlation between expected mean duration of speciation 
-and the nLTT statistic for the two different DNA alignment lengths. 
-The first subfigure uses a sample of all $n_s$ individuals nLTT statistics,
-where the second subfigure takes only the posterior's 
-nLTT statistic mean value, for the complete dataset.
-
-Both figures show again that there is an 
-increase in error for an increases expected duration
-of speciation, when assuming a linear relation, and
-a more complex relationship when using locally weighted smoothing. 
-The figures also show that, as expected, 
-the error is less with more information. Additionally,
-a longer DNA alignment length results in a slower accumulation of error
-for increased expected mean duration of speciation.
-
-A higher ESS for longer DNA alignments may be an additional
-explanation for the observed slower accumulation of error
-for increased expected mean duration of speciation.
-Figure \ref{fig:ess_expected_mean_dur_spec_alignment_length}
-shows that longer DNA alignments have a higher likelihood ESS, 
-but this trend is weaker than in figure 
-\ref{fig:error_expected_mean_dur_spec_alignment_length} and
-the 95\% confidence intervals of the linearily fitted trendlines only
-stop overlapping for the highest expected mean duration of speciation.
-The locally weighted smoothing trendlines show a different trend,
-suggesting that alignment length has no effect on the
-likelihood ESS.
-
-\todo{Figure \ref{fig:error_alignment_length} shows the same pattern, seperated
-for all combinations of PBD parameters.}
-
-For each incipient species tree, two species trees were created, by taking
-either the youngest or oldest subspecies to represent each species.
-The effect of this is shown in figure \ref{fig:error_expected_mean_dur_spec_sampling}.
-Both subfigures show the
-error made for different $t_{\mean{ds}}$, where the first subfigure uses a sample
-of all $\Delta_{nLTT}$ values, where the second subfigure uses the $\mean{\Delta_{nLTT}}$
-of each posterior. In both subfigures, the linear and locally weighted smoothing
-have overlapping 95\% confidence intervals, hinting that there is no strong effect
-of the way subspecies are selected to represent their species.
-
-\note{Figure \ref{fig:error_sampling_representative} shows the same data, ordered
-by the different PBD parameters instead}
-
-The effect of having a low or high tree likelihood ESS is shown in 
-figure \ref{fig:figure_error_expected_mean_dur_spec_low_high_ess}. We
-define a low ESS as being below median, where a high ESS is defined as 
-above-median. Plotting (a sample of or the means of) the nLTT statistic for different mean
-duration of speciation for a low an high ESS, shows that there is no clear
-difference in error made for posteriors with either low or high tree likelihood ESSes.
-
-It is expected that the bigger phylogeny is simulated, the less the error
-will be. Figure \ref{fig:figure_error_tree_size} shows the relation between the
-number of taxa and the mean nLTT statistic. Although most simulated
-phylogenies have less than 250 taxa, there is a clear positive trend, 
-which is the opposite of the
-expactation: the bigger the phylogeny, the bigger the error made.
-As the nLTT statistic normalized for the number of taxa, some other
-mechanism must be involved.
-
-For each species tree, $n_a$ different alignments are simulated. For each of
-these alignments, $n_b$ posteriors are simulated, using the same alignment. 
-It is assumed that two posteriors from the same alignment
-give rise to the same parameter estimates distributions.
-distribution. Figure \ref{fig:figure_posterior_distributions_likelihood} 
-shows the distribution of Mann-Whitney U test p-values for each pair of 
-posteriors working on a same alignment, 
-using their tree likelihood estimates.
-
-The figure shows that the tree likelihood posterior distributions 
-from a same alignment nearly always
-are significantly different. Subfigure (b) and (c) show the distribution of
-tree likelihood estimates for a posterior pair with the lowest (b) and 
-highest (c) p-value.
-
-Posteriors based on the same alignment are also (next to estimated tree likelihood) 
-assumed to have a similar nLTT statistic distribution. 
-Figure \ref{fig:figure_posterior_distribution_nltt} 
-shows the distribution of Mann-Whitney U test p-values for each pair of 
-posteriors working on a same alignment, 
-using their nLTT statistics.
-
-The figure shows that the nLTT statistic distributions 
-from a same alignment are commonly different and commonly the same. 
-Subfigures $(b)$ and $(c)$ show the distribution of
-nLTT statistics for a posterior pair with the lowest $(b)$ and 
-highest $(c)$ p-value.
-
-The nLTT statistic normalizes a phylogeny to have a crown age of 1.
-To prevent overlooking patterns in posterior crown ages, 
-figure \ref{fig:figure_posterior_distribution_crown_ages}
-shows the distribution of crown ages of all simulations.
-Almost all simulations have a crown age close to $0.15$, but there
-are rare cases with crown ages above $0.2$, ranging to $83.4$.
-
-\todo{I think it is very suspicious that the range centers on 0.15, when
-the crown age is 15. Check this!}
-
-Again, it is assumed that posteriors based on the exact same alignment
-have a similar crown age distribution. Of all posterior pairs, the Mann-Whitney
-U test p-values are shown in figure \ref{fig:figure_posterior_distribution_crown_ages_p_values}.
-It shows that crown age distributions being different is common and these
-being the same is uncommon.
-
-It is expected that if a tree is created by a Birth-Death process, BEAST2 is able
-to reasonably estimate those parameters.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Discussion}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-This article has quantified the error made in inference when (1) nature would follow the
-protracted speciation model, and (2) her DNA sequences would be analysed with a birth-death
-model.
-
-% The PBD model is just one implementation of speciation taking time
-
-The key feature of the PBD model is that it encompasses speciation taking time in the
-simplest way possible (that allows for mathematical analysis). The PBD model allows for
-multiple incipient species stages (where its earliest version had infinite
-stages \cite{rosindell2010protracted}). This research assumes that there is only one
-incipient species stage. One could argue that speciation takes place in more
-biological stages \todo{Add reference}, but we tried to use the simplest incarnation of the PBD possible.
-Also, the PBD model can also be viewed as a model of species recognition \todo{?Purvis et al., 2009},
-in which (human) recognition needs to build up, which we consider to be sufficiently
-encapsulated in one single stage. Future research may investigate if adding multiple incipient
-species stages for simulating nature changes the error made by (non-protracted) birth-death models.
-
-% Equal and constant rates
-
-The PBD model allows for multiple incipient species stages. 
-Each stage has its own extinction and speciation-initiation and 
-speciation-completion rate (except for good species, which lacks 
-a speciation-completion rate). These rates are assumed to be constant
-in time and equal for all stages. We chose to use the constant-rate
-birth-death model to keep this investigation as simple as possible.
-Future research may investigate whether a biologically more realistic
-rate model, for example, by adding diversity-dependence, results in
-the same inference error.
-
-We chose to set the extinction and speciation-initiation rates equal for all stages,
-again for choosing the simplest model. We expect these rates to differ in reality:
-incipient species have low initial abundances, which has an effect on extinction
-and speciation-initiation rates.
-
-In this research, we nevertheless set these rates equal. 
-Not only because it helps constrain the parameter space investigated,
-also paraphylies would pop up in the simulated trees. This would violate more assumptions
-of the BEAST2 model used as a prior and increase the complexity of the analysis.
-
-% Sampling species trees from incipient species trees
-
-An incipient species tree may have multiple incipient lineages from the
-same ancestor. We constructed two species trees from each incipient species
-tree, by selecting the youngest and oldest subspecies to represent a species.
-This selection maximizes the chance the species trees are different, where
-random sampling of representatives would also be a valid approach, closer
-to what would happen when scientists take measurements in the field. Our
-approach increases the error made.
-
-% Nucleotide substitution model
-
-For the simulation of DNA alignments, the Jukes-Cantor nucleotide substitution model
-is used. This substitution model is a simplification of the DNA mutation processes
-happening in reality \todo{Add reference}. In this simulation research, however, we could also
-state in BEAST2 that this nucleotide substitution model is used. Thus, we kept
-the simulation of the nucleotides in time as simple as possible, and used that
-knowledge in later steps. We expect that any nucleotide substitution model, when
-used in both alignment generation and in phylogeny inference, will do equally well,
-except that the full computational pipeline would take needlessly longer.
-
-% Allopatric speciation only
-
-When a speciation event is initiated, this investigation assumes no gene flow
-between the two incipient lineages. This follows the allopatric mode of speciation, 
-which is assumed to be most prominent of nature \todo{Add reference}. It would be interesting to
-allow gene flow between related incipient species, 
-so also peripatric and sympatric modes of speciation would be allowed for.
-We predict that this would increase the error in inference.
-
-% Why not put the tree prior into an MCMC tool directly?
-One could argue we should have written a PBD tree prior for an MCMC phylogenetic tool.
-We did not do that for two reasons. First, we think it is rational to first
-measure if there is need for a more complex tree prior. Second, 
-we could not have done so, as a phylogenetic tool like BEAST2 assumes species
-trees, where our input consists of incipient species trees. Before this
-research, we were unsure how to sample species representatives. This research
-has shown that any sampling is just as good, opening up avenues for
-implementing a PBD species tree prior.
-
-% BEAST2 is just one of many tools
-
-BEAST2 is one of many MCMC phylogenetic tools with its
-own internal technicalities. Similar tools are the R phytools
-package \cite{phytools}, BAli-Phy \cite{suchard2006bali}
-and BayesPhylogenies \cite{pagel2007bayesphylogenies}, to name a few.
-BEAST2 was chosen, because (1) it's FOSS, (2) its software
-architecture is modular, encouraging extension, (3) its development
-team follows good practices in software development.
-
-% BEAST2 parameter settings: strict clock
-
-\todo{elaborate}
-
-Use of strict clock is clean, but unrealistic.
-
-\todo{elaborate}
-
-% Alignment generation
-
-Use of JC69 is rational, but unrealistic.
-
-% Use for empiricists
-
-\todo{elaborate}
-
-This reseach shows that a DNA sequence length of 10k [is good].
-[must be effective DNA sequence length]
-[no indels generated]
-
-% Low number of incipient species trees
-
-\todo{elaborate}
-
-For each point in parameter space, twenty incipient species tree are created.
-Was fitted to 1000 jobs.
-
-% Failed simulations
-
-Not all parameter combinations resulted in posteriors with a high ESS,
-due to computational constraints. We think this is a valid approach:
-would a phylogeny get too big, we as scientists would cut it to
-pieces that are analytically tractable. Because these simulations
-are only a minority (as shown in the Results section), we do not
-expect this to influence our conclusions.
-
-% nLTT statistic is just one of many summary statistics
-
-We used the nLTT statistic for quantifying the error between the
-true and inferred phylogeny. There are simpler statistics that
-can do the same, like the gamma \cite{pagel1999inferring}, 
-delta-r \cite{pigot2010shape} \todo{check citation}, 
-tree balance \todo{Add reference} or
-Normalized Rooted Branch Score [Drummond] statistic, 
-or more complex, like the spectral analysis [Hendy and Penny, 1993].
-We consider those other summary statistics equally good, we just picked the
-one we thought of the most informative per unit of added computational time.
-
-% Add:
-%
-% Computational Performance and Statistical Accuracy of *BEAST and Comparisons with Other Methods (by Huw A. Ogilvie Joseph Heled Dong Xie Alexei J. Drummond)
-% cites rooted branch score, RBS, as being from Heled and Bouckaert 2013, `Looking for trees in the forest: Summary tree from posterior samples`
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Acknowledgements}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-We would like to thank the Center for Information Technology of the University of Groningen for their support
-and for providing access to the Peregrine high performance computing cluster.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Authors' contributions}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-RJCB and RSE conceived the idea for the package. RJCB created and tested the package, and wrote the first draft of the manuscript. RSE contributed substantially to revisions.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Bibliography
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% MEE style
-\bibliographystyle{mee}
-\bibliography{article}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Tables}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{table}
-  \centering 
-  \begin{tabular}{l l}
-    \hline
-    Symbol & Description \\
-    \hline
-    \hline
-    $b_g$ & Speciation initiation rate of a good species (probability per lineage, per million year) \\
-    $b_i$ & Speciation initiation rate of an incipient species (probability per lineage, per million year) \\
-    $\lambda$ & Speciation completion rate (probability per lineage, per million year) \\
-    $\mu_g$ & Extinction rate of a good species (probability per lineage, per million year) \\
-    $\mu_i$ & Extinction rate of an incipient species (probability per lineage, per million year) \\
-    \hline
-    $t_c$ & Crown age (million years) \\
-    $r$ & Mutation rate (probability of mutation per nucleotide per million years) \\
-    $l_a$ & DNA alignment length (base pairs) \\
-    \hline
-    $n_i$ & Number of incipient species trees per parameter set \\
-    $n_a$ & Number of alignments per species tree \\
-    $n_b$ & Number of BEAST2 runs per alignment \\
-    \hline
-    $n_s$ & Number of MCMC samples \\
-    $i_s$ & MCMC sampling interval \\
-    \hline
-    $n_{ti}$ & Number of taxa in an incipient species tree \\
-    $n_{t}$ & Number of taxa in a (good) species tree \\
-    $t_{\mean{ds}}$ & Average duration of speciation \\
-    $t_{sc}$ & Speciation completion time \\
-    $\Delta_{nLTT}$ & nLTT statistic value \\
-    $\mean{\Delta_{nLTT}}$ & Average nLTT statistic value of a posterior \\
-    $\median{\Delta_{nLTT}}$ & Median nLTT statistic value of a posterior \\
-    \hline
-  \end{tabular}
-  \caption{
-    Parameter descriptions. The sections are (1) the biological parameters, (2) the
-    biological tuning parameters, (3) parameters involved in replication, (4) MCMC
-    tuning parameters, and (5) other symbols.
-  }
-  \label{table:parameter_descriptions}
-\end{table}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{table}
-  \centering 
-  \begin{tabular}{l l}
-    \hline
-    Parameter             & Values \\
-    \hline
-    \hline
-    $b = b_g = b_i$       & 0.1, 0.5, 1.0 \\
-    $\lambda$             & 0.1, 0.3, 1.0, $10^6$ \\
-    $\mu = \mu_g = \mu_i$ & 0.0, 0.1, 0.2, 0.4 \\
-    \hline
-    $t_c$                 & 15 \\
-    $r$                   & 0.01 \\
-    $l_a$                 & $10^3$, $10^4$ \\
-    \hline
-    $n_i$                 & 20 \\
-    $n_a$                 & 2 \\
-    $n_b$                 & 2 \\
-    \hline
-    $n_s$                 & $10^3$ \\
-    $i_s$                 & $10^3$ \\
-    \hline
-  \end{tabular}
-  \caption{
-    Parameter values. The sections are (1) the biological parameters, (2) the
-    biological tuning parameters, (3) parameters involved in replication, 
-    and (4) MCMC tuning parameters.
-  }
-  \label{table:parameter_values}
-\end{table}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Figures}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{figure}
-  \centering
-  \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=4cm, semithick]   
-  \tikzstyle{every state}=[]
-  \node[state] (A)              {Good};   
-  \node[state] (B) [right of=A] {Incipient};   
-  \node[state] (C) [below of=B] {Extinct};   
-  \path (A) edge [bend right] node {$b_g$} (B)
-        (A) edge [bend right] node {$\mu_g$} (C)
-        (B) edge [loop above] node {$b_i$} (B)
-        (B) edge [bend right] node {$\lambda$} (A)
-        (B) edge [bend right] node {$\mu_i$} (C); 
-  \end{tikzpicture}
-
-  \caption{
-    The states and transitions of a species in the PBD model.
-    $b_i$: speciation-initiation rate of incipient species. 
-    $b_g$: speciation-initiation rate of good species. 
-    $\lambda$: speciation completion rate. 
-    $\mu_i$: extinction rate of incipient species. 
-    $\mu_g$: extinction rate of good species. 
-    Figure after Etienne et al, 2014, Evolution
-  }
-  \label{fig:pbd_states}
-\end{figure}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{figure}
-  \centering 
-  \resizebox {0.8\textwidth} {!} {
-    \CreateTikzFigureExperiment{} 
-  }
-  \caption{
-    Overview of the experiment: 
-    A: per parameter set, there are $n_i$ incipient species trees created. 
-    B1: constructing a species trees by using the youngest subspecies $A1$ to represent good species $A$.
-    B2: constructing a species trees by using the oldest subspecies $A2$ to represent good species $A$.
-    C1: constructing one of the $n_a$ alignments of B1. 
-    C2: constructing one of the $n_a$ alignments of B2. 
-    D1: reconstructing a species trees from alignment B1.
-    D2: reconstructing a species trees from alignment B2.
-    Comparisons are done between a species tree and its reconstructed species tree, 
-    for example, between B1 and D1.
-    All phylogenies are drawn at the same scale.
-  }
-  \label{fig:experiment}
-\end{figure}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{figure}
-  \centering 
-  \CreateTikzFigureSampling{}
-  \caption{
-    Sampling an incipient species tree. 
-    A) At stem age $t_0$, $A1$ is present.
-    At crown age $t_1$, its derived incipient species $A2$ is born. 
-    At $t_2$, $A2$ itself speciates into a new (unnamed) species, 
-    that becomes good species $B$ at $t_3$. At present time, $t_4$, there
-    exist two possible representatives of species $A$. 
-    B) Using the oldest representative, $A1$, to represent species $A$, the crown
-    age of the resulting species tree lies at $t_1$. 
-    C) Using the youngest representative, $A2$, to represent species $A$, the crown
-    age of the resulting species tree lies at $t_2$. 
-  }
-  \label{fig:sampling}
-\end{figure}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\listoftodos
-
-\end{document}

### 2018-03-29 20180326/appendix.tex
Start from scratch

\section{Appendix}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Other representations of the data in the Results section}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}
  %\includesvg[width=\textwidth]{figure_error_for_scr}
  \caption{
    Mean nLTT statistic distribution (each boxplot) 
    for different speciation completion rates (x axis boxplot)
    for different extinction rates (columns)
    and different speciation initiation rates (rows). 
  }
  \label{fig:error_for_scr}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[!htbp]
  \subfloat[]{
    %\includesvg[width=0.5\textwidth]{figure_error_sampling_representative}
  }
  \subfloat[]{
    %\includesvg[width=0.5\textwidth]{figure_error_sampling_representative_zoom}
  }

  \subfloat[]{
    %\includesvg[width=0.5\textwidth]{figure_error_sampling_representative_mean}
  }
  \subfloat[]{
    %\includesvg[width=0.5\textwidth]{figure_error_sampling_representative_mean_zoom}
  }

  \caption{
    The effect of sampling the youngest or oldest subspecies to represent a species
    on a sample of (subfigures a and b) or the mean (subfigures c and d) 
    nLTT statistic distributions (each boxplot) 
    for different speciation completion rates (x axis boxplot)
    when sampling either the youngest or oldest incipient species as a representative (fill of boxplot),
    for different extinction rates (columns)
    and different speciation initiation rates (rows). 
    The subfigures a and c show the full distribution, where subfigures
    b and d show only the values below the horizontal dotted line.
    See also figure \ref{fig:error_expected_mean_dur_spec_sampling}
  }
  \label{fig:error_sampling_representative}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}
  %\includesvg[width=\textwidth]{figure_error_low_high_ess}
  \caption{
    The effect of a low and high ESS
    on the mean nLTT statistic distribution (each boxplot) 
    for different speciation completion rates (x axis boxplot)
    for either low or high ESS (fill of boxplot),
    for different extinction rates (columns)
    and different speciation initiation rates (rows). 
  }
  \label{fig:error_low_high_ess}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}
  %\includesvg[width=\textwidth]{figure_error_alignment_length}
  \caption{
    The effect of alignment length
    on the nLTT statistic distribution (each boxplot) 
    for different speciation completion rates (x axis boxplot)
    for different DNA alignment lengths (fill of boxplot),
    for different extinction rates (columns)
    and different speciation initiation rates (rows). 
    Bars at median,
    boxes ends at 25 and 75,
    whiskers at 5 and 95
  }
  \label{fig:error_alignment_length}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}
  %\includesvg[width=\textwidth]{figure_error_alignment_length_mean}
  \caption{
    The effect of alignment length
    on the mean nLTT statistic distribution (each boxplot) 
    for different speciation completion rates (x axis boxplot)
    for different DNA alignment lengths (fill of boxplot),
    for different extinction rates (columns)
    and different speciation initiation rates (rows). 
    Bars at median,
    boxes ends at 25 and 75,
    whiskers at 5 and 95
  }
  \label{fig:error_alignment_length_mean}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}
  %\includesvg[width=\textwidth]{figure_error_alignment_length_median}
  \caption{
    The effect of alignment length
    on the median nLTT statistic distribution (each boxplot) 
    for different speciation completion rates (x axis boxplot)
    for different DNA alignment lengths (fill of boxplot),
    for different extinction rates (columns)
    and different speciation initiation rates (rows). 
    Bars at median,
    boxes ends at 25 and 75,
    whiskers at 5 and 95
  }
  \label{fig:error_alignment_length_median}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Diagnostics}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}
  %\includesvg[width=\textwidth]{figure_run_times}
  \caption{
    Distribution of simulation run times. All jobs finishing within 10 minutes are removed
  }
  \label{fig:run_times}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}
  %\includesvg[width=\textwidth]{figure_exit_statuses}
  \caption{
    Exit statuses of simulations
  }
  \label{fig:exit_statuses}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}
  %\includesvg[width=\textwidth]{figure_strees_different}
  \caption{
    Differences in the sampled species trees
  }
  \label{fig:strees_different}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}
  %\includesvg[width=\textwidth]{figure_alignment_qualities}
  \caption{
    Informativeness of alignments. 
    OK: All taxa have a different DNA sequence.
    zero: At least one taxon pair has an identical DNA sequence.
    na: At least one taxon pair has an incalculable distance; 75% or more of the nucleotides differ.
  }
  \label{fig:alignment_qualities}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}
  %\includesvg[width=\textwidth]{figure_exit_statuses_posteriors}
  \caption{
    Exit statuses of the posterior runs [Note: the be removed in final article]
  }
  \label{fig:exit_statuses_posteriors}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}
  %\includesvg[width=\textwidth]{figure_ess_distribution}
  \caption{
    Distribution of ESSes for all estimated parameters. 
    The dotted line indicates an 'acceptable' ESS
  }
  \label{fig:ess_distribution}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}
  %\includesvg[width=\textwidth]{figure_lowest_ess}
  \caption{
    Which parameter estimate has the lowest ESS?
  }
  \label{fig:lowest_ess}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}
  %\includesvg[width=\textwidth]{figure_acceptance_mcmc_operators}
  \caption{
    acceptance and rejection of MCMC operators. The horizontally
    dotted line is the target acceptance rate of 0.234 [BEAST book, page 89, quoting
    Gelman et al., 1996]. See table \ref{table:mcmc_operator_descriptions}
    for a description of these operators.
  }
  \label{fig:acceptance_mcmc_operators}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{table}
  \centering 
  \begin{tabular}{l l}
    \hline
    Name & Description \\
    \hline
    \hline
    BirthRateScaler & Modifies speciation rate $lambda$ \\
    DeathRateScaler & Modifies extinction rate $mu$ \\
    SubtreeSlide & Changes the topology of the tree \\
    treeRootScaler & Changes $t_0$ \\
    treeScaler & Rescales entire tree \\
    \hline
  \end{tabular}
  \caption{
    MCMC operators.
  }
  \label{table:mcmc_operator_descriptions}
\end{table}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2018-03-29 20180326/methods.tex
Start from scratch

--- 2018-03-26 article.tex
@@
-\documentclass{article}
-
-% Bibliography
-\usepackage{natbib}
-\bibpunct{(}{)}{;}{a}{}{;}
-
-% Use 'It was found that A is B (Name 1234)' style
-\setcitestyle{authoryear,open={},close={}}
-
-% Affiliations
-\usepackage{authblk}
-\title{babette: BEAUti 2, BEAST2 and Tracer for R}
-\author[1]{Rich\`el J.C. Bilderbeek}
-\author[1]{Rampal S. Etienne}
-\affil[1]{Groningen Institute for Evolutionary Life Sciences, University of Groningen, Groningen, The Netherlands}
-
-% Load my functions
-\input{functions.tex}
-
-% Use double spacing
-\usepackage{setspace}
-\doublespacing
-
-\usepackage{listings}
-\usepackage{hyperref}
-\usepackage{todonotes}
-\usepackage{verbatim}
-\usepackage{pgf}
-\usepackage{bm}
-
-% TikZ and friends
-\usepackage{tikz}
-\usepackage{tkz-graph}
-\usepackage{pgf}
-\usetikzlibrary{arrows,automata}
-
-% Style of listings
-% From http://r.789695.n4.nabble.com/How-to-nicely-display-R-code-with-the-LaTeX-package-listings-tp4648110.html
-\usepackage{fancyvrb} 
-\definecolor{codegreen}{rgb}{0,0.6,0}
-\definecolor{codegray}{rgb}{0.5,0.5,0.5}
-\definecolor{codepurple}{rgb}{0.58,0,0.82}
-\definecolor{backcolor}{rgb}{0.95,0.95,0.92}
-\lstdefinestyle{mystyle}{
-  language=R,% set programming language
-  basicstyle=\ttfamily\small,% basic font style
-  commentstyle=\color{gray},% comment style
-  % numbers=left,% display line numbers on the left side
-  numberstyle=\scriptsize,% use small line numbers
-  numbersep=10pt,% space between line numbers and code
-  tabsize=2,% sizes of tabs
-  showstringspaces=false,% do not replace spaces in strings by a certain character
-  captionpos=b,% positioning of the caption below
-  breaklines=true,% automatic line breaking
-  escapeinside={(*}{*)},% escaping to LaTeX
-  fancyvrb=true,% verbatim code is typset by listings
-  extendedchars=false,% prohibit extended chars (chars of codes 128--255)
-  alsoletter={.<-},% becomes a letter
-  alsoother={$},% becomes other
-  otherkeywords={!=, ~, $, \&, \%/\%, \%*\%, \%\%, <-, <<-, /},% other keywords
-  deletekeywords={c}% remove keywords 
-}
-\lstset{style=mystyle}
-
-% Adds numbered lines
-\usepackage{lineno}
-\linenumbers
-
-% Use TODO notes
-\usepackage{todonotes}
-
-% Rename 'Abstract' to 'Summary 
-\usepackage[english]{babel}
-\addto{\captionsenglish}{\renewcommand{\abstractname}{Summary}}
-
-\begin{document}
-
-\maketitle
-
-\begin{abstract}
-
-  The construction of phylogenies helps us answer evolutionary biological
-  questions.
-  Our current phylogenetic tools ignore the fact that speciation takes time,
-  which has an effect unknown in phylogeny reconstruction.
-  Here, we simulate true incipient species trees and their corresponding
-  DNA alignments, from which we measure the reconstruction of the phylogeny
-  using a standard birth-death model.
-  We measure the errors that the Bayesian phylogenetic software tool BEAST2
-  gives when recovering simulated phylogenies, for different times-to-speciate,
-  under a range of additional parameter settings.
-  It has been found that branch lengths are consistently and strongly 
-  underestimated for biologically relevant parameters.
-  This research shows that protractedness is a complexity of nature that
-  cannot always be ignored and should be incorporated in our phylogenetic
-  tools.
-
-\end{abstract}
-
-{\bf Keywords:} computational biology, evolution, phylogenetics, BEAST2, R
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Introduction}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{Introduction}
-
-% Biology: speciation takes time
-
-It takes time for new species-to-be to achieve the status of a new species.
-First, it takes time for the new species-to-be 
-to acquire sufficient different mutations to 
-establish reproductive isolation \cite{schluter2001ecology}. Second, 
-it takes time for us to recognize reproductive isolation
-and reclassify that one single species has become two 
-(or more) species.
-
-As an example, \cite{fennessy2016multi} 
-showed that Africa has four instead of only one
-giraffe species, with a common ancestor estimated 
-at around 2 million years ago.
-
-% The use of speciation models
-
-From DNA sequences obtained in the present, we can construct phylogenies
-to infer the evolutionary relationship between species and the time 
-from which ancestral species developed into separate.
-
-As we only sample DNA of extant species, 
-we can only infer a reconstructed tree, in which
-the lineages that have gone extinct are absent.
-Although most models assume extinctions (\cite{yule1925mathematical} being the
-classic exception, as at its time computation was expensive), 
-information on extinction times is lost due to our sampling.
-
-The time when reproductive isolation started to take off, is
-also lost in our sampling method, if the new species-to-be is not 
-recognized yet as such. Without fossil date, one cannot observe the
-possibly many times that reproductive isolation built up, but was lost again.
-
-% First model: constant rate birth death model
-
-Although we know that speciation takes time, we commonly ignore this when
-constructing a phylogeny, by choosing a constant-rate birth-death model
-as a speciation model. This can be justified, as for lineages 
-that have speciated far back in the past, there is no error. 
-Yet in the present, we do not recognize the future species-to-be
-that are already genetically converging, resulting in an underestimation
-of the number of species present today, which can only be 
-concluded in retrospect.
-
-The constant-rate birth-death model (as described in for example \cite{nee1994reconstructed}) 
-is among the simplest speciation models, and assumes a constant speciation
- rate $\lambda$ and constant extinction rate $\mu$.
-
-Additionally, it assumes speciation is instantaneous.
-The constant-rate birth-death model is popular for its simplicity, yet
-has also served as a starting point for more elaborate speciation models.
-
-% Other non-protracted speciation models
-
-Other speciation models may assume that speciation rate changes in 
-time \cite{rabosky2008explosive}, is dependent on the amount of species 
-present \cite{etienne2011diversity}, or is trait 
-dependent \cite{fitzjohn2009estimating}.
-
-In these examples, the original birth-death model has been extended by making 
-speciation rate dependent on time, diversity or trait value respectively.
-The assumption that all these extensions share is that speciation is instantaneous;
-that after a new branching event, both two lineages are (or are recognized as being) different
-species.
-
-% Protracted speciation model
-
-% Good and incipient states
-
-The protracted speciation model \cite{etienne2012prolonging} allows for speciation taking time.
-
-It adds an additional species state (see also figure \ref{fig:pbd_states}),
-coined the 'incipient' stage, which a lineages has to complete before becoming
-a 'good' species.
-
-One view is to say that good species have achieved reproductive isolation,
-where incipient species are in the process of achieving this \todo{Add reference}.
-
-As genetic drift is unlikely to be a relevant speciation mechanism,
-this is more likely to be ecological speciation [Sobel et al., 2009]. 
-\todo{RSE: Ben je het daarmee eens? Zou drift niet een belangrijke ooraak van speciation op islands kunnen zijn bijvoorbeeld?}
-
-Alternatively, an incipient species can be described as a good-species-to-be,
-yet not recognized as such \todo{?[Purvis et al., 2009]?}.
-
-% Biological mechanism of reproductive isolation
-
-The protracted speciation model makes no assumption about the exact biological
-mechanism by which reproductive isolation is 
-acquired \cite{etienne2012prolonging, etienne2014estimating,rosindell2010protracted}.
-
-A 'good' state can be viewed as a state in which there is convergent selection
-for that lineage.
-
-The 'incipient' state can be seen as a state in which there is the potential
-for speciation.
-
-A first example can be from the BDM \todo{Add reference} model, the state in which there
-are all three genotypes, before the intermediate genotype is lost.
-A second example can be in a sexual selection model [van Doorn and Weissing, 2002], 
-where there is the onset for convergent selection.
-In both cases, these intermediate states may fall back to the original
-state of one species, or complete to produce a new species.
-
-% Effects on topology
-
-A feature of the protracted birth-death model, is that it may result in
-paraphylies when the rates between good and incipient species differ \todo{Add reference}.
-Paraphylies may be more than an experimental artifact and are claimed to
-be an inherent feature of nature \cite{funk2003species}.
-
-\todo{Add something like: this research solves this by sampling from paraphyletic trees (at the subspecies level) to create non-paraphyletic trees at the species level}
-
-% Good and incipient rates
-
-Where the classic birth death model assumes constant speciation and extinction
- rates, the protracted birth-death model allows these rates to be state
- dependent.
-
-The only additional parameter is the rate at which incipient species become
-(recognized as) good species.
-When setting that parameter to infinity, incipient species become good
-species instantaneous and the model falls back to a constant-rate (non-protracted) 
-birth-death model \todo{Add reference}.
-
-% Relationship to constant-rate birth-death model
-
-\todo{?Wat bedoelde ik met deze onzin?}
-Although both standard and protracted birth-death models use a parameter 
-$\lambda$, the $\lambda$ in the standard birth death model denotes
-a speciation rates (initiation and completion), where in the PBD model
-$\lambda$ denotes only the speciation initiation rate. 
-There is no relationship known between the parameters of the constant-rate
-birth-death model and the protracted birth-death model \cite{etienne2014estimating} 
-\todo{refesh memory, then update paragraph}.
-
-% Good and incipient rates do not need to be constant
-
-The assumption that speciation takes time is independent of the dynamics
-of the speciation and extinction rates.
-The protracted speciation model in this research assumes these rates are
-constant, but these rates can easily be made to depend on time, diversity
-or trait.
-
-% Use of a speciation model in inferring a phylogeny
-
-Speciation models are widely used to make inferences from genetic data.
-There are multiple computer programs to create phylogenies and/or parameter
-estimates from aligned DNA sequences.
-BEAST2 is a widely used tool that allows for a Bayesian approach to 
-phylogenetics \cite{bouckaert2014beast}.
-
-BEAST2 supplies the user with multiple speciation models, that all assume
-instantaneous speciation.
-
-% This study
-
-This simulation investigates the consequence of BEAST2 using instantaneous
-speciation, by simulating a 'true' tree that is protracted, simulate DNA
-alignments following that tree, and seeing how well BEAST2 can recover
-the original phylogeny.
-
-% Novelty
-
-This study is the first to measure the error made when acknowledging that
-speciation takes time, yet using one of the many tools that ignores this
-fact.
-
-Part of the parameter space will create 'true' trees that are not protracted.
-This will result in a test of BEAST2 against itself, which is, to the best
-of our knowledge, not been done, probably due to the extensive computations
-that are needed \todo{RSE: Heb je dit uitgebreid getest? Misschien zou je Remco kunnen vragen}.
-
-% Analysis
-
-The 'true' and inferred trees are compared with the nLTT statistic, which
-is a novel and efficient summary statistic, that has been proven to have
-better performance than the gamma statistics in Approximate
-Bayesian Computation \cite{janzen2015approximate} (yet on par with the likelihood).
-
-The nLTT statistics is the summed absolute difference between two 
-Lineages-Through-Time plot lines that are normalized to 
-let both time and number of lineages
-range from zero to one.
-Additionally, this statistic allows to pinpoint the time span that contributed
-most to its value.
-
-% Expectations
-
-It is expected that for higher protractedness (thus deviation from BEAST2
-its assumptions), the error will increase.
-Trees are expected to be inferred better, when there is more information
-available to construct them, thus we expect less error when there are more
-taxa and when the simulation DNA alignments are longer.
-The error is expected to be biggest close to the present, as there will
-be the most incipient species present.
-
-It is unknown, however, under which biological settings this error is relevant.
-
-% Preview results
-
-This study shows that ... \todo{put result here}.
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\begin{figure}
+  \centering 
+  \resizebox {0.8\textwidth} {!} {
+    \CreateTikzFigureExperiment{} 
+  }
+  \caption{
+    Overview of the experiment: 
+    A: per parameter set, there are $n_i$ incipient species trees created. 
+    B1: constructing a species trees by using the youngest subspecies $A1$ to represent good species $A$.
+    B2: constructing a species trees by using the oldest subspecies $A2$ to represent good species $A$.
+    C1: constructing one of the $n_a$ alignments of B1. 
+    C2: constructing one of the $n_a$ alignments of B2. 
+    D1: reconstructing a species trees from alignment B1.
+    D2: reconstructing a species trees from alignment B2.
+    Comparisons are done between a species tree and its reconstructed species tree, 
+    for example, between B1 and D1.
+    All phylogenies are drawn at the same scale.
+  }
+  \label{fig:experiment}
+\end{figure}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsection{Parameter space}
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\begin{table}
+  \centering 
+  \begin{tabular}{l l}
+    \hline
+    Symbol & Description \\
+    \hline
+    \hline
+    $b_g$ & Speciation initiation rate of a good species (probability per lineage, per million year) \\
+    $b_i$ & Speciation initiation rate of an incipient species (probability per lineage, per million year) \\
+    $\lambda$ & Speciation completion rate (probability per lineage, per million year) \\
+    $\mu_g$ & Extinction rate of a good species (probability per lineage, per million year) \\
+    $\mu_i$ & Extinction rate of an incipient species (probability per lineage, per million year) \\
+    \hline
+    $t_c$ & Crown age (million years) \\
+    $r$ & Mutation rate (probability of mutation per nucleotide per million years) \\
+    $l_a$ & DNA alignment length (base pairs) \\
+    \hline
+    $n_i$ & Number of incipient species trees per parameter set \\
+    $n_a$ & Number of alignments per species tree \\
+    $n_b$ & Number of BEAST2 runs per alignment \\
+    \hline
+    $n_s$ & Number of MCMC samples \\
+    $i_s$ & MCMC sampling interval \\
+    \hline
+    $n_{ti}$ & Number of taxa in an incipient species tree \\
+    $n_{t}$ & Number of taxa in a (good) species tree \\
+    $t_{\mean{ds}}$ & Average duration of speciation \\
+    $t_{sc}$ & Speciation completion time \\
+    $\Delta_{nLTT}$ & nLTT statistic value \\
+    $\mean{\Delta_{nLTT}}$ & Average nLTT statistic value of a posterior \\
+    $\median{\Delta_{nLTT}}$ & Median nLTT statistic value of a posterior \\
+    \hline
+  \end{tabular}
+  \caption{
+    Parameter descriptions. The sections are (1) the biological parameters, (2) the
+    biological tuning parameters, (3) parameters involved in replication, (4) MCMC
+    tuning parameters, and (5) other symbols.
+  }
+  \label{table:parameter_descriptions}
+\end{table}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \todo{@RSE: would you prefer $t_{\mean{ds}}$ or $\mean{t_{ds}}$ for average duration of speciation?}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\begin{table}
+  \centering 
+  \begin{tabular}{l l}
+    \hline
+    Parameter             & Values \\
+    \hline
+    \hline
+    $b = b_g = b_i$       & 0.1, 0.5, 1.0 \\
+    $\lambda$             & 0.1, 0.3, 1.0, $10^6$ \\
+    $\mu = \mu_g = \mu_i$ & 0.0, 0.1, 0.2, 0.4 \\
+    \hline
+    $t_c$                 & 15 \\
+    $r$                   & 0.01 \\
+    $l_a$                 & $10^3$, $10^4$ \\
+    \hline
+    $n_i$                 & 20 \\
+    $n_a$                 & 2 \\
+    $n_b$                 & 2 \\
+    \hline
+    $n_s$                 & $10^3$ \\
+    $i_s$                 & $10^3$ \\
+    \hline
+  \end{tabular}
+  \caption{
+    Parameter values. The sections are (1) the biological parameters, (2) the
+    biological tuning parameters, (3) parameters involved in replication, 
+    and (4) MCMC tuning parameters.
+  }
+  \label{table:parameter_values}
+\end{table}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 species trees allows for clearer measuments on the effect of sampling.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\begin{figure}
+  \centering 
+  \CreateTikzFigureSampling{}
+  \caption{
+    Sampling an incipient species tree. 
+    A) At stem age $t_0$, $A1$ is present.
+    At crown age $t_1$, its derived incipient species $A2$ is born. 
+    At $t_2$, $A2$ itself speciates into a new (unnamed) species, 
+    that becomes good species $B$ at $t_3$. At present time, $t_4$, there
+    exist two possible representatives of species $A$. 
+    B) Using the oldest representative, $A1$, to represent species $A$, the crown
+    age of the resulting species tree lies at $t_1$. 
+    C) Using the youngest representative, $A2$, to represent species $A$, the crown
+    age of the resulting species tree lies at $t_2$. 
+  }
+  \label{fig:sampling}
+\end{figure}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 high performance computing cluster of the University of Groningen.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Results}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Creation of data set}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\todo{First focus on the real results, not their quality}
-
-% Simulation of incipient species trees
-
-\todo{redo with new data}
-
-\todo{create figure or table that shows this}
-
-957 out of 960 incipient species trees have been simulated successfully. 
-The 3 failed simulations created data files bigger than R can read, 
-and have been excluded from further analysis.
-
-The number of good species ranged from 2 to 14296, with a median of 87 and an average of 503.8. 
-There were 6 phylogenies with only 2 good species.
-
-% Sampling species trees}
-
-\todo{redo with new data}
-
-For each incipient species trees, two species trees were constructed. 
-In 653 out of 960 cases, these trees had different branch lengths.
-In 302 cases, the trees were identical. In the remaining 5 cases,
-there were computational problems (see figure \ref{fig:exit_statuses}).
-
-% Simulation of alignments
-
-\todo{redo with new data}
-
-For each species tree, $n_a$ alignments have been created, 
-resulting in $960 \cdot 2 \cdot 2 = 3840$ alignments. 
-$?$ of these were informative.
-
-\todo{May result in an overrepresentation of the two same species trees derived from one incipient species tree}
-
-% Simulation of posteriors}
-
-A BEAST2 posterior contains multiple parameter estimates (see figure \ref{fig:ess_distribution} for all),
-and each one has an ESS. An ESS of 200 is acceptable for any parameter estimate.
-The ESS of the tree likelihood should have such a decent ESS. Figure \ref{fig:ess_distribution_likelihood}
-shows the distribution of the likelihood ESS.
-
-\todo{How tp quote https://www.beast2.org/what-is-ess/ ? It states that not all variables
-need to have a high ESS, but 'The likelihoods (both of the tree and coalescent model) should have decent ESSs.}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Quantification of the error BEAST2 makes on all trees}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-To get a first global impression of the error BEAST2 makes, figure \ref{fig:error}
-shows a histogram of the nLTT statistics, colored by the expected mean
-duration of speciation. As BEAST2 assumes a mean duration of speciation of zero,
-it is expected that the nLTT statistic (a measure of error) is lowest. Zooming
-in on the left part of the distribution, there is no clean pattern.
-The right part of the tail of distribution shows that the biggest nLTT
-statistic values are found for high (but not highest) expected mean durations of
-speciation.
-
-\todo{figure \ref{fig:error_for_scr} shows the distribution of errors for all
-combinations of PBD parameters}
-
-As the expected correlation appears to be absent, figure 
-\ref{fig:error_expected_mean_dur_spec}, subfigure $a$, 
-shows the correlation between expected
-mean duration of speciation and the nLTT statistic. A linear fit shows
-that there is indeed an increase in error for an increases expected duration
-of speciation. A trendline using locally weighted smoothing, however,
-indicates a more complex relationship. Note that this figure only uses
-a subset of all measurements, consisting of $n_s$ nLTT statistic values 
-per posterior.
-
-Taking the average nLTT value per posterior frees us from computational
-constraints and allows us to use the full data set. The resulting
-plot, figure \ref{fig:error_expected_mean_dur_spec}, subfigure $b$, shows the same pattern:
-A linear fit shows
-that there is indeed an increase in error for an increases expected duration
-of speciation. But now, a trendline using locally weighted smoothing 
-follows this linear relationship closely.
-
-The inference error made is expected to be less for a longer DNA
-alignment length. 
-Figure \ref{fig:error_expected_mean_dur_spec_alignment_length} 
-show the correlation between expected mean duration of speciation 
-and the nLTT statistic for the two different DNA alignment lengths. 
-The first subfigure uses a sample of all $n_s$ individuals nLTT statistics,
-where the second subfigure takes only the posterior's 
-nLTT statistic mean value, for the complete dataset.
-
-Both figures show again that there is an 
-increase in error for an increases expected duration
-of speciation, when assuming a linear relation, and
-a more complex relationship when using locally weighted smoothing. 
-The figures also show that, as expected, 
-the error is less with more information. Additionally,
-a longer DNA alignment length results in a slower accumulation of error
-for increased expected mean duration of speciation.
-
-A higher ESS for longer DNA alignments may be an additional
-explanation for the observed slower accumulation of error
-for increased expected mean duration of speciation.
-Figure \ref{fig:ess_expected_mean_dur_spec_alignment_length}
-shows that longer DNA alignments have a higher likelihood ESS, 
-but this trend is weaker than in figure 
-\ref{fig:error_expected_mean_dur_spec_alignment_length} and
-the 95\% confidence intervals of the linearily fitted trendlines only
-stop overlapping for the highest expected mean duration of speciation.
-The locally weighted smoothing trendlines show a different trend,
-suggesting that alignment length has no effect on the
-likelihood ESS.
-
-\todo{Figure \ref{fig:error_alignment_length} shows the same pattern, seperated
-for all combinations of PBD parameters.}
-
-For each incipient species tree, two species trees were created, by taking
-either the youngest or oldest subspecies to represent each species.
-The effect of this is shown in figure \ref{fig:error_expected_mean_dur_spec_sampling}.
-Both subfigures show the
-error made for different $t_{\mean{ds}}$, where the first subfigure uses a sample
-of all $\Delta_{nLTT}$ values, where the second subfigure uses the $\mean{\Delta_{nLTT}}$
-of each posterior. In both subfigures, the linear and locally weighted smoothing
-have overlapping 95\% confidence intervals, hinting that there is no strong effect
-of the way subspecies are selected to represent their species.
-
-\note{Figure \ref{fig:error_sampling_representative} shows the same data, ordered
-by the different PBD parameters instead}
-
-The effect of having a low or high tree likelihood ESS is shown in 
-figure \ref{fig:figure_error_expected_mean_dur_spec_low_high_ess}. We
-define a low ESS as being below median, where a high ESS is defined as 
-above-median. Plotting (a sample of or the means of) the nLTT statistic for different mean
-duration of speciation for a low an high ESS, shows that there is no clear
-difference in error made for posteriors with either low or high tree likelihood ESSes.
-
-It is expected that the bigger phylogeny is simulated, the less the error
-will be. Figure \ref{fig:figure_error_tree_size} shows the relation between the
-number of taxa and the mean nLTT statistic. Although most simulated
-phylogenies have less than 250 taxa, there is a clear positive trend, 
-which is the opposite of the
-expactation: the bigger the phylogeny, the bigger the error made.
-As the nLTT statistic normalized for the number of taxa, some other
-mechanism must be involved.
-
-For each species tree, $n_a$ different alignments are simulated. For each of
-these alignments, $n_b$ posteriors are simulated, using the same alignment. 
-It is assumed that two posteriors from the same alignment
-give rise to the same parameter estimates distributions.
-distribution. Figure \ref{fig:figure_posterior_distributions_likelihood} 
-shows the distribution of Mann-Whitney U test p-values for each pair of 
-posteriors working on a same alignment, 
-using their tree likelihood estimates.
-
-The figure shows that the tree likelihood posterior distributions 
-from a same alignment nearly always
-are significantly different. Subfigure (b) and (c) show the distribution of
-tree likelihood estimates for a posterior pair with the lowest (b) and 
-highest (c) p-value.
-
-Posteriors based on the same alignment are also (next to estimated tree likelihood) 
-assumed to have a similar nLTT statistic distribution. 
-Figure \ref{fig:figure_posterior_distribution_nltt} 
-shows the distribution of Mann-Whitney U test p-values for each pair of 
-posteriors working on a same alignment, 
-using their nLTT statistics.
-
-The figure shows that the nLTT statistic distributions 
-from a same alignment are commonly different and commonly the same. 
-Subfigures $(b)$ and $(c)$ show the distribution of
-nLTT statistics for a posterior pair with the lowest $(b)$ and 
-highest $(c)$ p-value.
-
-The nLTT statistic normalizes a phylogeny to have a crown age of 1.
-To prevent overlooking patterns in posterior crown ages, 
-figure \ref{fig:figure_posterior_distribution_crown_ages}
-shows the distribution of crown ages of all simulations.
-Almost all simulations have a crown age close to $0.15$, but there
-are rare cases with crown ages above $0.2$, ranging to $83.4$.
-
-\todo{I think it is very suspicious that the range centers on 0.15, when
-the crown age is 15. Check this!}
-
-Again, it is assumed that posteriors based on the exact same alignment
-have a similar crown age distribution. Of all posterior pairs, the Mann-Whitney
-U test p-values are shown in figure \ref{fig:figure_posterior_distribution_crown_ages_p_values}.
-It shows that crown age distributions being different is common and these
-being the same is uncommon.
-
-It is expected that if a tree is created by a Birth-Death process, BEAST2 is able
-to reasonably estimate those parameters.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Discussion}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-This article has quantified the error made in inference when (1) nature would follow the
-protracted speciation model, and (2) her DNA sequences would be analysed with a birth-death
-model.
-
-% The PBD model is just one implementation of speciation taking time
-
-The key feature of the PBD model is that it encompasses speciation taking time in the
-simplest way possible (that allows for mathematical analysis). The PBD model allows for
-multiple incipient species stages (where its earliest version had infinite
-stages \cite{rosindell2010protracted}). This research assumes that there is only one
-incipient species stage. One could argue that speciation takes place in more
-biological stages \todo{Add reference}, but we tried to use the simplest incarnation of the PBD possible.
-Also, the PBD model can also be viewed as a model of species recognition \todo{?Purvis et al., 2009},
-in which (human) recognition needs to build up, which we consider to be sufficiently
-encapsulated in one single stage. Future research may investigate if adding multiple incipient
-species stages for simulating nature changes the error made by (non-protracted) birth-death models.
-
-% Equal and constant rates
-
-The PBD model allows for multiple incipient species stages. 
-Each stage has its own extinction and speciation-initiation and 
-speciation-completion rate (except for good species, which lacks 
-a speciation-completion rate). These rates are assumed to be constant
-in time and equal for all stages. We chose to use the constant-rate
-birth-death model to keep this investigation as simple as possible.
-Future research may investigate whether a biologically more realistic
-rate model, for example, by adding diversity-dependence, results in
-the same inference error.
-
-We chose to set the extinction and speciation-initiation rates equal for all stages,
-again for choosing the simplest model. We expect these rates to differ in reality:
-incipient species have low initial abundances, which has an effect on extinction
-and speciation-initiation rates.
-
-In this research, we nevertheless set these rates equal. 
-Not only because it helps constrain the parameter space investigated,
-also paraphylies would pop up in the simulated trees. This would violate more assumptions
-of the BEAST2 model used as a prior and increase the complexity of the analysis.
-
-% Sampling species trees from incipient species trees
-
-An incipient species tree may have multiple incipient lineages from the
-same ancestor. We constructed two species trees from each incipient species
-tree, by selecting the youngest and oldest subspecies to represent a species.
-This selection maximizes the chance the species trees are different, where
-random sampling of representatives would also be a valid approach, closer
-to what would happen when scientists take measurements in the field. Our
-approach increases the error made.
-
-% Nucleotide substitution model
-
-For the simulation of DNA alignments, the Jukes-Cantor nucleotide substitution model
-is used. This substitution model is a simplification of the DNA mutation processes
-happening in reality \todo{Add reference}. In this simulation research, however, we could also
-state in BEAST2 that this nucleotide substitution model is used. Thus, we kept
-the simulation of the nucleotides in time as simple as possible, and used that
-knowledge in later steps. We expect that any nucleotide substitution model, when
-used in both alignment generation and in phylogeny inference, will do equally well,
-except that the full computational pipeline would take needlessly longer.
-
-% Allopatric speciation only
-
-When a speciation event is initiated, this investigation assumes no gene flow
-between the two incipient lineages. This follows the allopatric mode of speciation, 
-which is assumed to be most prominent of nature \todo{Add reference}. It would be interesting to
-allow gene flow between related incipient species, 
-so also peripatric and sympatric modes of speciation would be allowed for.
-We predict that this would increase the error in inference.
-
-% Why not put the tree prior into an MCMC tool directly?
-One could argue we should have written a PBD tree prior for an MCMC phylogenetic tool.
-We did not do that for two reasons. First, we think it is rational to first
-measure if there is need for a more complex tree prior. Second, 
-we could not have done so, as a phylogenetic tool like BEAST2 assumes species
-trees, where our input consists of incipient species trees. Before this
-research, we were unsure how to sample species representatives. This research
-has shown that any sampling is just as good, opening up avenues for
-implementing a PBD species tree prior.
-
-% BEAST2 is just one of many tools
-
-BEAST2 is one of many MCMC phylogenetic tools with its
-own internal technicalities. Similar tools are the R phytools
-package \cite{phytools}, BAli-Phy \cite{suchard2006bali}
-and BayesPhylogenies \cite{pagel2007bayesphylogenies}, to name a few.
-BEAST2 was chosen, because (1) it's FOSS, (2) its software
-architecture is modular, encouraging extension, (3) its development
-team follows good practices in software development.
-
-% BEAST2 parameter settings: strict clock
-
-\todo{elaborate}
-
-Use of strict clock is clean, but unrealistic.
-
-\todo{elaborate}
-
-% Alignment generation
-
-Use of JC69 is rational, but unrealistic.
-
-% Use for empiricists
-
-\todo{elaborate}
-
-This reseach shows that a DNA sequence length of 10k [is good].
-[must be effective DNA sequence length]
-[no indels generated]
-
-% Low number of incipient species trees
-
-\todo{elaborate}
-
-For each point in parameter space, twenty incipient species tree are created.
-Was fitted to 1000 jobs.
-
-% Failed simulations
-
-Not all parameter combinations resulted in posteriors with a high ESS,
-due to computational constraints. We think this is a valid approach:
-would a phylogeny get too big, we as scientists would cut it to
-pieces that are analytically tractable. Because these simulations
-are only a minority (as shown in the Results section), we do not
-expect this to influence our conclusions.
-
-% nLTT statistic is just one of many summary statistics
-
-We used the nLTT statistic for quantifying the error between the
-true and inferred phylogeny. There are simpler statistics that
-can do the same, like the gamma \cite{pagel1999inferring}, 
-delta-r \cite{pigot2010shape} \todo{check citation}, 
-tree balance \todo{Add reference} or
-Normalized Rooted Branch Score [Drummond] statistic, 
-or more complex, like the spectral analysis [Hendy and Penny, 1993].
-We consider those other summary statistics equally good, we just picked the
-one we thought of the most informative per unit of added computational time.
-
-% Add:
-%
-% Computational Performance and Statistical Accuracy of *BEAST and Comparisons with Other Methods (by Huw A. Ogilvie Joseph Heled Dong Xie Alexei J. Drummond)
-% cites rooted branch score, RBS, as being from Heled and Bouckaert 2013, `Looking for trees in the forest: Summary tree from posterior samples`
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Acknowledgements}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-We would like to thank the Center for Information Technology of the University of Groningen for their support
-and for providing access to the Peregrine high performance computing cluster.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Authors' contributions}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-RJCB and RSE conceived the idea for the package. RJCB created and tested the package, and wrote the first draft of the manuscript. RSE contributed substantially to revisions.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Bibliography
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% MEE style
-\bibliographystyle{mee}
-\bibliography{article}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Tables}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{table}
-  \centering 
-  \begin{tabular}{l l}
-    \hline
-    Symbol & Description \\
-    \hline
-    \hline
-    $b_g$ & Speciation initiation rate of a good species (probability per lineage, per million year) \\
-    $b_i$ & Speciation initiation rate of an incipient species (probability per lineage, per million year) \\
-    $\lambda$ & Speciation completion rate (probability per lineage, per million year) \\
-    $\mu_g$ & Extinction rate of a good species (probability per lineage, per million year) \\
-    $\mu_i$ & Extinction rate of an incipient species (probability per lineage, per million year) \\
-    \hline
-    $t_c$ & Crown age (million years) \\
-    $r$ & Mutation rate (probability of mutation per nucleotide per million years) \\
-    $l_a$ & DNA alignment length (base pairs) \\
-    \hline
-    $n_i$ & Number of incipient species trees per parameter set \\
-    $n_a$ & Number of alignments per species tree \\
-    $n_b$ & Number of BEAST2 runs per alignment \\
-    \hline
-    $n_s$ & Number of MCMC samples \\
-    $i_s$ & MCMC sampling interval \\
-    \hline
-    $n_{ti}$ & Number of taxa in an incipient species tree \\
-    $n_{t}$ & Number of taxa in a (good) species tree \\
-    $t_{\mean{ds}}$ & Average duration of speciation \\
-    $t_{sc}$ & Speciation completion time \\
-    $\Delta_{nLTT}$ & nLTT statistic value \\
-    $\mean{\Delta_{nLTT}}$ & Average nLTT statistic value of a posterior \\
-    $\median{\Delta_{nLTT}}$ & Median nLTT statistic value of a posterior \\
-    \hline
-  \end{tabular}
-  \caption{
-    Parameter descriptions. The sections are (1) the biological parameters, (2) the
-    biological tuning parameters, (3) parameters involved in replication, (4) MCMC
-    tuning parameters, and (5) other symbols.
-  }
-  \label{table:parameter_descriptions}
-\end{table}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{table}
-  \centering 
-  \begin{tabular}{l l}
-    \hline
-    Parameter             & Values \\
-    \hline
-    \hline
-    $b = b_g = b_i$       & 0.1, 0.5, 1.0 \\
-    $\lambda$             & 0.1, 0.3, 1.0, $10^6$ \\
-    $\mu = \mu_g = \mu_i$ & 0.0, 0.1, 0.2, 0.4 \\
-    \hline
-    $t_c$                 & 15 \\
-    $r$                   & 0.01 \\
-    $l_a$                 & $10^3$, $10^4$ \\
-    \hline
-    $n_i$                 & 20 \\
-    $n_a$                 & 2 \\
-    $n_b$                 & 2 \\
-    \hline
-    $n_s$                 & $10^3$ \\
-    $i_s$                 & $10^3$ \\
-    \hline
-  \end{tabular}
-  \caption{
-    Parameter values. The sections are (1) the biological parameters, (2) the
-    biological tuning parameters, (3) parameters involved in replication, 
-    and (4) MCMC tuning parameters.
-  }
-  \label{table:parameter_values}
-\end{table}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Figures}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{figure}
-  \centering
-  \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=4cm, semithick]   
-  \tikzstyle{every state}=[]
-  \node[state] (A)              {Good};   
-  \node[state] (B) [right of=A] {Incipient};   
-  \node[state] (C) [below of=B] {Extinct};   
-  \path (A) edge [bend right] node {$b_g$} (B)
-        (A) edge [bend right] node {$\mu_g$} (C)
-        (B) edge [loop above] node {$b_i$} (B)
-        (B) edge [bend right] node {$\lambda$} (A)
-        (B) edge [bend right] node {$\mu_i$} (C); 
-  \end{tikzpicture}
-
-  \caption{
-    The states and transitions of a species in the PBD model.
-    $b_i$: speciation-initiation rate of incipient species. 
-    $b_g$: speciation-initiation rate of good species. 
-    $\lambda$: speciation completion rate. 
-    $\mu_i$: extinction rate of incipient species. 
-    $\mu_g$: extinction rate of good species. 
-    Figure after Etienne et al, 2014, Evolution
-  }
-  \label{fig:pbd_states}
-\end{figure}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{figure}
-  \centering 
-  \resizebox {0.8\textwidth} {!} {
-    \CreateTikzFigureExperiment{} 
-  }
-  \caption{
-    Overview of the experiment: 
-    A: per parameter set, there are $n_i$ incipient species trees created. 
-    B1: constructing a species trees by using the youngest subspecies $A1$ to represent good species $A$.
-    B2: constructing a species trees by using the oldest subspecies $A2$ to represent good species $A$.
-    C1: constructing one of the $n_a$ alignments of B1. 
-    C2: constructing one of the $n_a$ alignments of B2. 
-    D1: reconstructing a species trees from alignment B1.
-    D2: reconstructing a species trees from alignment B2.
-    Comparisons are done between a species tree and its reconstructed species tree, 
-    for example, between B1 and D1.
-    All phylogenies are drawn at the same scale.
-  }
-  \label{fig:experiment}
-\end{figure}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{figure}
-  \centering 
-  \CreateTikzFigureSampling{}
-  \caption{
-    Sampling an incipient species tree. 
-    A) At stem age $t_0$, $A1$ is present.
-    At crown age $t_1$, its derived incipient species $A2$ is born. 
-    At $t_2$, $A2$ itself speciates into a new (unnamed) species, 
-    that becomes good species $B$ at $t_3$. At present time, $t_4$, there
-    exist two possible representatives of species $A$. 
-    B) Using the oldest representative, $A1$, to represent species $A$, the crown
-    age of the resulting species tree lies at $t_1$. 
-    C) Using the youngest representative, $A2$, to represent species $A$, the crown
-    age of the resulting species tree lies at $t_2$. 
-  }
-  \label{fig:sampling}
-\end{figure}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\listoftodos
-
-\end{document}

### 2018-03-29 article.tex
Start from scratch

@@
 \usepackage{authblk}
-\title{babette: BEAUti 2, BEAST2 and Tracer for R}
+\title{The error when inferring phylogenies with incipient species by a Birth-Death model}
+% \subtitle{Should protracted speciation be incorporated in phylogenetic tree construction methods?}
+
 \author[1]{Rich\`el J.C. Bilderbeek}
@@
 
-\section{Introduction}
+What is the effect of sampling either the youngest or oldest species-to-be
+to represent a species?
 
-% Biology: speciation takes time
+What is the effect of protractedness on the inference error we make?
 
-It takes time for new species-to-be to achieve the status of a new species.
-First, it takes time for the new species-to-be 
-to acquire sufficient different mutations to 
-establish reproductive isolation \cite{schluter2001ecology}. Second, 
-it takes time for us to recognize reproductive isolation
-and reclassify that one single species has become two 
-(or more) species.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Pilot experiments}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-As an example, \cite{fennessy2016multi} 
-showed that Africa has four instead of only one
-giraffe species, with a common ancestor estimated 
-at around 2 million years ago.
+Do the DNA sequence length and mutation rate (as described in the Methods)
+give a good enough inference? Answer: pick the setting with the highest
+expected number of taxa (highest speciation initiation and completion rate, 
+no extinction). Simulate 1000 trees and obtain an nLTT distribution for a
+reference DNA sequence length and mutation rate. Increasing DNA sequence length
+or mutation rate increases the resolution of the experiment. This resolution
+must not be too low (this would cost accuracy), nor too high (this would cost
+too much computational resources). If doubling DNA length or DNA mutation rate
+results in an nLTT distribution with 90% overlap with the reference distribution,
+in 900 of 1000 cases, the (lower) reference resolution is considered high enough.
 
-% The use of speciation models
-
-From DNA sequences obtained in the present, we can construct phylogenies
-to infer the evolutionary relationship between species and the time 
-from which ancestral species developed into separate.
-
-As we only sample DNA of extant species, 
-we can only infer a reconstructed tree, in which
-the lineages that have gone extinct are absent.
-Although most models assume extinctions (\cite{yule1925mathematical} being the
-classic exception, as at its time computation was expensive), 
-information on extinction times is lost due to our sampling.
-
-The time when reproductive isolation started to take off, is
-also lost in our sampling method, if the new species-to-be is not 
-recognized yet as such. Without fossil date, one cannot observe the
-possibly many times that reproductive isolation built up, but was lost again.
-
-% First model: constant rate birth death model
-
-Although we know that speciation takes time, we commonly ignore this when
-constructing a phylogeny, by choosing a constant-rate birth-death model
-as a speciation model. This can be justified, as for lineages 
-that have speciated far back in the past, there is no error. 
-Yet in the present, we do not recognize the future species-to-be
-that are already genetically converging, resulting in an underestimation
-of the number of species present today, which can only be 
-concluded in retrospect.
-
-The constant-rate birth-death model (as described in for example \cite{nee1994reconstructed}) 
-is among the simplest speciation models, and assumes a constant speciation
- rate $\lambda$ and constant extinction rate $\mu$.
-
-Additionally, it assumes speciation is instantaneous.
-The constant-rate birth-death model is popular for its simplicity, yet
-has also served as a starting point for more elaborate speciation models.
-
-% Other non-protracted speciation models
-
-Other speciation models may assume that speciation rate changes in 
-time \cite{rabosky2008explosive}, is dependent on the amount of species 
-present \cite{etienne2011diversity}, or is trait 
-dependent \cite{fitzjohn2009estimating}.
-
-In these examples, the original birth-death model has been extended by making 
-speciation rate dependent on time, diversity or trait value respectively.
-The assumption that all these extensions share is that speciation is instantaneous;
-that after a new branching event, both two lineages are (or are recognized as being) different
-species.
-
-% Protracted speciation model
-
-% Good and incipient states
-
-The protracted speciation model \cite{etienne2012prolonging} allows for speciation taking time.
-
-It adds an additional species state (see also figure \ref{fig:pbd_states}),
-coined the 'incipient' stage, which a lineages has to complete before becoming
-a 'good' species.
-
-One view is to say that good species have achieved reproductive isolation,
-where incipient species are in the process of achieving this \todo{Add reference}.
-
-As genetic drift is unlikely to be a relevant speciation mechanism,
-this is more likely to be ecological speciation [Sobel et al., 2009]. 
-\todo{RSE: Ben je het daarmee eens? Zou drift niet een belangrijke ooraak van speciation op islands kunnen zijn bijvoorbeeld?}
-
-Alternatively, an incipient species can be described as a good-species-to-be,
-yet not recognized as such \todo{?[Purvis et al., 2009]?}.
-
-% Biological mechanism of reproductive isolation
-
-The protracted speciation model makes no assumption about the exact biological
-mechanism by which reproductive isolation is 
-acquired \cite{etienne2012prolonging, etienne2014estimating,rosindell2010protracted}.
-
-A 'good' state can be viewed as a state in which there is convergent selection
-for that lineage.
-
-The 'incipient' state can be seen as a state in which there is the potential
-for speciation.
-
-A first example can be from the BDM \todo{Add reference} model, the state in which there
-are all three genotypes, before the intermediate genotype is lost.
-A second example can be in a sexual selection model [van Doorn and Weissing, 2002], 
-where there is the onset for convergent selection.
-In both cases, these intermediate states may fall back to the original
-state of one species, or complete to produce a new species.
-
-% Effects on topology
-
-A feature of the protracted birth-death model, is that it may result in
-paraphylies when the rates between good and incipient species differ \todo{Add reference}.
-Paraphylies may be more than an experimental artifact and are claimed to
-be an inherent feature of nature \cite{funk2003species}.
-
-\todo{Add something like: this research solves this by sampling from paraphyletic trees (at the subspecies level) to create non-paraphyletic trees at the species level}
-
-% Good and incipient rates
-
-Where the classic birth death model assumes constant speciation and extinction
- rates, the protracted birth-death model allows these rates to be state
- dependent.
-
-The only additional parameter is the rate at which incipient species become
-(recognized as) good species.
-When setting that parameter to infinity, incipient species become good
-species instantaneous and the model falls back to a constant-rate (non-protracted) 
-birth-death model \todo{Add reference}.
-
-% Relationship to constant-rate birth-death model
-
-\todo{?Wat bedoelde ik met deze onzin?}
-Although both standard and protracted birth-death models use a parameter 
-$\lambda$, the $\lambda$ in the standard birth death model denotes
-a speciation rates (initiation and completion), where in the PBD model
-$\lambda$ denotes only the speciation initiation rate. 
-There is no relationship known between the parameters of the constant-rate
-birth-death model and the protracted birth-death model \cite{etienne2014estimating} 
-\todo{refesh memory, then update paragraph}.
-
-% Good and incipient rates do not need to be constant
-
-The assumption that speciation takes time is independent of the dynamics
-of the speciation and extinction rates.
-The protracted speciation model in this research assumes these rates are
-constant, but these rates can easily be made to depend on time, diversity
-or trait.
-
-% Use of a speciation model in inferring a phylogeny
-
-Speciation models are widely used to make inferences from genetic data.
-There are multiple computer programs to create phylogenies and/or parameter
-estimates from aligned DNA sequences.
-BEAST2 is a widely used tool that allows for a Bayesian approach to 
-phylogenetics \cite{bouckaert2014beast}.
-
-BEAST2 supplies the user with multiple speciation models, that all assume
-instantaneous speciation.
-
-% This study
-
-This simulation investigates the consequence of BEAST2 using instantaneous
-speciation, by simulating a 'true' tree that is protracted, simulate DNA
-alignments following that tree, and seeing how well BEAST2 can recover
-the original phylogeny.
-
-% Novelty
-
-This study is the first to measure the error made when acknowledging that
-speciation takes time, yet using one of the many tools that ignores this
-fact.
-
-Part of the parameter space will create 'true' trees that are not protracted.
-This will result in a test of BEAST2 against itself, which is, to the best
-of our knowledge, not been done, probably due to the extensive computations
-that are needed \todo{RSE: Heb je dit uitgebreid getest? Misschien zou je Remco kunnen vragen}.
-
-% Analysis
-
-The 'true' and inferred trees are compared with the nLTT statistic, which
-is a novel and efficient summary statistic, that has been proven to have
-better performance than the gamma statistics in Approximate
-Bayesian Computation \cite{janzen2015approximate} (yet on par with the likelihood).
-
-The nLTT statistics is the summed absolute difference between two 
-Lineages-Through-Time plot lines that are normalized to 
-let both time and number of lineages
-range from zero to one.
-Additionally, this statistic allows to pinpoint the time span that contributed
-most to its value.
-
-% Expectations
-
-It is expected that for higher protractedness (thus deviation from BEAST2
-its assumptions), the error will increase.
-Trees are expected to be inferred better, when there is more information
-available to construct them, thus we expect less error when there are more
-taxa and when the simulation DNA alignments are longer.
-The error is expected to be biggest close to the present, as there will
-be the most incipient species present.
-
-It is unknown, however, under which biological settings this error is relevant.
-
-% Preview results
-
-This study shows that ... \todo{put result here}.
+What is the MCMC chain length that gives a posterior ESS of 200 in all cases?
+Answer: run all simulations, measure the posterior's ESS. If at least one 
+ESS is lower than 200, elongate all MCMC chain lengths by a fraction 
+to predict to increase it to 220 (for example, would the lowest ESS be 165, 
+making the MCMC chains 33\% longer would bring it to (165\%*1.33=) 220. We use
+220 to prevent an asymptotic approaching to 200.
 
@@
 
-This research measures the error our current BD tools make, when
-the real process has been protracted. Figure \ref{fig:experiment} shows
-the following overview of the experiment graphically.
- An in silico data set is created for a wide
-range of parameters from which incipient species trees are simulated (A in figure \ref{fig:experiment}). 
-These phylogenies serve as 'the truth' or 'nature'. From that incipient species tree,
-species trees are created, by picking one representive per species.
-This is done twice, as either the youngest (B1) or oldest (B2) subspecies is chosen
-to represent a species. Per species tree, $n_a$ alignments are simulated (C1 and C2).
-Per alignment, $n_b$ posteriors are inferred, 
-each consisting of $n_s$ phylogenies and parameter estimates (D1 and D2). Each
-species tree is then compared with the species tree its alignment is based on (for
-example, B1 and D1).
+We simulate protracted birth-death trees. In a protracted birth-death tree,
+species-to-be are present, but not yet recognized as different from
+their ancestral species. We simulate these trees from the biological parameters 
+in a factorial fashion. Excluded from this are combinations of parameters in which
+speciation initiation rate is lower or equal the extinction rate, as these
+are biologically irrelevant. Out of computational necessity, 
+parameter combinations that result in an expected number of taxa above 1000 
+are removed as well.
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Parameter space}
-\label{subsec:parameter_space}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+From these PBD trees, we sample species tree in two different ways. 
+From all incipient species-to-be, we pick either the youngest or oldest
+species-to-be to represent the full species. This species tree we refer
+to as 'the true species tree'.
 
-\todo{@RSE: would you prefer $t_{\mean{ds}}$ or $\mean{t_{ds}}$ for average duration of speciation?}
+From a species tree, we simulate a DNA alignment that has the same history
+as the phylogeny. The nucleotides of the DNA alignment follow a JC69
+nucleotide substitution model, in which all nucleotide-to-nucleotide transitions
+are equally likely. Although this may seem as a simplification, in our Bayesian
+inference we can use this as a prior, which we know is the truth.
 
-% Biological parameters
+The DNA sequence length and mutation rate was chosen as such to provide one
+nucleotide change per 1000 years per lineage. As our time unit is in 
+millions of years, this equates to a rate of 1000 nucleotide substitutions 
+changes per time unit per lineage. To provide for each substitution being unique,
+we simulate as much DNA nucleotides as expected mutations, which equals the
+nucleotide substitution rate times the crown age, which equals 15K. As this
+mutation rate is constant over all branches, it in effect follows a strict 
+clock model.
 
-This study primarily investigates combinations of the five biological 
-parameters ($b_g$, $b_g$, $\lambda$, $\mu_g$ and $\mu_g$, 
-see also figure \ref{fig:pbd_states} and table \ref{table:parameter_descriptions}), 
-in a close to full factorial fashion, with two exceptions.
-First, following \cite{etienne2014estimating}, birth and extinction rates were each 
-set equal between good and incipient species ($b_g = b_i$, and $\mu_g = \mu_i$).
-Second, only parameter combinations with speciation initiation rates greater
-than extinction rates were used ($b_g > \mu_g$). The values 
-used in the experiment (see table \ref{table:parameter_values}) 
-were selected to have a reasonable expected number of incipient 
-lineages, $E(n_{ti})$, where \todo{?too basic to have a reference to? I assume Kendall, 1948, \cite{kendall1948generalized}, but could not look it up here and now}:
+The simulated alignment is the data that could be measured in the field.
+This alignment is used to infer a posterior of, using BEAST2. We assume a
+JC69 nucleotide substitution (as we do use that), a strict clock model (as
+we have that) and a Birth-Death tree prior (as this simplification is the goal
+of this research). Additionally, we assume a MRCA prior with a normal distribution
+with a mean of the crown age, and a standard deviation of 0.001 time units (which
+equates to 1000 years). After discarding a burn-in of 10\%, the MCMC algorithm must 
+result in an ESS of the posterior of at least 200 and the MCMC chain length is 
+chosen as such that this is the case for all simulations.
 
-\begin{equation}
-  E(n_{ti}) = e^{ (b - \mu) \cdot t_c }
-\end{equation}
+The similarity of each phylogeny in the posterior with 'the true species tree'
+is determined with the nLTT statistic. In this way, a distribution of phylogenies
+is transformed to a distribution of nLTT statistics.
 
-One value of speciation completion rate $\lambda$ is set (close to) infinity, as
-then the protracted speciation model falls back to a constant-rate 
-(non-protracted) birth-death model [?Etienne et al., 2012 or already Rosindell and Etienne 2011?].
+To measure if sampling (either the youngest or oldest species-to-be to represent
+a species) matters, for all parameter combinations that have incipient species,
+we simulate 1000 incipient species trees that result in different species trees
+when sampled either way. From each species tree, we compare the nLTT distributions
+derived from the posterior. We quantify the Kullback-Leibler divergence (more
+or less: the weighted fraction of overlapping nLTT 
+values, which is zero if the two distributions are completely seperate, and close
+to one if they overlap). For each parameter combination, we take the average
+divergence. We assign verbal labels to this average divergence as such: 
+0.0-0.25: clear effect, 0.25-0.5: strong effect, 0.5-0.75: mild effect, 
+0.75-1.0: no effect. Per parameter setting, we show the average divergence and the 
+verbal label. The distribution of these divergences is shown in the 
+supplementary information.
 
-% Biological tuning parameters
-
-Secondarily, there are three biological tuning parameters. The crown age of
-the incipient species tree, $t_c$, is set to 15 million years for all simulations, 
-following \cite{etienne2014estimating}. The per-nucleotide mutation rate, $r$, and the
-DNA aligment length, $l_a$, together must be tuned to result in informative
-DNA alignments (see section \ref{subsec:simulating_alignments}). $r$ is identical in all
-simulations. Increasing $l_a$ results in more informative alignments, and this
-effect is measured by using two values of $l_a$ 
-per biological parameter combination.
-
-% Replication parameters
-
-All processes involving stochasticity have replicates. 
-For each set of parameters, $n_i$ incipient species trees are simulated. 
-From each species tree, $n_a$ alignments are simulated. From each alignment, 
-$n_b$ posteriors are generated.
-
-% MCMC tuning parameters
-
-The posteriors are generated using BEAST2, which uses a Monte-Carlo Markov-Chain (MCMC)
-algorithm. This algorithm is tuned to result in a high effective sample size (ESS) by
-two constants. The first constant is the MCMC sampling interval, which is $i_s$ states.
-$i_s$ which must be high enough to avoid correlation between subsequently sampled posterior states.  
-The second constant is the number of MCMC state samples, $n_s$, which equals the ESS iff \todo{@RSE: Can I use iff?}
-$i_s$ is high enough to have no correlation between subsequently sampled posterior states.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Simulating incipient species trees}
-\label{subsec:simulating_incipient_species_trees}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-From each parameter set, $n_i$ incipient species trees with a crown age
-of $t_c$ and at least two extant taxa, 
-are created in R \cite{R} using the PBD package \cite{PBD}.
-The values of the biological parameters encourage that the number of tips in
-each phylogeny is kept within computational limits. However, due to
-the stochastic nature of the Doob-Gillespie algorithm used in the creation
-of these phylogenies, phylogenies can, by chance, become very big. 
-There is no maximum number of taxa set, yet imposed instead by the R 
-programming language itself, due to a limit on the amount of memory that can be 
-allocated. The incipient species trees that could not be simulated are
-counted and discarded from further analysis.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Sampling species trees}
-\label{subsec:sampling_species_trees}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-Per incipient species tree, two species trees are created. 
-To create a species tree (a tree in which each species is represented once)
-for an incipient species tree, one representative for each species is picked. 
-If a species has initiated one or more incipient species, either the youngest
-or the oldest representative is picked. See figure \ref{fig:sampling} for an example.
-
-An alternative sampling method would be to pick 
-random representatives of each species. This would decrease the variability
-between the two species tree. Maximizing variability between the two
-species trees allows for clearer measuments on the effect of sampling.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Simulating alignments}
-\label{subsec:simulating_alignments}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-From each species tree, $n_a$ DNA alignments were simulated that match the
-phylogeny's structure, using the phangorn R package \cite{phangorn}. 
-Each alignment contains one DNA sequence per species 
-of $l_a$ nucleotides. The alignment follows the Jukes-Cantor nucleotide 
-subsititution model \cite{cantor1969mammalian}, 
-with per-nucleotide rate of mutation, $r$.
-
-The Jukes-Cantor
-nucleotide substitution model is regarded as an oversimplification of 
-nature \cite{unknown}. In this in silico experiment, however, we simulate first
-a DNA alignment from a phylogeny using nucleotide substitution model $M_N$, 
-then infer a phylogeny from that DNA alignment, assuming nucleotide substitution model $M_N$.
-Too reduce unnecessary complexity, $M_N$ is chosen to be the simplest possible.
-
-The combination of $l_a$ and $r$ is tuned to result in informative alignments. 
-Within a maximally informative alignment, each species has its own unique DNA sequence.
-Additionally, the Jukes-Cantor distance \cite{unknown} between any 
-two sequences must be calculable, which is when the proportion of matching nucleotides is below 75\%.
-The tuning was done using an $l_a$ of $10^3$ nucleotides, 
-as this appears to be the current median of the sequenced genome lengths, as currently published.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Simulation of posteriors}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-From each DNA alignment, $n_b$ posteriors of phylogenies and parameter combinations 
-is constructed, using BEAST2, a Bayesian phylogenetic tool \cite{bouckaert2014beast}.
-The nucleotide substitution model is set to 
-be Jukes-Cantor (one nucleotide substition rate category, no invariant nucleotides) 
-and a strict molecular clock, which matches the procedure the DNA alignments are generated, 
-as described in section \ref{subsec:simulating_alignments}).
-The tree prior is set to the (non-protracted) birth-death model, 
-with an uniform prior for birth-rate of range $[0, 10^5]$ and 
-a uniform death-per-birth rate prior of range $[0, 1]$. 
-The MCMC creation parameters were set to have $n_s$ states, sampled at an interval of $i_s$ states.
-
-An MCMC must have an effective sample size (ESS) of at least 200 to result in a 
-posterior that is 'adequate' \cite{drummond2015bayesian}. Due to the multiple 
-sources of stochasticity in the preceding steps, this ESS may not always be achieved.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Analysis}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-To quantify the error, each species tree is compared with the posterior derived
-from its $n_a$ alignments (for example, B1 and D1 in figure \ref{fig:experiment}).
-This is by calculating the nLTT statistic \cite{janzen2015approximate}
-between the species tree and each of its posterior's trees (using 
-the nLTT package \cite{nLTT}), resulting in an nLTT statistic distribution.
-
-% BD trees
-
-The nLTT statistic distribution for all simulations with a high speciation 
-completion rate is visualized for the different parameter values. This allows
-to see the baseline error. Instead of using a complete
-nLTT statistic distribution, also the mean value of each posterior is 
-sometimes used to reduce variance in this.
-
-The effect of speciation completion time $\lambda$ is visualized, by showing
-nLTT statistic distributions for the different parameter values, for all values
-of $\lambda$.
-
-Instead of using the parameter value of $\lambda$ as a proxy for protractedness,
-the speciation completion time, $t_{sc}$, and average duration of speciation, $t_{\mean{ds}}$, 
-are used. Speciation completion time, $t_{sc}$, is simply the inverse of the
-speciation completion rate $\lambda$:
-
-\begin{equation}
-  t_{sc} = \frac{1}{\lambda}
-  \label{eq:speciation_completion_time}
-\end{equation}
-
-The speciation completion time differs from the observed average speciation time, 
-as incipient species may go extinct and may produce other incipient species 
-that complete speciation before the original incipient one. The average duration 
-of speciation, $t_{\mean{ds}}$, takes this into 
-account \todo{check formula ?19 of?20 in Etienne et al., 2012, then cite}:
-
-\begin{equation}
-    t_{\mean{ds}} = \mean{\rho} = \frac{2}{ (D - \lambda + b_i - \mu_i) \cdot \ln{ \frac{2}{(1 + (\lambda - b_i + \mu_i)/D)}  } }
-\end{equation}
-
-where $D$ is defined as:
-
-\begin{equation}
-    D = \sqrt{ (\lambda + b_i)^2 + 2 \cdot (\lambda - b_i) \cdot \mu_i + \mu_i^2 }
-\end{equation}
-
-For each simulation, the expected $t_{\mean{ds}}$ is calculated from its parameter 
-values (using the PBD package \cite{PBD}), after which the nLTT statistic 
-distributed is plotted, to see the effect of protractedness on the nLTT statistic
-distribution.
-
-As not all simulations result in posteriors with an adequate ESS, the
-effect that a below-median or above-median ESS has on the nLTT statistic
-distribution is displayed.
-
-To view the effect of alignment length, $l_a$, the nLTT statistic
-distributions are plotted side-by-side for all other parameter values.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Peripherals}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-Simulating incipient species trees, sampling species trees, simulating alignments
-and creating posteriors, are done with the wiritttes package \cite{wiritttes}. The
-analysis is done with the wiritttea package \cite{wiritttea}. Those packages rely on 
-functionality in the APE \cite{APE}, geiger \cite{GEIGER}, ggplot2 \cite{ggplot2}, 
-RBeast \cite{RBeast}. Quality of these packages is assured by 
-the covr \cite{covr}, devtools \cite{devtools}, 
-goodpractice \cite{goodpractice}, lintr \cite{lintr}, testit \cite{testit} 
-and testthat \cite{testthat} packages, using the Travis CI \cite{travis} continuous integration service. 
-Both packages are hosted at GitHub \cite{github}, a good practice for computational
-scientists \cite{perez2016ten} and helps improve transparency \cite{gorgolewski2016practical}.
-Code coverage, which correlates with code quality \cite{del1995correlation}, 
-is above 90\% for both packages.
-
-The simulated dataset can be downloaded from DataDyrad [but not yet]. The
-summarized dataset can be downloaded at \cite{wirittte_data}. The
-pre-publication of this article can be downloaded at \cite{wirittte_article}.
-
-Simulations and analyis was mostly performed on the Peregrine 
-high performance computing cluster of the University of Groningen.
+To measure if protractedness matters, we compare the nLTT distribution per
+combination of speciation initiation and extinction rate for a protracted
+setting with its non-protracted (with a high speciation completion 
+rate) setting. So, for each speciation initiation and extinction rate, 
+the nLTT statistic distribution of the 
+non-protracted setting is used as the baseline. That distribution is compared
+with each protracted setting. We quantify the Kullback-Leibler divergence 
+of these nLTT distributions as described above. Per protracted parameter setting, 
+we generate 1000 replicate incipient species trees, each sampled randomly. We show the 
+average divergence of these 1000 replicates and the verbal label.
+The distribution of these divergences is shown in the supplementary information.
 
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Creation of data set}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\begin{figure}[]
+  \includegraphics[width=0.9\textwidth]{fig_does_sampling_matter.png}
+\end{figure}
 
-\todo{First focus on the real results, not their quality}
-
-% Simulation of incipient species trees
-
-\todo{redo with new data}
-
-\todo{create figure or table that shows this}
-
-957 out of 960 incipient species trees have been simulated successfully. 
-The 3 failed simulations created data files bigger than R can read, 
-and have been excluded from further analysis.
-
-The number of good species ranged from 2 to 14296, with a median of 87 and an average of 503.8. 
-There were 6 phylogenies with only 2 good species.
-
-% Sampling species trees}
-
-\todo{redo with new data}
-
-For each incipient species trees, two species trees were constructed. 
-In 653 out of 960 cases, these trees had different branch lengths.
-In 302 cases, the trees were identical. In the remaining 5 cases,
-there were computational problems (see figure \ref{fig:exit_statuses}).
-
-% Simulation of alignments
-
-\todo{redo with new data}
-
-For each species tree, $n_a$ alignments have been created, 
-resulting in $960 \cdot 2 \cdot 2 = 3840$ alignments. 
-$?$ of these were informative.
-
-\todo{May result in an overrepresentation of the two same species trees derived from one incipient species tree}
-
-% Simulation of posteriors}
-
-A BEAST2 posterior contains multiple parameter estimates (see figure \ref{fig:ess_distribution} for all),
-and each one has an ESS. An ESS of 200 is acceptable for any parameter estimate.
-The ESS of the tree likelihood should have such a decent ESS. Figure \ref{fig:ess_distribution_likelihood}
-shows the distribution of the likelihood ESS.
-
-\todo{How tp quote https://www.beast2.org/what-is-ess/ ? It states that not all variables
-need to have a high ESS, but 'The likelihoods (both of the tree and coalescent model) should have decent ESSs.}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Quantification of the error BEAST2 makes on all trees}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-To get a first global impression of the error BEAST2 makes, figure \ref{fig:error}
-shows a histogram of the nLTT statistics, colored by the expected mean
-duration of speciation. As BEAST2 assumes a mean duration of speciation of zero,
-it is expected that the nLTT statistic (a measure of error) is lowest. Zooming
-in on the left part of the distribution, there is no clean pattern.
-The right part of the tail of distribution shows that the biggest nLTT
-statistic values are found for high (but not highest) expected mean durations of
-speciation.
-
-\todo{figure \ref{fig:error_for_scr} shows the distribution of errors for all
-combinations of PBD parameters}
-
-As the expected correlation appears to be absent, figure 
-\ref{fig:error_expected_mean_dur_spec}, subfigure $a$, 
-shows the correlation between expected
-mean duration of speciation and the nLTT statistic. A linear fit shows
-that there is indeed an increase in error for an increases expected duration
-of speciation. A trendline using locally weighted smoothing, however,
-indicates a more complex relationship. Note that this figure only uses
-a subset of all measurements, consisting of $n_s$ nLTT statistic values 
-per posterior.
-
-Taking the average nLTT value per posterior frees us from computational
-constraints and allows us to use the full data set. The resulting
-plot, figure \ref{fig:error_expected_mean_dur_spec}, subfigure $b$, shows the same pattern:
-A linear fit shows
-that there is indeed an increase in error for an increases expected duration
-of speciation. But now, a trendline using locally weighted smoothing 
-follows this linear relationship closely.
-
-The inference error made is expected to be less for a longer DNA
-alignment length. 
-Figure \ref{fig:error_expected_mean_dur_spec_alignment_length} 
-show the correlation between expected mean duration of speciation 
-and the nLTT statistic for the two different DNA alignment lengths. 
-The first subfigure uses a sample of all $n_s$ individuals nLTT statistics,
-where the second subfigure takes only the posterior's 
-nLTT statistic mean value, for the complete dataset.
-
-Both figures show again that there is an 
-increase in error for an increases expected duration
-of speciation, when assuming a linear relation, and
-a more complex relationship when using locally weighted smoothing. 
-The figures also show that, as expected, 
-the error is less with more information. Additionally,
-a longer DNA alignment length results in a slower accumulation of error
-for increased expected mean duration of speciation.
-
-A higher ESS for longer DNA alignments may be an additional
-explanation for the observed slower accumulation of error
-for increased expected mean duration of speciation.
-Figure \ref{fig:ess_expected_mean_dur_spec_alignment_length}
-shows that longer DNA alignments have a higher likelihood ESS, 
-but this trend is weaker than in figure 
-\ref{fig:error_expected_mean_dur_spec_alignment_length} and
-the 95\% confidence intervals of the linearily fitted trendlines only
-stop overlapping for the highest expected mean duration of speciation.
-The locally weighted smoothing trendlines show a different trend,
-suggesting that alignment length has no effect on the
-likelihood ESS.
-
-\todo{Figure \ref{fig:error_alignment_length} shows the same pattern, seperated
-for all combinations of PBD parameters.}
-
-For each incipient species tree, two species trees were created, by taking
-either the youngest or oldest subspecies to represent each species.
-The effect of this is shown in figure \ref{fig:error_expected_mean_dur_spec_sampling}.
-Both subfigures show the
-error made for different $t_{\mean{ds}}$, where the first subfigure uses a sample
-of all $\Delta_{nLTT}$ values, where the second subfigure uses the $\mean{\Delta_{nLTT}}$
-of each posterior. In both subfigures, the linear and locally weighted smoothing
-have overlapping 95\% confidence intervals, hinting that there is no strong effect
-of the way subspecies are selected to represent their species.
-
-\note{Figure \ref{fig:error_sampling_representative} shows the same data, ordered
-by the different PBD parameters instead}
-
-The effect of having a low or high tree likelihood ESS is shown in 
-figure \ref{fig:figure_error_expected_mean_dur_spec_low_high_ess}. We
-define a low ESS as being below median, where a high ESS is defined as 
-above-median. Plotting (a sample of or the means of) the nLTT statistic for different mean
-duration of speciation for a low an high ESS, shows that there is no clear
-difference in error made for posteriors with either low or high tree likelihood ESSes.
-
-It is expected that the bigger phylogeny is simulated, the less the error
-will be. Figure \ref{fig:figure_error_tree_size} shows the relation between the
-number of taxa and the mean nLTT statistic. Although most simulated
-phylogenies have less than 250 taxa, there is a clear positive trend, 
-which is the opposite of the
-expactation: the bigger the phylogeny, the bigger the error made.
-As the nLTT statistic normalized for the number of taxa, some other
-mechanism must be involved.
-
-For each species tree, $n_a$ different alignments are simulated. For each of
-these alignments, $n_b$ posteriors are simulated, using the same alignment. 
-It is assumed that two posteriors from the same alignment
-give rise to the same parameter estimates distributions.
-distribution. Figure \ref{fig:figure_posterior_distributions_likelihood} 
-shows the distribution of Mann-Whitney U test p-values for each pair of 
-posteriors working on a same alignment, 
-using their tree likelihood estimates.
-
-The figure shows that the tree likelihood posterior distributions 
-from a same alignment nearly always
-are significantly different. Subfigure (b) and (c) show the distribution of
-tree likelihood estimates for a posterior pair with the lowest (b) and 
-highest (c) p-value.
-
-Posteriors based on the same alignment are also (next to estimated tree likelihood) 
-assumed to have a similar nLTT statistic distribution. 
-Figure \ref{fig:figure_posterior_distribution_nltt} 
-shows the distribution of Mann-Whitney U test p-values for each pair of 
-posteriors working on a same alignment, 
-using their nLTT statistics.
-
-The figure shows that the nLTT statistic distributions 
-from a same alignment are commonly different and commonly the same. 
-Subfigures $(b)$ and $(c)$ show the distribution of
-nLTT statistics for a posterior pair with the lowest $(b)$ and 
-highest $(c)$ p-value.
-
-The nLTT statistic normalizes a phylogeny to have a crown age of 1.
-To prevent overlooking patterns in posterior crown ages, 
-figure \ref{fig:figure_posterior_distribution_crown_ages}
-shows the distribution of crown ages of all simulations.
-Almost all simulations have a crown age close to $0.15$, but there
-are rare cases with crown ages above $0.2$, ranging to $83.4$.
-
-\todo{I think it is very suspicious that the range centers on 0.15, when
-the crown age is 15. Check this!}
-
-Again, it is assumed that posteriors based on the exact same alignment
-have a similar crown age distribution. Of all posterior pairs, the Mann-Whitney
-U test p-values are shown in figure \ref{fig:figure_posterior_distribution_crown_ages_p_values}.
-It shows that crown age distributions being different is common and these
-being the same is uncommon.
-
-It is expected that if a tree is created by a Birth-Death process, BEAST2 is able
-to reasonably estimate those parameters.
+\begin{figure}[]
+  \includegraphics[width=0.9\textwidth]{fig_does_protractedness_matter.png}
+\end{figure}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-This article has quantified the error made in inference when (1) nature would follow the
-protracted speciation model, and (2) her DNA sequences would be analysed with a birth-death
-model.
-
-% The PBD model is just one implementation of speciation taking time
-
-The key feature of the PBD model is that it encompasses speciation taking time in the
-simplest way possible (that allows for mathematical analysis). The PBD model allows for
-multiple incipient species stages (where its earliest version had infinite
-stages \cite{rosindell2010protracted}). This research assumes that there is only one
-incipient species stage. One could argue that speciation takes place in more
-biological stages \todo{Add reference}, but we tried to use the simplest incarnation of the PBD possible.
-Also, the PBD model can also be viewed as a model of species recognition \todo{?Purvis et al., 2009},
-in which (human) recognition needs to build up, which we consider to be sufficiently
-encapsulated in one single stage. Future research may investigate if adding multiple incipient
-species stages for simulating nature changes the error made by (non-protracted) birth-death models.
-
-% Equal and constant rates
-
-The PBD model allows for multiple incipient species stages. 
-Each stage has its own extinction and speciation-initiation and 
-speciation-completion rate (except for good species, which lacks 
-a speciation-completion rate). These rates are assumed to be constant
-in time and equal for all stages. We chose to use the constant-rate
-birth-death model to keep this investigation as simple as possible.
-Future research may investigate whether a biologically more realistic
-rate model, for example, by adding diversity-dependence, results in
-the same inference error.
-
-We chose to set the extinction and speciation-initiation rates equal for all stages,
-again for choosing the simplest model. We expect these rates to differ in reality:
-incipient species have low initial abundances, which has an effect on extinction
-and speciation-initiation rates.
-
-In this research, we nevertheless set these rates equal. 
-Not only because it helps constrain the parameter space investigated,
-also paraphylies would pop up in the simulated trees. This would violate more assumptions
-of the BEAST2 model used as a prior and increase the complexity of the analysis.
-
-% Sampling species trees from incipient species trees
-
-An incipient species tree may have multiple incipient lineages from the
-same ancestor. We constructed two species trees from each incipient species
-tree, by selecting the youngest and oldest subspecies to represent a species.
-This selection maximizes the chance the species trees are different, where
-random sampling of representatives would also be a valid approach, closer
-to what would happen when scientists take measurements in the field. Our
-approach increases the error made.
-
-% Nucleotide substitution model
-
-For the simulation of DNA alignments, the Jukes-Cantor nucleotide substitution model
-is used. This substitution model is a simplification of the DNA mutation processes
-happening in reality \todo{Add reference}. In this simulation research, however, we could also
-state in BEAST2 that this nucleotide substitution model is used. Thus, we kept
-the simulation of the nucleotides in time as simple as possible, and used that
-knowledge in later steps. We expect that any nucleotide substitution model, when
-used in both alignment generation and in phylogeny inference, will do equally well,
-except that the full computational pipeline would take needlessly longer.
-
-% Allopatric speciation only
-
-When a speciation event is initiated, this investigation assumes no gene flow
-between the two incipient lineages. This follows the allopatric mode of speciation, 
-which is assumed to be most prominent of nature \todo{Add reference}. It would be interesting to
-allow gene flow between related incipient species, 
-so also peripatric and sympatric modes of speciation would be allowed for.
-We predict that this would increase the error in inference.
-
-% Why not put the tree prior into an MCMC tool directly?
-One could argue we should have written a PBD tree prior for an MCMC phylogenetic tool.
-We did not do that for two reasons. First, we think it is rational to first
-measure if there is need for a more complex tree prior. Second, 
-we could not have done so, as a phylogenetic tool like BEAST2 assumes species
-trees, where our input consists of incipient species trees. Before this
-research, we were unsure how to sample species representatives. This research
-has shown that any sampling is just as good, opening up avenues for
-implementing a PBD species tree prior.
-
-% BEAST2 is just one of many tools
-
-BEAST2 is one of many MCMC phylogenetic tools with its
-own internal technicalities. Similar tools are the R phytools
-package \cite{phytools}, BAli-Phy \cite{suchard2006bali}
-and BayesPhylogenies \cite{pagel2007bayesphylogenies}, to name a few.
-BEAST2 was chosen, because (1) it's FOSS, (2) its software
-architecture is modular, encouraging extension, (3) its development
-team follows good practices in software development.
-
-% BEAST2 parameter settings: strict clock
-
-\todo{elaborate}
-
-Use of strict clock is clean, but unrealistic.
-
-\todo{elaborate}
-
-% Alignment generation
-
-Use of JC69 is rational, but unrealistic.
-
-% Use for empiricists
-
-\todo{elaborate}
-
-This reseach shows that a DNA sequence length of 10k [is good].
-[must be effective DNA sequence length]
-[no indels generated]
-
-% Low number of incipient species trees
-
-\todo{elaborate}
-
-For each point in parameter space, twenty incipient species tree are created.
-Was fitted to 1000 jobs.
-
-% Failed simulations
-
-Not all parameter combinations resulted in posteriors with a high ESS,
-due to computational constraints. We think this is a valid approach:
-would a phylogeny get too big, we as scientists would cut it to
-pieces that are analytically tractable. Because these simulations
-are only a minority (as shown in the Results section), we do not
-expect this to influence our conclusions.
-
-% nLTT statistic is just one of many summary statistics
-
-We used the nLTT statistic for quantifying the error between the
-true and inferred phylogeny. There are simpler statistics that
-can do the same, like the gamma \cite{pagel1999inferring}, 
-delta-r \cite{pigot2010shape} \todo{check citation}, 
-tree balance \todo{Add reference} or
-Normalized Rooted Branch Score [Drummond] statistic, 
-or more complex, like the spectral analysis [Hendy and Penny, 1993].
-We consider those other summary statistics equally good, we just picked the
-one we thought of the most informative per unit of added computational time.
-
-% Add:
-%
-% Computational Performance and Statistical Accuracy of *BEAST and Comparisons with Other Methods (by Huw A. Ogilvie Joseph Heled Dong Xie Alexei J. Drummond)
-% cites rooted branch score, RBS, as being from Heled and Bouckaert 2013, `Looking for trees in the forest: Summary tree from posterior samples`
 
@@
 
-RJCB and RSE conceived the idea for the package. RJCB created and tested the package, and wrote the first draft of the manuscript. RSE contributed substantially to revisions.
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Tables}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{table}
-  \centering 
-  \begin{tabular}{l l}
-    \hline
-    Symbol & Description \\
-    \hline
-    \hline
-    $b_g$ & Speciation initiation rate of a good species (probability per lineage, per million year) \\
-    $b_i$ & Speciation initiation rate of an incipient species (probability per lineage, per million year) \\
-    $\lambda$ & Speciation completion rate (probability per lineage, per million year) \\
-    $\mu_g$ & Extinction rate of a good species (probability per lineage, per million year) \\
-    $\mu_i$ & Extinction rate of an incipient species (probability per lineage, per million year) \\
-    \hline
-    $t_c$ & Crown age (million years) \\
-    $r$ & Mutation rate (probability of mutation per nucleotide per million years) \\
-    $l_a$ & DNA alignment length (base pairs) \\
-    \hline
-    $n_i$ & Number of incipient species trees per parameter set \\
-    $n_a$ & Number of alignments per species tree \\
-    $n_b$ & Number of BEAST2 runs per alignment \\
-    \hline
-    $n_s$ & Number of MCMC samples \\
-    $i_s$ & MCMC sampling interval \\
-    \hline
-    $n_{ti}$ & Number of taxa in an incipient species tree \\
-    $n_{t}$ & Number of taxa in a (good) species tree \\
-    $t_{\mean{ds}}$ & Average duration of speciation \\
-    $t_{sc}$ & Speciation completion time \\
-    $\Delta_{nLTT}$ & nLTT statistic value \\
-    $\mean{\Delta_{nLTT}}$ & Average nLTT statistic value of a posterior \\
-    $\median{\Delta_{nLTT}}$ & Median nLTT statistic value of a posterior \\
-    \hline
-  \end{tabular}
-  \caption{
-    Parameter descriptions. The sections are (1) the biological parameters, (2) the
-    biological tuning parameters, (3) parameters involved in replication, (4) MCMC
-    tuning parameters, and (5) other symbols.
-  }
-  \label{table:parameter_descriptions}
-\end{table}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{table}
-  \centering 
-  \begin{tabular}{l l}
-    \hline
-    Parameter             & Values \\
-    \hline
-    \hline
-    $b = b_g = b_i$       & 0.1, 0.5, 1.0 \\
-    $\lambda$             & 0.1, 0.3, 1.0, $10^6$ \\
-    $\mu = \mu_g = \mu_i$ & 0.0, 0.1, 0.2, 0.4 \\
-    \hline
-    $t_c$                 & 15 \\
-    $r$                   & 0.01 \\
-    $l_a$                 & $10^3$, $10^4$ \\
-    \hline
-    $n_i$                 & 20 \\
-    $n_a$                 & 2 \\
-    $n_b$                 & 2 \\
-    \hline
-    $n_s$                 & $10^3$ \\
-    $i_s$                 & $10^3$ \\
-    \hline
-  \end{tabular}
-  \caption{
-    Parameter values. The sections are (1) the biological parameters, (2) the
-    biological tuning parameters, (3) parameters involved in replication, 
-    and (4) MCMC tuning parameters.
-  }
-  \label{table:parameter_values}
-\end{table}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Figures}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{figure}
-  \centering
-  \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=4cm, semithick]   
-  \tikzstyle{every state}=[]
-  \node[state] (A)              {Good};   
-  \node[state] (B) [right of=A] {Incipient};   
-  \node[state] (C) [below of=B] {Extinct};   
-  \path (A) edge [bend right] node {$b_g$} (B)
-        (A) edge [bend right] node {$\mu_g$} (C)
-        (B) edge [loop above] node {$b_i$} (B)
-        (B) edge [bend right] node {$\lambda$} (A)
-        (B) edge [bend right] node {$\mu_i$} (C); 
-  \end{tikzpicture}
-
-  \caption{
-    The states and transitions of a species in the PBD model.
-    $b_i$: speciation-initiation rate of incipient species. 
-    $b_g$: speciation-initiation rate of good species. 
-    $\lambda$: speciation completion rate. 
-    $\mu_i$: extinction rate of incipient species. 
-    $\mu_g$: extinction rate of good species. 
-    Figure after Etienne et al, 2014, Evolution
-  }
-  \label{fig:pbd_states}
-\end{figure}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{figure}
-  \centering 
-  \resizebox {0.8\textwidth} {!} {
-    \CreateTikzFigureExperiment{} 
-  }
-  \caption{
-    Overview of the experiment: 
-    A: per parameter set, there are $n_i$ incipient species trees created. 
-    B1: constructing a species trees by using the youngest subspecies $A1$ to represent good species $A$.
-    B2: constructing a species trees by using the oldest subspecies $A2$ to represent good species $A$.
-    C1: constructing one of the $n_a$ alignments of B1. 
-    C2: constructing one of the $n_a$ alignments of B2. 
-    D1: reconstructing a species trees from alignment B1.
-    D2: reconstructing a species trees from alignment B2.
-    Comparisons are done between a species tree and its reconstructed species tree, 
-    for example, between B1 and D1.
-    All phylogenies are drawn at the same scale.
-  }
-  \label{fig:experiment}
-\end{figure}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{figure}
-  \centering 
-  \CreateTikzFigureSampling{}
-  \caption{
-    Sampling an incipient species tree. 
-    A) At stem age $t_0$, $A1$ is present.
-    At crown age $t_1$, its derived incipient species $A2$ is born. 
-    At $t_2$, $A2$ itself speciates into a new (unnamed) species, 
-    that becomes good species $B$ at $t_3$. At present time, $t_4$, there
-    exist two possible representatives of species $A$. 
-    B) Using the oldest representative, $A1$, to represent species $A$, the crown
-    age of the resulting species tree lies at $t_1$. 
-    C) Using the youngest representative, $A2$, to represent species $A$, the crown
-    age of the resulting species tree lies at $t_2$. 
-  }
-  \label{fig:sampling}
-\end{figure}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\listoftodos
-
 \end{document}

### 2018-03-29 article.tex
Wrote introduction

@@
 \usepackage{authblk}
-\title{The error when inferring phylogenies with incipient species by a Birth-Death model}
+\title{Stage 1 Registered Report: The error when inferring phylogenies with incipient species by a Birth-Death model}
 % \subtitle{Should protracted speciation be incorporated in phylogenetic tree construction methods?}
@@
 
-\begin{abstract}
-
-  The construction of phylogenies helps us answer evolutionary biological
-  questions.
-  Our current phylogenetic tools ignore the fact that speciation takes time,
-  which has an effect unknown in phylogeny reconstruction.
-  Here, we simulate true incipient species trees and their corresponding
-  DNA alignments, from which we measure the reconstruction of the phylogeny
-  using a standard birth-death model.
-  We measure the errors that the Bayesian phylogenetic software tool BEAST2
-  gives when recovering simulated phylogenies, for different times-to-speciate,
-  under a range of additional parameter settings.
-  It has been found that branch lengths are consistently and strongly 
-  underestimated for biologically relevant parameters.
-  This research shows that protractedness is a complexity of nature that
-  cannot always be ignored and should be incorporated in our phylogenetic
-  tools.
-
-\end{abstract}
-
-{\bf Keywords:} computational biology, evolution, phylogenetics, BEAST2, R
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-What is the effect of sampling either the youngest or oldest species-to-be
-to represent a species?
+Phylogenetics enables us to infer which species lived when and
+estimate how long it took for a species to give rise to a new
+lineage. Phylogenies show us taxa that have completed speciation, yet
+those taxa only: species that are in the process of speciation are 
+overlooked.
 
-What is the effect of protractedness on the inference error we make?
+There are multiple reasons why a speciation event, during
+or after its course, goes unnoticed. It may be that the build-up
+of reproductive isolation is still in an intermediate stage. But it
+may also be that speciation has completed, but we, humans, have not
+yet observed it. One example are the four African giraffe species, 
+that have been thought to have been one single species up until 2016,
+where the MRCA of the two most recent species goes back 1.6 million years.
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Pilot experiments}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+It took some time until the Birth-Death model, a phylogenetic model
+assuming constant rates of speciation and extinction, got extended
+by a version that incorporates the fact that speciation takes time.
+This extension is called the Protracted Birth-Death model. In this
+model, a branching event does not give rise to a new species, but to
+a new species-to-be, called an incipient species. Such an incipient
+species may go extinct, or finish its speciation to become a good species.
 
-Do the DNA sequence length and mutation rate (as described in the Methods)
-give a good enough inference? Answer: pick the setting with the highest
-expected number of taxa (highest speciation initiation and completion rate, 
-no extinction). Simulate 1000 trees and obtain an nLTT distribution for a
-reference DNA sequence length and mutation rate. Increasing DNA sequence length
-or mutation rate increases the resolution of the experiment. This resolution
-must not be too low (this would cost accuracy), nor too high (this would cost
-too much computational resources). If doubling DNA length or DNA mutation rate
-results in an nLTT distribution with 90% overlap with the reference distribution,
-in 900 of 1000 cases, the (lower) reference resolution is considered high enough.
+Although we know speciation takes time, most of our theoretical 
+phylogenetic modeling ignores this fact. To the best of our knowledge,
+there is no tool incorporating this fact. One reason may be to ease
+computation, enabling researchers to work on bigger phylogenies (with
+more power statistical power) instead. Additionally, speciation being
+protracted is only playing a role in the present. As seen from the present,
+all older species have finished speciation, as we -by definition- recognize
+these species as such.
 
-What is the MCMC chain length that gives a posterior ESS of 200 in all cases?
-Answer: run all simulations, measure the posterior's ESS. If at least one 
-ESS is lower than 200, elongate all MCMC chain lengths by a fraction 
-to predict to increase it to 220 (for example, would the lowest ESS be 165, 
-making the MCMC chains 33\% longer would bring it to (165\%*1.33=) 220. We use
-220 to prevent an asymptotic approaching to 200.
+There are problems with ignoring protracted speciation. A very pragmatic
+example is the conservation of the now four smaller groups of giraffes,
+instead of one big group. But from a more theoretical perspective, the
+observed decline of speciation rate in time (excluding the contemporary human-driven
+mass extinction event) may be explained by these new species not being 
+recognized yet. Another problem is that the MRCA of young species is estimated 
+incorrectly, as in the BD model a species can be ultra-young (as speciation
+is assumed to be instantaneous), where the PBD model will place these MRCAs
+further back in time (as two events need to have finished completion).
+
+From a theoretical point of view, there is another reason why protracted
+speciation has yet become a member of our toolbox: there is no framework
+where it fits in: a framework assumes either an analysis at the species or
+subspecies level. When picking a tool that does its analysis at the
+species level, there is no protocol how to select which incipient species
+gets to represent the full species it is still a member of. When picking
+a tool at the sub-species level, the incipient state disappears as a
+sub-species lineage, and the process degrades to an ordinary birth-death 
+process.
+
+This research's goals are two-fold, which are (1) to quantify the
+inference error we make with contemporary tools assuming a Birth-Death
+process, and (2) to establish a protocol how to sample incipient species
+to represent a species in such a tool.
+
+To quantify the inference error, we simulate 'true' incipient species
+trees, sample a 'true' species trees from those, simulate a DNA alignments 
+that follows a species tree's ontogeny, then infer a posterior of
+phylogenies from each alignment. The 'true' species tree is compared to
+the posterior's phylogenies with the nLTT statistic, which can be viewed
+as a measure of similarity. We expect this error to increase with an
+increasing protractednedness (that is, a low speciation completion rate),
+but as the extent of this error has never been measured, we will be the
+first to measure it.
+
+To establish a protocol how to sample incipient species to represent species,
+we compare two sampling methods that are opposites of one another: one
+method is to select the oldest sub-species member to represent a species,
+the other method picks the youngest sub-species member. As we do not have
+a way to determine which sampling method will be better, all we determine
+is if the sampling methods are different. We expect these
+sampling methods to be equivalent. This research measures the extent to
+which this is true.
 
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Pilot experiments}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+Do the DNA sequence length and mutation rate (as described in the Methods)
+give a good enough inference? Answer: pick the setting with the highest
+expected number of taxa (highest speciation initiation and completion rate, 
+no extinction). Simulate 1000 trees and obtain an nLTT distribution for a
+reference DNA sequence length and mutation rate. Increasing DNA sequence length
+or mutation rate increases the resolution of the experiment. This resolution
+must not be too low (this would cost accuracy), nor too high (this would cost
+too much computational resources). If doubling DNA length or DNA mutation rate
+results in an nLTT distribution with 90% overlap with the reference distribution,
+in 900 of 1000 cases, the (lower) reference resolution is considered high enough.
+
+What is the MCMC chain length that gives a posterior ESS of 200 in all cases?
+Answer: run all simulations, measure the posterior's ESS. If at least one 
+ESS is lower than 200, elongate all MCMC chain lengths by a fraction 
+to predict to increase it to 220 (for example, would the lowest ESS be 165, 
+making the MCMC chains 33\% longer would bring it to (165\%*1.33=) 220. We use
+220 to prevent an asymptotic approaching to 200.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2018-04-03 article.tex
No pilot study

@@
 of this research). Additionally, we assume a MRCA prior with a normal distribution
-with a mean of the crown age, and a standard deviation of 0.001 time units (which
-equates to 1000 years). After discarding a burn-in of 10\%, the MCMC algorithm must 
+with a mean of the crown age, and a standard deviation of 0.001 time units. 0.001
+time units equates to 1000 years, which is the same resolution as which the
+alignment is created. After discarding a burn-in of 10\%, the MCMC algorithm must 
 result in an ESS of the posterior of at least 200 and the MCMC chain length is 
-chosen as such that this is the case for all simulations.
+chosen as such that this is the case for all simulations. For simulations
+with an ESS below 200, their MCMC chain length is elongated. An
+MCMC chain is elongated by the proportion that is expected to increase it to 220. 
+For example, would an ESS be 165, increasing its MCMC chain length by 33\% 
+would bring it to (165\%*1.33=) 220. We use
+220 as a safeguard to prevent an asymptotic approach to 200.
 
-The similarity of each phylogeny in the posterior with 'the true species tree'
-is determined with the nLTT statistic. In this way, a distribution of phylogenies
-is transformed to a distribution of nLTT statistics.
+The nLTT statistic is used to quantify the similarity of each posterior's phylogeny 
+with 'the true species tree'. In this way, a distribution of phylogenies
+is transformed to a distribution of nLTT statistics. With this, the generative
+model of this experiment has been completely described, from parameter estimates
+to data (nLTT statistics) generated.
 
-To measure if sampling (either the youngest or oldest species-to-be to represent
-a species) matters, for all parameter combinations that have incipient species,
+To quantify the effect of sampling (either the youngest or oldest species-to-be to represent
+a species), for each parameter set that allows for incipient species,
 we simulate 1000 incipient species trees that result in different species trees
 when sampled either way. From each species tree, we compare the nLTT distributions
-derived from the posterior. We quantify the Kullback-Leibler divergence (more
-or less: the weighted fraction of overlapping nLTT 
-values, which is zero if the two distributions are completely seperate, and close
-to one if they overlap). For each parameter combination, we take the average
-divergence. We assign verbal labels to this average divergence as such: 
-0.0-0.25: clear effect, 0.25-0.5: strong effect, 0.5-0.75: mild effect, 
-0.75-1.0: no effect. Per parameter setting, we show the average divergence and the 
-verbal label. The distribution of these divergences is shown in the 
-supplementary information.
+for the two sampling methods. These distributions are compared 
+using the method described in [Kruschke 2013], with a non-informative
+prior. In brief: if the 95\% Highest Density Interval of the differences between 
+these two distributions excludes zero, it is credible the distributions are different.
 
@@
 non-protracted setting is used as the baseline. That distribution is compared
-with each protracted setting. We quantify the Kullback-Leibler divergence 
-of these nLTT distributions as described above. Per protracted parameter setting, 
-we generate 1000 replicate incipient species trees, each sampled randomly. We show the 
-average divergence of these 1000 replicates and the verbal label.
-The distribution of these divergences is shown in the supplementary information.
+with each protracted setting. This is done the same way as with the two
+different sampling methods. [NOTE: we expect nLTT to go up for increasing SCR,
+this should be in the method].
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Pilot experiments}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-Do the DNA sequence length and mutation rate (as described in the Methods)
-give a good enough inference? Answer: pick the setting with the highest
-expected number of taxa (highest speciation initiation and completion rate, 
-no extinction). Simulate 1000 trees and obtain an nLTT distribution for a
-reference DNA sequence length and mutation rate. Increasing DNA sequence length
-or mutation rate increases the resolution of the experiment. This resolution
-must not be too low (this would cost accuracy), nor too high (this would cost
-too much computational resources). If doubling DNA length or DNA mutation rate
-results in an nLTT distribution with 90% overlap with the reference distribution,
-in 900 of 1000 cases, the (lower) reference resolution is considered high enough.
-
-What is the MCMC chain length that gives a posterior ESS of 200 in all cases?
-Answer: run all simulations, measure the posterior's ESS. If at least one 
-ESS is lower than 200, elongate all MCMC chain lengths by a fraction 
-to predict to increase it to 220 (for example, would the lowest ESS be 165, 
-making the MCMC chains 33\% longer would bring it to (165\%*1.33=) 220. We use
-220 to prevent an asymptotic approaching to 200.
+[NOTE: We do have more info, from Rampal's earlier articles. Perhaps this 
+info can be put in]
 

### 2018-04-04 issue_6.R
40K rows in dataset, fix #6

@@ -1,0 +1,17 @@
+ncols <- 1100
+nrows <- 40000
+ncells <- ncols*nrows
+testit::assert(ncells < 2^32-1)
+mem_use <- ncells * object.size(3.14)
+format(mem_use, units = "Gb")
+
+# Computer will freeze if you ignore this warning ..
+testit::assert(mem_use < 2^32-1)
+
+df <- data.frame(matrix(nrow = nrows, ncol = ncols, data = seq(1,ncells)))
+df[2,1:10]
+filename <- tempfile()
+write.csv(x = df, file = filename, row.names = FALSE)
+df2 <- read.csv(file = filename)
+all.equal(df2, df, tolerance = 0.1)
+file.info(filename)

### 2018-04-04 article.tex
EOD

@@
 
+This exploratorary research will produce a data set that maps the biological
+parameters to the error made. The data set will consist out of two files:
+one containing raw data, the other summarized data. 
+A data file should be convenient to handle, thus as an upper limit for this data set's size, 
+we use the maximum size of a table (called a data frame) of the R programming 
+language. Within R, such a table may have maximally $2^{32}-1$ cells and occupy
+less then 2 gigabytes in memory. 
+The raw data file will contain parameters and 1000 nLTT statistics values.
+Each row will contain the parameter set and these nLTT statistics.
+From the constraints of the R programming language, this allows for 40K rows. 
+The file with the summarized data will provide summary statistics of the raw
+data. Each row will contain a parameter set, summary statistics and other
+measurements. See the appendix, chapter 'Data set specifications', for the full list.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
+[NOTE: as we know the expected number of lineages through time for both BD and PBD
+model, we expect it to fall with the 95\% Highest Density Interval. It would also
+increase support for Rampal's choice of parameter values (next to their 
+computational feasability)]
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\appendix
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Data set specifications}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+The raw data file will consist of 40K rows. On each row there will be 1012 values:
+
+\begin{itemize}  
+  \item 12 simulation parameters:
+  \begin{enumerate}  
+    \item speciation initiation rate
+    \item speciation completion rate
+    \item extinction rate
+    \item crown age
+    \item crown age sigma
+    \item sampling method
+    \item mutation rate
+    \item sequence length
+    \item mcmc length
+    \item tree sim rng seed
+    \item alignment rng seed
+    \item beast2 rnd seed
+  \end{enumerate}
+  \item 1000 nLTT statistics
+\end{itemize}
+
+The summarized data file will consist of 40K rows. On each row there will be 29 values:
+
+\begin{itemize}  
+  \item 12 simulation parameters:
+  \begin{enumerate}  
+    \item speciation initiation rate
+    \item speciation completion rate
+    \item extinction rate
+    \item crown age
+    \item crown age sigma
+    \item sampling method
+    \item mutation rate
+    \item sequence length
+    \item mcmc length
+    \item tree sim rng seed
+    \item alignment rng seed
+    \item beast2 rnd seed
+  \end{enumerate}
+  \item number of taxa in incipient species tree (also: number of subspecies)
+  \item number of taxa in species tree (also: number of species)
+  \item mean nLTT statistic
+  \item median nLTT statistic
+  \item lower bound of 95\% HPD value of nLTT statistic
+  \item upper bound of 95\% HPD value of nLTT statistic
+  \item 11 ESSes of all parameters estimated by BEAST2:
+  \begin{enumerate}  
+    \item posterior
+    \item likelihood 
+    \item prior 
+    \item treeLikelihood 
+    \item TreeHeight 
+    \item BirthDeath 
+    \item BDBirthRate 
+    \item BDDeathRate 
+    \item logP.mrca 
+    \item mrcatime 
+    \item clockRate
+  \end{enumerate}
+\end{itemize}
+
 \end{document}

### 2018-04-04 issue_6.R

@@ -1,12 +1,14 @@
-ncols <- 1100
+ncols <- 1050
 nrows <- 40000
 ncells <- ncols*nrows
 testit::assert(ncells < 2^32-1)
 mem_use <- ncells * object.size(3.14)
-format(mem_use, units = "Gb")
+format(mem_use, units = "Mb")
 
 # Computer will freeze if you ignore this warning ..
-testit::assert(mem_use < 2^32-1)
+testit::assert(mem_use < ((2^32)-1))
+
+((2^32)-1) / as.double(object.size(3.14)) / ncols
 
 df <- data.frame(matrix(nrow = nrows, ncol = ncols, data = seq(1,ncells)))
 df[2,1:10]

### 2018-04-05 article.tex
EOD

@@
 
-This research's goals are two-fold, which are (1) to quantify the
-inference error we make with contemporary tools assuming a Birth-Death
-process, and (2) to establish a protocol how to sample incipient species
-to represent a species in such a tool.
+This exploratory research's goal is to produce a data set that 
+quantifies the inference error we make from choosing an overly
+simplistic prior. In this context, we measure the error made on
+phylogenies in which speciation takes time, when using a Birth-Death
+prior.
 
@@
 that follows a species tree's ontogeny, then infer a posterior of
-phylogenies from each alignment. The 'true' species tree is compared to
+phylogenies from each alignment, using an overly simplistic Birth-Death
+prior. The 'true' species tree is compared to
 the posterior's phylogenies with the nLTT statistic, which can be viewed
-as a measure of similarity. We expect this error to increase with an
+as a measure of similarity. We predict this error to increase with an
 increasing protractednedness (that is, a low speciation completion rate),
-but as the extent of this error has never been measured, we will be the
-first to measure it.
+but we cannot predict the the extent of this error, as it has never been 
+measured.
 
-To establish a protocol how to sample incipient species to represent species,
-we compare two sampling methods that are opposites of one another: one
-method is to select the oldest sub-species member to represent a species,
-the other method picks the youngest sub-species member. As we do not have
-a way to determine which sampling method will be better, all we determine
-is if the sampling methods are different. We expect these
-sampling methods to be equivalent. This research measures the extent to
-which this is true.
-
-This exploratorary research will produce a data set that maps the biological
-parameters to the error made. The data set will consist out of two files:
+The data set produced by the exploratory reseach will consist out of two files:
 one containing raw data, the other summarized data. 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Abbreviations}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\begin{table}
+  \centering 
+  \begin{tabular}{l l}
+    \hline
+    Symbol & Description \\
+    \hline
+    \hline
+    $b_g$ & Speciation initiation rate of a good species (probability per lineage, per million year) \\
+    $b_i$ & Speciation initiation rate of an incipient species (probability per lineage, per million year) \\
+    $\lambda$ & Speciation completion rate (probability per lineage, per million year) \\
+    $\mu_g$ & Extinction rate of a good species (probability per lineage, per million year) \\
+    $\mu_i$ & Extinction rate of an incipient species (probability per lineage, per million year) \\
+    \hline
+    $t_c$ & Crown age (million years) \\
+    $r$ & Mutation rate (probability of mutation per nucleotide per million years) \\
+    $l_a$ & DNA alignment length (base pairs) \\
+    \hline
+    $n_i$ & Number of incipient species trees per parameter set \\
+    $n_a$ & Number of alignments per species tree \\
+    $n_b$ & Number of BEAST2 runs per alignment \\
+    \hline
+    $n_s$ & Number of MCMC samples \\
+    $i_s$ & MCMC sampling interval \\
+    \hline
+    $n_{ti}$ & Number of taxa in an incipient species tree \\
+    $n_{t}$ & Number of taxa in a (good) species tree \\
+    $t_{\mean{ds}}$ & Average duration of speciation \\
+    $t_{sc}$ & Speciation completion time \\
+    $\Delta_{nLTT}$ & nLTT statistic value \\
+    $\mean{\Delta_{nLTT}}$ & Average nLTT statistic value of a posterior \\
+    $\median{\Delta_{nLTT}}$ & Median nLTT statistic value of a posterior \\
+    \hline
+  \end{tabular}
+  \caption{
+    Parameter descriptions. The sections are (1) the biological parameters, (2) the
+    biological tuning parameters, (3) parameters involved in replication, (4) MCMC
+    tuning parameters, and (5) other symbols.
+  }
+  \label{table:parameter_descriptions}
+\end{table}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Parameters}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\begin{table}
+  \centering 
+  \begin{tabular}{l l}
+    \hline
+    Parameter             & Values \\
+    \hline
+    \hline
+    $b = b_g = b_i$       & 0.1, 0.5, 1.0 \\
+    $\lambda$             & 0.1, 0.3, 1.0, $10^6$ \\
+    $\mu = \mu_g = \mu_i$ & 0.0, 0.1, 0.2, 0.4 \\
+    \hline
+    $t_c$                 & 15 \\
+    $r$                   & 1000 \\
+    $l_a$                 & $15000$ \\
+    \hline
+    $n_i$                 & 178 \\
+    $n_a$                 & 2 per species tree\\
+    $n_b$                 & 2 per alignment\\
+    \hline
+    $n_s$                 & $10^3$ states (sampled)\\
+    $i_s$                 & $10^3$ states (between samples)\\
+    \hline
+  \end{tabular}
+  \caption{
+    Parameter values. The sections are (1) the biological parameters, (2) the
+    biological tuning parameters, (3) parameters involved in replication, 
+    and (4) MCMC tuning parameters.
+  }
+  \label{table:parameter_values}
+\end{table}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Data set specifications}
@@
   \item number of taxa in incipient species tree (also: number of subspecies)
+  \item would there be a difference in brach lengths between the two sampling methods?
   \item number of taxa in species tree (also: number of species)

### 2018-04-05 article.tex
Exploratory

@@
 
-This exploratory research's goal is to produce a data set that 
-quantifies the inference error we make from choosing an overly
-simplistic prior. In this context, we measure the error made on
-phylogenies in which speciation takes time, when using a Birth-Death
-prior.
+This exploratory research's goal is to produce a 
+usefull and balanced data set, that quantifies the inference error 
+we make from choosing an overly simplistic prior. In this context, 
+we measure the error made on phylogenies in which speciation takes 
+time, when using a Birth-Death prior.
 
-To quantify the inference error, we simulate 'true' incipient species
-trees, sample a 'true' species trees from those, simulate a DNA alignments 
-that follows a species tree's ontogeny, then infer a posterior of
-phylogenies from each alignment, using an overly simplistic Birth-Death
-prior. The 'true' species tree is compared to
-the posterior's phylogenies with the nLTT statistic, which can be viewed
-as a measure of similarity. We predict this error to increase with an
-increasing protractednedness (that is, a low speciation completion rate),
+The parameter space the data set explores is inspired on 
+[Etienne et al, 2014] for two reasons: (1) the academic
+reason of being able to compare results, and (2) the practical
+reason that the simulated phylogies will have enough taxa to
+be informative, but not too big to be computationally overdemanding.
+This research uses a complete superset of those parameter settings.
+For the speciation initiation rate, we'll use $0.1$, $0.5$ and $1.0$ 
+speciation initiation events per (good species) lineage per time unit, 
+extending the set of
+values used from one ($0.5$) to three. 
+The speciation completion
+rates used are $0.1$, $0.3$, $1.0$ and $10^6$ speciation completion
+events per (incipient species) lineage per time unit, in which only
+the last value is added as a computational proxy for infinity, which
+makes the experiment fall back to a Birth-Death model (as speciation
+is again instantaneous), allowing to measure the baseline error.
+The extinction rates used are $0.0$, $0.1$, $0.2$ and $0.4$ 
+extinction events per (good or incipient) lineage per time unit,
+in which $0.4$ is a new extinction rate.
+We investigate these parameters in a factorial fashion, except for
+(1) those combinations in which the extinction rate equals or exceeds
+the speciation initiation rate, and (2) the difference between
+speciation initiation rate and extinction rate exceeds 0.8. This is
+done to prevent overly taxon-poor and taxon-rich phylogenies respectively.
+
+From each biological parameter set, a protracted birth-death tree is simulated,
+with the same crown age as [Etienne et al., 2014] of 15 million years. 
+Each protracted birth-death tree uses a different random number
+generatior seed, and thus will be unique, resulting in a balanced 
+data set.
+
+From an incipient species tree, we sample a species tree
+by picking either the youngest or oldest
+species-to-be to represent the full species.
+We use either sampling method in 50\% of all cases. We chose not to use
+a random species-to-be to represent a full species as to maximize variance
+in this aspect. Note that only in few cases (when speciation completion rate
+is low) the sampling method has an effect on the branch length distribution
+of the sampled species tree, as this requires a very specific scenario.
+
+From a species tree, we simulate a DNA alignment that has the same history
+as the phylogeny. The nucleotides of the DNA alignment follow a JC69
+nucleotide substitution model, in which all nucleotide-to-nucleotide transitions
+are equally likely. Although this may seem as a simplification, in our Bayesian
+inference (see below) we can use this exact site model as a prior, which we 
+know is the truth.
+The DNA sequence length and mutation rate was chosen as such to provide one
+nucleotide change per 1000 years per lineage. As our time unit is in 
+millions of years, this equates to a rate of 1000 nucleotide substitutions 
+changes per time unit per lineage. To provide for each substitution being unique,
+we simulate as much DNA nucleotides as expected mutations, which equals the
+nucleotide substitution rate times the crown age, which equals 15K. As this
+mutation rate is constant over all branches, it in effect follows a strict 
+clock model, which we will specify as the known clock model prior later.
+
+From an alignment, we run a Bayesian analysis and create a posterior, 
+using the phylogetic tool BEAST2. We assume a
+JC69 nucleotide substitution (as we do use that), a strict clock model (as
+we have that) and a Birth-Death tree prior (as this simplification is the goal
+of this research). Additionally, we assume a MRCA prior with a normal distribution
+with a mean of the crown age, and a standard deviation of 0.001 time units. 0.001
+time units equates to 1000 years, which is the same resolution as which the
+alignment is created. After discarding a burn-in of 10\%, the MCMC algorithm must 
+result in an ESS of the posterior of at least 200 and the MCMC chain length is 
+chosen as such that this is the case for all simulations. For simulations
+with an ESS below 200, their MCMC chain length is elongated. An
+MCMC chain is elongated by the proportion that is expected to increase it to 220. 
+For example, would an ESS be 165, increasing its MCMC chain length by 33\% 
+would bring it to (165\%*1.33=) 220. We use 220 as a safeguard to prevent 
+an asymptotic approach to 200.
+
+We realize there will be an unavoidable bias in our data set: the MCMC
+algorithm will produce phylogenies that are representative for the joint
+likelihood of (correct) site model, (correct) clock model and 
+(incorrect) (Birth-Death) speciation model. This will
+warp the distribution of phylogenies somewhat towards the incorrect
+speciation model: branch length distributions that are likely in the
+Birth-Death model will be relatively overrepresented. We expect
+this bias to be systematic, equally affecting distributions
+of identical speciation completion rates, and symptomatic
+of using contemporary tools.
+
+Each posterior's phylogeny is compared to the 'true' species tree.
+To quantify the difference between two phylogenies, the nLTT statistic
+is used. The nLTT statistic equates to the area between the normalized
+lineages-through-time-plots of two phylogenies, which will be a value between 
+zero (for identical phylogenies) and one. As the nLTT statistic increases
+when phylogenies get increasingly different, we use the nLTT statistic
+as a measure of inference error. Comparing the one 'true' species tree
+with each of the posterior's species trees, a distribution nLTT statistics
+is created.
+
+We predict that nLTT statistic values increase 
+with an increasing protractednedness (that is, a low speciation completion rate),
 but we cannot predict the the extent of this error, as it has never been 
-measured.
+measured [NOTE: this may be easy to change: Etienne & Rosindel 2012 show the equations
+for the expected number of lineages through time. I think I may be able to pull
+of a prediction here, by fitting a BD curve on a PBD curve with a minimized error,
+but research would be needed to investigate that].
 
-The data set produced by the exploratory reseach will consist out of two files:
+The data set produced by this exploratory research will consist out of two files:
 one containing raw data, the other summarized data. 
@@
 measurements. See the appendix, chapter 'Data set specifications', for the full list.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Methods}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-We simulate protracted birth-death trees. In a protracted birth-death tree,
-species-to-be are present, but not yet recognized as different from
-their ancestral species. We simulate these trees from the biological parameters 
-in a factorial fashion. Excluded from this are combinations of parameters in which
-speciation initiation rate is lower or equal the extinction rate, as these
-are biologically irrelevant. Out of computational necessity, 
-parameter combinations that result in an expected number of taxa above 1000 
-are removed as well.
-
-From these PBD trees, we sample species tree in two different ways. 
-From all incipient species-to-be, we pick either the youngest or oldest
-species-to-be to represent the full species. This species tree we refer
-to as 'the true species tree'.
-
-From a species tree, we simulate a DNA alignment that has the same history
-as the phylogeny. The nucleotides of the DNA alignment follow a JC69
-nucleotide substitution model, in which all nucleotide-to-nucleotide transitions
-are equally likely. Although this may seem as a simplification, in our Bayesian
-inference we can use this as a prior, which we know is the truth.
-
-The DNA sequence length and mutation rate was chosen as such to provide one
-nucleotide change per 1000 years per lineage. As our time unit is in 
-millions of years, this equates to a rate of 1000 nucleotide substitutions 
-changes per time unit per lineage. To provide for each substitution being unique,
-we simulate as much DNA nucleotides as expected mutations, which equals the
-nucleotide substitution rate times the crown age, which equals 15K. As this
-mutation rate is constant over all branches, it in effect follows a strict 
-clock model.
-
-The simulated alignment is the data that could be measured in the field.
-This alignment is used to infer a posterior of, using BEAST2. We assume a
-JC69 nucleotide substitution (as we do use that), a strict clock model (as
-we have that) and a Birth-Death tree prior (as this simplification is the goal
-of this research). Additionally, we assume a MRCA prior with a normal distribution
-with a mean of the crown age, and a standard deviation of 0.001 time units. 0.001
-time units equates to 1000 years, which is the same resolution as which the
-alignment is created. After discarding a burn-in of 10\%, the MCMC algorithm must 
-result in an ESS of the posterior of at least 200 and the MCMC chain length is 
-chosen as such that this is the case for all simulations. For simulations
-with an ESS below 200, their MCMC chain length is elongated. An
-MCMC chain is elongated by the proportion that is expected to increase it to 220. 
-For example, would an ESS be 165, increasing its MCMC chain length by 33\% 
-would bring it to (165\%*1.33=) 220. We use
-220 as a safeguard to prevent an asymptotic approach to 200.
-
-The nLTT statistic is used to quantify the similarity of each posterior's phylogeny 
-with 'the true species tree'. In this way, a distribution of phylogenies
-is transformed to a distribution of nLTT statistics. With this, the generative
-model of this experiment has been completely described, from parameter estimates
-to data (nLTT statistics) generated.
-
-To quantify the effect of sampling (either the youngest or oldest species-to-be to represent
-a species), for each parameter set that allows for incipient species,
-we simulate 1000 incipient species trees that result in different species trees
-when sampled either way. From each species tree, we compare the nLTT distributions
-for the two sampling methods. These distributions are compared 
-using the method described in [Kruschke 2013], with a non-informative
-prior. In brief: if the 95\% Highest Density Interval of the differences between 
-these two distributions excludes zero, it is credible the distributions are different.
-
-To measure if protractedness matters, we compare the nLTT distribution per
-combination of speciation initiation and extinction rate for a protracted
-setting with its non-protracted (with a high speciation completion 
-rate) setting. So, for each speciation initiation and extinction rate, 
-the nLTT statistic distribution of the 
-non-protracted setting is used as the baseline. That distribution is compared
-with each protracted setting. This is done the same way as with the two
-different sampling methods. [NOTE: we expect nLTT to go up for increasing SCR,
-this should be in the method].
-
-[NOTE: We do have more info, from Rampal's earlier articles. Perhaps this 
-info can be put in]
-
-[NOTE: as we know the expected number of lineages through time for both BD and PBD
-model, we expect it to fall with the 95\% Highest Density Interval. It would also
-increase support for Rampal's choice of parameter values (next to their 
-computational feasability)]
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Discussion}
+\section{Glossary}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+% Please supply, as a separate list, the definitions of field-specific terms used in your article.
 
@@
   \end{enumerate}
+  \item expected speciation completion time (from Etienne & Rosindell 2012)
+  \item expected duration of speciation (from Etienne & Rosindell 2012)
   \item number of taxa in incipient species tree (also: number of subspecies)
@@
   \end{enumerate}
+  \item [some] MCMC operator acceptances
 \end{itemize}

### 2018-04-05 article.tex
Should I?

@@
   \end{enumerate}
-  \item expected speciation completion time (from Etienne & Rosindell 2012)
-  \item expected duration of speciation (from Etienne & Rosindell 2012)
+  \item [NOTE: add yes/no?] expected speciation completion time (from Etienne & Rosindell 2012)
+  \item [NOTE: add yes/no?] expected duration of speciation (from Etienne & Rosindell 2012)
   \item number of taxa in incipient species tree (also: number of subspecies)
-  \item would there be a difference in brach lengths between the two sampling methods?
+  \item [NOTE: add yes/no?] would there be a difference in brach lengths between the two sampling methods?
   \item number of taxa in species tree (also: number of species)

### 2018-04-05 article.tex
Can create PDF

@@
 \usepackage{authblk}
-\title{Stage 1 Registered Report: The error when inferring phylogenies with incipient species by a Birth-Death model}
+\title{The error when inferring phylogenies with incipient species by a Birth-Death model}
 % \subtitle{Should protracted speciation be incorporated in phylogenetic tree construction methods?}
@@
 but we cannot predict the the extent of this error, as it has never been 
-measured [NOTE: this may be easy to change: Etienne & Rosindel 2012 show the equations
+measured [NOTE: this may be easy to change: Etienne \& Rosindel 2012 show the equations
 for the expected number of lineages through time. I think I may be able to pull
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\begin{figure}[]
-  \includegraphics[width=0.9\textwidth]{fig_does_sampling_matter.png}
-\end{figure}
-
-\begin{figure}[]
-  \includegraphics[width=0.9\textwidth]{fig_does_protractedness_matter.png}
-\end{figure}
 
@@
   \end{enumerate}
-  \item [NOTE: add yes/no?] expected speciation completion time (from Etienne & Rosindell 2012)
-  \item [NOTE: add yes/no?] expected duration of speciation (from Etienne & Rosindell 2012)
+  \item [NOTE: add yes/no?] expected speciation completion time (from Etienne \& Rosindell 2012)
+  \item [NOTE: add yes/no?] expected duration of speciation (from Etienne \& Rosindell 2012)
   \item number of taxa in incipient species tree (also: number of subspecies)

### 2018-04-06 article.tex
EOD

@@
 \maketitle
+
+\begin{abstract}
+
+  The tools for reconstructing phylogenetic relationships between taxonomic units (e.g.
+species) have become very advanced in the last two decades.
+Among the most popular tools are Bayesian approaches, such as BEAST, MrBayes and RevBayes, that use efficient tree sampling routines to provide a posterior probability distribution of the phylogenetic tree.
+A feature of these approaches is the possibility to incorporate known or hypothesized structure of the phylogenetic tree through the tree prior.
+It has ebe shown that the effect of the prior on the posterior distribution of trees can be substantial.
+Currently implemented tree priors assume that speciation is instantaneous.
+However, we know that speciation can be a gradual process and this idea has been termed protracted speciation.
+Here we explore the effects of ignoring the protractedness of the speciation process with an extensive simulation study.
+We simulate protracted phylogenies using the protracted birth-death process and then simulate sequence alignments according to a substitution model and clock model.
+We then use BEAST to infer the phylogeny from these alignments using the standard birth-death prior that assumes instantaneous speciation.
+We compare the inferred tree to the simulated tree, and find that ....
+
+  Furthermore, we identify an important issue related to protracted speciation.
+Because the tree produced by the protracted birth-death process is not necessarily monophyletic, we cannot speak of "the" species tree.
+Instead we have to sample among the incipient species to represent species.
+We show that different ways of sampling the representative species results in ....
+
+  We conclude that ...
+
+\end{abstract}
+
+{\bf Keywords:} computational biology, evolution, phylogenetics, BEAST2, R
 
@@
 
+To visualize the data set, we will produce two figures.
+The first figure will display the distribution 
+of median nLTT statistics per combination of speciation initiation, speciation
+completion and extinction rate. We will use the median nLTT statistic over 
+summing nLTT statistic distributions, as we think this will give the cleanest
+visualization [NOTE: I am unconvinced, wouldn't a violin plot would
+be better?]. We will use the median nLTT statistic, instead of the mean, as
+we expect the distribution of nLTT statistics not to be symmetrical. The distribution
+of median nLTT statistics is displayed per biological parameter set in a box-plot,
+to shown of that distribution as needed, without showing an overly amount of detail
+regarding that distribution. 
+The second figure will show the distribution of median nLTT statistics per expected
+mean duration of speciation. The expected mean duration of parameter can be calculated
+with equation 19 of [Etienne \& Rosidell 2012] and is a combination of the three
+biological parameters combined to their relevant biological effect. Through these
+points, we'll fit a straight line (we have no reason to expect an effect of
+higher order) and a locally weighted scatterplot smoothing (LOESS) 
+curve, which is a non-parametric fit to expose potential higher order effects.
+Additionally, we'll draw a horizontal line to the average median nLTT value 
+[NOTE: here the median of all distributions would feel better] of
+those simulations with instantanous speciation, which serves as a baseline of
+the inference error made.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{figure}[!htbp]
+  \includegraphics[width=0.8\textwidth]{nltt_stats_per_setup.png}
+  \caption{
+    median nLTT statistic distribution per parameter setting
+  }
+\end{figure}
+
+\begin{figure}[!htbp]
+  \includegraphics[width=0.8\textwidth]{nllt_stats_per_mean_dur_spec.png}
+  \caption{
+    median nLTT statistic distribution per mean duration of speciation
+  }
+\end{figure}
 
@@
   \item number of taxa in incipient species tree (also: number of subspecies)
-  \item [NOTE: add yes/no?] would there be a difference in brach lengths between the two sampling methods?
+  \item [NOTE: add yes/no?] would there be a difference in branch lengths between the two sampling methods?
   \item number of taxa in species tree (also: number of species)

### 2018-04-06 article.tex
Data set specified

@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Abbreviations}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\begin{table}
+  \centering 
+  \begin{tabular}{l l l}
+    \hline
+    Parameter             & Description & Values \\
+    \hline
+    \hline
+    $b = b_g = b_i$       & Speciation initiation rate & 0.1, 0.5, 1.0 \\
+    $\lambda$             & Speciation completion rate & 0.1, 0.3, 1.0, $10^6$ \\
+    $\mu = \mu_g = \mu_i$ & Extinction rate & 0.0, 0.1, 0.2, 0.4 \\
+    $t_c$                 & Crown age & 15 \\
+    $\sigma_c$            & Standard deviation around crown age & 0.001 \\
+    M                     & Sampling method & Youngest, oldest \\
+    $r$                   & Mutation rate & 1000 \\
+    $l_a$                 & DNA alignment length & $15000$ \\
+    L                     & MCMC chain length & 1M \\
+    $R_i$                 & RNG seed incipient tree & 1 to 40000 \\
+    $R_a$                 & RNG seed alignment simulation & 1 to 40000 \\
+    $R_b$                 & RNG seed BEAST2 & 1 to 40000 \\
+    \hline
+  \end{tabular}
+  \caption{
+    Overview of the 12 simulation parameters.
+  }
+  \label{table:simulation_parameters}
+\end{table}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
     \hline
-    Symbol & Description \\
+    $n$ & Description \\
     \hline
     \hline
-    $b_g$ & Speciation initiation rate of a good species (probability per lineage, per million year) \\
-    $b_i$ & Speciation initiation rate of an incipient species (probability per lineage, per million year) \\
-    $\lambda$ & Speciation completion rate (probability per lineage, per million year) \\
-    $\mu_g$ & Extinction rate of a good species (probability per lineage, per million year) \\
-    $\mu_i$ & Extinction rate of an incipient species (probability per lineage, per million year) \\
+    $12$   & simulation parameters \\
+    $1000$ & nLTT statistic \\
     \hline
-    $t_c$ & Crown age (million years) \\
-    $r$ & Mutation rate (probability of mutation per nucleotide per million years) \\
-    $l_a$ & DNA alignment length (base pairs) \\
     \hline
-    $n_i$ & Number of incipient species trees per parameter set \\
-    $n_a$ & Number of alignments per species tree \\
-    $n_b$ & Number of BEAST2 runs per alignment \\
-    \hline
-    $n_s$ & Number of MCMC samples \\
-    $i_s$ & MCMC sampling interval \\
-    \hline
-    $n_{ti}$ & Number of taxa in an incipient species tree \\
-    $n_{t}$ & Number of taxa in a (good) species tree \\
-    $t_{\mean{ds}}$ & Average duration of speciation \\
-    $t_{sc}$ & Speciation completion time \\
-    $\Delta_{nLTT}$ & nLTT statistic value \\
-    $\mean{\Delta_{nLTT}}$ & Average nLTT statistic value of a posterior \\
-    $\median{\Delta_{nLTT}}$ & Median nLTT statistic value of a posterior \\
+    $12$   & simulation parameters \\
+    $1$ & [NOTE: add yes/no?] expected speciation completion time (from Etienne \& Rosindell 2012) \\
+    $1$ & [NOTE: add yes/no?] expected duration of speciation (from Etienne \& Rosindell 2012) \\
+    $1$ & number of taxa in incipient species tree (also: number of subspecies) \\
+    $1$ & [NOTE: add yes/no?] difference in branch lengths between the two sampling methods? \\
+    $1$ & number of taxa in species tree (also: number of species) \\
+    $1$ & mean nLTT statistic \\
+    $1$ & median nLTT statistic \\
+    $1$ & lower bound of 95\% HPD value of nLTT statistic \\
+    $1$ & upper bound of 95\% HPD value of nLTT statistic \\
+    $11$ & ESSes of all parameters estimated by BEAST2 (see specs below) \\
+    $?10$ & MCMC operator acceptances\\
     \hline
@@
   \caption{
-    Parameter descriptions. The sections are (1) the biological parameters, (2) the
-    biological tuning parameters, (3) parameters involved in replication, (4) MCMC
-    tuning parameters, and (5) other symbols.
+    Specification of the data sets. $n$ denotes the number
+    of columns a certain item will occupy. Above the double horizontal line,
+    the content of the raw data file is specified, resulting in a table of 
+    1012 columns and 40K rows. Below the double horizontal line,
+    the content of the summary data file specified, resulting in a table of
+    42 columns and 40K rows.
   }
-  \label{table:parameter_descriptions}
+  \label{table:raw_data_set_specs}
 \end{table}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Parameters}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
     \hline
-    Parameter             & Values \\
+    \# & Description \\
     \hline
     \hline
-    $b = b_g = b_i$       & 0.1, 0.5, 1.0 \\
-    $\lambda$             & 0.1, 0.3, 1.0, $10^6$ \\
-    $\mu = \mu_g = \mu_i$ & 0.0, 0.1, 0.2, 0.4 \\
-    \hline
-    $t_c$                 & 15 \\
-    $r$                   & 1000 \\
-    $l_a$                 & $15000$ \\
-    \hline
-    $n_i$                 & 178 \\
-    $n_a$                 & 2 per species tree\\
-    $n_b$                 & 2 per alignment\\
-    \hline
-    $n_s$                 & $10^3$ states (sampled)\\
-    $i_s$                 & $10^3$ states (between samples)\\
+    1 & posterior \\
+    2 & likelihood \\
+    3 & prior \\
+    4 & treeLikelihood \\
+    5 & TreeHeight \\
+    6 & BirthDeath \\
+    7 & BDBirthRate \\
+    8 & BDDeathRate \\
+    9 & logP.mrca \\
+    10 & mrcatime \\
+    11 & clockRate \\
     \hline
@@
   \caption{
-    Parameter values. The sections are (1) the biological parameters, (2) the
-    biological tuning parameters, (3) parameters involved in replication, 
-    and (4) MCMC tuning parameters.
+    Overview of the 11 BEAST2 estimated parameters
   }
-  \label{table:parameter_values}
+  \label{table:estimated_parameters}
 \end{table}
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Data set specifications}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-The raw data file will consist of 40K rows. On each row there will be 1012 values:
-
-\begin{itemize}  
-  \item 12 simulation parameters:
-  \begin{enumerate}  
-    \item speciation initiation rate
-    \item speciation completion rate
-    \item extinction rate
-    \item crown age
-    \item crown age sigma
-    \item sampling method
-    \item mutation rate
-    \item sequence length
-    \item mcmc length
-    \item tree sim rng seed
-    \item alignment rng seed
-    \item beast2 rnd seed
-  \end{enumerate}
-  \item 1000 nLTT statistics
-\end{itemize}
-
-The summarized data file will consist of 40K rows. On each row there will be 29 values:
-
-\begin{itemize}  
-  \item 12 simulation parameters:
-  \begin{enumerate}  
-    \item speciation initiation rate
-    \item speciation completion rate
-    \item extinction rate
-    \item crown age
-    \item crown age sigma
-    \item sampling method
-    \item mutation rate
-    \item sequence length
-    \item mcmc length
-    \item tree sim rng seed
-    \item alignment rng seed
-    \item beast2 rnd seed
-  \end{enumerate}
-  \item [NOTE: add yes/no?] expected speciation completion time (from Etienne \& Rosindell 2012)
-  \item [NOTE: add yes/no?] expected duration of speciation (from Etienne \& Rosindell 2012)
-  \item number of taxa in incipient species tree (also: number of subspecies)
-  \item [NOTE: add yes/no?] would there be a difference in branch lengths between the two sampling methods?
-  \item number of taxa in species tree (also: number of species)
-  \item mean nLTT statistic
-  \item median nLTT statistic
-  \item lower bound of 95\% HPD value of nLTT statistic
-  \item upper bound of 95\% HPD value of nLTT statistic
-  \item 11 ESSes of all parameters estimated by BEAST2:
-  \begin{enumerate}  
-    \item posterior
-    \item likelihood 
-    \item prior 
-    \item treeLikelihood 
-    \item TreeHeight 
-    \item BirthDeath 
-    \item BDBirthRate 
-    \item BDDeathRate 
-    \item logP.mrca 
-    \item mrcatime 
-    \item clockRate
-  \end{enumerate}
-  \item [some] MCMC operator acceptances
-\end{itemize}
-
 \end{document}

### 2018-04-09 article.tex
Add 'How to construct a Nature summary paragraph'
Use @rsetienne's suggested introduction

@@
 \documentclass{article}
+
+% Short communications of approximately 3000 words are also accepted. 
+% These papers should contain no more than two figures, two tables, and thirty references. 
+% A short abstract of fewer than 200 words is acceptable.
 
@@
 
-% Use 'It was found that A is B (Name 1234)' style
+% Use 'It was found that something is something (Name 1234)' style
 \setcitestyle{authoryear,open={},close={}}
@@
 
-  The tools for reconstructing phylogenetic relationships between taxonomic units (e.g.
-species) have become very advanced in the last two decades.
-Among the most popular tools are Bayesian approaches, such as BEAST, MrBayes and RevBayes, that use efficient tree sampling routines to provide a posterior probability distribution of the phylogenetic tree.
-A feature of these approaches is the possibility to incorporate known or hypothesized structure of the phylogenetic tree through the tree prior.
-It has ebe shown that the effect of the prior on the posterior distribution of trees can be substantial.
-Currently implemented tree priors assume that speciation is instantaneous.
-However, we know that speciation can be a gradual process and this idea has been termed protracted speciation.
-Here we explore the effects of ignoring the protractedness of the speciation process with an extensive simulation study.
-We simulate protracted phylogenies using the protracted birth-death process and then simulate sequence alignments according to a substitution model and clock model.
-We then use BEAST to infer the phylogeny from these alignments using the standard birth-death prior that assumes instantaneous speciation.
-We compare the inferred tree to the simulated tree, and find that ....
+  % From 'How to construct a Nature summary paragraph'
 
-  Furthermore, we identify an important issue related to protracted speciation.
-Because the tree produced by the protracted birth-death process is not necessarily monophyletic, we cannot speak of "the" species tree.
-Instead we have to sample among the incipient species to represent species.
-We show that different ways of sampling the representative species results in ....
+  % A short abstract of fewer than 200 words is acceptable.
 
+  % One or two sentences providing a basic
+  % introduction to the field,
+  % comprehensible to a scientist in any discipline.
+  The tools for reconstructing phylogenetic relationships between taxonomic 
+  units (e.g. species) have become very advanced in the last two decades.
+
+  % Two to three sentences of
+  % more detailed background, comprehensible to
+  % scientists in related disciplines.
+  Among the most popular tools are Bayesian approaches, such as BEAST, MrBayes and RevBayes, 
+  that use efficient tree sampling routines to create a posterior probability distribution 
+  of the phylogenetic tree. 
+  A feature of these approaches is the possibility to incorporate 
+  known or hypothesized structure of the phylogenetic tree through the tree prior. 
+  It has been shown that the effect of the prior on the posterior distribution 
+  of trees can be substantial.
+
+  % One sentence clearly stating the general
+  % problem being addressed by this particular
+  % study.
+  Currently implemented tree priors assume that speciation is instantaneous,
+  where we know that speciation can be a gradual process.
+
+  % One sentence summarising the main
+  % result (with the words “here we show”
+  % their equivalent).
+  Here we explore the effects of ignoring 
+  the protractedness of the speciation process with an extensive simulation study.
+
+  % Two or three sentences explaining what
+  % the main result reveals in direct
+  % comparison to what was thought to be the case
+  % previously, or how the main result adds to
+  % previous knowledge.
+
+  % One or two sentences to put the results into a
+  % more general context.
+
+  % Two or three sentences to provide a
+  % broader perspective, readily comprehensible
+  % to a scientist in any discipline, may be included
+  % in the first paragraph if the editor considers that the accessibility of the paper is significantly enhanced
+  % by their inclusion. Under these circumstances, the length of the paragraph can be up to 300 words.
+  % (The above example is 190 words without the final section, and 250 words with it).
   We conclude that ...
+
+  % Does not seem to fit here: 
+  %  
+  % We simulate protracted phylogenies using the protracted birth-death process 
+  % and then simulate sequence alignments according to a substitution model and clock model. 
+  % We then use BEAST to infer the phylogeny from these alignments using the standard 
+  % birth-death prior that assumes instantaneous speciation. We compare the 
+  % inferred tree to the simulated tree, and find that ....
+  % 
+  % Furthermore, we identify an important issue related to protracted speciation. 
+  % Because the tree produced by the protracted birth-death process 
+  % is not necessarily monophyletic, we cannot speak of "the" species tree. 
+  % Instead we have to sample among the incipient species to represent species. 
+  % We show that different ways of sampling the representative species results in ....
 
@@
 
-{\bf Keywords:} computational biology, evolution, phylogenetics, BEAST2, R
+{\bf Keywords:} computational biology, evolution, phylogenetics, prior choice
 
@@
 
-Phylogenetics enables us to infer which species lived when and
-estimate how long it took for a species to give rise to a new
-lineage. Phylogenies show us taxa that have completed speciation, yet
-those taxa only: species that are in the process of speciation are 
-overlooked.
+The computational tools a contemporary phylogeneticist has at his/her disposal
+goes beyond the wildest imagination of those living two decades ago. 
+Advances in computational power allowed the first cladograms to be inferred 
+from DNA alignments in [some year, some ref]. The first time-trees were
+published in [some year, some ref]. Thanks to the addition of effective sampling
+methods, the first Bayesian tools appeared in [some year, some ref], allowing
+for unprecedented flexibility in the setup of a phylogenetic model.
 
-There are multiple reasons why a speciation event, during
-or after its course, goes unnoticed. It may be that the build-up
-of reproductive isolation is still in an intermediate stage. But it
-may also be that speciation has completed, but we, humans, have not
-yet observed it. One example are the four African giraffe species, 
-that have been thought to have been one single species up until 2016,
-where the MRCA of the two most recent species goes back 1.6 million years.
+Currently, the most popular Bayesian phylogenetics tools are BEAST [ref],  
+MrBayes [ref] and RevBayes [ref]. Each of these use efficient tree sampling 
+routines to create a posterior probability distribution of the to-be-inferred 
+phylogenetic tree. But next to the computational speed, they also allow
+to incorporate known or hypothesized structure of the phylogenetic 
+tree-to-be-inferred through the model priors.
 
-It took some time until the Birth-Death model, a phylogenetic model
-assuming constant rates of speciation and extinction, got extended
-by a version that incorporates the fact that speciation takes time.
-This extension is called the Protracted Birth-Death model. In this
-model, a branching event does not give rise to a new species, but to
+The model prior most relevant here, is the tree prior. The tree prior
+embodies the speciation model assumed. Each speciation model has a
+different expected branch length distribution. It has been shown 
+[NOTE: where?] that the effect of the prior on the posterior distribution 
+of trees can be substantial.
+
+All of the currently implemented tree priors assume that speciation is 
+instantaneous, where we know that speciation can be a gradual process.
+The protracted birth-death (PBD) model, an extension of 
+the (constant-rate) birth-death (BD) model, incorporates this idea.
+In the PBD model, a branching event does not give rise to a new species, but to
 a new species-to-be, called an incipient species. Such an incipient
@@
 
-Although we know speciation takes time, most of our theoretical 
-phylogenetic modeling ignores this fact. To the best of our knowledge,
-there is no tool incorporating this fact. One reason may be to ease
+Although we know speciation takes time, all of our theoretical 
+phylogenetic modeling tools ignore this fact. One reason may be to ease
 computation, enabling researchers to work on bigger phylogenies (with
@@
 There are problems with ignoring protracted speciation. A very pragmatic
-example is the conservation of the now four smaller groups of giraffes,
-instead of one big group. But from a more theoretical perspective, the
+example is the conservation of an underestimated amount of species. 
+But from a more theoretical perspective, the
 observed decline of speciation rate in time (excluding the contemporary human-driven
@@
 This exploratory research's goal is to produce a 
-usefull and balanced data set, that quantifies the inference error 
-we make from choosing an overly simplistic prior. In this context, 
-we measure the error made on phylogenies in which speciation takes 
-time, when using a Birth-Death prior.
+useful and balanced data set, that quantifies the inference error 
+we make from choosing the incorrect prior. In this context, 
+we measure the error made on PBD phylogenies, when using a BD prior.
 
@@
 alignment is created. After discarding a burn-in of 10\%, the MCMC algorithm must 
-result in an ESS of the posterior of at least 200 and the MCMC chain length is 
+result in an ESS of the posterior of at least 200 and the MCMC sampling interval is 
 chosen as such that this is the case for all simulations. For simulations
-with an ESS below 200, their MCMC chain length is elongated. An
-MCMC chain is elongated by the proportion that is expected to increase it to 220. 
-For example, would an ESS be 165, increasing its MCMC chain length by 33\% 
-would bring it to (165\%*1.33=) 220. We use 220 as a safeguard to prevent 
-an asymptotic approach to 200.
+with an ESS below 200, their MCMC sampling interval is doubled up until
+the ESS exceeds 200.
 
@@
 
-Each posterior's phylogeny is compared to the 'true' species tree.
+Each posterior's phylogeny is compared to the (sampled) species tree.
 To quantify the difference between two phylogenies, the nLTT statistic
@@
 when phylogenies get increasingly different, we use the nLTT statistic
-as a measure of inference error. Comparing the one 'true' species tree
+as a measure of inference error. Comparing the one (sampled) species tree
 with each of the posterior's species trees, a distribution nLTT statistics
@@
 curve, which is a non-parametric fit to expose potential higher order effects.
-Additionally, we'll draw a horizontal line to the average median nLTT value 
-[NOTE: here the median of all distributions would feel better] of
-those simulations with instantanous speciation, which serves as a baseline of
+Additionally, we'll draw a horizontal line to the average nLTT 
+value (over all replicates) of those simulations with instantaneous speciation, 
+which serves as a baseline of
 the inference error made.
@@
     $\sigma_c$            & Standard deviation around crown age & 0.001 \\
-    M                     & Sampling method & Youngest, oldest \\
-    $r$                   & Mutation rate & 1000 \\
-    $l_a$                 & DNA alignment length & $15000$ \\
-    L                     & MCMC chain length & 1M \\
-    $R_i$                 & RNG seed incipient tree & 1 to 40000 \\
-    $R_a$                 & RNG seed alignment simulation & 1 to 40000 \\
-    $R_b$                 & RNG seed BEAST2 & 1 to 40000 \\
+    $M$                   & Sampling method & Youngest for odd $R_i$, oldest for even $R_i$ \\
+    $r$                   & Mutation rate & 1K \\
+    $l_a$                 & DNA alignment length & $15K$ \\
+    $n_s$                 & Number of MCMC states & 1K \\
+    $f_i$                 & MCMC sampling interval & 1K or more \\
+    $R_i$                 & RNG seed incipient tree & 1 to 40K \\
+    $R_a$                 & RNG seed alignment simulation & 1 to 40K \\
+    $R_b$                 & RNG seed BEAST2 & 1 to 40K \\
     \hline
@@
     $12$ & simulation parameters \\
-    $1$ & [NOTE: add yes/no?] expected speciation completion time (from Etienne \& Rosindell 2012) \\
-    $1$ & [NOTE: add yes/no?] expected duration of speciation (from Etienne \& Rosindell 2012) \\
+    $1$ & expected duration of speciation (from Etienne \& Rosindell 2012) \\
     $1$ & number of taxa in incipient species tree (also: number of subspecies) \\
-    $1$ & [NOTE: add yes/no?] difference in branch lengths between the two sampling methods? \\
     $1$ & number of taxa in species tree (also: number of species) \\
@@
     $11$ & ESSes of all parameters estimated by BEAST2 (see specs below) \\
-    $?10$ & MCMC operator acceptances\\
     \hline

### 2018-04-09 article.tex
EOD

@@
     $l_a$                 & DNA alignment length & $15K$ \\
-    $n_s$                 & Number of MCMC states & 1K \\
     $f_i$                 & MCMC sampling interval & 1K or more \\
@@
     \hline
-    \hline
     $12$ & simulation parameters \\
@@
     Specification of the data sets. $n$ denotes the number
-    of columns a certain item will occupy. Above the double horizontal line,
+    of columns a certain item will occupy. Above the horizontal line,
     the content of the raw data file is specified, resulting in a table of 
-    1012 columns and 40K rows. Below the double horizontal line,
+    1012 columns and 40K rows. Below the horizontal line,
     the content of the summary data file specified, resulting in a table of

### 2018-04-09 article.tex
EOD

@@
 \bibpunct{(}{)}{;}{a}{}{;}
+
+\usepackage[english]{babel}
 
@@
 \linenumbers
-
-% Use TODO notes
-\usepackage{todonotes}
-
-% Rename 'Abstract' to 'Summary 
-\usepackage[english]{babel}
-\addto{\captionsenglish}{\renewcommand{\abstractname}{Summary}}
 
@@
 The model prior most relevant here, is the tree prior. The tree prior
-embodies the speciation model assumed. Each speciation model has a
+embodies the assumed speciation model. Each speciation model has a
 different expected branch length distribution. It has been shown 
@@
 mass extinction event) may be explained by these new species not being 
-recognized yet. Another problem is that the MRCA of young species is estimated 
-incorrectly, as in the BD model a species can be ultra-young (as speciation
-is assumed to be instantaneous), where the PBD model will place these MRCAs
+recognized yet. Another problem is that the most recent common 
+ancestor (MRCA) of a young species duo is estimated 
+too close to the present, as in the BD model a species can be 
+ultra-young (as speciation is assumed to be instantaneous), 
+where the PBD model will place these MRCAs
 further back in time (as two events need to have finished completion).
@@
 From a theoretical point of view, there is another reason why protracted
-speciation has yet become a member of our toolbox: there is no framework
+speciation has not yet become a member of our toolbox: there is no framework
 where it fits in: a framework assumes either an analysis at the species or
@@
     the content of the summary data file specified, resulting in a table of
-    42 columns and 40K rows.
+    40 columns and 40K rows.
   }

### 2018-04-10 article.tex
Fix bug

@@
 nucleotide change per 1000 years per lineage. As our time unit is in 
-millions of years, this equates to a rate of 1000 nucleotide substitutions 
+millions of years, and our crown age is 15 million years,
+this equates to a rate of $\frac{1000}{15}$ nucleotide substitutions 
 changes per time unit per lineage. To provide for each substitution being unique,

### 2018-04-10 article.tex
EOD

@@
     $M$                   & Sampling method & Youngest for odd $R_i$, oldest for even $R_i$ \\
-    $r$                   & Mutation rate & 1K \\
+    $r$                   & Mutation rate & $frac{1000}{15}$ \\
     $l_a$                 & DNA alignment length & $15K$ \\

### 2018-04-10 article.tex
Fix bug

--- 2018-04-10 article.tex
@@
     $M$                   & Sampling method & Youngest for odd $R_i$, oldest for even $R_i$ \\
-    $r$                   & Mutation rate & 1K \\
+    $r$                   & Mutation rate & $\frac{1000}{15}$ \\
     $l_a$                 & DNA alignment length & $15K$ \\

### 2018-04-11 article.tex
Visualize difference in sampling

@@
 different expected branch length distribution. It has been shown 
-[NOTE: where?] that the effect of the prior on the posterior distribution 
+[Moeller, du Plessis, Stadler. 2018] that the effect of the prior on the posterior distribution 
 of trees can be substantial.
@@
 using the phylogetic tool BEAST2. We assume a
-JC69 nucleotide substitution (as we do use that), a strict clock model (as
+JC69 nucleotide substitution (as we do use that), a strict clock model with 
+the same fixed rate as used in the simulation of the alignment (as
 we have that) and a Birth-Death tree prior (as this simplification is the goal

### 2018-04-11 article.tex
Violin plot

@@
 
-The data set produced by this exploratory research will consist out of two files:
-one containing raw data, the other summarized data. 
-A data file should be convenient to handle, thus as an upper limit for this data set's size, 
-we use the maximum size of a table (called a data frame) of the R programming 
-language. Within R, such a table may have maximally $2^{32}-1$ cells and occupy
-less then 2 gigabytes in memory. 
+There will be two data sets produced by this research.
+The first data set will be a general balanced data set to chart
+the effect of the biological parameters on the nLTT statistic
+distribution. The second data set charts the effect of sampling
+subspecies (young versus old) and only uses phylogenies in which this sampling
+has an effect.
+
+Each data set will consist out of two files, containing raw and summarized 
+data respectively. We placed 
+an upper limit for this data set's size at the maximum size of a table (called 
+a data frame) of the R programming language, in which such a table may 
+have maximally $2^{32}-1$ cells and occupy less then 2 gigabytes in memory.
 The raw data file will contain parameters and 1000 nLTT statistics values.

### 2018-04-11 article.tex
Feedback, thanks @rsetienne

@@
   % (The above example is 190 words without the final section, and 250 words with it).
-  We conclude that ...
+  We conclude that in general [something], and the effect of sampling is [something].
 
@@
 we measure the error made on PBD phylogenies, when using a BD prior.
+In brief, we simulate protracted phylogenies using the protracted birth-death process,
+from those phylogenies sample a species tree. From that species tree, 
+we simulate a DNA sequence alignment. Then we let BEAST infer a phylogeny 
+from these alignments using the standard birth-death prior. 
+We compare the posterior phylogenies to the species tree using the nLTT statistic.
 
@@
 the speciation initiation rate, and (2) the difference between
-speciation initiation rate and extinction rate exceeds 0.8. This is
+speciation initiation rate and extinction rate exceeds $0.8$. This is
 done to prevent overly taxon-poor and taxon-rich phylogenies respectively.
@@
 know is the truth.
-The DNA sequence length and mutation rate was chosen as such to provide one
-nucleotide change per 1000 years per lineage. As our time unit is in 
+The DNA sequence length and mutation rate was chosen as such to provide a
+resolution of a 1000 years, that is, to have one expected nucleotide change 
+per 1000 years per lineage on average. As our time unit is in 
 millions of years, and our crown age is 15 million years,
@@
 
-To visualize the data set, we will produce two figures.
-The first figure will display the distribution 
-of median nLTT statistics per combination of speciation initiation, speciation
-completion and extinction rate. We will use the median nLTT statistic over 
-summing nLTT statistic distributions, as we think this will give the cleanest
-visualization [NOTE: I am unconvinced, wouldn't a violin plot would
-be better?]. We will use the median nLTT statistic, instead of the mean, as
-we expect the distribution of nLTT statistics not to be symmetrical. The distribution
-of median nLTT statistics is displayed per biological parameter set in a box-plot,
-to shown of that distribution as needed, without showing an overly amount of detail
-regarding that distribution. 
-The second figure will show the distribution of median nLTT statistics per expected
-mean duration of speciation. The expected mean duration of parameter can be calculated
-with equation 19 of [Etienne \& Rosidell 2012] and is a combination of the three
-biological parameters combined to their relevant biological effect. Through these
-points, we'll fit a straight line (we have no reason to expect an effect of
-higher order) and a locally weighted scatterplot smoothing (LOESS) 
-curve, which is a non-parametric fit to expose potential higher order effects.
-Additionally, we'll draw a horizontal line to the average nLTT 
-value (over all replicates) of those simulations with instantaneous speciation, 
-which serves as a baseline of
-the inference error made.
+Our results will show the general effect of the biological parameters (speciation initiation, 
+speciation completion and extinction rate) using the balanced data set,
+and the effect of sampling using the data set conditioned on sampling having an effect.
+In both cases, the nLTT statistics distribution will be plotted per
+biological parameter set using a violin plot. A violin plot is chosen to
+maintain information about the distribution. When displaying the effect of a 
+sampling regime, per regime, the nLTT statistics distribution is shown per
+biological parameter set.
 
@@
 \begin{figure}[!htbp]
-  \includegraphics[width=0.8\textwidth]{nltt_stats_per_setup.png}
+  \includegraphics[width=0.8\textwidth]{fig_nltt_stats_per_setup.png}
   \caption{
-    median nLTT statistic distribution per parameter setting
+    nLTT statistic distribution per biological parameter set, using the
+    balanced data set
   }
@@
 \begin{figure}[!htbp]
-  \includegraphics[width=0.8\textwidth]{nllt_stats_per_mean_dur_spec.png}
+  \includegraphics[width=0.8\textwidth]{fig_sampling.png}
   \caption{
-    median nLTT statistic distribution per mean duration of speciation
+    nLTT statistic distribution per biological parameter set per sampling
+    regime, using the data set conditioned on sampling regime having an effect 
   }
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+RJCB and RSE conceived the idea for this experiment and package. 
+RJCB created and tested the experiment and package, 
+and wrote the first draft of the manuscript. 
+RSE contributed substantially to revisions.
+
 % Bibliography
@@
     $\mu = \mu_g = \mu_i$ & Extinction rate & 0.0, 0.1, 0.2, 0.4 \\
+    \hline
     $t_c$                 & Crown age & 15 \\
@@
   \caption{
-    Overview of the 12 simulation parameters.
+    Overview of the 12 simulation parameters. Above the horizontal line is 
+    the biological parameter set.
   }

### 2018-04-11 article.tex
Create extra figures in seperate folder

@@
 \affil[1]{Groningen Institute for Evolutionary Life Sciences, University of Groningen, Groningen, The Netherlands}
-
-% Load my functions
-\input{functions.tex}
 
@@
   % (The above example is 190 words without the final section, and 250 words with it).
-  We conclude that in general [something], and the effect of sampling is [something].
+
+  We compare the inferred tree to the simulated tree, and find that ....
+
+  Furthermore, we identify an important issue related to protracted speciation:
+  because the tree produced by the protracted birth-death process 
+  is not necessarily monophyletic, we cannot speak of "the" species tree, but we
+  have to sample among the incipient species to represent species. 
+  % We show that different ways of sampling the representative species results in ....
 
@@
   %  
-  % We simulate protracted phylogenies using the protracted birth-death process 
-  % and then simulate sequence alignments according to a substitution model and clock model. 
-  % We then use BEAST to infer the phylogeny from these alignments using the standard 
-  % birth-death prior that assumes instantaneous speciation. We compare the 
-  % inferred tree to the simulated tree, and find that ....
   % 
@@
     $R_i$                 & RNG seed incipient tree & 1 to 40K \\
-    $R_a$                 & RNG seed alignment simulation & 1 to 40K \\
-    $R_b$                 & RNG seed BEAST2 & 1 to 40K \\
+    $R_a$                 & RNG seed alignment simulation & $R_i$ \\
+    $R_b$                 & RNG seed BEAST2 & $R_i$ \\
     \hline

### 2018-04-11 article.tex
Depend on less packages

@@
 
-\usepackage{listings}
+\usepackage{pgf}
 \usepackage{hyperref}
-\usepackage{todonotes}
 \usepackage{verbatim}
-\usepackage{pgf}
-\usepackage{bm}
-
-% TikZ and friends
-\usepackage{tikz}
-\usepackage{tkz-graph}
-\usepackage{pgf}
-\usetikzlibrary{arrows,automata}
-
-% Style of listings
-% From http://r.789695.n4.nabble.com/How-to-nicely-display-R-code-with-the-LaTeX-package-listings-tp4648110.html
-\usepackage{fancyvrb} 
-\definecolor{codegreen}{rgb}{0,0.6,0}
-\definecolor{codegray}{rgb}{0.5,0.5,0.5}
-\definecolor{codepurple}{rgb}{0.58,0,0.82}
-\definecolor{backcolor}{rgb}{0.95,0.95,0.92}
-\lstdefinestyle{mystyle}{
-  language=R,% set programming language
-  basicstyle=\ttfamily\small,% basic font style
-  commentstyle=\color{gray},% comment style
-  % numbers=left,% display line numbers on the left side
-  numberstyle=\scriptsize,% use small line numbers
-  numbersep=10pt,% space between line numbers and code
-  tabsize=2,% sizes of tabs
-  showstringspaces=false,% do not replace spaces in strings by a certain character
-  captionpos=b,% positioning of the caption below
-  breaklines=true,% automatic line breaking
-  escapeinside={(*}{*)},% escaping to LaTeX
-  fancyvrb=true,% verbatim code is typset by listings
-  extendedchars=false,% prohibit extended chars (chars of codes 128--255)
-  alsoletter={.<-},% becomes a letter
-  alsoother={$},% becomes other
-  otherkeywords={!=, ~, $, \&, \%/\%, \%*\%, \%\%, <-, <<-, /},% other keywords
-  deletekeywords={c}% remove keywords 
-}
-\lstset{style=mystyle}
 

### 2018-04-11 20180411.R
EOD

@@ -1,0 +1,10 @@
+# Can R handle violin plots of the size I need?
+library(ggplot2)
+
+n_repeats <- 100
+ncols <- 1000
+
+as <- rep(seq(1,3), each = 9, times = 1)
+bs <- rep(seq(1,3), each = 3, times = 3)
+cs <- rep(seq(1,3), each = 1, times = 9)
+

### 2018-04-12 article.tex
Refs in intro

@@
 Advances in computational power allowed the first cladograms to be inferred 
-from DNA alignments in [some year, some ref]. The first time-trees were
-published in [some year, some ref]. Thanks to the addition of effective sampling
-methods, the first Bayesian tools appeared in [some year, some ref], allowing
+from DNA alignments in 1981 \cite{felsenstein1981}. The first Bayesian tools 
+appeared in 1996 \cite{rannala1996}, allowing
 for unprecedented flexibility in the setup of a phylogenetic model.
 
-Currently, the most popular Bayesian phylogenetics tools are BEAST [ref],  
+Currently, the most popular Bayesian phylogenetics tools are BEAST \cite{beast},  
 MrBayes [ref] and RevBayes [ref]. Each of these use efficient tree sampling 

### 2018-04-12 article.tex
Note about Moeller et al., 2018 article

@@
   The tools for reconstructing phylogenetic relationships between taxonomic 
-  units (e.g. species) have become very advanced in the last two decades.
+  units (e.g. species) have become very advanced in the last three decades.
 
@@
 The computational tools a contemporary phylogeneticist has at his/her disposal
-goes beyond the wildest imagination of those living two decades ago. 
+goes beyond the wildest imagination of those living three decades ago. 
 Advances in computational power allowed the first cladograms to be inferred 
@@
 Currently, the most popular Bayesian phylogenetics tools are BEAST \cite{beast},  
-MrBayes [ref] and RevBayes [ref]. Each of these use efficient tree sampling 
+MrBayes \cite{mrbayes} and RevBayes \cite{revbayes}. 
+Each of these use efficient tree sampling 
 routines to create a posterior probability distribution of the to-be-inferred 
@@
 
-The model prior most relevant here, is the tree prior. The tree prior
-embodies the assumed speciation model. Each speciation model has a
-different expected branch length distribution. It has been shown 
-[Moeller, du Plessis, Stadler. 2018] that the effect of the prior on the posterior distribution 
-of trees can be substantial.
+In a Bayesian analysis, the model priors are explicitly specified.
+Those model priors can be grouped in three groups: (1) site model, which
+governs the nucleotide subsititution model, (2) clock model, specifying
+the rate of mutation per lineage in time, and (3) tree prior, embodying
+the speciation model behind branching events (speciation) 
+and branch termination (extinction).
+
+The effect of choosing a (potentially wrong) prior will have its effect on
+the posterior. Recently, for example, \cite{moller2018} show that the effect
+of choosing a tree prior biases the estimation of the molecular clock rate, 
+for DNA sequences of 100-1000 base pairs.
+
+This bias has its implications in the research and policy regarding rapidly 
+evolving viruses. [NOTE: I would say that same study concludes 
+that it does not matter for DNA alignments of 15K, which is the DNA 
+alignment length that I use].
 
@@
 we simulate as much DNA nucleotides as expected mutations, which equals the
-nucleotide substitution rate times the crown age, which equals 15K. As this
+nucleotide substitution rate times the crown age, which equals 15K (which is
+coincidentally the same as \cite{moller2018}). As this
 mutation rate is constant over all branches, it in effect follows a strict 

### 2018-04-12 article.tex
Intro gets better
Introduction

@@
 Advances in computational power allowed the first cladograms to be inferred 
-from DNA alignments in 1981 \cite{felsenstein1981}. The first Bayesian tools 
-appeared in 1996 \cite{rannala1996}, allowing
+from DNA alignments in 1981 (\cite{felsenstein1981}). The first Bayesian tools 
+appeared in 1996 (\cite{rannala1996}), allowing
 for unprecedented flexibility in the setup of a phylogenetic model.
 
-Currently, the most popular Bayesian phylogenetics tools are BEAST \cite{beast},  
-MrBayes \cite{mrbayes} and RevBayes \cite{revbayes}. 
+Currently, the most popular Bayesian phylogenetics tools are BEAST (\cite{beast})
+and its successor BEAST2 (\cite{beast2}),  
+MrBayes (\cite{mrbayes}) and RevBayes (\cite{revbayes}). 
 Each of these use efficient tree sampling 
@@
 and branch termination (extinction).
+The effect of choosing a (potentially wrong) prior will have its effect on
+the posterior. For example, recently, it was shown that the effect
+of choosing a tree prior biases the estimation of the molecular clock rate, 
+for DNA sequences of 100-1000 base pairs (\cite{moller2018}).
 
-The effect of choosing a (potentially wrong) prior will have its effect on
-the posterior. Recently, for example, \cite{moller2018} show that the effect
-of choosing a tree prior biases the estimation of the molecular clock rate, 
-for DNA sequences of 100-1000 base pairs.
+At the present time, for sexually producing animals, all we have at our
+disposal are tree priors that assume that speciation
+is instantaneous, where we know that speciation is a gradual process.
+When assuming big populations sizes (thus the effect of sampling to 
+be small), the (constant-rate) birth-death (BD) model is a commonly 
+used tree prior, which does not incorporate the temporal aspect of speciation.
+The protracted birth-death (PBD) model, an extension of 
+the BD model, does incorporate the idea that speciation takes time.
+In this model, a branching event does not give rise to a new species, but to
+a new species-to-be, called an incipient species. Such an incipient
+species may go extinct, finish its speciation to become a good species, or give
+rise to new incipient species.
 
-This bias has its implications in the research and policy regarding rapidly 
-evolving viruses. [NOTE: I would say that same study concludes 
-that it does not matter for DNA alignments of 15K, which is the DNA 
-alignment length that I use].
+The effect of using the (incorrect) BD tree prior for a PBD process is unknown.
+A potential problem in species conservation is that the number of 
+species is underestimated (see \cite{fennessy2016} for a clear example). 
+Additionally, protracted speciation may be one 
+explanation in the observed decline of speciation rates in 
+time (\cite{etienne2012}). Also, a BD model places the most recent common 
+ancestor (MRCA) of a young species duo closer to the present, as the
+BD model allows for a speciation event being recognized immediately,
+where the PBD model acknowledges that this will take time.
 
-All of the currently implemented tree priors assume that speciation is 
-instantaneous, where we know that speciation can be a gradual process.
-The protracted birth-death (PBD) model, an extension of 
-the (constant-rate) birth-death (BD) model, incorporates this idea.
-In the PBD model, a branching event does not give rise to a new species, but to
-a new species-to-be, called an incipient species. Such an incipient
-species may go extinct, or finish its speciation to become a good species.
+There are multiple possiblities why the PBD model is relatively unexplored.
+Biologically, the PBD model is predicted to have an effect strongest in
+the present (as earlier speciation events are nearly always recognized),
+so in research that investigates (mostly) older species, a BD model would suffice. 
+Computationally, the BD model is simpler, thus more convenient, model. 
+Methodological, there is no computational tool where 
+the PBD model fits in: every contemporary framework assumes 
+either an analysis at the species or subspecies level. In the PBD
+model, incipient species are the cause there is no such thing 
+as a 'true' species tree, as incipient species may give rise to paraphylies.
+It may be possible to establish a protocol to sample a species tree from an 
+incipient species tree, but this has not yet been explored [NOTE: except by Simonet].
 
-Although we know speciation takes time, all of our theoretical 
-phylogenetic modeling tools ignore this fact. One reason may be to ease
-computation, enabling researchers to work on bigger phylogenies (with
-more power statistical power) instead. Additionally, speciation being
-protracted is only playing a role in the present. As seen from the present,
-all older species have finished speciation, as we -by definition- recognize
-these species as such.
+This research's goal is to explore the effect of using an overly simplistic
+BD prior on a PBD simulated reality. We provide a data set, 
+that quantifies the inference error made in general, and explores
+the effect of the way species trees are sampled from an incipient species tree.  
+In brief, we simulate protracted phylogenies using the PBD process,
+from which we sample a species tree. From that species tree, 
+we simulate a DNA sequence alignment. Then, BEAST2 uses these alignments
+to infer a posterior of phylogenies, using a BD prior. The difference
+between the (BD) posterior phylogenies and (PBD) species tree is quantified.
 
-There are problems with ignoring protracted speciation. A very pragmatic
-example is the conservation of an underestimated amount of species. 
-But from a more theoretical perspective, the
-observed decline of speciation rate in time (excluding the contemporary human-driven
-mass extinction event) may be explained by these new species not being 
-recognized yet. Another problem is that the most recent common 
-ancestor (MRCA) of a young species duo is estimated 
-too close to the present, as in the BD model a species can be 
-ultra-young (as speciation is assumed to be instantaneous), 
-where the PBD model will place these MRCAs
-further back in time (as two events need to have finished completion).
-
-From a theoretical point of view, there is another reason why protracted
-speciation has not yet become a member of our toolbox: there is no framework
-where it fits in: a framework assumes either an analysis at the species or
-subspecies level. When picking a tool that does its analysis at the
-species level, there is no protocol how to select which incipient species
-gets to represent the full species it is still a member of. When picking
-a tool at the sub-species level, the incipient state disappears as a
-sub-species lineage, and the process degrades to an ordinary birth-death 
-process.
-
-This exploratory research's goal is to produce a 
-useful and balanced data set, that quantifies the inference error 
-we make from choosing the incorrect prior. In this context, 
-we measure the error made on PBD phylogenies, when using a BD prior.
-In brief, we simulate protracted phylogenies using the protracted birth-death process,
-from those phylogenies sample a species tree. From that species tree, 
-we simulate a DNA sequence alignment. Then we let BEAST infer a phylogeny 
-from these alignments using the standard birth-death prior. 
-We compare the posterior phylogenies to the species tree using the nLTT statistic.
+% Methods
 

### 2018-04-12 article.tex
Method

@@
 This research's goal is to explore the effect of using an overly simplistic
-BD prior on a PBD simulated reality. We provide a data set, 
+BD prior on PBD simulated phylogenies. We provide a data set, 
 that quantifies the inference error made in general, and explores
@@
 
-The parameter space the data set explores is inspired on 
-[Etienne et al, 2014] for two reasons: (1) the academic
-reason of being able to compare results, and (2) the practical
-reason that the simulated phylogies will have enough taxa to
-be informative, but not too big to be computationally overdemanding.
-This research uses a complete superset of those parameter settings.
-For the speciation initiation rate, we'll use $0.1$, $0.5$ and $1.0$ 
-speciation initiation events per (good species) lineage per time unit, 
-extending the set of
-values used from one ($0.5$) to three. 
-The speciation completion
-rates used are $0.1$, $0.3$, $1.0$ and $10^6$ speciation completion
-events per (incipient species) lineage per time unit, in which only
-the last value is added as a computational proxy for infinity, which
-makes the experiment fall back to a Birth-Death model (as speciation
-is again instantaneous), allowing to measure the baseline error.
+The PBD model has five biological parameters (see \ref{table:parameters}), 
+which we explore in a factorial fashion, excluding some combinations. 
+We assume that the speciation initiation and extinction rates
+of an incipient and good species are equal ($b_i = b_g$ and $\mu_i = \mu_g$),
+as this will result in monophyletic incipient species trees.
+We only simulate a PBD process for phylogenies in which
+speciation initiation exceeds extinction rate ($b > \mu$),
+and in which their difference is not too big ($b - \mu < 0.8$), 
+to prevent overly taxon-poor and taxon-rich phylogenies respectively.
+The parameter values chosen are a 
+superset of \cite{etienne2014}, as these parameters result in reasonably
+sized phylogenies and allows us to compare results. 
+For the speciation initiation rate $b$, we'll use $0.1$, $0.5$ and $1.0$ 
+speciation initiation events per (good species) lineage per time unit.
+The speciation completion rates used are $0.1$, $0.3$, $1.0$ and $\infty$ speciation completion
+events per (incipient species) lineage per time unit. For $\lambda = \infty$, 
+the PBD model equals a BD model, which allows us to measure the baseline error.
 The extinction rates used are $0.0$, $0.1$, $0.2$ and $0.4$ 
-extinction events per (good or incipient) lineage per time unit,
-in which $0.4$ is a new extinction rate.
-We investigate these parameters in a factorial fashion, except for
-(1) those combinations in which the extinction rate equals or exceeds
-the speciation initiation rate, and (2) the difference between
-speciation initiation rate and extinction rate exceeds $0.8$. This is
-done to prevent overly taxon-poor and taxon-rich phylogenies respectively.
+extinction events per (good or incipient) lineage per time unit.
 
@@
   }
-  \label{table:simulation_parameters}
+  \label{table:parameters}
 \end{table}

### 2018-04-12 article.tex
Ready for @rsetienne to review the methods

@@
 
-% Methods
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Methods (but we are not allowed to keep this header)}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 From each biological parameter set, a protracted birth-death tree is simulated,
-with the same crown age as [Etienne et al., 2014] of 15 million years. 
+using the PBD package (\cite{pbd}) in the R programming language (\cite{r}), 
+with the same crown age as \cite{etienne2014} of 15 million years. 
 Each protracted birth-death tree uses a different random number
@@
 
-From an incipient species tree, we sample a species tree
-by picking either the youngest or oldest
-species-to-be to represent the full species.
-We use either sampling method in 50\% of all cases. We chose not to use
-a random species-to-be to represent a full species as to maximize variance
-in this aspect. Note that only in few cases (when speciation completion rate
-is low) the sampling method has an effect on the branch length distribution
-of the sampled species tree, as this requires a very specific scenario.
+From an incipient species tree, we sample a species tree. To do
+so, from each species a sub-species must be chosen to represent
+the good species as a whole. To clarify, it may be that an
+incipient species branched of from a good species. Both of these
+subspecies are recognized as the good species, and both can be 
+picked as a representative. Here we use three sampling scenario's,
+in which we pick the youngest (most recent incipient speciation), 
+oldest or random subspecies. Note that only in few 
+cases (when speciation completion rate is low) the incipient species
+chosen will have an effect on the branch length distribution of species 
+tree, as this requires a very specific scenario.
 
 From a species tree, we simulate a DNA alignment that has the same history
-as the phylogeny. The nucleotides of the DNA alignment follow a JC69
+as the phylogeny, using the phangorn package (\cite{phangorn}). 
+The nucleotides of the DNA alignment follow a Jukes-Cantor (\cite{jc69})
 nucleotide substitution model, in which all nucleotide-to-nucleotide transitions
 are equally likely. Although this may seem as a simplification, in our Bayesian
-inference (see below) we can use this exact site model as a prior, which we 
-know is the truth.
-The DNA sequence length and mutation rate was chosen as such to provide a
+inference (see below) we use this exact site model as the (obviously correct) site model prior.
+The DNA sequence length and mutation rate are chosen as such to provide a
 resolution of a 1000 years, that is, to have one expected nucleotide change 
-per 1000 years per lineage on average. As our time unit is in 
-millions of years, and our crown age is 15 million years,
+per 1000 years per lineage on average. As we use a crown age is 15 million years,
 this equates to a rate of $\frac{1000}{15}$ nucleotide substitutions 
-changes per time unit per lineage. To provide for each substitution being unique,
-we simulate as much DNA nucleotides as expected mutations, which equals the
-nucleotide substitution rate times the crown age, which equals 15K (which is
+changes per million years per lineage. To provide for each substitution being unique,
+we simulate as much DNA nucleotides as the expected number of mutations (including those
+that do not modify the nucleotide), which equals the
+nucleotide substitution rate times the crown age, which equals 15000 nucleotides (which is
 coincidentally the same as \cite{moller2018}). As this
 mutation rate is constant over all branches, it in effect follows a strict 
-clock model, which we will specify as the known clock model prior later.
+clock model, which we will specify as the known clock model prior in the Bayesian inference.
 
 From an alignment, we run a Bayesian analysis and create a posterior, 
-using the phylogetic tool BEAST2. We assume a
-JC69 nucleotide substitution (as we do use that), a strict clock model with 
-the same fixed rate as used in the simulation of the alignment (as
-we have that) and a Birth-Death tree prior (as this simplification is the goal
-of this research). Additionally, we assume a MRCA prior with a normal distribution
+using the phylogetic tool BEAST2 \cite{beast2} using the 
+pirouette (\cite{pirouette}) package. As our site model, we assume a Jukes-Cantor 
+nucleotide substitution model, as we used that in the simulation of the alignment.
+For our clock model, we assume a strict clock with the same fixed rate as 
+used in the simulation of the alignment. The tree prior assumed is the BD model, 
+as this simplification is the goal of this research. 
+Additionally, we assume a MRCA prior with a normal distribution
 with a mean of the crown age, and a standard deviation of 0.001 time units. 0.001
@@
 alignment is created. After discarding a burn-in of 10\%, the MCMC algorithm must 
-result in an ESS of the posterior of at least 200 and the MCMC sampling interval is 
-chosen as such that this is the case for all simulations. For simulations
+result in an effective sample size (ESS) of the posterior of at least 200 
+to have a strong enough inference (\cite{beastbook}). The MCMC sampling interval is 
+chosen in such a way that this is the case for all simulations. For simulations
 with an ESS below 200, their MCMC sampling interval is doubled up until
@@
 
-Each posterior's phylogeny is compared to the (sampled) species tree.
-To quantify the difference between two phylogenies, the nLTT statistic
-is used. The nLTT statistic equates to the area between the normalized
+Each posterior's phylogeny is compared to the (sampled) species tree
+by the nLTT statistic (\cite{janzen2015}), using the nLTT package (\cite{nltt}). 
+The nLTT statistic equates to the area between the normalized
 lineages-through-time-plots of two phylogenies, which will be a value between 
@@
 
-We predict that nLTT statistic values increase 
-with an increasing protractednedness (that is, a low speciation completion rate),
-but we cannot predict the the extent of this error, as it has never been 
-measured [NOTE: this may be easy to change: Etienne \& Rosindel 2012 show the equations
-for the expected number of lineages through time. I think I may be able to pull
-of a prediction here, by fitting a BD curve on a PBD curve with a minimized error,
-but research would be needed to investigate that].
-
 There will be two data sets produced by this research.
@@
 the effect of the biological parameters on the nLTT statistic
-distribution. The second data set charts the effect of sampling
-subspecies (young versus old) and only uses phylogenies in which this sampling
-has an effect.
+distribution. In this data set, incipient species are sampled 
+randomly to represent a good species.
+The second data set charts the effect of sampling
+subspecies and only uses PBD trees in which this sampling
+has an effect. For each of these trees, we sample both youngest and
+oldest subspecies. We predict that these two most extreme sampling methods
+result in the most pronounced differences.
 
-Each data set will consist out of two files, containing raw and summarized 
-data respectively. We placed 
+Each data set will be stored as a comma-seperated file. We placed 
 an upper limit for this data set's size at the maximum size of a table (called 
@@
 have maximally $2^{32}-1$ cells and occupy less then 2 gigabytes in memory.
-The raw data file will contain parameters and 1000 nLTT statistics values.
-Each row will contain the parameter set and these nLTT statistics.
-From the constraints of the R programming language, this allows for 40K rows. 
-The file with the summarized data will provide summary statistics of the raw
-data. Each row will contain a parameter set, summary statistics and other
-measurements. See the appendix, chapter 'Data set specifications', for the full list.
+Each row will contain the parameter set, nLTT statistics and some 
+diagnostics [NOTE: should I keep the diagnostics?] (see \ref{table:specs} for the 
+complete list). From the constraints of the R programming language, this allows for 40K rows.
 
@@
 biological parameter set.
+We predict that nLTT statistic values increase 
+with an increasing protractednedness (that is, a low speciation completion rate),
+but we cannot predict the the extent of this error, as it has never been 
+measured.
 
@@
     $b = b_g = b_i$       & Speciation initiation rate & 0.1, 0.5, 1.0 \\
-    $\lambda$             & Speciation completion rate & 0.1, 0.3, 1.0, $10^6$ \\
+    $\lambda$             & Speciation completion rate & 0.1, 0.3, 1.0, $\infty$ \\
     $\mu = \mu_g = \mu_i$ & Extinction rate & 0.0, 0.1, 0.2, 0.4 \\
@@
     $\sigma_c$            & Standard deviation around crown age & 0.001 \\
-    $M$                   & Sampling method & Youngest for odd $R_i$, oldest for even $R_i$ \\
+    $M$                   & Sampling method & Random \\
     $r$                   & Mutation rate & $\frac{1000}{15}$ \\
@@
     \hline
-    $12$   & simulation parameters \\
-    $1000$ & nLTT statistic \\
-    \hline
-    $12$ & simulation parameters \\
-    $1$ & expected duration of speciation (from Etienne \& Rosindell 2012) \\
-    $1$ & number of taxa in incipient species tree (also: number of subspecies) \\
-    $1$ & number of taxa in species tree (also: number of species) \\
-    $1$ & mean nLTT statistic \\
-    $1$ & median nLTT statistic \\
-    $1$ & lower bound of 95\% HPD value of nLTT statistic \\
-    $1$ & upper bound of 95\% HPD value of nLTT statistic \\
+    $12$   & simulation parameters, see table \ref{table:parameters} \\
+    $1000$ & nLTT statistic values \\
+    $1$ & [NOTE: keep this in?] expected duration of speciation (from Etienne \& Rosindell 2012) \\
+    $1$ & [NOTE: keep this in?] number of taxa in incipient species tree (also: number of subspecies) \\
+    $1$ & [NOTE: keep this in?] number of taxa in species tree (also: number of species) \\
+    $1$ & [NOTE: keep this in?] mean nLTT statistic \\
+    $1$ & [NOTE: keep this in?] median nLTT statistic \\
+    $1$ & [NOTE: keep this in?] lower bound of 95\% HPD value of nLTT statistic \\
+    $1$ & [NOTE: keep this in?] upper bound of 95\% HPD value of nLTT statistic \\
     $11$ & ESSes of all parameters estimated by BEAST2 (see specs below) \\
@@
     Specification of the data sets. $n$ denotes the number
-    of columns a certain item will occupy. Above the horizontal line,
-    the content of the raw data file is specified, resulting in a table of 
-    1012 columns and 40K rows. Below the horizontal line,
-    the content of the summary data file specified, resulting in a table of
-    40 columns and 40K rows.
+    of columns a certain item will occupy, resulting in a table of 
+    1030 columns and 40K rows.
   }
-  \label{table:raw_data_set_specs}
+  \label{table:specs}
 \end{table}

### 2018-04-15 article.tex
EOD

@@
 time units equates to 1000 years, which is the same resolution as which the
-alignment is created. After discarding a burn-in of 10\%, the MCMC algorithm must 
-result in an effective sample size (ESS) of the posterior of at least 200 
-to have a strong enough inference (\cite{beastbook}). The MCMC sampling interval is 
+alignment is created. The MCMC chain is run to generate 1100 states,
+of which the first 10\% (also called the 'burn-in') is removed. Of the remaining
+1000 MCMC states, the effective sample size (ESS) of the posterior must at least be 200
+for a strong enough inference (\cite{beastbook}). The MCMC sampling interval is 
 chosen in such a way that this is the case for all simulations. For simulations

### 2018-04-16 article.tex
Implemented feedback, thanks @rsetienne

@@
 chosen will have an effect on the branch length distribution of species 
-tree, as this requires a very specific scenario.
+tree. The simplest scenario requires a species, with one good and
+one incipient lineage, after which either lineages gives 
+rise to a new good species, before the first incipient lineage
+completes speciation itself.
 
@@
 of which the first 10\% (also called the 'burn-in') is removed. Of the remaining
-1000 MCMC states, the effective sample size (ESS) of the posterior must at least be 200
-for a strong enough inference (\cite{beastbook}). The MCMC sampling interval is 
-chosen in such a way that this is the case for all simulations. For simulations
-with an ESS below 200, their MCMC sampling interval is doubled up until
-the ESS exceeds 200.
-
-We realize there will be an unavoidable bias in our data set: the MCMC
-algorithm will produce phylogenies that are representative for the joint
-likelihood of (correct) site model, (correct) clock model and 
-(incorrect) (Birth-Death) speciation model. This will
-warp the distribution of phylogenies somewhat towards the incorrect
-speciation model: branch length distributions that are likely in the
-Birth-Death model will be relatively overrepresented. We expect
-this bias to be systematic, equally affecting distributions
-of identical speciation completion rates, and symptomatic
-of using contemporary tools.
+1000 MCMC states [NOTE: Why 1000? Why not 250?], 
+the effective sample size (ESS) of the posterior must at least be 200
+for a strong enough inference (\cite{beastbook}). An ESS can be increased by increasing
+the number of samples or decreasing the autocorrelation between samples. 
+Would the ESS be less than 200, we decrease autocorrelation by doubling 
+the MCMC sampling interval of that simulation, until the ESS exceeds 200.
 

### 2018-04-16 article.tex
Solved https://github.com/richelbilderbeek/raket/issues/4

@@
 speciation initiation events per (good species) lineage per time unit.
-The speciation completion rates used are $0.1$, $0.3$, $1.0$ and $\infty$ speciation completion
-events per (incipient species) lineage per time unit. For $\lambda = \infty$, 
+The speciation completion rates used are $0.1$, $0.3$, $1.0$ and $10^9$ speciation completion
+events per (incipient species) lineage per time unit. For 
+$\lambda = \infty$ (where we assume that in this context $10^9 \approx \infty$), 
 the PBD model equals a BD model, which allows us to measure the baseline error.

### 2018-04-17 article.tex
Fixed flaw

@@
 picked as a representative. Here we use three sampling scenario's,
-in which we pick the youngest (most recent incipient speciation), 
-oldest or random subspecies. Note that only in few 
+in which we pick the most recent common ancestor (MRCA), 
+most distant common ancestor (MDCA) or random subspecies. Note that only in few 
 cases (when speciation completion rate is low) the incipient species
@@
 subspecies and only uses PBD trees in which this sampling
-has an effect. For each of these trees, we sample both youngest and
-oldest subspecies. We predict that these two most extreme sampling methods
+has an effect. For each of these trees, we sample both MRCA and
+MDCA subspecies. We predict that these two most extreme sampling methods
 result in the most pronounced differences.
@@
     $\sigma_c$            & Standard deviation around crown age & 0.001 \\
-    $M$                   & Sampling method & Random \\
+    $M$                   & Sampling method & MRCA, MDCA or random \\
     $r$                   & Mutation rate & $\frac{1000}{15}$ \\
@@
     Overview of the 12 simulation parameters. Above the horizontal line is 
-    the biological parameter set.
+    the biological parameter set. Sampling method $M$ is random for the general
+    data set. For the data set exploring the effect of sampling, MRCA is
+    used for odd values of $R_i$, and MDCA is used for even values of $R_i$.
+    $R_i$ is 1 for the first simulation, 2 for the next, etcetera.
   }
@@
     $1000$ & nLTT statistic values \\
-    $1$ & [NOTE: keep this in?] expected duration of speciation (from Etienne \& Rosindell 2012) \\
-    $1$ & [NOTE: keep this in?] number of taxa in incipient species tree (also: number of subspecies) \\
-    $1$ & [NOTE: keep this in?] number of taxa in species tree (also: number of species) \\
-    $1$ & [NOTE: keep this in?] mean nLTT statistic \\
-    $1$ & [NOTE: keep this in?] median nLTT statistic \\
-    $1$ & [NOTE: keep this in?] lower bound of 95\% HPD value of nLTT statistic \\
-    $1$ & [NOTE: keep this in?] upper bound of 95\% HPD value of nLTT statistic \\
     $11$ & ESSes of all parameters estimated by BEAST2 (see specs below) \\
@@
     of columns a certain item will occupy, resulting in a table of 
-    1030 columns and 40K rows.
+    1023 columns and 40K rows.
   }

### 2018-04-17 issue_6.R
Use correct MCMC chain length

@@ -1,19 +1,0 @@
-ncols <- 1050
-nrows <- 40000
-ncells <- ncols*nrows
-testit::assert(ncells < 2^32-1)
-mem_use <- ncells * object.size(3.14)
-format(mem_use, units = "Mb")
-
-# Computer will freeze if you ignore this warning ..
-testit::assert(mem_use < ((2^32)-1))
-
-((2^32)-1) / as.double(object.size(3.14)) / ncols
-
-df <- data.frame(matrix(nrow = nrows, ncol = ncols, data = seq(1,ncells)))
-df[2,1:10]
-filename <- tempfile()
-write.csv(x = df, file = filename, row.names = FALSE)
-df2 <- read.csv(file = filename)
-all.equal(df2, df, tolerance = 0.1)
-file.info(filename)

### 2018-04-17 article.tex
Use correct MCMC chain length
Use half of max size, 20K rows

@@
 time units equates to 1000 years, which is the same resolution as which the
-alignment is created. The MCMC chain is run to generate 1100 states,
+alignment is created. The MCMC chain is run to generate 1111 states,
 of which the first 10\% (also called the 'burn-in') is removed. Of the remaining
@@
 have maximally $2^{32}-1$ cells and occupy less then 2 gigabytes in memory.
-Each row will contain the parameter set, nLTT statistics and some 
-diagnostics [NOTE: should I keep the diagnostics?] (see \ref{table:specs} for the 
-complete list). From the constraints of the R programming language, this allows for 40K rows.
+To provide for copying such a table (thus, allowing it to be present in
+memory twice), the size of the table has an upper limit of half its maximal size.
+Each row will contain the parameter set and the generated nLTT 
+statistics (see \ref{table:specs} for the exact data specification).
+From these constraints, this allows for 20K rows.
 

### 2018-04-23 article.tex
EOD

@@
 \usepackage{authblk}
-\title{The error when inferring phylogenies with incipient species by a Birth-Death model}
+\title{The error when inferring phylogenies with incipient species by a birth-death model}
 % \subtitle{Should protracted speciation be incorporated in phylogenetic tree construction methods?}
@@
 inference (see below) we use this exact site model as the (obviously correct) site model prior.
-The DNA sequence length and mutation rate are chosen as such to provide a
-resolution of a 1000 years, that is, to have one expected nucleotide change 
-per 1000 years per lineage on average. As we use a crown age is 15 million years,
-this equates to a rate of $\frac{1000}{15}$ nucleotide substitutions 
-changes per million years per lineage. To provide for each substitution being unique,
-we simulate as much DNA nucleotides as the expected number of mutations (including those
-that do not modify the nucleotide), which equals the
-nucleotide substitution rate times the crown age, which equals 15000 nucleotides (which is
-coincidentally the same as \cite{moller2018}). As this
-mutation rate is constant over all branches, it in effect follows a strict 
-clock model, which we will specify as the known clock model prior in the Bayesian inference.
+The mutation rate is set in such a way to maximize chronologic information.
+To do so, the mutation rate is set to expect on average 1 (possibly silent) mutation per nucleotide
+between crown age and present, which equates to $\frac{1}{15}$ mutations
+per million year.
+The DNA sequence length is chosen to provide a
+resolution of a 1K years, that is, to have one expected nucleotide change 
+per 1k years per lineage on average. As 1 nucleotide is expected 
+to have on average 1 (possibly silent) mutation per 15 million years, 15K
+nucleotides provides for 1 mutation per alignment per 1K years (which is
+coincidentally the same as \cite{moller2018}). 
+The simulation of these DNA aligment follows a strict clock model, 
+which we will specify as the known clock model prior in the Bayesian inference.
 
@@
     $M$                   & Sampling method & MRCA, MDCA or random \\
-    $r$                   & Mutation rate & $\frac{1000}{15}$ \\
+    $r$                   & Mutation rate & $\frac{1}{15}$ \\
     $l_a$                 & DNA alignment length & $15K$ \\
     $f_i$                 & MCMC sampling interval & 1K or more \\
-    $R_i$                 & RNG seed incipient tree & 1 to 40K \\
+    $R_i$                 & RNG seed incipient tree & 1 to 20K \\
     $R_a$                 & RNG seed alignment simulation & $R_i$ \\
@@
   \caption{
-    Specification of the data sets. $n$ denotes the number
+    Specification of the data sets. Each row will contain one experiment,
+    where the columns contain parameters, measurements and diagnostics.
+    This table displays the content of the columns. 
+    $n$ denotes the number
     of columns a certain item will occupy, resulting in a table of 
-    1023 columns and 40K rows.
+    1023 columns and 20K rows.
   }

### 2018-04-23 article.tex
Feedback and rewrite, thanks @rsetienne

@@
 Advances in computational power allowed the first cladograms to be inferred 
-from DNA alignments in 1981 (\cite{felsenstein1981}). The first Bayesian tools 
-appeared in 1996 (\cite{rannala1996}), allowing
-for unprecedented flexibility in the setup of a phylogenetic model.
+from DNA alignments in 1981 (\cite{felsenstein1981}), where  
+the first Bayesian tools emerged in 1996 (\cite{rannala1996}), the latter
+providing for unprecedented flexibility in the setup of a phylogenetic model.
 
 Currently, the most popular Bayesian phylogenetics tools are BEAST (\cite{beast})
-and its successor BEAST2 (\cite{beast2}),  
-MrBayes (\cite{mrbayes}) and RevBayes (\cite{revbayes}). 
-Each of these use efficient tree sampling 
-routines to create a posterior probability distribution of the to-be-inferred 
-phylogenetic tree. But next to the computational speed, they also allow
-to incorporate known or hypothesized structure of the phylogenetic 
-tree-to-be-inferred through the model priors.
+and its successor BEAST2 (\cite{beast2}), MrBayes (\cite{mrbayes}) and RevBayes (\cite{revbayes}). 
+They allow to incorporate known or hypothesized structure of a phylogenetic 
+tree-to-be-inferred through model priors. From these priors and an alignment 
+of DNA, RNA or protein sequences, a posterior is created. A posterior
+is a collection of phylogenies and parameter estimates (of the model prior), 
+in which likelier combinations are present more often.
+Each of these tools use efficient tree sampling routines to create an 
+informative posterior fast.
 
@@
 and branch termination (extinction).
-The effect of choosing a (potentially wrong) prior will have its effect on
+The effect of choosing a (potentially wrong) prior affects
 the posterior. For example, recently, it was shown that the effect
@@
 
-At the present time, for sexually producing animals, all we have at our
-disposal are tree priors that assume that speciation
-is instantaneous, where we know that speciation is a gradual process.
-When assuming big populations sizes (thus the effect of sampling to 
+The contemporary phylogenetic tools provide only for tree priors 
+that assume speciation is instantaneous, where we know that, 
+in animals, speciation is a gradual process.
+When big populations sizes can be assumed big (thus the effect of sampling to 
 be small), the (constant-rate) birth-death (BD) model is a commonly 
-used tree prior, which does not incorporate the temporal aspect of speciation.
+used tree prior, which ignores the temporal aspect of speciation.
 The protracted birth-death (PBD) model, an extension of 
@@
 BD model allows for a speciation event being recognized immediately,
-where the PBD model acknowledges that this will take time.
+where the PBD model accounts for speciation needing time.
 
@@
 so in research that investigates (mostly) older species, a BD model would suffice. 
-Computationally, the BD model is simpler, thus more convenient, model. 
+Computationally, the BD model is simpler, thus more light-weight, model. 
 Methodological, there is no computational tool where 
@@
 as a 'true' species tree, as incipient species may give rise to paraphylies.
-It may be possible to establish a protocol to sample a species tree from an 
-incipient species tree, but this has not yet been explored [NOTE: except by Simonet].
 
@@
 We assume that the speciation initiation and extinction rates
-of an incipient and good species are equal ($b_i = b_g$ and $\mu_i = \mu_g$),
-as this will result in monophyletic incipient species trees.
+of an incipient and good species are equal ($b = b_i = b_g$ and $\mu = \mu_i = \mu_g$),
+as this enables use to do more replicates 
+\textbf{[NOTE: I am unconvinced. I think we should also explore $b_i \neq b_g$ and $\mu_i \neq \mu_g$]}.
 We only simulate a PBD process for phylogenies in which
@@
 From an incipient species tree, we sample a species tree. To do
-so, from each species a sub-species must be chosen to represent
+so, from each species a sub-species is chosen to represent
 the good species as a whole. To clarify, it may be that an
-incipient species branched of from a good species. Both of these
-subspecies are recognized as the good species, and both can be 
-picked as a representative. Here we use three sampling scenario's,
+incipient species branched of from its mother lineage. 
+Both of these subspecies are recognized as the good species of the 
+mother lineage, and both can be picked as an (equally good) representative of the 
+good mother species. 
+In this research, we use three sampling scenario's,
 in which we pick the most recent common ancestor (MRCA), 
-most distant common ancestor (MDCA) or random subspecies. Note that only in few 
-cases (when speciation completion rate is low) the incipient species
-chosen will have an effect on the branch length distribution of species 
-tree. The simplest scenario requires a species, with one good and
-one incipient lineage, after which either lineages gives 
-rise to a new good species, before the first incipient lineage
-completes speciation itself.
+most distant common ancestor (MDCA) or random subspecies. 
+The scenario in which sampling has an effect on the branch length
+distributions of the species tree, is when a species in the proces
+of speciation, gives rise to a new incipient lineage that finishes
+speciation before the ancestral completes speciation itself.
 
@@
 nucleotide substitution model, in which all nucleotide-to-nucleotide transitions
-are equally likely. Although this may seem as a simplification, in our Bayesian
-inference (see below) we use this exact site model as the (obviously correct) site model prior.
+are equally likely. Although this may seem as a simplification, 
+in our Bayesian inference (see below) we use this exact site model as the (obviously correct) site model prior.
 The mutation rate is set in such a way to maximize chronologic information.
-To do so, the mutation rate is set to expect on average 1 (possibly silent) mutation per nucleotide
+To do so, the mutation rate is set to expect on average one (possibly silent) mutation per nucleotide
 between crown age and present, which equates to $\frac{1}{15}$ mutations
@@
 The DNA sequence length is chosen to provide a
-resolution of a 1K years, that is, to have one expected nucleotide change 
-per 1k years per lineage on average. As 1 nucleotide is expected 
-to have on average 1 (possibly silent) mutation per 15 million years, 15K
-nucleotides provides for 1 mutation per alignment per 1K years (which is
+resolution of $10^3$ years, that is, to have one expected nucleotide change 
+per $10^3$ years per lineage on average. As one nucleotide is expected 
+to have on average one (possibly silent) mutation per 15 million years, $15 \cdot 10^3$
+nucleotides results in 1 mutation per alignment per $10^3$ years (which is
 coincidentally the same as \cite{moller2018}). 
@@
 using the phylogetic tool BEAST2 \cite{beast2} using the 
-pirouette (\cite{pirouette}) package. As our site model, we assume a Jukes-Cantor 
+pirouette (\cite{pirouette}) package. For our site model, we assume a Jukes-Cantor 
 nucleotide substitution model, as we used that in the simulation of the alignment.
 For our clock model, we assume a strict clock with the same fixed rate as 
-used in the simulation of the alignment. The tree prior assumed is the BD model, 
+used in the simulation of the alignment \textbf{[NOTE: \cite{moller2018} did not use a
+fixed clock rate, I do not see why]}. The tree prior assumed is the BD model, 
 as this simplification is the goal of this research. 
 Additionally, we assume a MRCA prior with a normal distribution
-with a mean of the crown age, and a standard deviation of 0.001 time units. 0.001
-time units equates to 1000 years, which is the same resolution as which the
-alignment is created. The MCMC chain is run to generate 1111 states,
+with a mean of the crown age, and a standard deviation of $0.5 \cdot 10^{-3}$ time units,
+resulting in 95\% of the crown ages used have the same resolution (of $10^{-3}$ time 
+units) as the alignment. The MCMC chain is run to generate 1111 states,
 of which the first 10\% (also called the 'burn-in') is removed. Of the remaining
-1000 MCMC states [NOTE: Why 1000? Why not 250?], 
-the effective sample size (ESS) of the posterior must at least be 200
+1000 MCMC states \textbf{[NOTE: Why 1000? Why not 250? I would say 250 is preferable, as
+the information will be more dense]}, the effective sample size (ESS) of the posterior must at least be 200
 for a strong enough inference (\cite{beastbook}). An ESS can be increased by increasing
@@
 The nLTT statistic equates to the area between the normalized
-lineages-through-time-plots of two phylogenies, which will be a value between 
-zero (for identical phylogenies) and one. As the nLTT statistic increases
-when phylogenies get increasingly different, we use the nLTT statistic
-as a measure of inference error. Comparing the one (sampled) species tree
+lineages-through-time-plots of two phylogenies, which has a range 
+from zero (for identical phylogenies) to one. We use inference error 
+and nLTT statistic synonymously. Comparing the one (sampled) species tree
 with each of the posterior's species trees, a distribution nLTT statistics
@@
 
-There will be two data sets produced by this research.
-The first data set will be a general balanced data set to chart
+Two data sets are produced by this research.
+The first data set is a general balanced data set to chart
 the effect of the biological parameters on the nLTT statistic
@@
 
-Each data set will be stored as a comma-seperated file. We placed 
-an upper limit for this data set's size at the maximum size of a table (called 
-a data frame) of the R programming language, in which such a table may 
-have maximally $2^{32}-1$ cells and occupy less then 2 gigabytes in memory.
-To provide for copying such a table (thus, allowing it to be present in
-memory twice), the size of the table has an upper limit of half its maximal size.
-Each row will contain the parameter set and the generated nLTT 
+Each data set is stored as a comma-seperated file. 
+As a theoretical study such as this could theoretically (pun intended)
+produce an infinitely big data set, we placed an upper limit 
+for this data set's size. This size is chosen as such the 
+the R programming language (\cite{r}) can contain a data file in
+memory twice. Each row will contain a parameter set and the generated nLTT 
 statistics (see \ref{table:specs} for the exact data specification).
-From these constraints, this allows for 20K rows.
+From these constraints, this allows for $2 \cdot 10^3$ rows. As there are 48 combinations
+of biological parameter \textbf{[NOTE: calculate exact number when Rampal decides
+for $b_i = b_g$ only or for also allowing $b_i \neq b_g$]}, there will be 
+168 {[NOTE: recalculate, as before]} replicates per biological parameter set.
 
-Our results will show the general effect of the biological parameters (speciation initiation, 
-speciation completion and extinction rate) using the balanced data set,
+Our results show the general effect of the biological 
+parameters ($b$, $\lambda$, $\mu$) using the balanced data set,
 and the effect of sampling using the data set conditioned on sampling having an effect.
-In both cases, the nLTT statistics distribution will be plotted per
-biological parameter set using a violin plot. A violin plot is chosen to
-maintain information about the distribution. When displaying the effect of a 
-sampling regime, per regime, the nLTT statistics distribution is shown per
-biological parameter set.
+In both cases, the nLTT statistics distribution is plotted per
+biological parameter set using a violin plot, as such a plot 
+maintains information about distribution.
 We predict that nLTT statistic values increase 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
 % Please supply, as a separate list, the definitions of field-specific terms used in your article.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\begin{table}
+  \centering 
+  \begin{tabular}{l p{0.8\textwidth}}
+    \hline
+    Term                  & Definition \\
+    \hline
+    \hline
+    Phylogenetics         & The inference of evolutionary relationships of groups of organisms using genetics \\
+    Model prior           & Knowledge or assumptions about the onotogeny of evolutionary histories \\
+    Posterior             & A collection of phylogenies and parameter estimates, in which likelier combinations are present more \\
+    Protracted speciation & The process in which speciation takes two events to complete: a speciation initiation and a speciation completion event  \\
+    Speciation initiation & The start of a speciatiation event, in which initially the new species-to-be is not recognized as such \\
+    Speciation completion & The end of a speciatiation event, in which the new species is recognized as such \\
+    \hline
+  \end{tabular}
+  \caption{
+    Glossary [NOTE: this is requested by the journal]
+  }
+  \label{table:glossary}
+\end{table}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
+\textbf{[NOTE: journal does not request for this. Suggest to remove, but how to acknowledge Peregrine otherwise?]}
 We would like to thank the Center for Information Technology of the University of Groningen for their support
@@
 
+\textbf{[NOTE: journal does not request for this]}
 RJCB and RSE conceived the idea for this experiment and package. 

### 2018-04-26 article.tex
EOD

@@
 which we explore in a factorial fashion, excluding some combinations. 
-We assume that the speciation initiation and extinction rates
-of an incipient and good species are equal ($b = b_i = b_g$ and $\mu = \mu_i = \mu_g$),
-as this enables use to do more replicates 
-\textbf{[NOTE: I am unconvinced. I think we should also explore $b_i \neq b_g$ and $\mu_i \neq \mu_g$]}.
 We only simulate a PBD process for phylogenies in which
-speciation initiation exceeds extinction rate ($b > \mu$),
-and in which their difference is not too big ($b - \mu < 0.8$), 
+speciation initiation exceeds extinction rate ($b_i > \mu_i$ and $b_g > \mu_g$),
+and in which the expected number of taxa is less than 1000:
+
+```
+EQUATION SHOWING THE EXPECTED NUMBER OF TAXA < 1000
+```
+
 to prevent overly taxon-poor and taxon-rich phylogenies respectively.
@@
 of which the first 10\% (also called the 'burn-in') is removed. Of the remaining
-1000 MCMC states \textbf{[NOTE: Why 1000? Why not 250? I would say 250 is preferable, as
-the information will be more dense]}, the effective sample size (ESS) of the posterior must at least be 200
+1000 MCMC states, the effective sample size (ESS) of the posterior must at least be 200
 for a strong enough inference (\cite{beastbook}). An ESS can be increased by increasing
@@
 \textbf{[NOTE: journal does not request for this]}
-RJCB and RSE conceived the idea for this experiment and package. 
-RJCB created and tested the experiment and package, 
+RSE conceived the idea for this experiment. 
+RJCB created and tested the experiment, 
 and wrote the first draft of the manuscript. 
@@
     \hline
-    $b = b_g = b_i$       & Speciation initiation rate & 0.1, 0.5, 1.0 \\
+    $b_g$                 & Speciation initiation rate of a good species & 0.1, 0.5, 1.0 \\
+    $b_i$                 & Speciation initiation rate of an incipient species & 0.1, 0.5, 1.0 \\
     $\lambda$             & Speciation completion rate & 0.1, 0.3, 1.0, $\infty$ \\
-    $\mu = \mu_g = \mu_i$ & Extinction rate & 0.0, 0.1, 0.2, 0.4 \\
+    $\mu_g$               & Extinction rate of a good species & 0.0, 0.1, 0.2, 0.4 \\
+    $\mu_i$               & Extinction rate of an incipient species & 0.0, 0.1, 0.2, 0.4 \\
     \hline

### 2018-04-30 article.tex
Implemented feedback, thanks @rsetienne

@@
 
+% Target journal: Molecular Phylogenetics and Evolution
 % Short communications of approximately 3000 words are also accepted. 
@@
 \usepackage{authblk}
-\title{The error when inferring phylogenies with incipient species by a birth-death model}
-% \subtitle{Should protracted speciation be incorporated in phylogenetic tree construction methods?}
+\title{The error in Bayesian phylogenetic reconstruction when speciation is not instantaneous}
 
@@
 
-  Furthermore, we identify an important issue related to protracted speciation:
-  because the tree produced by the protracted birth-death process 
-  is not necessarily monophyletic, we cannot speak of "the" species tree, but we
-  have to sample among the incipient species to represent species. 
-  % We show that different ways of sampling the representative species results in ....
-
-  % Does not seem to fit here: 
-  %  
-  % 
-  % Furthermore, we identify an important issue related to protracted speciation. 
-  % Because the tree produced by the protracted birth-death process 
-  % is not necessarily monophyletic, we cannot speak of "the" species tree. 
-  % Instead we have to sample among the incipient species to represent species. 
-  % We show that different ways of sampling the representative species results in ....
-
 \end{abstract}
@@
 
-The computational tools a contemporary phylogeneticist has at his/her disposal
-goes beyond the wildest imagination of those living three decades ago. 
+The computational tools that are currently available go beyond the wildest 
+imagination of those living four decades ago. 
 Advances in computational power allowed the first cladograms to be inferred 
-from DNA alignments in 1981 (\cite{felsenstein1981}), where  
-the first Bayesian tools emerged in 1996 (\cite{rannala1996}), the latter
-providing for unprecedented flexibility in the setup of a phylogenetic model.
+from DNA alignments in 1981 (\cite{felsenstein1981}), and  
+the first Bayesian tools emerged in 1996 (\cite{rannala1996}),
+providing unprecedented flexibility in the setup of a phylogenetic model.
 
@@
 and its successor BEAST2 (\cite{beast2}), MrBayes (\cite{mrbayes}) and RevBayes (\cite{revbayes}). 
-They allow to incorporate known or hypothesized structure of a phylogenetic 
-tree-to-be-inferred through model priors. From these priors and an alignment 
-of DNA, RNA or protein sequences, a posterior is created. A posterior
-is a collection of phylogenies and parameter estimates (of the model prior), 
-in which likelier combinations are present more often.
-Each of these tools use efficient tree sampling routines to create an 
-informative posterior fast.
+They allow to known or hypothesized structure of a phylogenetic 
+tree-to-be-inferred to be incorporated through model priors. 
+From these priors and an alignment 
+of DNA, RNA or protein sequences, they create a posterior distribution
+of phylogenies and parameter estimates (of the models used as a prior), 
+in which more probable combinations are represented more often.
+Each of these tools use efficient tree sampling routines to rapidly create an 
+informative posterior.
 
-In a Bayesian analysis, the model priors are explicitly specified.
-Those model priors can be grouped in three groups: (1) site model, which
-governs the nucleotide subsititution model, (2) clock model, specifying
+These model priors in Bayesian phylogenetic reconstruction 
+can be grouped in three groups: (1) site model, specifying 
+nucleotide substitutions, (2) clock model, specifying
 the rate of mutation per lineage in time, and (3) tree prior, embodying
-the speciation model behind branching events (speciation) 
+the speciation model underlying branching events (speciation) 
 and branch termination (extinction).
-The effect of choosing a (potentially wrong) prior affects
-the posterior. For example, recently, it was shown that the effect
-of choosing a tree prior biases the estimation of the molecular clock rate, 
-for DNA sequences of 100-1000 base pairs (\cite{moller2018}).
+The choice of a wrong site model (\cite{posada_and_bucklet_2004}), 
+clock model (\cite{baele_et_al_2012}) 
+or tree prior (\cite{moller2018, yang_and_ranalla_2005}) is known to affect
+the posterior.
 
-The contemporary phylogenetic tools provide only for tree priors 
-that assume speciation is instantaneous, where we know that, 
-in animals, speciation is a gradual process.
-When big populations sizes can be assumed big (thus the effect of sampling to 
-be small), the (constant-rate) birth-death (BD) model is a commonly 
-used tree prior, which ignores the temporal aspect of speciation.
+Current phylogenetic tools use tree priors 
+that assume speciation is instantaneous, whilst we know that, 
+speciation is often a gradual process.
+The (constant-rate) birth-death (BD) model is a commonly 
+used tree prior, but it ignores this temporal aspect of speciation.
 The protracted birth-death (PBD) model, an extension of 
@@
 species may go extinct, finish its speciation to become a good species, or give
-rise to new incipient species.
+rise to new incipient species. Protracted speciation may explain observed 
+declines in lineage accumulation (\cite{etienne_and_rosindell_2012}).
 
-The effect of using the (incorrect) BD tree prior for a PBD process is unknown.
-A potential problem in species conservation is that the number of 
-species is underestimated (see \cite{fennessy2016} for a clear example). 
-Additionally, protracted speciation may be one 
-explanation in the observed decline of speciation rates in 
-time (\cite{etienne2012}). Also, a BD model places the most recent common 
-ancestor (MRCA) of a young species duo closer to the present, as the
-BD model allows for a speciation event being recognized immediately,
-where the PBD model accounts for speciation needing time.
+% RSE: perhaps move to discussion
+% The effect of using the (incorrect) BD tree prior for a PBD process is unknown.
+% A potential problem in species conservation is that the number of 
+% species is underestimated (see \cite{fennessy2016} for a clear example). 
+% Also, a BD model places the most recent common 
+% ancestor (MRCA) of a young species duo closer to the present, as the
+% BD model allows for a speciation event being recognized immediately,
+% where the PBD model accounts for speciation needing time.
 
-There are multiple possiblities why the PBD model is relatively unexplored.
-Biologically, the PBD model is predicted to have an effect strongest in
-the present (as earlier speciation events are nearly always recognized),
-so in research that investigates (mostly) older species, a BD model would suffice. 
-Computationally, the BD model is simpler, thus more light-weight, model. 
-Methodological, there is no computational tool where 
-the PBD model fits in: every contemporary framework assumes 
-either an analysis at the species or subspecies level. In the PBD
-model, incipient species are the cause there is no such thing 
-as a 'true' species tree, as incipient species may give rise to paraphylies.
+Unfortunately, a tree prior according to this model, 
+providing the probability of a species tree under the PBD model, 
+is unavailable in current Bayesian phylogenetic tools. 
+Whilst an approximate formula for this probability has been derived (\cite{lambert_et_al_2015}) 
+and the approximation is very good (\cite{simonet_et_al_2018}), 
+it has not been implemented as tree prior yet. 
+There are various reasons for this. 
+First, the computation of this probability involves solving a set of 
+non-linear differential equations, and while this computation is quite fast, 
+it still takes much more time than the corresponding probability 
+of the BD model which is a simple analytical formula. 
+In a Bayesian MCMC chain, the tree prior probability must be calculated many times, 
+and hence the total computation will take considerably longer with a PBD tree prior. 
+Furthermore, the approximate probability is a probability for the species tree 
+assuming an underlying incipient species tree. 
+It can be safely used as tree prior when only one individual per species is sampled, 
+but if one has multiple samples per species -which is currently often the case- the methods 
+to account for this such as the multi-species coalescent (\cite{heled_and_drummond_2009}) 
+may not be compatible with the underlying incipient species tree. 
+More precisely, the phylogeny under the PBD model may contain paraphylies, 
+while the multi-species coalescent was developed exactly to avoid 
+this by explaining them as incomplete lineage sorting. 
+Because of the paraphylies there is no such thing as a true species tree in the PBD model. 
+To get a species-level tree one must sample one incipient species per species. 
+Which incipient species is sampled may therefore have an impact on the species tree.
 
-This research's goal is to explore the effect of using an overly simplistic
-BD prior on PBD simulated phylogenies. We provide a data set, 
-that quantifies the inference error made in general, and explores
-the effect of the way species trees are sampled from an incipient species tree.  
+Here we aim to explore the effect of using the
+BD prior on PBD simulated phylogenies, taking into account possible sampling effects. 
 In brief, we simulate protracted phylogenies using the PBD process,
-from which we sample a species tree. From that species tree, 
-we simulate a DNA sequence alignment. Then, BEAST2 uses these alignments
-to infer a posterior of phylogenies, using a BD prior. The difference
-between the (BD) posterior phylogenies and (PBD) species tree is quantified.
+from which we sample a species tree in two very different ways. Given this species tree, 
+we simulate a DNA sequence alignment. Then, we use BEAST2 on these alignments
+to infer a posterior of phylogenies, using a BD prior. We quantify the difference
+between the (BD) posterior phylogenies and the simulated (PBD) species tree.
 
@@
 speciation initiation exceeds extinction rate ($b_i > \mu_i$ and $b_g > \mu_g$),
-and in which the expected number of taxa is less than 1000:
-
-```
-EQUATION SHOWING THE EXPECTED NUMBER OF TAXA < 1000
-```
+and in which the expected number of taxa is less than 1000.
+As the analytic solution for the expected number of lineages 
+is too complex to be derived (\cite{etienne_and_rosindell_2012}),
+we use a numerical approximation instead \textbf{NOTE: maybe Lambert et al., 2014, 
+but the (closed access) article is unavailable from home}.
 
 to prevent overly taxon-poor and taxon-rich phylogenies respectively.
-The parameter values chosen are a 
-superset of \cite{etienne2014}, as these parameters result in reasonably
-sized phylogenies and allows us to compare results. 
-For the speciation initiation rate $b$, we'll use $0.1$, $0.5$ and $1.0$ 
+The parameter values chosen are based on the parameter sets used by 
+\cite{etienne2014}, as these parameters were shown to result in reasonably
+sized phylogenies and using the same set allows us to compare results. 
+For the speciation initiation rate $b$, we use $0.1$, $0.5$ and $1.0$ 
 speciation initiation events per (good species) lineage per time unit.
 The speciation completion rates used are $0.1$, $0.3$, $1.0$ and $10^9$ speciation completion
-events per (incipient species) lineage per time unit. For 
-$\lambda = \infty$ (where we assume that in this context $10^9 \approx \infty$), 
-the PBD model equals a BD model, which allows us to measure the baseline error.
+events per (incipient species) lineage per time unit. We used $10^9 \approx \infty$
+to mimic the BD model, because the PBD model reduces to the BD model for $\lambda = \infty$.
+This allows us to measure the baseline error, 
+which is the difference between inferred tree and true species tree that arises purely due to noise 
+because the generating model and the model used in inference are identical.
 The extinction rates used are $0.0$, $0.1$, $0.2$ and $0.4$ 
@@
 
-From each biological parameter set, a protracted birth-death tree is simulated,
+From each biological parameter set, we simulated a protracted birth-death tree,
 using the PBD package (\cite{pbd}) in the R programming language (\cite{r}), 
-with the same crown age as \cite{etienne2014} of 15 million years. 
+all with a crown age of 15 million years. 
 Each protracted birth-death tree uses a different random number
-generatior seed, and thus will be unique, resulting in a balanced 
+generatior seed, which makes all runs independent, resulting in a balanced 
 data set.
 
-From an incipient species tree, we sample a species tree. To do
-so, from each species a sub-species is chosen to represent
-the good species as a whole. To clarify, it may be that an
-incipient species branched of from its mother lineage. 
-Both of these subspecies are recognized as the good species of the 
-mother lineage, and both can be picked as an (equally good) representative of the 
-good mother species. 
-In this research, we use three sampling scenario's,
-in which we pick the most recent common ancestor (MRCA), 
-most distant common ancestor (MDCA) or random subspecies. 
-The scenario in which sampling has an effect on the branch length
-distributions of the species tree, is when a species in the proces
-of speciation, gives rise to a new incipient lineage that finishes
-speciation before the ancestral completes speciation itself.
+From each incipient species tree, we construct a species tree,
+by sampling one incipient/good species per good species. 
+For example, when an
+incipient species branched off from its mother lineage, 
+both of these subspecies are recognized as representing the species, 
+and hence both can be picked as an (equally good) representative of the species. 
+Here, we use three sampling scenario's,
+in which we pick the representative randomly or in such a way that this
+results in either the shortest or longest branch lengths. 
+See the supplementary information for a visualization of these sampling methods.
 
-From a species tree, we simulate a DNA alignment that has the same history
-as the phylogeny, using the phangorn package (\cite{phangorn}). 
-The nucleotides of the DNA alignment follow a Jukes-Cantor (\cite{jc69})
+Based on the sampled species tree, we simulate a DNA alignment that has the same history
+as this species tree, using the phangorn package (\cite{phangorn}). 
+We assume that the nucleotides of the DNA alignment follow a Jukes-Cantor (\cite{jc69})
 nucleotide substitution model, in which all nucleotide-to-nucleotide transitions
-are equally likely. Although this may seem as a simplification, 
-in our Bayesian inference (see below) we use this exact site model as the (obviously correct) site model prior.
-The mutation rate is set in such a way to maximize chronologic information.
-To do so, the mutation rate is set to expect on average one (possibly silent) mutation per nucleotide
+are equally likely. 
+In our Bayesian inference (see below) 
+we use the same site model as the (obviously correct) site model prior.
+One could explore other substitution models in the simulations 
+and in the Bayesian inference, 
+but we chose this simple model because we are primarily interested 
+in the effect of the choice of tree prior. 
+If anything, our results are conservative: with a more complex substitution model, 
+there will be more noise and hence our inference error will increase.
+We set the mutation rate in such a way to maximize the information contained in the alignment.
+To do so, we set the mutation rate such that we expect on average one (possibly silent) mutation per nucleotide
 between crown age and present, which equates to $\frac{1}{15}$ mutations
-per million year.
+per million years.
 The DNA sequence length is chosen to provide a
@@
 to have on average one (possibly silent) mutation per 15 million years, $15 \cdot 10^3$
-nucleotides results in 1 mutation per alignment per $10^3$ years (which is
+nucleotides result in 1 mutation per alignment per $10^3$ years (which is
 coincidentally the same as \cite{moller2018}). 
-The simulation of these DNA aligment follows a strict clock model, 
+The simulation of these DNA aligments follows a strict clock model, 
 which we will specify as the known clock model prior in the Bayesian inference.
 
-From an alignment, we run a Bayesian analysis and create a posterior, 
-using the phylogetic tool BEAST2 \cite{beast2} using the 
-pirouette (\cite{pirouette}) package. For our site model, we assume a Jukes-Cantor 
-nucleotide substitution model, as we used that in the simulation of the alignment.
+From an alignment, we run a Bayesian analysis and 
+create a posterior distribution of trees and parameters
+using the babette (\cite{babette}) package
+that sets the input parameters similar to BEAUti 2 and then runs BEAST2. 
+For our site model, we assume a Jukes-Cantor 
+nucleotide substitution model, as used in the simulation of the alignment.
 For our clock model, we assume a strict clock with the same fixed rate as 
 used in the simulation of the alignment \textbf{[NOTE: \cite{moller2018} did not use a
-fixed clock rate, I do not see why]}. The tree prior assumed is the BD model, 
-as this simplification is the goal of this research. 
-Additionally, we assume a MRCA prior with a normal distribution
-with a mean of the crown age, and a standard deviation of $0.5 \cdot 10^{-3}$ time units,
+fixed clock rate, I do not see why]}. The tree prior assumed in inference is the BD model, 
+because studying the effect of this assumption is the goal of this study. 
+We assumed an MRCA prior with a tight normal distribution
+around the crown age, by choosing the crown age as mean, and a standard deviation 
+of $0.5 \cdot 10^{-3}$ time units,
 resulting in 95\% of the crown ages used have the same resolution (of $10^{-3}$ time 
-units) as the alignment. The MCMC chain is run to generate 1111 states,
-of which the first 10\% (also called the 'burn-in') is removed. Of the remaining
+units) as the alignment. 
+We ran the MCMC chain to generate 1111 states,
+of which we removed the first 10\% (also called the 'burn-in'). 
+Of the remaining
 1000 MCMC states, the effective sample size (ESS) of the posterior must at least be 200
@@
 
-Each posterior's phylogeny is compared to the (sampled) species tree
+We compared each posterior phylogeny to the (sampled) species tree
 by the nLTT statistic (\cite{janzen2015}), using the nLTT package (\cite{nltt}). 
-The nLTT statistic equates to the area between the normalized
+The nLTT statistic is equal to the area between the normalized
 lineages-through-time-plots of two phylogenies, which has a range 
 from zero (for identical phylogenies) to one. We use inference error 
-and nLTT statistic synonymously. Comparing the one (sampled) species tree
-with each of the posterior's species trees, a distribution nLTT statistics
-is created.
+and nLTT statistic interchangeably. Comparing the one (sampled) species tree
+with each of the posterior species trees yields a distribution of nLTT statistics.
 
-Two data sets are produced by this research.
-The first data set is a general balanced data set to chart
-the effect of the biological parameters on the nLTT statistic
-distribution. In this data set, incipient species are sampled 
-randomly to represent a good species.
-The second data set charts the effect of sampling
-subspecies and only uses PBD trees in which this sampling
-has an effect. For each of these trees, we sample both MRCA and
-MDCA subspecies. We predict that these two most extreme sampling methods
-result in the most pronounced differences.
+We produce two data sets as a comma-seperated file.
+We set the number of replicates for each parameter combination such,
+that this file and a possible copy can be handled in R's memory.
+Each row will then contain a parameter set and the generated nLTT 
+statistics (see \ref{table:specs} for the exact data specification).
+The abovementioned memory constraints allows for $2 \cdot 10^3$ rows. 
+With 48 \textbf{[NOTE: recalculate]} combinations of biological parameter, there will be 
+168 \textbf{[NOTE: recalculate]} replicates per parameter set.
 
-Each data set is stored as a comma-seperated file. 
-As a theoretical study such as this could theoretically (pun intended)
-produce an infinitely big data set, we placed an upper limit 
-for this data set's size. This size is chosen as such the 
-the R programming language (\cite{r}) can contain a data file in
-memory twice. Each row will contain a parameter set and the generated nLTT 
-statistics (see \ref{table:specs} for the exact data specification).
-From these constraints, this allows for $2 \cdot 10^3$ rows. As there are 48 combinations
-of biological parameter \textbf{[NOTE: calculate exact number when Rampal decides
-for $b_i = b_g$ only or for also allowing $b_i \neq b_g$]}, there will be 
-168 {[NOTE: recalculate, as before]} replicates per biological parameter set.
-
-Our results show the general effect of the biological 
-parameters ($b$, $\lambda$, $\mu$) using the balanced data set,
-and the effect of sampling using the data set conditioned on sampling having an effect.
-In both cases, the nLTT statistics distribution is plotted per
-biological parameter set using a violin plot, as such a plot 
-maintains information about distribution.
-We predict that nLTT statistic values increase 
-with an increasing protractednedness (that is, a low speciation completion rate),
-but we cannot predict the the extent of this error, as it has never been 
-measured.
+For both data sets, we plot the nLTT statistics distribution per
+parameter set using a violin plot, as such a plot 
+maintains information about the distribution.
 
@@
     Phylogenetics         & The inference of evolutionary relationships of groups of organisms using genetics \\
-    Model prior           & Knowledge or assumptions about the onotogeny of evolutionary histories \\
-    Posterior             & A collection of phylogenies and parameter estimates, in which likelier combinations are present more \\
-    Protracted speciation & The process in which speciation takes two events to complete: a speciation initiation and a speciation completion event  \\
-    Speciation initiation & The start of a speciatiation event, in which initially the new species-to-be is not recognized as such \\
-    Speciation completion & The end of a speciatiation event, in which the new species is recognized as such \\
+    Model prior           & Knowledge or assumptions about the ontogeny of evolutionary histories \\
+    Posterior             & A collection of phylogenies and parameter estimates, in which more probable combinations (determined by the data and the model prior) are presented more frequently \\
+    Protracted speciation & The process in which speciation takes two events to complete: a speciation-initiation event and a speciation-completion event  \\
+    Speciation initiation & The start of a speciation event creating an incipient species \\
+    Speciation completion & The end of a speciation event, in which an incipient species is recognized as a good species \\
     \hline
@@
     $\sigma_c$            & Standard deviation around crown age & 0.001 \\
-    $M$                   & Sampling method & MRCA, MDCA or random \\
+    $M$                   & Sampling method & 'shortest', 'longest' or random \\
     $r$                   & Mutation rate & $\frac{1}{15}$ \\

### 2018-05-01 article.tex
Update figure
Improved text, recreate PDF and proprietary formats

@@
 and its successor BEAST2 (\cite{beast2}), MrBayes (\cite{mrbayes}) and RevBayes (\cite{revbayes}). 
-They allow to known or hypothesized structure of a phylogenetic 
-tree-to-be-inferred to be incorporated through model priors. 
-From these priors and an alignment 
-of DNA, RNA or protein sequences, they create a posterior distribution
-of phylogenies and parameter estimates (of the models used as a prior), 
+They allow to incorporate known or hypothesized structure of a phylogenetic 
+tree-to-be-inferred through model priors. 
+From these priors and an alignment of DNA, RNA or protein sequences, they create a posterior distribution
+of parameter estimates (of the models used as a prior) and phylogenies, 
 in which more probable combinations are represented more often.
@@
 
-% RSE: perhaps move to discussion
-% The effect of using the (incorrect) BD tree prior for a PBD process is unknown.
-% A potential problem in species conservation is that the number of 
-% species is underestimated (see \cite{fennessy2016} for a clear example). 
-% Also, a BD model places the most recent common 
-% ancestor (MRCA) of a young species duo closer to the present, as the
-% BD model allows for a speciation event being recognized immediately,
-% where the PBD model accounts for speciation needing time.
-
 Unfortunately, a tree prior according to this model, 
@@
 this by explaining them as incomplete lineage sorting. 
-Because of the paraphylies there is no such thing as a true species tree in the PBD model. 
+Because of these paraphylies there is no such thing as a true species tree in the PBD model. 
 To get a species-level tree one must sample one incipient species per species. 
@@
 speciation initiation exceeds extinction rate ($b_i > \mu_i$ and $b_g > \mu_g$),
-and in which the expected number of taxa is less than 1000.
-As the analytic solution for the expected number of lineages 
+and in which the expected number of extant good species is less than 1000.
+As the analytic solution for the expected number of extant good species 
 is too complex to be derived (\cite{etienne_and_rosindell_2012}),
-we use a numerical approximation instead \textbf{NOTE: maybe Lambert et al., 2014, 
-but the (closed access) article is unavailable from home}.
-
-to prevent overly taxon-poor and taxon-rich phylogenies respectively.
+we a numerical approximation \textbf{NOTE: I will suggest somewhen/somewhere else if this
+may become a function of the PBD package}.
+We use 1000 good species as a threshold, to prevent overly taxon-poor and taxon-rich phylogenies respectively.
 The parameter values chosen are based on the parameter sets used by 
@@
 sized phylogenies and using the same set allows us to compare results. 
-For the speciation initiation rate $b$, we use $0.1$, $0.5$ and $1.0$ 
-speciation initiation events per (good species) lineage per time unit.
+For the speciation initiation rates of good and incipient species, $b_g$ and $b_i$ 
+respectively, we use $0.1$, $0.5$ and $1.0$ speciation initiation events per good/incipient species per time unit.
 The speciation completion rates used are $0.1$, $0.3$, $1.0$ and $10^9$ speciation completion
-events per (incipient species) lineage per time unit. We used $10^9 \approx \infty$
+events per (incipient species) species per time unit. We used $10^9 \approx \infty$
 to mimic the BD model, because the PBD model reduces to the BD model for $\lambda = \infty$.
@@
 because the generating model and the model used in inference are identical.
-The extinction rates used are $0.0$, $0.1$, $0.2$ and $0.4$ 
-extinction events per (good or incipient) lineage per time unit.
+The extinction rates of good and incipient species, $\mu_g$ and $\mu_i$ respectively, 
+that we used are $0.0$, $0.1$, $0.2$ and $0.4$ 
+extinction events per good/incipient species per time unit.
 
@@
 generatior seed, which makes all runs independent, resulting in a balanced 
-data set.
+data set. \textbf{NOTE: Rampal suggested to re-use same seeds. I see in Progress
+Report of week 50 of 2015 this was already abandoned in favor of a more smooth data set.
+I do check that nLTT values of multiple random alignments on the same species tree
+result in same distribution, but that may not be written down here}
 
@@
 are equally likely. 
-In our Bayesian inference (see below) 
-we use the same site model as the (obviously correct) site model prior.
+In our Bayesian inference (see below) we use the same site model as the (obviously correct) site model prior.
 One could explore other substitution models in the simulations 
@@
 For our clock model, we assume a strict clock with the same fixed rate as 
-used in the simulation of the alignment \textbf{[NOTE: \cite{moller2018} did not use a
-fixed clock rate, I do not see why]}. The tree prior assumed in inference is the BD model, 
+used in the simulation of the alignment. 
+The tree prior assumed in inference is the BD model, 
 because studying the effect of this assumption is the goal of this study. 
@@
 Of the remaining
-1000 MCMC states, the effective sample size (ESS) of the posterior must at least be 200
+1000 MCMC states, the effective sample size (ESS) of the posterior \textbf{NOTE: there
+is a parameter estimate called 'posterior'. I choose to pick that one, and I assume
+it is the wiser choice over 'prior' and parameter estimates. Agree?} must at least be 200
 for a strong enough inference (\cite{beastbook}). An ESS can be increased by increasing
@@
 by the nLTT statistic (\cite{janzen2015}), using the nLTT package (\cite{nltt}). 
-The nLTT statistic is equal to the area between the normalized
+The nLTT statistic equals the area between the normalized
 lineages-through-time-plots of two phylogenies, which has a range 
@@
 parameter set using a violin plot, as such a plot 
-maintains information about the distribution.
+maintains information about the distribution. To simplify the interpretation
+of these plots, only nLTT statistics distribution are shown 
+for $\lambda_g = \lambda_i$ and $\mu_g = \mu_i$.
 

### 2018-05-07 article.tex
Fixed grammar mistake

@@
 is too complex to be derived (\cite{etienne_and_rosindell_2012}),
-we a numerical approximation \textbf{NOTE: I will suggest somewhen/somewhere else if this
+we use a numerical approximation \textbf{NOTE: I will suggest somewhen/somewhere else if this
 may become a function of the PBD package}.

### 2018-05-09 article.tex
Implemented feedback, thanks @rsetienne

--- 2018-05-01 article.tex
@@
 
-The computational tools that are currently available go beyond the wildest 
+The computational tools that are currently available 
+to the phylogeneticists go beyond the wildest 
 imagination of those living four decades ago.
@@
 
-These model priors in Bayesian phylogenetic reconstruction 
-can be grouped in three groups: (1) site model, specifying 
+The model priors in Bayesian phylogenetic reconstruction 
+can be grouped into three categories: (1) site model, specifying 
 nucleotide substitutions, (2) clock model, specifying
-the rate of mutation per lineage in time, and (3) tree prior, embodying
-the speciation model underlying branching events (speciation) 
+the rate of mutation per lineage in time, and (3) tree model, 
+constituting the speciation model underlying branching events (speciation) 
 and branch termination (extinction).
@@
 that assume speciation is instantaneous, whilst we know that, 
-speciation is often a gradual process.
+speciation is often a gradual process \notebf{[REF]}.
 The (constant-rate) birth-death (BD) model is a commonly 
@@
 while the multi-species coalescent was developed exactly to avoid 
-this by explaining them as incomplete lineage sorting. 
+these by explaining them as arising from incomplete lineage sorting. 
 Because of these paraphylies there is no such thing as a true species tree in the PBD model. 
@@
 and in which the expected number of extant good species is less than 1000.
-As the analytic solution for the expected number of extant good species 
-is too complex to be derived (\cite{etienne_and_rosindell_2012}),
-we a numerical approximation \textbf{NOTE: I will suggest somewhen/somewhere else if this
-may become a function of the PBD package}.
+{NOTE: accoring to Rampal this has been solved analytically (for the Protracted Birth and Death model). Where?}.
 We use 1000 good species as a threshold, to prevent overly taxon-poor and taxon-rich phylogenies respectively.
@@
 respectively, we use $0.1$, $0.5$ and $1.0$ speciation initiation events per good/incipient species per time unit.
-The speciation completion rates used are $0.1$, $0.3$, $1.0$ and $10^9$ speciation completion
-events per (incipient species) species per time unit. We used $10^9 \approx \infty$
+The speciation completion rates we use are $0.1$, $0.3$, $1.0$ and $10^9$ speciation completion
+events per (incipient species) species per time unit. We use $10^9 \approx \infty$
 to mimic the BD model, because the PBD model reduces to the BD model for $\lambda = \infty$.
@@
 which is the difference between inferred tree and true species tree that arises purely due to noise 
-because the generating model and the model used in inference are identical.
+because the generating model and the model used in inference are identical in this case.
 The extinction rates of good and incipient species, $\mu_g$ and $\mu_i$ respectively, 
-that we used are $0.0$, $0.1$, $0.2$ and $0.4$ 
+that we use are $0.0$, $0.1$, $0.2$ and $0.4$ 
 extinction events per good/incipient species per time unit.
 
-From each biological parameter set, we simulated a protracted birth-death tree,
+From each biological parameter set, we simulate a protracted birth-death tree,
 using the PBD package (\cite{pbd}) in the R programming language (\cite{r}), 
@@
 generatior seed, which makes all runs independent, resulting in a balanced 
-data set. \textbf{NOTE: Rampal suggested to re-use same seeds. I see in Progress
-Report of week 50 of 2015 this was already abandoned in favor of a more smooth data set.
-I do check that nLTT values of multiple random alignments on the same species tree
-result in same distribution, but that may not be written down here}
+data set. \textbf{NOTE: maybe discuss with Rampal to remove confusion. I see if I can get the
+students to make a histogram of number of taxa in the incipient species trees, as Rampal
+requires a geometric distribution}
 
@@
 and hence both can be picked as an (equally good) representative of the species. 
-Here, we use three sampling scenario's,
+Here, we use three sampling scenarios,
 in which we pick the representative randomly or in such a way that this
@@
 Based on the sampled species tree, we simulate a DNA alignment that has the same history
-as this species tree, using the phangorn package (\cite{phangorn}). 
+as this species tree, using the \verb;phangorn; package (\cite{phangorn}). 
 We assume that the nucleotides of the DNA alignment follow a Jukes-Cantor (\cite{jc69})
@@
 create a posterior distribution of trees and parameters
-using the babette (\cite{babette}) package
+using the \verb;babette; (\cite{babette}) package
 that sets the input parameters similar to BEAUti 2 and then runs BEAST2. 
@@
 because studying the effect of this assumption is the goal of this study. 
-We assumed an MRCA prior with a tight normal distribution
+We assume an MRCA prior with a tight normal distribution
 around the crown age, by choosing the crown age as mean, and a standard deviation 
 of $0.5 \cdot 10^{-3}$ time units,
-resulting in 95\% of the crown ages used have the same resolution (of $10^{-3}$ time 
+resulting in 95\% of the crown ages inferred have the same resolution (of $10^{-3}$ time 
 units) as the alignment. 
 We ran the MCMC chain to generate 1111 states,
-of which we removed the first 10\% (also called the 'burn-in'). 
+of which we remove the first 10\% (also called the 'burn-in'). 
 Of the remaining
@@
 is a parameter estimate called 'posterior'. I choose to pick that one, and I assume
-it is the wiser choice over 'prior' and parameter estimates. Agree?} must at least be 200
+it is the wiser choice over 'prior' and parameter estimates. Must discuss} must at least be 200
 for a strong enough inference (\cite{beastbook}). An ESS can be increased by increasing
 the number of samples or decreasing the autocorrelation between samples. 
-Would the ESS be less than 200, we decrease autocorrelation by doubling 
+If the ESS is less than 200, we decrease autocorrelation by doubling 
 the MCMC sampling interval of that simulation, until the ESS exceeds 200.
 
-We compared each posterior phylogeny to the (sampled) species tree
+We compare each posterior phylogeny to the (sampled) species tree
 by the nLTT statistic (\cite{janzen2015}), using the nLTT package (\cite{nltt}). 
@@
 from zero (for identical phylogenies) to one. We use inference error 
-and nLTT statistic interchangeably. Comparing the one (sampled) species tree
+and nLTT statistic interchangeably. Comparing the simulated species tree
 with each of the posterior species trees yields a distribution of nLTT statistics.
@@
     $f_i$                 & MCMC sampling interval & 1K or more \\
-    $R_i$                 & RNG seed incipient tree & 1 to 20K \\
+    $R_i$                 & RNG seed incipient tree and randomly sampled species tree & 1 to 20K \\
     $R_a$                 & RNG seed alignment simulation & $R_i$ \\
@@
     the biological parameter set. Sampling method $M$ is random for the general
-    data set. For the data set exploring the effect of sampling, MRCA is
-    used for odd values of $R_i$, and MDCA is used for even values of $R_i$.
+    data set. For the data set exploring the effect of sampling, we use 'shortest'
+    for odd values of $R_i$, and 'longest' for even values of $R_i$.
     $R_i$ is 1 for the first simulation, 2 for the next, etcetera.

### 2018-05-14 article.tex
Feedback

@@
 that assume speciation is instantaneous, whilst we know that, 
-speciation is often a gradual process \notebf{[REF]}.
+speciation is often a gradual process (\cite{schluter2009}).
 The (constant-rate) birth-death (BD) model is a commonly 
@@
 and in which the expected number of extant good species is less than 1000.
-{NOTE: accoring to Rampal this has been solved analytically (for the Protracted Birth and Death model). Where?}.
+\textbf{[NOTE: accoring to Rampal this has been solved analytically (for the Protracted Birth and Death model). Where?]}.
 We use 1000 good species as a threshold, to prevent overly taxon-poor and taxon-rich phylogenies respectively.
@@
 generatior seed, which makes all runs independent, resulting in a balanced 
-data set. \textbf{NOTE: maybe discuss with Rampal to remove confusion. I see if I can get the
-students to make a histogram of number of taxa in the incipient species trees, as Rampal
-requires a geometric distribution}
+data set. \textbf{[NOTE: maybe discuss with Rampal to remove confusion. Rampal requests a histogram of branch lengths to follow a geometric distribution]}
 
@@
 Of the remaining
-1000 MCMC states, the effective sample size (ESS) of the posterior \textbf{NOTE: there
+1000 MCMC states, the effective sample size (ESS) of the posterior \textbf{[NOTE: there
 is a parameter estimate called 'posterior'. I choose to pick that one, and I assume
-it is the wiser choice over 'prior' and parameter estimates. Must discuss} must at least be 200
+it is the wiser choice over 'prior' and parameter estimates. Must discuss]} must at least be 200
 for a strong enough inference (\cite{beastbook}). An ESS can be increased by increasing

### 2018-05-22 article.tex
EOD

@@
 that sets the input parameters similar to BEAUti 2 and then runs BEAST2. 
-For our site model, we assume a Jukes-Cantor 
-nucleotide substitution model, as used in the simulation of the alignment.
-For our clock model, we assume a strict clock with the same fixed rate as 
-used in the simulation of the alignment. 
-The tree prior assumed in inference is the BD model, 
-because studying the effect of this assumption is the goal of this study. 
+For our site and clock model, we assume a Jukes-Cantor 
+nucleotide substitution model and strict clock model, 
+as those are also used for simulating that alignment.
+We set the BD model as a tree prior, 
+as gauging the effect of this incorrect assumption is the goal of this study. 
 We assume an MRCA prior with a tight normal distribution

### 2018-05-23 article.tex
Update figures
Update text

@@
 
+\textbf{[NOTE: I will propose different parameter combinations, due to computational and aesthetic problems]}
 The PBD model has five biological parameters (see \ref{table:parameters}), 
@@
 \begin{figure}[!htbp]
-  \includegraphics[width=0.8\textwidth]{fig_nltt_stats_per_setup.png}
+  \includegraphics[width=0.8\textwidth]{fig_general.png}
   \caption{

### 2018-05-28 article.tex
EOD

--- 2018-05-22 article.tex
@@
 which we explore in a factorial fashion, excluding some combinations.
-We only simulate a PBD process for phylogenies in which
-speciation initiation exceeds extinction rate ($b_i > \mu_i$ and $b_g > \mu_g$),
-and in which the expected number of extant good species is less than 1000.
-\textbf{[NOTE: accoring to Rampal this has been solved analytically (for the Protracted Birth and Death model). Where?]}.
+We only simulate a PBD process for those combinations in which 
+95\% of all simulated phylogenies are expected to have less than 1000 extant good species.
+\textbf{[NOTE: use Rampals newest code. That new code assumes sirg = siri?]}.
 We use 1000 good species as a threshold, to prevent overly taxon-poor and taxon-rich phylogenies respectively.
@@
 For the speciation initiation rates of good and incipient species, $b_g$ and $b_i$ 
-respectively, we use $0.1$, $0.5$ and $1.0$ speciation initiation events per good/incipient species per time unit.
+respectively, we use $0.3$ and $0.5$ speciation initiation events per good/incipient species per time unit.
 The speciation completion rates we use are $0.1$, $0.3$, $1.0$ and $10^9$ speciation completion
@@
 The extinction rates of good and incipient species, $\mu_g$ and $\mu_i$ respectively, 
-that we use are $0.0$, $0.1$, $0.2$ and $0.4$ 
+that we use are $0.0$, $0.1$ and $0.2$ 
 extinction events per good/incipient species per time unit.
@@
 generatior seed, which makes all runs independent, resulting in a balanced 
-data set. \textbf{[NOTE: maybe discuss with Rampal to remove confusion. Rampal requests a histogram of branch lengths to follow a geometric distribution]}
+data set. \textbf{[NOTE: Rampal assumed runs with close seeds were related. I hope I have convinced him otherwise]}
 
@@
 is a parameter estimate called 'posterior'. I choose to pick that one, and I assume
-it is the wiser choice over 'prior' and parameter estimates. Must discuss]} must at least be 200
+it is the wisest choice of all BEAST2 parameter estimates, as displayed in table 4]} must at least be 200
 for a strong enough inference (\cite{beastbook}). An ESS can be increased by increasing
@@
   \caption{
-    Glossary [NOTE: this is requested by the journal]
+    Glossary
   }
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Acknowledgements}
+%\section{Acknowledgements}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-\textbf{[NOTE: journal does not request for this. Suggest to remove, but how to acknowledge Peregrine otherwise?]}
-We would like to thank the Center for Information Technology of the University of Groningen for their support
-and for providing access to the Peregrine high performance computing cluster.
+%\textbf{[NOTE: journal does not request for this. Suggest to remove, but how to acknowledge Peregrine otherwise?]}
+%We would like to thank the Center for Information Technology of the University of Groningen for their support
+%and for providing access to the Peregrine high performance computing cluster.
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Authors' contributions}
+%\section{Authors' contributions}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-\textbf{[NOTE: journal does not request for this]}
-RSE conceived the idea for this experiment. 
-RJCB created and tested the experiment, 
-and wrote the first draft of the manuscript. 
-RSE contributed substantially to revisions.
+%\textbf{[NOTE: journal does not request for this]}
+%RSE conceived the idea for this experiment. 
+%RJCB created and tested the experiment, 
+%and wrote the first draft of the manuscript. 
+%RSE contributed substantially to revisions.
 
@@
     \hline
-    $b_g$                 & Speciation initiation rate of a good species & 0.1, 0.5, 1.0 \\
-    $b_i$                 & Speciation initiation rate of an incipient species & 0.1, 0.5, 1.0 \\
+    $b_g$                 & Speciation initiation rate of a good species & 0.3, 0.5 \\
+    $b_i$                 & Speciation initiation rate of an incipient species & 0.3, 0.5 \\
     $\lambda$             & Speciation completion rate & 0.1, 0.3, 1.0, $\infty$ \\
-    $\mu_g$               & Extinction rate of a good species & 0.0, 0.1, 0.2, 0.4 \\
+    $\mu_g$               & Extinction rate of a good species & 0.0, 0.1, 0.2 \\
     $\mu_i$               & Extinction rate of an incipient species & 0.0, 0.1, 0.2, 0.4 \\

### 2018-05-29 article.tex
Fix overfull warnings
Fix all warnings

@@
 
+% Use multi-line cell
+\newcommand{\specialcell}[2][c]{%
+  \begin{tabular}[#1]{@{}c@{}}#2\end{tabular}}
+
 % Use double spacing
@@
 \linenumbers
+
+\hyphenation{
+  BEAST
+  Pa-ra-me-ter
+  Drum-mond 
+  Bayes-ian 
+  Mr-Bayes 
+  ap-proach-es 
+  Rev-Bayes 
+  cre-ate
+ }
 
@@
   % scientists in related disciplines.
-  Among the most popular tools are Bayesian approaches, such as BEAST, MrBayes and RevBayes, 
+  Among the most popular tools are Bayesian approaches, 
+  such as \\ BEAST, MrBayes and RevBayes, 
   that use efficient tree sampling routines to create a posterior probability distribution 
@@
 
-Currently, the most popular Bayesian phylogenetics tools are BEAST (\cite{beast})
-and its successor BEAST2 (\cite{beast2}), MrBayes (\cite{mrbayes}) and RevBayes (\cite{revbayes}). 
+Currently, the most popular Bayesian phylogenetics tools are \\ 
+BEAST (\cite{beast}) and its successor BEAST2 (\cite{beast2}), 
+MrBayes (\cite{mrbayes}) and RevBayes (\cite{revbayes}). 
 They allow to incorporate known or hypothesized structure of a phylogenetic 
 tree-to-be-inferred through model priors. 
-From these priors and an alignment of DNA, RNA or protein sequences, they create a posterior distribution
+From these priors and an alignment of DNA, RNA or protein sequences, 
+they create a posterior distribution
 of parameter estimates (of the models used as a prior) and phylogenies, 
@@
   \centering 
-  \begin{tabular}{l p{0.8\textwidth}}
+  \begin{tabular}{l p{0.65\textwidth}}
     \hline
@@
   \centering 
-  \begin{tabular}{l l l}
+  \begin{tabular}{p{0.05\textwidth} p{0.6\textwidth} p{0.2\textwidth}}
     \hline
-    Parameter             & Description & Values \\
+                          & Description & Values \\
     \hline
@@
     $\sigma_c$            & Standard deviation around crown age & 0.001 \\
-    $M$                   & Sampling method & 'shortest', 'longest' or random \\
+    $M$                   & Sampling method & S, L, R \\
     $r$                   & Mutation rate & $\frac{1}{15}$ \\
@@
     Overview of the 12 simulation parameters. Above the horizontal line is 
-    the biological parameter set. Sampling method $M$ is random for the general
+    the biological parameter set. The sampling methods are abbreviated as such: R denotes random
+    sampling, 'S' is 'shortest' and 'L' is 'longest'. Sampling method $M$ is random for the general
     data set. For the data set exploring the effect of sampling, we use 'shortest'

### 2018-05-30 article.tex
Use more site and clock models

@@
 % Target journal: Molecular Phylogenetics and Evolution
+%
+% From author guidelines:
+%
 % Short communications of approximately 3000 words are also accepted. 
@@
 % A short abstract of fewer than 200 words is acceptable.
+
+% @rsetienne:
+% At the moment, there are 87 parameter combinations for one site and one clock model
+% This allows for 40K / 87 = 460 replicates [*]
+% Using two site and two clock models would result in approx 100 replicates
+% Using four site and four clock models would result in approx 25 replicates
+%
+% I personally think the broadest data set will be most interesting,
+% because it gives a rough idea, but -more importantly- shows the scientific
+% community it can easily be done nowadays.
+%
+% [*] the number of replicates can only be determined by actually running the
+% experiment, as there are multiple constraints that I cannot test for
 
@@
 and branch termination (extinction).
-The choice of a wrong site model (\cite{posada_and_bucklet_2004}), 
+The choice of a wrong site model (\cite{posada_and_buckley_2004}), 
 clock model (\cite{baele_et_al_2012}) 
@@
 between the (BD) posterior phylogenies and the simulated (PBD) species tree.
+To gain more insight of incorrect prior choice, we also explore the effect of 
+two clock and site priors.
 
@@
 
-The PBD model has five biological parameters (see \ref{table:parameters}), 
+The PBD model has five biological parameters, depicted in table \ref{table:parameters}, 
 which we explore in a factorial fashion, excluding some combinations.
-We only simulate a PBD process for those combinations in which 
-95\% of all simulated phylogenies are expected to have less than 1000 extant good species.
-\textbf{[NOTE: use Rampals newest code. That new code assumes sirg = siri?]}.
-We use 1000 good species as a threshold, to prevent overly taxon-poor and taxon-rich phylogenies respectively.
+% HIERO
+We simulate a PBD process for those combinations in which the 
+95\% quantile has less than 1250 extant good species. The quantile is
+calculated with a recently added function to the PBD package,
+inspired on equation 6 of \cite{etienne2014}. This calculation
+assumes $b = b_g = b_i$, we used $b = \max(b_g, b_i)$.
+We use 1250 good species as a threshold, to prevent overly taxon-poor and taxon-rich phylogenies respectively.
 The parameter values chosen are based on the parameter sets used by 
@@
 generatior seed, which makes all runs independent, resulting in a balanced 
-data set. \textbf{[NOTE: Rampal assumed runs with close seeds were related. I hope I have convinced him otherwise]}
+data set. \textbf{[NOTE: Rampal assumed runs with close seeds were related. 
+I assume I have convinced him otherwise]}
 
@@
 as this species tree, using the \verb;phangorn; package (\cite{phangorn}). 
-We assume that the nucleotides of the DNA alignment follow a Jukes-Cantor (\cite{jc69})
+We set the nucleotides of the DNA alignment to follo a Jukes-Cantor (\cite{jc69})
 nucleotide substitution model, in which all nucleotide-to-nucleotide transitions
 are equally likely.
-In our Bayesian inference (see below) we use the same site model as the (obviously correct) site model prior.
-One could explore other substitution models in the simulations 
-and in the Bayesian inference, 
-but we chose this simple model because we are primarily interested 
-in the effect of the choice of tree prior. 
-If anything, our results are conservative: with a more complex substitution model, 
-there will be more noise and hence our inference error will increase.
+In our Bayesian inference (see below) we use the same site model as the (obviously correct) site model prior,
+but we also explore the effect of assuming a more complex site model prior.
+We predict with the more complex substitution model, 
+that there will be more noise and hence our inference error will increase.
 We set the mutation rate in such a way to maximize the information contained in the alignment.
@@
 The DNA sequence length is chosen to provide a
-resolution of $10^3$ years, that is, to have one expected nucleotide change 
+resolution of $10^3$ years, 
+% If sims take too long, we will decrease this resolution
+that is, to have one expected nucleotide change 
 per $10^3$ years per lineage on average. As one nucleotide is expected 
@@
 that sets the input parameters similar to BEAUti 2 and then runs BEAST2. 
-For our site and clock model, we assume a Jukes-Cantor 
-nucleotide substitution model and strict clock model, 
-as those are also used for simulating that alignment.
+For our site model, we assume either a Jukes-Cantor or GTR nucleotide substitution model.
+The Jukes-Cantor model is the correct one, as it is used for simulating that alignment,
+but the GTR model is the site model with the fewest assumptions and most flexibility.
+\textbf{[NOTE: Why then not use all four site models?]}
+For our clock model, we assume either a strict or relaxed log-normal \textbf{[NOTE: 
+Why that one? Why then not use all clock models?]} clock model. The strict clock
+model is the correct one, as it is used for simulating that alignment,
+but the relaxed log-normal clock model is the model with fewer assumptions and
+more flexibility.
 We set the BD model as a tree prior, 
@@
 Of the remaining
-1000 MCMC states, the effective sample size (ESS) of the posterior \textbf{[NOTE: there
-is a parameter estimate called 'posterior'. I choose to pick that one, and I assume
-it is the wisest choice of all BEAST2 parameter estimates, as displayed in table 4]} must at least be 200
+1000 MCMC states, the effective sample size (ESS) of the posterior \textbf{[NOTE: I chose 
+the ESS of the posterior over the ESS of the tree likelihood (and the 
+others displayed in table 4). For both something can be said. Agree on this choice?]} must at least be 200
 for a strong enough inference (\cite{beastbook}). An ESS can be increased by increasing
@@
 The abovementioned memory constraints allows for $2 \cdot 10^3$ rows. 
-With 48 \textbf{[NOTE: recalculate]} combinations of biological parameter, there will be 
-168 \textbf{[NOTE: recalculate]} replicates per parameter set.
+With 87 \textbf{[NOTE: volatile value, recalculate]} combinations of biological parameter, there will be 
+168 \textbf{[NOTE: volatile value, recalculate]} replicates per parameter set.
 
@@
 of these plots, only nLTT statistics distribution are shown 
-for $\lambda_g = \lambda_i$ and $\mu_g = \mu_i$.
+for $\lambda_g = \lambda_i$ and $\mu_g = \mu_i$. The general parameter set plot
+does not make an additional seperation between site model and clock model, but these
+are shown in the appendix. The sampling parameter set plot does seperate 
+on the sampling method used.
 
@@
     nLTT statistic distribution per biological parameter set, using the
-    balanced data set
+    balanced data set, for all clock and site models.
   }
@@
     nLTT statistic distribution per biological parameter set per sampling
-    regime, using the data set conditioned on sampling regime having an effect 
+    regime, using the data set conditioned on sampling regime having an effect.
   }
@@
 
+\begin{figure}[!htbp]
+  \includegraphics[width=0.8\textwidth]{fig_clock_model.png}
+  \caption{
+    nLTT statistic distribution per biological parameter set per clock model,
+    using the balanced data set 
+  }
+\end{figure}
+
+\begin{figure}[!htbp]
+  \includegraphics[width=0.8\textwidth]{fig_site_model.png}
+  \caption{
+    nLTT statistic distribution per biological parameter set per site model,
+    using the balanced data set 
+  }
+\end{figure}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
     $\sigma_c$            & Standard deviation around crown age & 0.001 \\
-    $M$                   & Sampling method & S, L, R \\
+    $M_s$                 & Sampling method & S, L, R \\
+    $M_c$                 & Clock model & S, RLN \\
+    $M_t$                 & Site model & JC69, GTR \\
     $r$                   & Mutation rate & $\frac{1}{15}$ \\
@@
     the biological parameter set. The sampling methods are abbreviated as such: R denotes random
-    sampling, 'S' is 'shortest' and 'L' is 'longest'. Sampling method $M$ is random for the general
-    data set. For the data set exploring the effect of sampling, we use 'shortest'
+    sampling, 'S' is 'shortest' and 'L' is 'longest'. Sampling method $M_s$ is random for the general
+    data set. 
+    The clock models are abbreviated as 'S' is a strict and 'RLN' is a relaxed log-normal model.
+    The site models are abbreviated as 'JC69' for Jukes-Cantor and 'GTR' for the generalised 
+    time-reversible model.
+    For the data set exploring the effect of sampling, we use 'shortest'
     for odd values of $R_i$, and 'longest' for even values of $R_i$.

### 2018-05-30 article.tex
Ready for feedback

@@
 \affil[1]{Groningen Institute for Evolutionary Life Sciences, University of Groningen, Groningen, The Netherlands}
-
-% Use multi-line cell
-\newcommand{\specialcell}[2][c]{%
-  \begin{tabular}[#1]{@{}c@{}}#2\end{tabular}}
 
@@
 which we explore in a factorial fashion, excluding some combinations.
-% HIERO
 We simulate a PBD process for those combinations in which the 
-95\% quantile has less than 1250 extant good species. The quantile is
+95\% quantile of the expected number of good species is less than 1250.
+The quantile is
 calculated with a recently added function to the PBD package,
-inspired on equation 6 of \cite{etienne2014}. This calculation
+based on equation 6 of \cite{etienne2014}. This calculation
 assumes $b = b_g = b_i$, we used $b = \max(b_g, b_i)$.

### 2018-05-31 article.tex
After meeting @rsetienne

@@
 % A short abstract of fewer than 200 words is acceptable.
-
-% @rsetienne:
-% At the moment, there are 87 parameter combinations for one site and one clock model
-% This allows for 40K / 87 = 460 replicates [*]
-% Using two site and two clock models would result in approx 100 replicates
-% Using four site and four clock models would result in approx 25 replicates
-%
-% I personally think the broadest data set will be most interesting,
-% because it gives a rough idea, but -more importantly- shows the scientific
-% community it can easily be done nowadays.
-%
-% [*] the number of replicates can only be determined by actually running the
-% experiment, as there are multiple constraints that I cannot test for
 
@@
 The Jukes-Cantor model is the correct one, as it is used for simulating that alignment,
-but the GTR model is the site model with the fewest assumptions and most flexibility.
-\textbf{[NOTE: Why then not use all four site models?]}
-For our clock model, we assume either a strict or relaxed log-normal \textbf{[NOTE: 
-Why that one? Why then not use all clock models?]} clock model. The strict clock
-model is the correct one, as it is used for simulating that alignment,
-but the relaxed log-normal clock model is the model with fewer assumptions and
-more flexibility.
+where the GTR model is the site model that is picked as a default by most users.
+For our clock model, we assume either a strict or relaxed log-normal 
+clock model. 
+Also here, the strict clock model is the correct one, as it is used for simulating that alignment,
+but the relaxed log-normal clock model is the the one most commonly used.
 We set the BD model as a tree prior, 

### 2018-06-05 article.tex
EOD
Updates from ShareLaTeX

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Methods (but we are not allowed to keep this header)}
+% Methods (but we are not allowed to use this header)
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 as this species tree, using the \verb;phangorn; package (\cite{phangorn}). 
-We set the nucleotides of the DNA alignment to follo a Jukes-Cantor (\cite{jc69})
+We set the nucleotides of the DNA alignment to follow a Jukes-Cantor (\cite{jc69})
 nucleotide substitution model, in which all nucleotide-to-nucleotide transitions
@@
 
-We produce two data sets as a comma-seperated file.
+We produce two data sets as a comma-separated file.
+Each parameter combination has ten replicates.
+
 We set the number of replicates for each parameter combination such,
@@
 statistics (see \ref{table:specs} for the exact data specification).
-The abovementioned memory constraints allows for $2 \cdot 10^3$ rows. 
+The above-mentioned memory constraints allows for $2 \cdot 10^3$ rows. 
 With 87 \textbf{[NOTE: volatile value, recalculate]} combinations of biological parameter, there will be 
@@
 for $\lambda_g = \lambda_i$ and $\mu_g = \mu_i$. The general parameter set plot
-does not make an additional seperation between site model and clock model, but these
-are shown in the appendix. The sampling parameter set plot does seperate 
+does not make an additional separation between site model and clock model, but these
+are shown in the appendix. The sampling parameter set plot does separate 
 on the sampling method used.
@@
   \caption{
-    Overview of the 12 simulation parameters. Above the horizontal line is 
-    the biological parameter set. The sampling methods are abbreviated as such: R denotes random
+    Overview of the simulation parameters. Above the horizontal line is 
+    the biological parameter set. 
+    The RNG seed $R_i$ is 1 for the first simulation, 2 for the next, etc.
+    The sampling methods are abbreviated as such: R denotes random
     sampling, 'S' is 'shortest' and 'L' is 'longest'. Sampling method $M_s$ is random for the general
-    data set. 
+    data set. For the data set exploring the effect of sampling, we use 'shortest'
+    for odd values of $R_i$, and 'longest' for even values of $R_i$.
     The clock models are abbreviated as 'S' is a strict and 'RLN' is a relaxed log-normal model.
-    The site models are abbreviated as 'JC69' for Jukes-Cantor and 'GTR' for the generalised 
+    The site models are abbreviated as 'JC69' for Jukes-Cantor and 'GTR' for the generalized 
     time-reversible model.
-    For the data set exploring the effect of sampling, we use 'shortest'
-    for odd values of $R_i$, and 'longest' for even values of $R_i$.
-    $R_i$ is 1 for the first simulation, 2 for the next, etcetera.
   }

### 2018-06-05 article.tex
Updates from ShareLaTeX
Improve wording

@@
 The simulation of these DNA aligments follows a strict clock model, 
-which we will specify as the known clock model prior in the Bayesian inference.
+which we will specify as one of the two clock models assumed in the Bayesian inference.
 
@@
 Each parameter combination has ten replicates.
-
 We set the number of replicates for each parameter combination such,

### 2018-06-06 article.tex
Add 'rampal' command
Reword

@@
 
-For both data sets, we plot the nLTT statistics distribution per
+For both data sets, we display the nLTT statistics distribution per
+biological parameter combination as a violin plot.
+We only show combinations for which $\lambda_g = \lambda_i$ and $\mu_g = \mu_i$,
+to simplify the interpretation of these results.
+
 parameter set using a violin plot, as such a plot 
-maintains information about the distribution. To simplify the interpretation
-of these plots, only nLTT statistics distribution are shown 
-for $\lambda_g = \lambda_i$ and $\mu_g = \mu_i$. The general parameter set plot
-does not make an additional separation between site model and clock model, but these
+maintains information about the distribution. 
+does not make an additional separation between site model and clock model. but these
 are shown in the appendix. The sampling parameter set plot does separate 

### 2018-06-06 article.tex
Improve text

@@
 % A short abstract of fewer than 200 words is acceptable.
+
+% Create the 'rampal' command
+% Usage:
+%
+% One plus one equals two\rampal{brilliant!}.
+%
+\newcommand*\rampal[1]{\textcolor{red}{\textbf{[RSE: #1]}}}
+
+% Create the 'richel' command
+\newcommand*\richel[1]{\textcolor{orange}{\textbf{[RJCB: #1]}}}
 
@@
   units (e.g. species) have become very advanced in the last three decades. 
+  \rampal{test comment, so Rampal can see this in action}
 
@@
 generatior seed, which makes all runs independent, resulting in a balanced 
-data set. \textbf{[NOTE: Rampal assumed runs with close seeds were related. 
+data set. \richel{Rampal assumed runs with close seeds were related. 
 I assume I have convinced him otherwise]}
@@
 Of the remaining
-1000 MCMC states, the effective sample size (ESS) of the posterior \textbf{[NOTE: I chose 
+1000 MCMC states, the effective sample size (ESS) of the posterior \richel{I chose 
 the ESS of the posterior over the ESS of the tree likelihood (and the 
-others displayed in table 4). For both something can be said. Agree on this choice?]} must at least be 200
+others displayed in table 4). For both something can be said. Agree on this choice?} must at least be 200
 for a strong enough inference (\cite{beastbook}). An ESS can be increased by increasing
@@
 We produce two data sets as a comma-separated file.
-Each parameter combination has ten replicates.
-We set the number of replicates for each parameter combination such,
-that this file and a possible copy can be handled in R's memory.
-Each row will then contain a parameter set and the generated nLTT 
-statistics (see \ref{table:specs} for the exact data specification).
-The above-mentioned memory constraints allows for $2 \cdot 10^3$ rows. 
-With 87 \textbf{[NOTE: volatile value, recalculate]} combinations of biological parameter, there will be 
-168 \textbf{[NOTE: volatile value, recalculate]} replicates per parameter set.
+The general data set has 348 different combinations
+of biological parameter combinations, site and clock models.
+The data set to investigate sampling has 496 different combinations
+of biological parameter combinations, site models, clock models 
+and sampling methods. The experiment is computationally intensive:
+pilot experiments show that the experiment takes roughly 100 days
+of CPU time and 20 days of wall clock time per replicate. 
+Due to this, we choose to perform ten replicates, so that the complete
+experiment will take an acceptable time of roughly seven months.
 
@@
 to simplify the interpretation of these results. 
-
-parameter set using a violin plot, as such a plot 
-maintains information about the distribution. 
-does not make an additional separation between site model and clock model. but these
-are shown in the appendix. The sampling parameter set plot does separate 
-on the sampling method used.
+Additionally, we only show the nLTT distributions
+that were generated under the (correct) assumptions of a Jukes-Cantor site model
+and a strict clock model. We display the nLTT statistic distributions for both
+site and clock models in the supplementary information. We show the effect of sampling,
+by seperating the nLTT statistics distribution per sampling method used.
 
@@
     nLTT statistic distribution per biological parameter set, using the
-    balanced data set, for all clock and site models.
+    balanced data set, 
+    for the subset of combinations in which $\lambda_g = \lambda_i$, $\mu_g = \mu_i$, 
+    under the (correct) assumptions of a strict clock and Jukes-Cantor site model.
   }
@@
     nLTT statistic distribution per biological parameter set per sampling
-    regime, using the data set conditioned on sampling regime having an effect.
+    regime, using the data set conditioned on sampling regime having an effect, 
+    for the subset of combinations in which $\lambda_g = \lambda_i$, $\mu_g = \mu_i$, 
+    under the (correct) assumptions of a strict clock and Jukes-Cantor site model.
   }
@@
 
-%\textbf{[NOTE: journal does not request for this. Suggest to remove, but how to acknowledge Peregrine otherwise?]}
+%\richel{journal does not request for this. Suggest to remove, but how to acknowledge Peregrine otherwise?}
 %We would like to thank the Center for Information Technology of the University of Groningen for their support
@@
 
-%\textbf{[NOTE: journal does not request for this]}
+%\richel{journal does not request for this}
 %RSE conceived the idea for this experiment. 
@@
     nLTT statistic distribution per biological parameter set per clock model,
-    using the balanced data set 
+    using the balanced data set, 
+    for the subset of combinations in which $\lambda_g = \lambda_i$, $\mu_g = \mu_i$, 
+    under the (correct) assumption of a Jukes-Cantor site model.
   }
@@
     nLTT statistic distribution per biological parameter set per site model,
-    using the balanced data set 
+    using the balanced data set, 
+    for the subset of combinations in which $\lambda_g = \lambda_i$, $\mu_g = \mu_i$, 
+    under the (correct) assumption of a strict clock model.
   }

### 2018-06-13 article.tex
Use 1K good species max
Feedback rsetienne

@@
 %
-\newcommand*\rampal[1]{\textcolor{red}{\textbf{[RSE: #1]}}}
+\newcommand*\rampal[1]{\textcolor{red}{\textbf{#1}}}
+\newcommand*\RAMPAL[1]{\textcolor{blue}{\textbf{[RSE: #1]}}}
 
@@
   units (e.g. species) have become very advanced in the last three decades. 
-  \rampal{test comment, so Rampal can see this in action}
-
     % Two to three sentences of
@@
   of the phylogenetic tree. 
-  A feature of these approaches is the possibility to incorporate 
-  known or hypothesized structure of the phylogenetic tree through the tree prior. 
+  A feature of these approaches is the possibility to incorporate known or hypothesized structure of the phylogenetic tree through the tree prior. 
   It has been shown that the effect of the prior on the posterior distribution 
@@
 
-{\bf Keywords:} computational biology, evolution, phylogenetics, prior choice
+{\bf Keywords:} computational biology, evolution, phylogenetics, Bayesian analysis, tree prior
 
@@
 Currently, the most popular Bayesian phylogenetics tools are \\ 
-BEAST (\cite{beast}) and its successor BEAST2 (\cite{beast2}), 
+BEAST (\cite{beast}) and its extension \RAMPAL{it is not really a successor, more a branch} BEAST2 (\cite{beast2}), 
 MrBayes (\cite{mrbayes}) and RevBayes (\cite{revbayes}). 
@@
 tree-to-be-inferred through model priors. 
-From these priors and an alignment of DNA, RNA or protein sequences, 
-they create a posterior distribution
-of parameter estimates (of the models used as a prior) and phylogenies, 
+With these priors and an alignment of DNA, RNA or protein sequences, 
+they create \rampal{a sample of} the posterior distribution
+of \rampal{phylogenies and parameter estimates} (of the models used as a prior), 
 in which more probable combinations are represented more often.
@@
 and branch termination (extinction).
-The choice of a wrong site model (\cite{posada_and_buckley_2004}), 
+The choice of \RAMPAL{delete: a wrong} site model (\cite{posada_and_buckley_2004}), 
 clock model (\cite{baele_et_al_2012}) 
@@
 It can be safely used as tree prior when only one individual per species is sampled, 
-but if one has multiple samples per species -which is currently often the case- the methods 
+but if one has multiple samples per species - which is currently often the case - the methods 
 to account for this such as the multi-species coalescent (\cite{heled_and_drummond_2009}) 
@@
 between the (BD) posterior phylogenies and the simulated (PBD) species tree.
-To gain more insight of incorrect prior choice, we also explore the effect of 
-two clock and site priors.
+\rampal{Furthermore, while we evidently know the clock and site models used in the simulation, using a different clock and/or site model prior in inference may compensate or increase this difference between inferred and simulated tree.
+To study this, we also explore the effect of
+a different clock and site model prior in inference.}
 
@@
 The PBD model has five biological parameters, depicted in table \ref{table:parameters}, 
-which we explore in a factorial fashion, excluding some combinations.
-We simulate a PBD process for those combinations in which the 
-95\% quantile of the expected number of good species is less than 1250.
-The quantile is
-calculated with a recently added function to the PBD package,
-based on equation 6 of \cite{etienne2014}. This calculation
+which we explore in a factorial fashion, excluding - for computational reasons - the combinations in which the 
+95\% quantile of the expected number of good species is more than 1250.
+This quantile is
+calculated with \rampal{the pbd\_numspec\_quantile function we added to the} \verb;PBD; package (\cite{pbd}).
+\RAMPAL{I think we should release the new version with this function with this manuscript.
+I deleted the reference to equation 6, because it is incorrect.
+Do you want to put the derivation here?}.
+This calculation
 assumes $b = b_g = b_i$, we used $b = \max(b_g, b_i)$.
-We use 1250 good species as a threshold, to prevent overly taxon-poor and taxon-rich phylogenies respectively.
+We use 1000 good species as a threshold, to prevent overly taxon-poor and taxon-rich phylogenies respectively.
 The parameter values chosen are based on the parameter sets used by 
@@
 sized phylogenies and using the same set allows us to compare results. 
-For the speciation initiation rates of good and incipient species, $b_g$ and $b_i$ 
-respectively, we use $0.3$ and $0.5$ speciation initiation events per good/incipient species per time unit.
-The speciation completion rates we use are $0.1$, $0.3$, $1.0$ and $10^9$ speciation completion
-events per (incipient species) species per time unit. We use $10^9 \approx \infty$
+For the speciation initiation rates of good and incipient species, we use $b_g = 0.35$ and $b_i = 0.55$, 
+respectively.\RAMPAL{Jouw units zijn niet helemaal correct, want het zijn probability rates.}
+The speciation completion rates we use are $\lambda = 0.1$, $0.3$, $1.0$ and $10^9$. We use $10^9 \approx \infty$
 to mimic the BD model, because the PBD model reduces to the BD model for $\lambda = \infty$.
@@
 because the generating model and the model used in inference are identical in this case.
-The extinction rates of good and incipient species, $\mu_g$ and $\mu_i$ respectively, 
-that we use are $0.0$, $0.1$ and $0.2$ 
-extinction events per good/incipient species per time unit.
+The extinction rates of good and incipient species used 
+are $\mu_g = \mu_i = 0.0$, $0.1$ and $0.2$.\RAMPAL{I made some changes here as well}
 
@@
 using the PBD package (\cite{pbd}) in the R programming language (\cite{r}), 
-all with a crown age of 15 million years.
+all with a crown age of 15 million years \rampal{as used in \cite{etienne2014}.}
 Each protracted birth-death tree uses a different random number
-generatior seed, which makes all runs independent, resulting in a balanced 
+\rampal{generator} seed, which makes all runs independent, resulting in a balanced 
 data set. \richel{Rampal assumed runs with close seeds were related. 
-I assume I have convinced him otherwise]}
+I assume I have convinced him otherwise]} \rampal{I am not saying that close seeds mean correlation; we simply don't know whether two lists of pseudo-random numbers generated from the two different seeds are not correlated, but perhaps the easiest way to resolve this is to check this for a fairly long list of numbers}
 
@@
 results in either the shortest or longest branch lengths. 
-See the supplementary information for a visualization of these sampling methods.
+See the supplementary information for a visualization of these sampling methods.\RAMPAL{Why not include that in the paper here; I think it is essential}
 
@@
 We predict with the more complex substitution model, 
-that there will be more noise and hence our inference error will increase.
+that there will be more noise and hence our inference error will increase.\RAMPAL{I am not sure: perhaps we compensate for the error made by adding more flexibility in another prior}
 We set the mutation rate in such a way to maximize the information contained in the alignment.
@@
 The simulation of these DNA aligments follows a strict clock model, 
-which we will specify as one of the two clock models assumed in the Bayesian inference.
+which we will specify as one of the two clock models assumed in the Bayesian inference \rampal{(see below)}.
 
@@
 clock model. 
-Also here, the strict clock model is the correct one, as it is used for simulating that alignment,
-but the relaxed log-normal clock model is the the one most commonly used.
+Also here, the strict clock model is the correct one, as it is used for simulating \rampal{the} alignment,
+but the relaxed log-normal clock model is the one most commonly used.
 We set the BD model as a tree prior, 
@@
 We compare each posterior phylogeny to the (sampled) species tree
-by the nLTT statistic (\cite{janzen2015}), using the nLTT package (\cite{nltt}). 
+\rampal{using} the nLTT statistic (\cite{janzen2015}), \rampal{from} the nLTT package (\cite{nltt}). 
 The nLTT statistic equals the area between the normalized
@@
 Due to this, we choose to perform ten replicates, so that the complete
-experiment will take an acceptable time of roughly seven months.
+experiment will take an acceptable time of roughly seven months. \RAMPAL{So this includes queue time? Perhaps mention this.}
 
@@
 biological parameter combination as a violin plot.
-We only show combinations for which $\lambda_g = \lambda_i$ and $\mu_g = \mu_i$,
-to simplify the interpretation of these results. 
+We only show combinations for which $\lambda_g = \lambda_i$ and $\mu_g = \mu_i$ \RAMPAL{So where are the other results? All results should at least be in the supplementary material},
+to simplify the interpretation of \rampal{the} results. 
 Additionally, we only show the nLTT distributions
 that were generated under the (correct) assumptions of a Jukes-Cantor site model
-and a strict clock model. We display the nLTT statistic distributions for both
-site and clock models in the supplementary information. We show the effect of sampling,
-by seperating the nLTT statistics distribution per sampling method used.
+and a strict clock model. We display the nLTT statistic distributions for \RAMPAL{replace 'both' by 'other'?}both
+site and clock models in the supplementary information. {Finally}, we show the effect of sampling,
+by \rampal{separating} the nLTT statistics distribution per sampling method used.
 
@@
     Posterior             & A collection of phylogenies and parameter estimates, in which more probable combinations (determined by the data and the model prior) are presented more frequently \\
-    Protracted speciation & The process in which speciation takes two events to complete: a speciation-initiation event and a speciation-completion event  \\
+    Protracted speciation & The process in which speciation takes two events \RAMPAL{delete: to complete}: a speciation-initiation event and a speciation-completion event  \\
     Speciation initiation & The start of a speciation event creating an incipient species \\
-    Speciation completion & The end of a speciation event, in which an incipient species is recognized as a good species \\
+    Speciation completion & The end of a speciation event, in which an incipient species \rampal{becomes or }is recognized as a good species \\
     \hline
@@
     nLTT statistic distribution per biological parameter set per clock model,
-    using the balanced data set, 
+    using the balanced data set\RAMPAL{what is a balanced data set? Did you define that anywhere?}, 
     for the subset of combinations in which $\lambda_g = \lambda_i$, $\mu_g = \mu_i$, 
@@
     The RNG seed $R_i$ is 1 for the first simulation, 2 for the next, etc.
-    The sampling methods are abbreviated as such: R denotes random
+    The sampling methods are abbreviated as such: \rampal{'R'} denotes random
     sampling, 'S' is 'shortest' and 'L' is 'longest'. Sampling method $M_s$ is random for the general
@@
     for odd values of $R_i$, and 'longest' for even values of $R_i$.
-    The clock models are abbreviated as 'S' is a strict and 'RLN' is a relaxed log-normal model.
-    The site models are abbreviated as 'JC69' for Jukes-Cantor and 'GTR' for the generalized 
+    The clock models are abbreviated as 'S' \rampal{for} a strict and 'RLN' \rampal{for} a relaxed log-normal model.
+    The site models are abbreviated as 'JC69' for Jukes-Cantor \RAMPAL{cite the JC 1969 paper?} and 'GTR' for the generalized 
     time-reversible model.
@@
   \caption{
-    Overview of the 11 BEAST2 estimated parameters
+    Overview of the 11 \rampal{parameters estimated by BEAST2}
   }

### 2018-06-13 article.tex
Implemented feedback, thanks @rsetienne

@@
 
-% Create the 'rampal' command
-% Usage:
-%
-% One plus one equals two\rampal{brilliant!}.
-%
+% Annotation/feedback commands
 \newcommand*\rampal[1]{\textcolor{red}{\textbf{#1}}}
 \newcommand*\RAMPAL[1]{\textcolor{blue}{\textbf{[RSE: #1]}}}
-
-% Create the 'richel' command
 \newcommand*\richel[1]{\textcolor{orange}{\textbf{[RJCB: #1]}}}
@@
   cre-ate
+  spe-ci-a-tion-com-ple-tion
+  pro-trac-ted
  }
@@
 Currently, the most popular Bayesian phylogenetics tools are \\ 
-BEAST (\cite{beast}) and its extension \RAMPAL{it is not really a successor, more a branch} BEAST2 (\cite{beast2}), 
+BEAST (\cite{beast}) and its offshoot BEAST2 (\cite{beast2}), 
 MrBayes (\cite{mrbayes}) and RevBayes (\cite{revbayes}). 
@@
 With these priors and an alignment of DNA, RNA or protein sequences, 
-they create \rampal{a sample of} the posterior distribution
-of \rampal{phylogenies and parameter estimates} (of the models used as a prior), 
+they create a sample of the posterior distribution
+of phylogenies and parameter estimates (of the models used as a prior), 
 in which more probable combinations are represented more often.
@@
 and branch termination (extinction).
-The choice of \RAMPAL{delete: a wrong} site model (\cite{posada_and_buckley_2004}), 
+The choice of site model (\cite{posada_and_buckley_2004}), 
 clock model (\cite{baele_et_al_2012}) 
@@
 between the (BD) posterior phylogenies and the simulated (PBD) species tree.
-\rampal{Furthermore, while we evidently know the clock and site models used in the simulation, using a different clock and/or site model prior in inference may compensate or increase this difference between inferred and simulated tree.
+Furthermore, while we evidently know the clock and site models used in the simulation, 
+using a different clock and/or site model prior in inference 
+may compensate or increase this difference between inferred and simulated tree. 
 To study this, we also explore the effect of 
-a different clock and site model prior in inference.}
+a different clock and site model prior in inference.
 
@@
 This quantile is
-calculated with \rampal{the pbd\_numspec\_quantile function we added to the} \verb;PBD; package (\cite{pbd}).
-\RAMPAL{I think we should release the new version with this function with this manuscript.
-I deleted the reference to equation 6, because it is incorrect.
-Do you want to put the derivation here?}.
-This calculation
+calculated with the \verb;pbd_numspec_quantile; function we added to the \verb;PBD; package (\cite{pbd}).
+\RAMPAL{I think we should release the new version with this function with this manuscript}\richel{I would personally prefer 'release early, release often', but as lead maintainer you get to decide}\RAMPAL{Do you want to put the derivation here?}\richel{Yes please, either here or at the PBD package (the functions pbd\_geom and
+pbd\_numspec\_quantile) documentation. Wherever you put it, I will transfer it to the other place}. This calculation
 assumes $b = b_g = b_i$, we used $b = \max(b_g, b_i)$.
@@
 sized phylogenies and using the same set allows us to compare results. 
-For the speciation initiation rates of good and incipient species, we use $b_g = 0.35$ and $b_i = 0.55$, 
-respectively.\RAMPAL{Jouw units zijn niet helemaal correct, want het zijn probability rates.}
+We use a set of speciation initiation rates, $B = \{0.3, 0.5\}$, of which the speciation initiation
+rate of good species $b_g \in B$ and incipient species $b_i \in B$.
+\richel{these rates can differ, although we do not plot the combinations in which they differ}.\RAMPAL{your units are not entirely correct, as these are probability rates.}\richel{I though I was correct in my units.
+If I am wrong, what are the correct units then?
+In Etienne and Rosindell 2012, there is never 'probability rate' written.
+To be explicit, I think these sentences are all valid: (1) the extinction rate of good species, mu\_g is 1.2 extinction events per time unit (2) the speciation initiation rate of good species, b\_g is 2.3 speciation events per time unit, (3) the speciation completion rate, lambda is 3.4 speciation completion events per time unit.
+What would be the better unit?}
 The speciation completion rates we use are $\lambda = 0.1$, $0.3$, $1.0$ and $10^9$. We use $10^9 \approx \infty$
@@
 because the generating model and the model used in inference are identical in this case.
-The extinction rates of good and incipient species used 
-are $\mu_g = \mu_i = 0.0$, $0.1$ and $0.2$.\RAMPAL{I made some changes here as well}
+We use a set of extinction rates, $M = \{0.0, 0.1, 0.2, \infty\}$, of which the extinction
+rate of good species $\mu_g \in M$ and incipient species $\mu_i \in M$. 
+\richel{Also here: these rates can differ}
 
@@
 using the PBD package (\cite{pbd}) in the R programming language (\cite{r}), 
-all with a crown age of 15 million years \rampal{as used in \cite{etienne2014}.}
+all with a crown age of 15 million years as used in \cite{etienne2014}.
 Each protracted birth-death tree uses a different random number
-\rampal{generator} seed, which makes all runs independent, resulting in a balanced 
+generator seed, which makes all runs independent, resulting in a balanced 
 data set. \richel{Rampal assumed runs with close seeds were related. 
-I assume I have convinced him otherwise]} \rampal{I am not saying that close seeds mean correlation; we simply don't know whether two lists of pseudo-random numbers generated from the two different seeds are not correlated, but perhaps the easiest way to resolve this is to check this for a fairly long list of numbers}
+I assume I have convinced him otherwise]} \rampal{I am not saying that close seeds mean correlation; we simply don't know whether two lists of pseudo-random numbers generated from the two different seeds are not correlated, but perhaps the easiest way to resolve this is to check this for a fairly long list of numbers}\richel{(1) I checked reproducing one random number from two different (adjacent) seeds, see/run code below in comments.
+(2) I quote from https://stat.ethz.ch/R-manual/R-patched/library/base/html/Random.html: 'There is no guarantee that different values of seed will seed the RNG differently, although any exceptions would be extremely rare'.
+The full quote: 'set.seed uses a single integer argument to set as many seeds as are required.
+It is intended as a simple way to get quite different seeds by specifying small integer arguments, and also as a way to get valid seed sets for the more complicated methods (especially "Mersenne-Twister" [see below, ...]).
+There is no guarantee that different values of seed will seed the RNG differently, although any exceptions would be extremely rare.
+"Mersenne-Twister": From Matsumoto and Nishimura (1998).
+A twisted GFSR with period 2 to the power of 19937 - 1 and equidistribution in 623 consecutive dimensions (over the whole period).
+The 'seed' is a 624-dimensional set of 32-bit integers plus a current position in that set.'
+(3) I've added a more elaborate example below that shows there is no overlapping sequence  
+}
+
+% From Progress Report of week 21 2018, May 21 - May 25 2018:
+% # If a random number if drawn with one seed,
+% # how long until it is drawn by differently seeded
+% # RNG?
+%
+% Pick any starting seed
+% set.seed(2)
+% value <- runif(n = 1)
+%
+% Pick any other seed
+% set.seed(1)
+% i <- 0
+% while (value != runif(n = 1)) {
+%   i <- i + 1
+% }  
+% print(i)
+
+% More elaborate example:
+% 
+% # Goal:
+% # Proof that runs with different RNG seeds result in completely different sequences
+% 
+% # SETUP
+% n_seeds <- 10000
+% sequence_length <- 10000
+% 
+% # START
+% runs <- list()
+% for (seed in seq(1, n_seeds)) {
+%   set.seed(seed)
+%   runs[[seed]] <- runif(n = sequence_length)
+% }
+% 
+% numbers <- unlist(runs)
+% duplicate_numbers <- numbers[duplicated(numbers)]
+% 
+% if (length(duplicate_numbers) == 0) {
+%   print("all numbers are unique")
+% } else {
+%   percentage <- 100 * length(duplicate_numbers) / length(numbers)
+%   print(paste(length(duplicate_numbers), "of", length(numbers), "numbers (", percentage ,"% ) occur multiple times"))
+%   run_length_encoding <- rle(duplicated(numbers))
+%   repeated_sequence_lengths <- run_length_encoding$lengths[ run_length_encoding$values == TRUE ]
+%   longest_repeated_sequence <- max(repeated_sequence_lengths)
+%   print(paste("Longest sequence of numbers that occurred multiple times (not necessarily in that same order) is", longest_repeated_sequence))
+%   ggplot2::ggplot(
+%     data.frame(length = repeated_sequence_lengths),
+%     ggplot2::aes(x = length)
+%   ) + ggplot2::geom_histogram(binwidth = 1) + ggplot2::coord_cartesian(ylim=c(0, 20)) +
+%     ggplot2::ggtitle(
+%       paste0(
+%         "Lengths of sequence of numbers that occurred multiple times (not necessarily in that same order)\n",
+%         "sequence length: ", sequence_length, ", n_seeds: ", n_seeds
+%       )
+%     )
+%  }
 
@@
 results in either the shortest or longest branch lengths. 
-See the supplementary information for a visualization of these sampling methods.\RAMPAL{Why not include that in the paper here; I think it is essential}
+See the supplementary information for a visualization of these sampling methods.\RAMPAL{Why not include that in the paper here; I think it is essential}\richel{Because we are limited to two figures}
 
@@
 We predict with the more complex substitution model, 
-that there will be more noise and hence our inference error will increase.\RAMPAL{I am not sure: perhaps we compensate for the error made by adding more flexibility in another prior}
+that there will be more noise and hence our inference error will increase.
+On the other hand, we dare not rule out that the inference error will decrease,
+due to more flexibility in the more complex prior.
 We set the mutation rate in such a way to maximize the information contained in the alignment.
@@
 The simulation of these DNA aligments follows a strict clock model, 
-which we will specify as one of the two clock models assumed in the Bayesian inference \rampal{(see below)}.
+which we will specify as one of the two clock models assumed in the Bayesian inference (see below).
 
@@
 clock model. 
-Also here, the strict clock model is the correct one, as it is used for simulating \rampal{the} alignment,
+Also here, the strict clock model is the correct one, as it is used for simulating the alignment,
 but the relaxed log-normal clock model is the one most commonly used.
@@
 Of the remaining
-1000 MCMC states, the effective sample size (ESS) of the posterior \richel{I chose 
-the ESS of the posterior over the ESS of the tree likelihood (and the 
-others displayed in table 4). For both something can be said. Agree on this choice?} must at least be 200
+1000 MCMC states, the effective sample size (ESS) of the posterior 
+must at least be 200
 for a strong enough inference (\cite{beastbook}). An ESS can be increased by increasing
@@
 We compare each posterior phylogeny to the (sampled) species tree
-\rampal{using} the nLTT statistic (\cite{janzen2015}), \rampal{from} the nLTT package (\cite{nltt}). 
+using the nLTT statistic (\cite{janzen2015}), from the \verb;nLTT; package (\cite{nltt}). 
 The nLTT statistic equals the area between the normalized
@@
 pilot experiments show that the experiment takes roughly 100 days
-of CPU time and 20 days of wall clock time per replicate. 
+of CPU time and 20 days of wall clock time (which includes the queued 
+waiting for computational resources) per replicate. 
 Due to this, we choose to perform ten replicates, so that the complete
-experiment will take an acceptable time of roughly seven months. \RAMPAL{So this includes queue time? Perhaps mention this.}
+experiment will take an acceptable time of roughly seven months.
 
@@
 biological parameter combination as a violin plot.
-We only show combinations for which $\lambda_g = \lambda_i$ and $\mu_g = \mu_i$ \RAMPAL{So where are the other results? All results should at least be in the supplementary material},
-to simplify the interpretation of \rampal{the} results. 
+We show combinations for which $b_g = b_i$ and $\mu_g = \mu_i$,
+to simplify the interpretation of the results, where the other combinations
+are shown in the supplementary material.
 Additionally, we only show the nLTT distributions
 that were generated under the (correct) assumptions of a Jukes-Cantor site model
-and a strict clock model. We display the nLTT statistic distributions for \RAMPAL{replace 'both' by 'other'?}both
-site and clock models in the supplementary information. {Finally}, we show the effect of sampling,
-by \rampal{separating} the nLTT statistics distribution per sampling method used.
+and a strict clock model, separated per sampling method used. 
+We display the nLTT statistic distributions separated per site or clock model 
+in the supplementary information.
 
@@
     nLTT statistic distribution per biological parameter set, using the
-    balanced data set, 
-    for the subset of combinations in which $\lambda_g = \lambda_i$, $\mu_g = \mu_i$, 
+    general data set, 
+    for the subset of combinations in which $b_g = b_i$, $\mu_g = \mu_i$, 
     under the (correct) assumptions of a strict clock and Jukes-Cantor site model.
@@
     regime, using the data set conditioned on sampling regime having an effect, 
-    for the subset of combinations in which $\lambda_g = \lambda_i$, $\mu_g = \mu_i$, 
+    for the subset of combinations in which $b_g = b_i$, $\mu_g = \mu_i$, 
     under the (correct) assumptions of a strict clock and Jukes-Cantor site model.
@@
     Posterior             & A collection of phylogenies and parameter estimates, in which more probable combinations (determined by the data and the model prior) are presented more frequently \\
-    Protracted speciation & The process in which speciation takes two events \RAMPAL{delete: to complete}: a speciation-initiation event and a speciation-completion event  \\
+    Protracted speciation & The process in which speciation takes two events: a speciation-initiation event and a speciation-completion event  \\
     Speciation initiation & The start of a speciation event creating an incipient species \\
-    Speciation completion & The end of a speciation event, in which an incipient species \rampal{becomes or }is recognized as a good species \\
+    Speciation completion & The end of a speciation event, in which an incipient species becomes or is recognized as a good species \\
     \hline
@@
 \begin{figure}[!htbp]
+  \includegraphics[width=0.8\textwidth]{fig_sampling_methods.png}
+  \caption{
+    Sampling a species tree from an incipient species tree. At the top left,
+    an incipient species tree is shown, of three different good species (the
+    first and second number in the taxon label) and four different subspecies (the
+    third number in the taxon tabel). The other four trees are species trees,
+    that use a different sampling method to determine which sub-species is picked to
+    represent a good species. These are: 'Youngest', 'Oldest', 'Shortest' and
+    'Longest'. With 'Youngest' the youngest sub-species is picked to represent the
+    good species. With 'Oldest' the oldest sub-species is picked to represent the
+    good species. 'Shortest' is the sampling method in which the sub-species are
+    picked to assure the shortest branch lengths. 'Longest' is the sampling method 
+    in which the sub-species are picked to assure the longest branch lengths.
+  }
+\end{figure}
+
+\begin{figure}[!htbp]
   \includegraphics[width=0.8\textwidth]{fig_clock_model.png}
@@
     nLTT statistic distribution per biological parameter set per clock model,
-    using the balanced data set\RAMPAL{what is a balanced data set? Did you define that anywhere?}, 
-    for the subset of combinations in which $\lambda_g = \lambda_i$, $\mu_g = \mu_i$, 
+    using the general data set, 
+    for the subset of combinations in which $b_g = b_i$, $\mu_g = \mu_i$, 
     under the (correct) assumption of a Jukes-Cantor site model.
@@
     nLTT statistic distribution per biological parameter set per site model,
-    using the balanced data set, 
-    for the subset of combinations in which $\lambda_g = \lambda_i$, $\mu_g = \mu_i$, 
+    using the general data set, 
+    for the subset of combinations in which $b_g = b_i$, $\mu_g = \mu_i$, 
     under the (correct) assumption of a strict clock model.
+  }
+\end{figure}
+
+\begin{figure}[!htbp]
+  \includegraphics[width=0.8\textwidth]{fig_general_1.png}
+  \caption{
+    nLTT statistic distribution per biological parameter set, using the
+    general data set, 
+    for $b_i = 0.3$ and $\mu_i = 0.0$, 
+    under the (correct) assumptions of a strict clock and Jukes-Cantor site model.
+  }
+\end{figure}
+
+\begin{figure}[!htbp]
+  \includegraphics[width=0.8\textwidth]{fig_general_2.png}
+  \caption{
+    nLTT statistic distribution per biological parameter set, using the
+    general data set, 
+    for $b_i = 0.3$ and $\mu_i = 0.1$, 
+    under the (correct) assumptions of a strict clock and Jukes-Cantor site model.
+  }
+\end{figure}
+
+\begin{figure}[!htbp]
+  \includegraphics[width=0.8\textwidth]{fig_general_3.png}
+  \caption{
+    nLTT statistic distribution per biological parameter set, using the
+    general data set, 
+    for $b_i = 0.3$ and $\mu_i = 0.2$, 
+    under the (correct) assumptions of a strict clock and Jukes-Cantor site model.
+  }
+\end{figure}
+
+\begin{figure}[!htbp]
+  \includegraphics[width=0.8\textwidth]{fig_general_4.png}
+  \caption{
+    nLTT statistic distribution per biological parameter set, using the
+    general data set, 
+    for $b_i = 0.5$ and $\mu_i = 0.0$, 
+    under the (correct) assumptions of a strict clock and Jukes-Cantor site model.
+  }
+\end{figure}
+
+\begin{figure}[!htbp]
+  \includegraphics[width=0.8\textwidth]{fig_general_5.png}
+  \caption{
+    nLTT statistic distribution per biological parameter set, using the
+    general data set, 
+    for $b_i = 0.5$ and $\mu_i = 0.1$, 
+    under the (correct) assumptions of a strict clock and Jukes-Cantor site model.
+  }
+\end{figure}
+
+\begin{figure}[!htbp]
+  \includegraphics[width=0.8\textwidth]{fig_general_6.png}
+  \caption{
+    nLTT statistic distribution per biological parameter set, using the
+    general data set, 
+    for $b_i = 0.5$ and $\mu_i = 0.2$, 
+    under the (correct) assumptions of a strict clock and Jukes-Cantor site model.
   }
@@
     $\mu_g$               & Extinction rate of a good species & 0.0, 0.1, 0.2 \\
-    $\mu_i$               & Extinction rate of an incipient species & 0.0, 0.1, 0.2, 0.4 \\
+    $\mu_i$               & Extinction rate of an incipient species & 0.0, 0.1, 0.2 \\
     \hline
@@
     $f_i$                 & MCMC sampling interval & 1K or more \\
-    $R_i$                 & RNG seed incipient tree and randomly sampled species tree & 1 to 20K \\
+    $R_i$                 & RNG seed incipient tree and randomly sampled species tree & 1, 2, ... \\
     $R_a$                 & RNG seed alignment simulation & $R_i$ \\
@@
     the biological parameter set. 
-    The RNG seed $R_i$ is 1 for the first simulation, 2 for the next, etc.
-    The sampling methods are abbreviated as such: \rampal{'R'} denotes random
+    The RNG seed $R_i$ is 1 for the first simulation of the general data set, 2 for the next,
+    and so on, up to and including 3480. The RNG seeds for the data set investigating the effect 
+    of sampling continues from there.
+    The sampling methods are abbreviated as such: 'R' denotes random
     sampling, 'S' is 'shortest' and 'L' is 'longest'. Sampling method $M_s$ is random for the general
@@
     for odd values of $R_i$, and 'longest' for even values of $R_i$.
-    The clock models are abbreviated as 'S' \rampal{for} a strict and 'RLN' \rampal{for} a relaxed log-normal model.
-    The site models are abbreviated as 'JC69' for Jukes-Cantor \RAMPAL{cite the JC 1969 paper?} and 'GTR' for the generalized 
-    time-reversible model.
+    The clock models are abbreviated as 'S' for a strict and 'RLN' for a relaxed log-normal model.
+    The site models are abbreviated as 'JC69' for Jukes-Cantor (\cite{jc69}) and 'GTR' for the generalized 
+    time-reversible model (\cite{gtr}).
   }
@@
   \caption{
-    Overview of the 11 \rampal{parameters estimated by BEAST2}
+    Overview of the 11 parameters estimated by BEAST2
   }

### 2018-06-14 article.tex
Add feedback from rsetienne

@@
 rate of good species $b_g \in B$ and incipient species $b_i \in B$.
-\richel{these rates can differ, although we do not plot the combinations in which they differ}.\RAMPAL{your units are not entirely correct, as these are probability rates.}\richel{I though I was correct in my units.
+\richel{these rates can differ, although we do not plot the combinations in which they differ}\RAMPAL{Prima om die verschillend te nemen.
+Het ms was hier idd niet duidelijk over.}.\RAMPAL{your units are not entirely correct, as these are probability rates.}\richel{I though I was correct in my units.
 If I am wrong, what are the correct units then?
@@
 To be explicit, I think these sentences are all valid: (1) the extinction rate of good species, mu\_g is 1.2 extinction events per time unit (2) the speciation initiation rate of good species, b\_g is 2.3 speciation events per time unit, (3) the speciation completion rate, lambda is 3.4 speciation completion events per time unit.
-What would be the better unit?}
+What would be the better unit?}\RAMPAL{
+Je hebt het over de eenheden van de rates.
+Dat is subtiel omdat het om een stochastisch model gaat en dus om kansen.
+Voorbeeld: mu * dt is de kans op een extinction event per species in een tijdsinterval dt.
+1/mu geeft de verwachte tijd tot een extinctie event van een soort.
+mu is dus een probability rate.
+}
 The speciation completion rates we use are $\lambda = 0.1$, $0.3$, $1.0$ and $10^9$. We use $10^9 \approx \infty$
@@
 }
+\RAMPAL{M.i. is het theoretisch nog niet evident, maar empirisch in orde, dus prima zo.}
 

### 2018-06-14 article.tex
Rates can differ

@@
 We use a set of speciation initiation rates, $B = \{0.3, 0.5\}$, of which the speciation initiation
-rate of good species $b_g \in B$ and incipient species $b_i \in B$.
-\richel{these rates can differ, although we do not plot the combinations in which they differ}\RAMPAL{Prima om die verschillend te nemen.
-Het ms was hier idd niet duidelijk over.}.\RAMPAL{your units are not entirely correct, as these are probability rates.}\richel{I though I was correct in my units.
+rate of good species $b_g \in B$ and incipient species $b_i \in B$.\RAMPAL{your units are not entirely correct, as these are probability rates.}\richel{I though I was correct in my units.
 If I am wrong, what are the correct units then?
@@
 rate of good species $\mu_g \in M$ and incipient species $\mu_i \in M$.
-\richel{Also here: these rates can differ}
 

### 2018-06-14 article.tex
Starting with different RNG seed is good enough

@@
 generator seed, which makes all runs independent, resulting in a balanced 
-data set. \richel{Rampal assumed runs with close seeds were related. 
-I assume I have convinced him otherwise]} \rampal{I am not saying that close seeds mean correlation; we simply don't know whether two lists of pseudo-random numbers generated from the two different seeds are not correlated, but perhaps the easiest way to resolve this is to check this for a fairly long list of numbers}\richel{(1) I checked reproducing one random number from two different (adjacent) seeds, see/run code below in comments.
-(2) I quote from https://stat.ethz.ch/R-manual/R-patched/library/base/html/Random.html: 'There is no guarantee that different values of seed will seed the RNG differently, although any exceptions would be extremely rare'.
-The full quote: 'set.seed uses a single integer argument to set as many seeds as are required.
-It is intended as a simple way to get quite different seeds by specifying small integer arguments, and also as a way to get valid seed sets for the more complicated methods (especially "Mersenne-Twister" [see below, ...]).
-There is no guarantee that different values of seed will seed the RNG differently, although any exceptions would be extremely rare.
-"Mersenne-Twister": From Matsumoto and Nishimura (1998).
-A twisted GFSR with period 2 to the power of 19937 - 1 and equidistribution in 623 consecutive dimensions (over the whole period).
-The 'seed' is a 624-dimensional set of 32-bit integers plus a current position in that set.'
-(3) I've added a more elaborate example below that shows there is no overlapping sequence  
-}
-\RAMPAL{M.i. is het theoretisch nog niet evident, maar empirisch in orde, dus prima zo.}
-
-% From Progress Report of week 21 2018, May 21 - May 25 2018:
-% # If a random number if drawn with one seed,
-% # how long until it is drawn by differently seeded
-% # RNG?
-%
-% Pick any starting seed
-% set.seed(2)
-% value <- runif(n = 1)
-%
-% Pick any other seed
-% set.seed(1)
-% i <- 0
-% while (value != runif(n = 1)) {
-%   i <- i + 1
-% }  
-% print(i)
-
-% More elaborate example:
-% 
-% # Goal:
-% # Proof that runs with different RNG seeds result in completely different sequences
-% 
-% # SETUP
-% n_seeds <- 10000
-% sequence_length <- 10000
-% 
-% # START
-% runs <- list()
-% for (seed in seq(1, n_seeds)) {
-%   set.seed(seed)
-%   runs[[seed]] <- runif(n = sequence_length)
-% }
-% 
-% numbers <- unlist(runs)
-% duplicate_numbers <- numbers[duplicated(numbers)]
-% 
-% if (length(duplicate_numbers) == 0) {
-%   print("all numbers are unique")
-% } else {
-%   percentage <- 100 * length(duplicate_numbers) / length(numbers)
-%   print(paste(length(duplicate_numbers), "of", length(numbers), "numbers (", percentage ,"% ) occur multiple times"))
-%   run_length_encoding <- rle(duplicated(numbers))
-%   repeated_sequence_lengths <- run_length_encoding$lengths[ run_length_encoding$values == TRUE ]
-%   longest_repeated_sequence <- max(repeated_sequence_lengths)
-%   print(paste("Longest sequence of numbers that occurred multiple times (not necessarily in that same order) is", longest_repeated_sequence))
-%   ggplot2::ggplot(
-%     data.frame(length = repeated_sequence_lengths),
-%     ggplot2::aes(x = length)
-%   ) + ggplot2::geom_histogram(binwidth = 1) + ggplot2::coord_cartesian(ylim=c(0, 20)) +
-%     ggplot2::ggtitle(
-%       paste0(
-%         "Lengths of sequence of numbers that occurred multiple times (not necessarily in that same order)\n",
-%         "sequence length: ", sequence_length, ", n_seeds: ", n_seeds
-%       )
-%     )
-%  }
+data set.
 

### 2018-06-14 article.tex
Describe RNG seed selectiion better

@@
     and so on, up to and including 3480. The RNG seeds for the data set investigating the effect 
-    of sampling continues from there.
+    of sampling continue from there, but only those RNG seeds are used in which sampling has an effect.
     The sampling methods are abbreviated as such: 'R' denotes random
@@
     data set. For the data set exploring the effect of sampling, we use 'shortest'
-    for odd values of $R_i$, and 'longest' for even values of $R_i$.
+    and 'longest' for each value of $R_i$ (which are random seeds in which sampling has an effect).
     The clock models are abbreviated as 'S' for a strict and 'RLN' for a relaxed log-normal model.

### 2018-06-14 article.tex
Sampling experiment has 552 instead of 496 combinations

@@
 of biological parameter combinations, site and clock models.
-The data set to investigate sampling has 496 different combinations
+The data set to investigate sampling has 552 different combinations
 of biological parameter combinations, site models, clock models 

### 2018-09-03 article.tex
Explicitly state that root sequence has 25 percent of each nucleotide

@@
 calculated with the \verb;pbd_numspec_quantile; function we added to the \verb;PBD; package (\cite{pbd}).
-\RAMPAL{I think we should release the new version with this function with this manuscript}\richel{I would personally prefer 'release early, release often', but as lead maintainer you get to decide}\RAMPAL{Do you want to put the derivation here?}\richel{Yes please, either here or at the PBD package (the functions pbd\_geom and
-pbd\_numspec\_quantile) documentation. Wherever you put it, I will transfer it to the other place}. This calculation
-assumes $b = b_g = b_i$, we used $b = \max(b_g, b_i)$.
+\RAMPAL{I think we should release the new version with this function with this manuscript}
+\richel{I would personally prefer 'release early, release often', but as lead maintainer you get to decide}
+\RAMPAL{Do you want to put the derivation here?}
+\richel{No, as you've added it to the PBD::pbd\_geom and PBD::pbd\_numspec\_quantile) documentation, so I will transfer it. Thanks!}
+This calculation assumes $b = b_g = b_i$, we used $b = \max(b_g, b_i)$.
 We use 1000 good species as a threshold, to prevent overly taxon-poor and taxon-rich phylogenies respectively.
@@
 We use a set of speciation initiation rates, $B = \{0.3, 0.5\}$, of which the speciation initiation
-rate of good species $b_g \in B$ and incipient species $b_i \in B$.\RAMPAL{your units are not entirely correct, as these are probability rates.}\richel{I though I was correct in my units.
+rate of good species $b_g \in B$ and incipient species $b_i \in B$.
+\RAMPAL{your units are not entirely correct, as these are probability rates.}
+\richel{I (perhaps naively) think I am correct in my units.
 If I am wrong, what are the correct units then?
 In Etienne and Rosindell 2012, there is never 'probability rate' written.
-To be explicit, I think these sentences are all valid: (1) the extinction rate of good species, mu\_g is 1.2 extinction events per time unit (2) the speciation initiation rate of good species, b\_g is 2.3 speciation events per time unit, (3) the speciation completion rate, lambda is 3.4 speciation completion events per time unit.
-What would be the better unit?}\RAMPAL{
-Je hebt het over de eenheden van de rates.
+To be explicit, I think theis sentence is valid: the extinction rate of good species, mu\_g has an expected 1.2 extinction events per time unit.
+What would be the better unit?}
+\RAMPAL{Je hebt het over de eenheden van de rates.
 Dat is subtiel omdat het om een stochastisch model gaat en dus om kansen.
@@
 1/mu geeft de verwachte tijd tot een extinctie event van een soort.
-mu is dus een probability rate.
-}
+mu is dus een probability rate.}
+\richel{'mu is dus een probability rate' with what unit? You can redirect me to an article with the proper wording, or finish this sentence: 'the extinction rate of good species, mu\_g ...'. Thanks!}
 The speciation completion rates we use are $\lambda = 0.1$, $0.3$, $1.0$ and $10^9$. We use $10^9 \approx \infty$
@@
 results in either the shortest or longest branch lengths. 
-See the supplementary information for a visualization of these sampling methods.\RAMPAL{Why not include that in the paper here; I think it is essential}\richel{Because we are limited to two figures}
-
+See the supplementary information for a visualization of these sampling methods.
+% \RAMPAL{Why not include that in the paper here; I think it is essential}
+% \richel{Because we are limited to two figures}
 Based on the sampled species tree, we simulate a DNA alignment that has the same history
@@
 are equally likely. 
+\richel{New:}
+The DNA sequence used for the root ancestor has an equal amount of each nucleotide.
+\richel{End of new}
 In our Bayesian inference (see below) we use the same site model as the (obviously correct) site model prior,

### 2018-09-03 article.tex
Implemented feedback, thanks @rsetienne

@@
 are equally likely. 
-\richel{New:}
-The DNA sequence used for the root ancestor has an equal amount of each nucleotide.
-\richel{End of new}
+The DNA sequence of the root ancestor consists of four equally sized single-nucleotide blocks of adenine, cytosine, guanine and thymine respectively. 
+For example, for a DNA sequence length of 12, this would be AAACCCGGGTTT. 
+The order of nucletides does not matter in this study, 
+because we do not consider several partitions of the sequence with their own parameters. 
+Only the frequency of occurrence matters.
 In our Bayesian inference (see below) we use the same site model as the (obviously correct) site model prior,

### 2018-09-05 article.tex
Use number of species as rows in figures

@@
 
-\begin{figure}[!htbp]
-  \includegraphics[width=0.8\textwidth]{fig_general_1.png}
-  \caption{
-    nLTT statistic distribution per biological parameter set, using the
-    general data set, 
-    for $b_i = 0.3$ and $\mu_i = 0.0$, 
-    under the (correct) assumptions of a strict clock and Jukes-Cantor site model.
-  }
-\end{figure}
-
-\begin{figure}[!htbp]
-  \includegraphics[width=0.8\textwidth]{fig_general_2.png}
-  \caption{
-    nLTT statistic distribution per biological parameter set, using the
-    general data set, 
-    for $b_i = 0.3$ and $\mu_i = 0.1$, 
-    under the (correct) assumptions of a strict clock and Jukes-Cantor site model.
-  }
-\end{figure}
-
-\begin{figure}[!htbp]
-  \includegraphics[width=0.8\textwidth]{fig_general_3.png}
-  \caption{
-    nLTT statistic distribution per biological parameter set, using the
-    general data set, 
-    for $b_i = 0.3$ and $\mu_i = 0.2$, 
-    under the (correct) assumptions of a strict clock and Jukes-Cantor site model.
-  }
-\end{figure}
-
-\begin{figure}[!htbp]
-  \includegraphics[width=0.8\textwidth]{fig_general_4.png}
-  \caption{
-    nLTT statistic distribution per biological parameter set, using the
-    general data set, 
-    for $b_i = 0.5$ and $\mu_i = 0.0$, 
-    under the (correct) assumptions of a strict clock and Jukes-Cantor site model.
-  }
-\end{figure}
-
-\begin{figure}[!htbp]
-  \includegraphics[width=0.8\textwidth]{fig_general_5.png}
-  \caption{
-    nLTT statistic distribution per biological parameter set, using the
-    general data set, 
-    for $b_i = 0.5$ and $\mu_i = 0.1$, 
-    under the (correct) assumptions of a strict clock and Jukes-Cantor site model.
-  }
-\end{figure}
-
-\begin{figure}[!htbp]
-  \includegraphics[width=0.8\textwidth]{fig_general_6.png}
-  \caption{
-    nLTT statistic distribution per biological parameter set, using the
-    general data set, 
-    for $b_i = 0.5$ and $\mu_i = 0.2$, 
-    under the (correct) assumptions of a strict clock and Jukes-Cantor site model.
-  }
-\end{figure}
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2018-09-05 article.tex
Rewrote picking trees with desired number of good species

@@
 
-The PBD model has five biological parameters, depicted in table \ref{table:parameters}, 
-which we explore in a factorial fashion, excluding - for computational reasons - the combinations in which the 
-95\% quantile of the expected number of good species is more than 1250.
-This quantile is
-calculated with the \verb;pbd_numspec_quantile; function we added to the \verb;PBD; package (\cite{pbd}).
-\RAMPAL{I think we should release the new version with this function with this manuscript}
-\richel{I would personally prefer 'release early, release often', but as lead maintainer you get to decide}
-\RAMPAL{Do you want to put the derivation here?}
-\richel{No, as you've added it to the PBD::pbd\_geom and PBD::pbd\_numspec\_quantile) documentation, so I will transfer it. Thanks!}
-This calculation assumes $b = b_g = b_i$, we used $b = \max(b_g, b_i)$.
-We use 1000 good species as a threshold, to prevent overly taxon-poor and taxon-rich phylogenies respectively.
-The parameter values chosen are based on the parameter sets used by 
-\cite{etienne2014}, as these parameters were shown to result in reasonably
-sized phylogenies and using the same set allows us to compare results. 
-We use a set of speciation initiation rates, $B = \{0.3, 0.5\}$, of which the speciation initiation
-rate of good species $b_g \in B$ and incipient species $b_i \in B$.
-\RAMPAL{your units are not entirely correct, as these are probability rates.}
-\richel{I (perhaps naively) think I am correct in my units.
-If I am wrong, what are the correct units then?
-In Etienne and Rosindell 2012, there is never 'probability rate' written.
-To be explicit, I think theis sentence is valid: the extinction rate of good species, mu\_g has an expected 1.2 extinction events per time unit.
-What would be the better unit?}
-\RAMPAL{Je hebt het over de eenheden van de rates.
-Dat is subtiel omdat het om een stochastisch model gaat en dus om kansen.
-Voorbeeld: mu * dt is de kans op een extinction event per species in een tijdsinterval dt.
-1/mu geeft de verwachte tijd tot een extinctie event van een soort.
-mu is dus een probability rate.}
-\richel{'mu is dus een probability rate' with what unit? You can redirect me to an article with the proper wording, or finish this sentence: 'the extinction rate of good species, mu\_g ...'. Thanks!}
-The speciation completion rates we use are $\lambda = 0.1$, $0.3$, $1.0$ and $10^9$. We use $10^9 \approx \infty$
-to mimic the BD model, because the PBD model reduces to the BD model for $\lambda = \infty$.
-This allows us to measure the baseline error, 
-which is the difference between inferred tree and true species tree that arises purely due to noise 
-because the generating model and the model used in inference are identical in this case.
-We use a set of extinction rates, $M = \{0.0, 0.1, 0.2, \infty\}$, of which the extinction
-rate of good species $\mu_g \in M$ and incipient species $\mu_i \in M$.
+\richel{Start new}
+The PBD model has five parameters, depicted in table \ref{table:parameters}. 
+The speciation completion rates $\lambda$ we use are $0.1$, $0.3$, $1.0$ and $10^9$ probability of occurence per time unit. 
+\richel{Is the unit correct unit now?}
+The extinction rates $\mu = \mu_g = \mu_i$ we use are $0.0$, $0.1$ and $0.2$ probability of occurence per time unit.
+We use expected mean tree sizes $n$ of 50, 100 and 200 good taxa.
+From each combination of $\lambda$, $\mu$ and $n$, we derive a speciation initiation rate $b = b_i = b_g$, shown in table \ref{table:pbd_parameters}.
+Our parameters are inspired on existing work \cite{etienne_and_rosindell_2012}\cite{etienne2014}.
+We use $\lambda = 10^9 \approx \infty$ to let the PBD model reduce to the BD model.
 
-From each biological parameter set, we simulate a protracted birth-death tree,
-using the PBD package (\cite{pbd}) in the R programming language (\cite{r}), 
-all with a crown age of 15 million years as used in \cite{etienne2014}.
-Each protracted birth-death tree uses a different random number
-generator seed, which makes all runs independent, resulting in a balanced 
-data set.
+We simulate protracted birth-death trees, using the PBD package (\cite{pbd}) in the R programming language (\cite{r}).
+For each combination of ${\lambda, \mu, b, n}$, we generate incipient species trees with a crown age of 15 million years.
+Only trees with the desired number of good taxa are kept.
+
+This research creates two data sets: a general one, to explore parameter space, 
+and one to investigate the effect of sampling incipient species (see below).
+For the general data set, all the trees with the correct number of good species are kept.
+For the data set to investigate sampling, only trees with the additional constraint of sampling having an effect are kept.
+As sampling does not have an effect for $\lambda = \infty$, this parameter value is absent in that data set.
+\richel{End new}
 
@@
 results in either the shortest or longest branch lengths.
+
 See the supplementary information for a visualization of these sampling methods.
-% \RAMPAL{Why not include that in the paper here; I think it is essential}
-% \richel{Because we are limited to two figures}
 Based on the sampled species tree, we simulate a DNA alignment that has the same history
@@
 with each of the posterior species trees yields a distribution of nLTT statistics. 
+\richel{Start new}
+The input trees generated with a $\lambda = 10^9$ allow us to measure the noise of the experiment.
+For $\lambda = \infty$, the PBD model that generates the starting trees reduces to a BD model.
+In the following steps, sampling will have no effect, BEAST2 will assume the correct speciation model,
+and the difference between inferred tree and true species tree are explained purely due to this experimental noise.
+\richel{End new}
 
 We produce two data sets as a comma-separated file.
-The general data set has 348 different combinations
+The general data set has 144 \richel{recalc} different combinations
 of biological parameter combinations, site and clock models.
-The data set to investigate sampling has 552 different combinations
+The data set to investigate sampling has ?552 \richel{recalc} different combinations
 of biological parameter combinations, site models, clock models 
@@
 biological parameter combination as a violin plot.
-We show combinations for which $b_g = b_i$ and $\mu_g = \mu_i$,
-to simplify the interpretation of the results, where the other combinations
-are shown in the supplementary material.
-Additionally, we only show the nLTT distributions
+\richel{Start new}
+We only show the nLTT distributions
 that were generated under the (correct) assumptions of a Jukes-Cantor site model
 and a strict clock model, separated per sampling method used. 
+\richel{End new}
 We display the nLTT statistic distributions separated per site or clock model 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%\section{Acknowledgements}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%\richel{journal does not request for this. Suggest to remove, but how to acknowledge Peregrine otherwise?}
-%We would like to thank the Center for Information Technology of the University of Groningen for their support
-%and for providing access to the Peregrine high performance computing cluster.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%\section{Authors' contributions}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%\richel{journal does not request for this}
-%RSE conceived the idea for this experiment. 
-%RJCB created and tested the experiment, 
-%and wrote the first draft of the manuscript. 
-%RSE contributed substantially to revisions.
 
@@
     \hline
-    $b_g$                 & Speciation initiation rate of a good species & 0.3, 0.5 \\
-    $b_i$                 & Speciation initiation rate of an incipient species & 0.3, 0.5 \\
-    $\lambda$             & Speciation completion rate & 0.1, 0.3, 1.0, $\infty$ \\
+    $b_g$                 & Speciation initiation rate of a good species & derived, see \ref{table:pbd_parameters} \\
+    $b_i$                 & Speciation initiation rate of an incipient species & derived, see \ref{table:pbd_parameters} \\
+    $\lambda$             & Speciation completion rate & 0.1, 0.3, 1.0, $10^9$ \\
     $\mu_g$               & Extinction rate of a good species & 0.0, 0.1, 0.2 \\
@@
     \hline
+    $n$                   & Number of good taxa & 50, 100, 200 \\
     $t_c$                 & Crown age & 15 \\
@@
 \begin{table}
+  \begin{tabular}{|l|l|l|l|l|}
+    \hline
+       & $\mu$ & $n$ & $\lambda$  & $b$ \\
+    \hline
+    1  & 0   & 50  & 0.1   & 0.30944 \\
+    2  & 0.1 & 50  & 0.1   & 0.39674 \\
+    3  & 0.2 & 50  & 0.1   & 0.48667 \\
+    4  & 0   & 100 & 0.1   & 0.36344 \\
+    5  & 0.1 & 100 & 0.1   & 0.45283 \\
+    6  & 0.2 & 100 & 0.1   & 0.54425 \\
+    7  & 0   & 200 & 0.1   & 0.41669 \\
+    8  & 0.1 & 200 & 0.1   & 0.50759 \\
+    9  & 0.2 & 200 & 0.1   & 0.6001  \\
+    10 & 0   & 50  & 0.3   & 0.25717 \\
+    11 & 0.1 & 50  & 0.3   & 0.34003 \\
+    12 & 0.2 & 50  & 0.3   & 0.42648 \\
+    13 & 0   & 100 & 0.3   & 0.30862 \\
+    14 & 0.1 & 100 & 0.3   & 0.39455 \\
+    15 & 0.2 & 100 & 0.3   & 0.48328 \\
+    16 & 0   & 200 & 0.3   & 0.35991 \\
+    17 & 0.1 & 200 & 0.3   & 0.44804 \\
+    18 & 0.2 & 200 & 0.3   & 0.53841 \\
+    19 & 0   & 50  & 1     & 0.2297  \\
+    20 & 0.1 & 50  & 1     & 0.30759 \\
+    21 & 0.2 & 50  & 1     & 0.38984 \\
+    22 & 0   & 100 & 1     & 0.2778  \\
+    23 & 0.1 & 100 & 1     & 0.35961 \\
+    24 & 0.2 & 100 & 1     & 0.44481 \\
+    25 & 0   & 200 & 1     & 0.32617 \\
+    26 & 0.1 & 200 & 1     & 0.41078 \\
+    27 & 0.2 & 200 & 1     & 0.49818 \\
+    28 & 0   & 50  & $10^9$ & 0.21589 \\
+    29 & 0.1 & 50  & $10^9$ & 0.28896 \\
+    30 & 0.2 & 50  & $10^9$ & 0.36635 \\
+    31 & 0   & 100 & $10^9$ & 0.26146 \\
+    32 & 0.1 & 100 & $10^9$ & 0.33872 \\
+    33 & 0.2 & 100 & $10^9$ & 0.41945 \\
+    34 & 0   & 200 & $10^9$ & 0.30733 \\
+    35 & 0.1 & 200 & $10^9$ & 0.38768 \\
+    36 & 0.2 & 200 & $10^9$ & 0.47099 \\
+    \hline
+  \end{tabular}
+  \caption{
+    The speciation parameters used. Starting from extinction rate $\mu$ ($\mu = \mu_g = \mu_i$), the expected
+    mean number of good species $n$, speciation completion rate $\lambda$, the speciation initation
+    rate $b$ ($b = b_g = b_i$) follows.  
+  }
+  \label{table:pbd_parameters}
+\end{table}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\begin{table}
   \centering 
@@
     of columns a certain item will occupy, resulting in a table of 
-    1023 columns and 20K rows.
+    1023 \richel{recalc} columns and 20K rows.
   }
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Acknowledgements}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\richel{put this section here, as the journal does not request for this}
+We would like to thank the Center for Information Technology of the University of Groningen for their support
+and for providing access to the Peregrine high performance computing cluster.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Authors' contributions}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\richel{put this section here, as the journal does not request for this}
+RSE conceived the idea for this experiment. 
+RJCB created and tested the experiment, and wrote the first draft of the manuscript. 
+RSE contributed substantially to revisions.
+
 \end{document}

### 2018-09-05 article.tex
Explicitly mention the incipient tree RNG seeds used

@@
 We simulate protracted birth-death trees, using the PBD package (\cite{pbd}) in the R programming language (\cite{r}).
+The first tree has a random number generator seed of 1, which is incremented by 1 for each simulated tree.
 For each combination of ${\lambda, \mu, b, n}$, we generate incipient species trees with a crown age of 15 million years.
@@
 \begin{figure}[!htbp]
-  \includegraphics[width=0.8\textwidth]{fig_general.png}
+  \includegraphics[width=\textwidth]{fig_general.png}
   \caption{
@@
     general data set, 
-    for the subset of combinations in which $b_g = b_i$, $\mu_g = \mu_i$, 
     under the (correct) assumptions of a strict clock and Jukes-Cantor site model.
@@
 \begin{figure}[!htbp]
-  \includegraphics[width=0.8\textwidth]{fig_sampling.png}
+  \includegraphics[width=\textwidth]{fig_sampling.png}
   \caption{
@@
     regime, using the data set conditioned on sampling regime having an effect, 
-    for the subset of combinations in which $b_g = b_i$, $\mu_g = \mu_i$, 
     under the (correct) assumptions of a strict clock and Jukes-Cantor site model.
@@
 \begin{figure}[!htbp]
-  \includegraphics[width=0.8\textwidth]{fig_sampling_methods.png}
+  \includegraphics[width=\textwidth]{fig_sampling_methods.png}
   \caption{
@@
 \begin{figure}[!htbp]
-  \includegraphics[width=0.8\textwidth]{fig_clock_model.png}
+  \includegraphics[width=\textwidth]{fig_clock_model.png}
   \caption{
@@
     using the general data set, 
-    for the subset of combinations in which $b_g = b_i$, $\mu_g = \mu_i$, 
     under the (correct) assumption of a Jukes-Cantor site model.
@@
 \begin{figure}[!htbp]
-  \includegraphics[width=0.8\textwidth]{fig_site_model.png}
+  \includegraphics[width=\textwidth]{fig_site_model.png}
   \caption{
@@
     using the general data set, 
-    for the subset of combinations in which $b_g = b_i$, $\mu_g = \mu_i$, 
     under the (correct) assumption of a strict clock model.
@@
     $f_i$                 & MCMC sampling interval & 1K or more \\
-    $R_i$                 & RNG seed incipient tree and randomly sampled species tree & 1, 2, ... \\
+    $R_i$                 & RNG seed incipient tree and randomly sampled species tree & 1, 2, etc. \\
     $R_a$                 & RNG seed alignment simulation & $R_i$ \\
@@
     the biological parameter set. 
+    \richel{Start new}
     The RNG seed $R_i$ is 1 for the first simulation of the general data set, 2 for the next,
-    and so on, up to and including 3480. The RNG seeds for the data set investigating the effect 
+    and so on.  
+    The RNG seeds for the data set investigating the effect 
     of sampling continue from there, but only those RNG seeds are used in which sampling has an effect.
+    \richel{End new}
     The sampling methods are abbreviated as such: 'R' denotes random

### 2018-09-17 article.tex
Add Maturana text temporarily

@@
 
+From Maturana et al:
+Although the marginal likelihood is generally ignored in parameter inference, it plays a key role in model
+selection: it is a measure of the goodness of fit. Indeed, it is the probability of the data given the model,
+i.e., it is by definition a measure of model fit. The marginal likelihood acts as the normalisation constant in
+the posterior distribution making it a probability density function. Thus, this quantity is a multidimensional
+integral of the prior distribution times the likelihood function over the parameter space. MCMC methods
+used for parameter estimation within a model use only ratios of posterior densities, and are therefore unable
+to measure its normalisation in general.
+Unlike maximum likelihood, which represents the model fit at a single point, this quantity stands for an average
+of how well the model fits the data. By being an average of the likelihood function with respect to the prior,
+the model with the greatest evidence might be different from the model with the highest likelihood because
+the prior could down-weight some regions of parameter space. Also, the marginal likelihood is sensitive to the
+size of the region over which the likelihood is high. As a result, both methods could favour different models.
+Despite its important role in model selection, the marginal likelihood is usually analytically intractable and
+has to be approximated by numerical methods.
+
 We produce two data sets as a comma-separated file.

### 2018-09-17 article.tex
Changes of @rsetienne

--- 2018-09-05 article.tex
@@
 The PBD model has five parameters, depicted in table \ref{table:parameters}. 
-The speciation completion rates $\lambda$ we use are $0.1$, $0.3$, $1.0$ and $10^9$ probability of occurence per time unit. 
-\richel{Is the unit correct unit now?}
-The extinction rates $\mu = \mu_g = \mu_i$ we use are $0.0$, $0.1$ and $0.2$ probability of occurence per time unit.
-We use expected mean tree sizes $n$ of 50, 100 and 200 good taxa.
-From each combination of $\lambda$, $\mu$ and $n$, we derive a speciation initiation rate $b = b_i = b_g$, shown in table \ref{table:pbd_parameters}.
-Our parameters are inspired on existing work \cite{etienne_and_rosindell_2012}\cite{etienne2014}.
-We use $\lambda = 10^9 \approx \infty$ to let the PBD model reduce to the BD model.
+The per species speciation completion rates $\lambda$ we use are $0.1$, $0.3$, $1.0$ and $10^9$.
+This means that there is a $0.15 dt$ ($0.35 dt$, $1.0 dt$, $10^9 dt$) probability of speciation completion occurring in an infinitesimal time $dt$.
+We use per species extinction rates of $\mu = \mu_g = \mu_i$ $0.0$, $0.1$ and $0.2$.
+For each combination of $\lambda$, $\mu$ and $n$, we use a speciation initiation rate $b = b_i = b_g$ that gives a given value of the expected number of species.
+We use expected mean tree sizes $n$ of 50, 100 and 200 good taxa and the corresponding speciation initiation rates for all parameter combinations are shown in table \ref{table:pbd_parameters}.
+Our parameters are inspired by existing work on protracted speciation \cite{etienne_and_rosindell_2012}\cite{etienne2014}.
+We use $\lambda = 10^9 \approx \infty$ as our control for which the PBD model reduces to the BD model.
 
@@
 
-This research creates two data sets: a general one, to explore parameter space, 
+We create two data sets: a general one, to explore parameter space, 
 and one to investigate the effect of sampling incipient species (see below).
@@
 For $\lambda = \infty$, the PBD model that generates the starting trees reduces to a BD model.
-In the following steps, sampling will have no effect, BEAST2 will assume the correct speciation model,
-and the difference between inferred tree and true species tree are explained purely due to this experimental noise.
+In the subsequent steps, sampling will have no effect in this case, because BEAST2 will assume the correct speciation model,
+so the difference between inferred tree and true species tree are explained purely due to this experimental noise.
 \richel{End new}

### 2018-10-11 article.tex
Fix typo
Methods first paragraph

@@
 \usepackage[english]{babel}
+
+\newcommand*\mean[1]{\overline{#1}}
 
@@
 
-\richel{Start new}
 The PBD model has five parameters, depicted in table \ref{table:parameters}. 
 The per species speciation completion rates $\lambda$ we use are $0.1$, $0.3$, $1.0$ and $10^9$. 
-This means that there is a $0.15 dt$ ($0.35 dt$, $1.0 dt$, $10^9 dt$) probability of speciation completion occurring in an infinitesimal time $dt$.
+This means that there is a $0.15\thinspace dt$ ($0.35\thinspace dt$, $1.0\thinspace dt$, $10^9\thinspace dt$) 
+probability of speciation completion occurring in an infinitesimal time $dt$. 
 We use per species extinction rates of $\mu = \mu_g = \mu_i$ $0.0$, $0.1$ and $0.2$.
-For each combination of $\lambda$, $\mu$ and $n$, we use a speciation initiation rate $b = b_i = b_g$ that gives a given value of the expected number of species.
-We use expected mean tree sizes $n$ of 50, 100 and 200 good taxa and the corresponding speciation initiation rates for all parameter combinations are shown in table \ref{table:pbd_parameters}.
-Our parameters are inspired by existing work on protracted speciation \cite{etienne_and_rosindell_2012}\cite{etienne2014}.
+We use tree sizes of $n$ of 50, 100 and 200 good taxa.
+For each combination of $\lambda$, $\mu$ and $n$, 
+we use a speciation initiation rate $b = b_i = b_g$ 
+so that the expected mean number of species $\mathbf{E}(\mean{n}; b, \lambda, \mu)$,
+given a $b$, $\lambda$ and $\mu$,
+equals the desired number of species $n$.
+$b$ is calculated using the PBD R package (\cite{pbd})
+for each parameter combination and shown in table \ref{table:pbd_parameters}.
 We use $\lambda = 10^9 \approx \infty$ as our control for which the PBD model reduces to the BD model.
@@
 For example, for a DNA sequence length of 12, this would be AAACCCGGGTTT. 
-The order of nucletides does not matter in this study, 
+The order of nucleotides does not matter in this study, 
 because we do not consider several partitions of the sequence with their own parameters. 

### 2018-10-11 article.tex
Describe model selection

@@
 For the general data set, all the trees with the correct number of good species are kept.
-For the data set to investigate sampling, only trees with the additional constraint of sampling having an effect are kept.
-As sampling does not have an effect for $\lambda = \infty$, this parameter value is absent in that data set.
-\richel{End new}
+There is an additional selection criterion, for the data set to investigate sampling:
+to generate that data set, only incipient species trees are kept, 
+on which the two sampling methods (see below) result in different species trees.
+As sampling will never have an effect for $\lambda = \infty$, 
+this parameter value is absent in that data set.
 
@@
 with each of the posterior species trees yields a distribution of nLTT statistics. 
-\richel{Start new}
 The input trees generated with a $\lambda = 10^9$ allow us to measure the noise of the experiment.
@@
 so the difference between inferred tree and true species tree are explained purely due to this experimental noise.
-\richel{End new}
+
+As described above, per alignment, we do four different Bayian analyses,
+as we use two different site models and two different clock models.
+We know the generative site model (which is JC69) and clock 
+model (a strict clock), but want to explore how often the correct model
+is indeed preferred.
 
@@
 \begin{figure}[!htbp]
-  \includegraphics[width=\textwidth]{fig_general.png}
+  \includegraphics[width=\textwidth]{fig_sampling.png}
   \caption{
-    nLTT statistic distribution per biological parameter set, using the
-    general data set, 
+    nLTT statistic distribution per biological parameter set per sampling
+    regime, using the data set conditioned on sampling regime having an effect, 
     under the (correct) assumptions of a strict clock and Jukes-Cantor site model.
@@
 \begin{figure}[!htbp]
-  \includegraphics[width=\textwidth]{fig_sampling.png}
+  \includegraphics[width=\textwidth]{fig_model_selection.png}
   \caption{
-    nLTT statistic distribution per biological parameter set per sampling
-    regime, using the data set conditioned on sampling regime having an effect, 
-    under the (correct) assumptions of a strict clock and Jukes-Cantor site model.
+    Model preference on the general parameter set.
   }
@@
     in which the sub-species are picked to assure the longest branch lengths.
+  }
+\end{figure}
+
+\begin{figure}[!htbp]
+  \includegraphics[width=\textwidth]{fig_general.png}
+  \caption{
+    nLTT statistic distribution per biological parameter set, using the
+    general data set, 
+    under the (correct) assumptions of a strict clock and Jukes-Cantor site model.
   }

### 2018-10-11 article.tex
Describe NS

@@
 model (a strict clock), but want to explore how often the correct model
-is indeed preferred.
-
-From Maturana et al:
-Although the marginal likelihood is generally ignored in parameter inference, it plays a key role in model
-selection: it is a measure of the goodness of fit. Indeed, it is the probability of the data given the model,
-i.e., it is by definition a measure of model fit. The marginal likelihood acts as the normalisation constant in
-the posterior distribution making it a probability density function. Thus, this quantity is a multidimensional
-integral of the prior distribution times the likelihood function over the parameter space. MCMC methods
-used for parameter estimation within a model use only ratios of posterior densities, and are therefore unable
-to measure its normalisation in general.
-Unlike maximum likelihood, which represents the model fit at a single point, this quantity stands for an average
-of how well the model fits the data. By being an average of the likelihood function with respect to the prior,
-the model with the greatest evidence might be different from the model with the highest likelihood because
-the prior could down-weight some regions of parameter space. Also, the marginal likelihood is sensitive to the
-size of the region over which the likelihood is high. As a result, both methods could favour different models.
-Despite its important role in model selection, the marginal likelihood is usually analytically intractable and
-has to be approximated by numerical methods.
+is indeed preferred. Per alignment, we measure the estimated marginal 
+likelihood (which is the probability of the data given the model)
+for each of the four models and measure their relative propertions.
+To estimate the marginal likelihood, we use the novel 
+Nested Sampling approach (\cite{maturana_et_al_2018}),
+which we configure to have an effective sample size (in estimating
+the marginal likelihood) of at least 200, by using a single-particle
+sub-chain with a length equal to the MCMC interval.
+\richel{
+  There are no rules described in the setup of the NS
+  (for example, what should the relation between MCMC chain length
+  and NS sub-chain length be?),
+  but one can experiment with a sub-chain length and number of particles.
+  I will need to do some pilot studies to verify if this is
+  the correct way to do and describe this.  
+}
 

### 2018-10-11 article.tex
New version

@@
 
+\richel{Start new}
 The PBD model has five parameters, depicted in table \ref{table:parameters}. 
@@
 this parameter value is absent in that data set.
+\richel{End new}
 
@@
 
+\richel{
+  Note there is no twinning, as in the razzo package.
+  Already sent an email to discuss this with RSE.
+}
+
+\richel{Start new}
 As described above, per alignment, we do four different Bayian analyses,
@@
 }
+\richel{End new}
 
@@
 
-For both data sets, we display the nLTT statistics distribution per
-biological parameter combination as a violin plot.
 \richel{Start new}
-We only show the nLTT distributions
-that were generated under the (correct) assumptions of a Jukes-Cantor site model
-and a strict clock model, separated per sampling method used. 
+In this article, we showcase the effect of sampling
+and the certainty when selecting a (site and clock) model. 
+We show the effect sampling has on the inference error,
+for the nLTT distributions generated from assuming the (correct) Jukes-Cantor 
+site model and a strict clock model. We do so per parameter combination,
+displaying the distribution as a violin plot.
+In the supplementary information, we display these error distributions 
+of the general data set all combined, or separated per site or clock model.
+The certainty in the model selection is again shown per parameter setting,
+as a stacked bar. 
+\richel{
+  I am unsure how to depict that stacked bar and its uncertainty.
+  Sure, the stacked bars (without uncertainty) will work,
+  but information about uncertainty is lost. 
+  See figure 'Showing a stacked bar with uncertainty'
+  for three equivalent ways to display a stacked bar with uncertainty.
+}
+
+\begin{figure}[!htbp]
+  \includegraphics[width=\textwidth]{fig_stacked_bar_with_uncertainty.png}
+  \caption{
+    Showing a stacked bar with uncertainty. These three different ways
+    display the same underlying data.
+  }
+\end{figure}
+
 \richel{End new}
-We display the nLTT statistic distributions separated per site or clock model 
-in the supplementary information.
 
@@
   \caption{
-    Model preference on the general parameter set. 
+    Model preference on the general data set.
   }
@@
     $11$ & ESSes of all parameters estimated by BEAST2 (see specs below) \\
+    $1$ & Marginal likelihood estimate \\
+    $1$ & Marginal likelihood estimation uncertainty \\
+    $1$ & Marginal likelihood ESS \\
     \hline

### 2019-05-06 article.tex
EOD
Move content to content.tex

--- 2018-10-11 article.tex
@@
 
-\begin{abstract}
-
-  % From 'How to construct a Nature summary paragraph'
-
-  % A short abstract of fewer than 200 words is acceptable.
-
-  % One or two sentences providing a basic
-  % introduction to the field,
-  % comprehensible to a scientist in any discipline.
-  The tools for reconstructing phylogenetic relationships between taxonomic 
-  units (e.g. species) have become very advanced in the last three decades. 
-    % Two to three sentences of
-  % more detailed background, comprehensible to
-  % scientists in related disciplines.
-  Among the most popular tools are Bayesian approaches, 
-  such as \\ BEAST, MrBayes and RevBayes, 
-  that use efficient tree sampling routines to create a posterior probability distribution 
-  of the phylogenetic tree. 
-  A feature of these approaches is the possibility to incorporate known or hypothesized structure of the phylogenetic tree through the tree prior. 
-  It has been shown that the effect of the prior on the posterior distribution 
-  of trees can be substantial.
-
-  % One sentence clearly stating the general
-  % problem being addressed by this particular
-  % study.
-  Currently implemented tree priors assume that speciation is instantaneous,
-  where we know that speciation can be a gradual process.
-
-  % One sentence summarising the main
-  % result (with the words “here we show”
-  % their equivalent).
-  Here we explore the effects of ignoring 
-  the protractedness of the speciation process with an extensive simulation study.
-
-  % Two or three sentences explaining what
-  % the main result reveals in direct
-  % comparison to what was thought to be the case
-  % previously, or how the main result adds to
-  % previous knowledge.
-
-  % One or two sentences to put the results into a
-  % more general context.
-
-  % Two or three sentences to provide a
-  % broader perspective, readily comprehensible
-  % to a scientist in any discipline, may be included
-  % in the first paragraph if the editor considers that the accessibility of the paper is significantly enhanced
-  % by their inclusion. Under these circumstances, the length of the paragraph can be up to 300 words.
-  % (The above example is 190 words without the final section, and 250 words with it).
-
-  We compare the inferred tree to the simulated tree, and find that ....
-
-\end{abstract}
-
-{\bf Keywords:} computational biology, evolution, phylogenetics, Bayesian analysis, tree prior
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Introduction}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-The computational tools that are currently available 
-to the phylogeneticists go beyond the wildest 
-imagination of those living four decades ago.
-Advances in computational power allowed the first cladograms to be inferred 
-from DNA alignments in 1981 (\cite{felsenstein1981}), and  
-the first Bayesian tools emerged in 1996 (\cite{rannala1996}),
-providing unprecedented flexibility in the setup of a phylogenetic model.
-
-Currently, the most popular Bayesian phylogenetics tools are \\ 
-BEAST (\cite{beast}) and its offshoot BEAST2 (\cite{beast2}), 
-MrBayes (\cite{mrbayes}) and RevBayes (\cite{revbayes}). 
-They allow to incorporate known or hypothesized structure of a phylogenetic 
-tree-to-be-inferred through model priors. 
-With these priors and an alignment of DNA, RNA or protein sequences, 
-they create a sample of the posterior distribution
-of phylogenies and parameter estimates (of the models used as a prior), 
-in which more probable combinations are represented more often.
-Each of these tools use efficient tree sampling routines to rapidly create an 
-informative posterior.
-
-The model priors in Bayesian phylogenetic reconstruction 
-can be grouped into three categories: (1) site model, specifying 
-nucleotide substitutions, (2) clock model, specifying
-the rate of mutation per lineage in time, and (3) tree model, 
-constituting the speciation model underlying branching events (speciation) 
-and branch termination (extinction).
-The choice of site model (\cite{posada_and_buckley_2004}), 
-clock model (\cite{baele_et_al_2012}) 
-or tree prior (\cite{moller2018, yang_and_ranalla_2005}) is known to affect
-the posterior.
-
-Current phylogenetic tools use tree priors 
-that assume speciation is instantaneous, whilst we know that, 
-speciation is often a gradual process (\cite{schluter2009}).
-The (constant-rate) birth-death (BD) model is a commonly 
-used tree prior, but it ignores this temporal aspect of speciation.
-The protracted birth-death (PBD) model, an extension of 
-the BD model, does incorporate the idea that speciation takes time.
-In this model, a branching event does not give rise to a new species, but to
-a new species-to-be, called an incipient species. Such an incipient
-species may go extinct, finish its speciation to become a good species, or give
-rise to new incipient species. Protracted speciation may explain observed 
-declines in lineage accumulation (\cite{etienne_and_rosindell_2012}).
-
-Unfortunately, a tree prior according to this model, 
-providing the probability of a species tree under the PBD model, 
-is unavailable in current Bayesian phylogenetic tools. 
-Whilst an approximate formula for this probability has been derived (\cite{lambert_et_al_2015}) 
-and the approximation is very good (\cite{simonet_et_al_2018}), 
-it has not been implemented as tree prior yet. 
-There are various reasons for this. 
-First, the computation of this probability involves solving a set of 
-non-linear differential equations, and while this computation is quite fast, 
-it still takes much more time than the corresponding probability 
-of the BD model which is a simple analytical formula. 
-In a Bayesian MCMC chain, the tree prior probability must be calculated many times, 
-and hence the total computation will take considerably longer with a PBD tree prior. 
-Furthermore, the approximate probability is a probability for the species tree 
-assuming an underlying incipient species tree. 
-It can be safely used as tree prior when only one individual per species is sampled, 
-but if one has multiple samples per species - which is currently often the case - the methods 
-to account for this such as the multi-species coalescent (\cite{heled_and_drummond_2009}) 
-may not be compatible with the underlying incipient species tree. 
-More precisely, the phylogeny under the PBD model may contain paraphylies, 
-while the multi-species coalescent was developed exactly to avoid 
-these by explaining them as arising from incomplete lineage sorting. 
-Because of these paraphylies there is no such thing as a true species tree in the PBD model. 
-To get a species-level tree one must sample one incipient species per species. 
-Which incipient species is sampled may therefore have an impact on the species tree.
-
-Here we aim to explore the effect of using the
-BD prior on PBD simulated phylogenies, taking into account possible sampling effects.
-In brief, we simulate protracted phylogenies using the PBD process,
-from which we sample a species tree in two very different ways. Given this species tree, 
-we simulate a DNA sequence alignment. Then, we use BEAST2 on these alignments
-to infer a posterior of phylogenies, using a BD prior. We quantify the difference
-between the (BD) posterior phylogenies and the simulated (PBD) species tree.
-Furthermore, while we evidently know the clock and site models used in the simulation, 
-using a different clock and/or site model prior in inference 
-may compensate or increase this difference between inferred and simulated tree. 
-To study this, we also explore the effect of 
-a different clock and site model prior in inference.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Methods (but we are not allowed to use this header)
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-The PBD model has five parameters, depicted in table \ref{table:parameters}. 
-The per species speciation completion rates $\lambda$ we use are $0.1$, $0.3$, $1.0$ and $10^9$. 
-This means that there is a $0.15\thinspace dt$ ($0.35\thinspace dt$, $1.0\thinspace dt$, $10^9\thinspace dt$) 
-probability of speciation completion occurring in an infinitesimal time $dt$. 
-We use per species extinction rates of $\mu = \mu_g = \mu_i$ $0.0$, $0.1$ and $0.2$.
-We use tree sizes of $n$ of 50, 100 and 200 good taxa.
-For each combination of $\lambda$, $\mu$ and $n$, 
-we use a speciation initiation rate $b = b_i = b_g$ 
-so that the expected mean number of species $\mathbf{E}(\mean{n}; b, \lambda, \mu)$,
-given a $b$, $\lambda$ and $\mu$,
-equals the desired number of species $n$.
-$b$ is calculated using the PBD R package (\cite{pbd})
-for each parameter combination and shown in table \ref{table:pbd_parameters}.
-We use $\lambda = 10^9 \approx \infty$ as our control for which the PBD model reduces to the BD model.
-
-We simulate protracted birth-death trees, using the PBD package (\cite{pbd}) in the R programming language (\cite{r}).
-The first tree has a random number generator seed of 1, which is incremented by 1 for each simulated tree.
-For each combination of ${\lambda, \mu, b, n}$, we generate incipient species trees with a crown age of 15 million years.
-Only trees with the desired number of good taxa are kept.
-
-We create two data sets: a general one, to explore parameter space, 
-and one to investigate the effect of sampling incipient species (see below).
-For the general data set, all the trees with the correct number of good species are kept.
-There is an additional selection criterion, for the data set to investigate sampling:
-to generate that data set, only incipient species trees are kept, 
-on which the two sampling methods (see below) result in different species trees.
-As sampling will never have an effect for $\lambda = \infty$, 
-this parameter value is absent in that data set.
-
-From each incipient species tree, we construct a species tree,
-by sampling one incipient/good species per good species. 
-For example, when an
-incipient species branched off from its mother lineage, 
-both of these subspecies are recognized as representing the species, 
-and hence both can be picked as an (equally good) representative of the species. 
-Here, we use three sampling scenarios,
-in which we pick the representative randomly or in such a way that this
-results in either the shortest or longest branch lengths.
-
-See the supplementary information for a visualization of these sampling methods.
-Based on the sampled species tree, we simulate a DNA alignment that has the same history
-as this species tree, using the \verb;phangorn; package (\cite{phangorn}). 
-We set the nucleotides of the DNA alignment to follow a Jukes-Cantor (\cite{jc69})
-nucleotide substitution model, in which all nucleotide-to-nucleotide transitions
-are equally likely. 
-The DNA sequence of the root ancestor consists of four equally sized single-nucleotide blocks of adenine, cytosine, guanine and thymine respectively. 
-For example, for a DNA sequence length of 12, this would be AAACCCGGGTTT. 
-The order of nucleotides does not matter in this study, 
-because we do not consider several partitions of the sequence with their own parameters. 
-Only the frequency of occurrence matters.
-In our Bayesian inference (see below) we use the same site model as the (obviously correct) site model prior,
-but we also explore the effect of assuming a more complex site model prior.
-We predict with the more complex substitution model, 
-that there will be more noise and hence our inference error will increase.
-On the other hand, we dare not rule out that the inference error will decrease,
-due to more flexibility in the more complex prior.
-We set the mutation rate in such a way to maximize the information contained in the alignment.
-To do so, we set the mutation rate such that we expect on average one (possibly silent) mutation per nucleotide
-between crown age and present, which equates to $\frac{1}{15}$ mutations
-per million years.
-The DNA sequence length is chosen to provide a
-resolution of $10^3$ years, 
-% If sims take too long, we will decrease this resolution
-that is, to have one expected nucleotide change 
-per $10^3$ years per lineage on average. As one nucleotide is expected 
-to have on average one (possibly silent) mutation per 15 million years, $15 \cdot 10^3$
-nucleotides result in 1 mutation per alignment per $10^3$ years (which is
-coincidentally the same as \cite{moller2018}). 
-The simulation of these DNA aligments follows a strict clock model, 
-which we will specify as one of the two clock models assumed in the Bayesian inference (see below).
-
-From an alignment, we run a Bayesian analysis and 
-create a posterior distribution of trees and parameters
-using the \verb;babette; (\cite{babette}) package
-that sets the input parameters similar to BEAUti 2 and then runs BEAST2. 
-For our site model, we assume either a Jukes-Cantor or GTR nucleotide substitution model.
-The Jukes-Cantor model is the correct one, as it is used for simulating that alignment,
-where the GTR model is the site model that is picked as a default by most users.
-For our clock model, we assume either a strict or relaxed log-normal 
-clock model. 
-Also here, the strict clock model is the correct one, as it is used for simulating the alignment,
-but the relaxed log-normal clock model is the one most commonly used.
-We set the BD model as a tree prior, 
-as gauging the effect of this incorrect assumption is the goal of this study. 
-We assume an MRCA prior with a tight normal distribution
-around the crown age, by choosing the crown age as mean, and a standard deviation 
-of $0.5 \cdot 10^{-3}$ time units,
-resulting in 95\% of the crown ages inferred have the same resolution (of $10^{-3}$ time 
-units) as the alignment. 
-We ran the MCMC chain to generate 1111 states,
-of which we remove the first 10\% (also called the 'burn-in'). 
-Of the remaining
-1000 MCMC states, the effective sample size (ESS) of the posterior 
-must at least be 200
-for a strong enough inference (\cite{beastbook}). An ESS can be increased by increasing
-the number of samples or decreasing the autocorrelation between samples. 
-If the ESS is less than 200, we decrease autocorrelation by doubling 
-the MCMC sampling interval of that simulation, until the ESS exceeds 200.
-
-We compare each posterior phylogeny to the (sampled) species tree
-using the nLTT statistic (\cite{janzen2015}), from the \verb;nLTT; package (\cite{nltt}). 
-The nLTT statistic equals the area between the normalized
-lineages-through-time-plots of two phylogenies, which has a range 
-from zero (for identical phylogenies) to one. We use inference error 
-and nLTT statistic interchangeably. Comparing the simulated species tree
-with each of the posterior species trees yields a distribution of nLTT statistics. 
-The input trees generated with a $\lambda = 10^9$ allow us to measure the noise of the experiment.
-For $\lambda = \infty$, the PBD model that generates the starting trees reduces to a BD model.
-In the subsequent steps, sampling will have no effect in this case, because BEAST2 will assume the correct speciation model,
-so the difference between inferred tree and true species tree are explained purely due to this experimental noise.
-
-As described above, per alignment, we do four different Bayian analyses,
-as we use two different site models and two different clock models.
-We know the generative site model (which is JC69) and clock 
-model (a strict clock), but want to explore how often the correct model
-is indeed preferred. Per alignment, we measure the estimated marginal 
-likelihood (which is the probability of the data given the model)
-for each of the four models and measure their relative propertions.
-To estimate the marginal likelihood, we use the novel 
-Nested Sampling approach (\cite{maturana_et_al_2018}),
-which we configure to have an effective sample size (in estimating
-the marginal likelihood) of at least 200, by using a single-particle
-sub-chain with a length equal to the MCMC interval.
-\richel{
-  There are no rules described in the setup of the NS
-  (for example, what should the relation between MCMC chain length
-  and NS sub-chain length be?),
-  but one can experiment with a sub-chain length and number of particles.
-  I will need to do some pilot studies to verify if this is
-  the correct way to do and describe this.  
-}
-
-We produce two data sets as a comma-separated file.
-The general data set has 144 \richel{recalc} different combinations
-of biological parameter combinations, site and clock models.
-The data set to investigate sampling has ?552 \richel{recalc} different combinations
-of biological parameter combinations, site models, clock models 
-and sampling methods. The experiment is computationally intensive:
-pilot experiments show that the experiment takes roughly 100 days
-of CPU time and 20 days of wall clock time (which includes the queued 
-waiting for computational resources) per replicate. 
-Due to this, we choose to perform ten replicates, so that the complete
-experiment will take an acceptable time of roughly seven months.
-
-For both data sets, we display the nLTT statistics distribution per
-biological parameter combination as a violin plot.
-\richel{Start new}
-We only show the nLTT distributions
-that were generated under the (correct) assumptions of a Jukes-Cantor site model
-and a strict clock model, separated per sampling method used. 
-\richel{End new}
-We display the nLTT statistic distributions separated per site or clock model 
-in the supplementary information.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Results}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\begin{figure}[!htbp]
-  \includegraphics[width=\textwidth]{fig_sampling.png}
-  \caption{
-    nLTT statistic distribution per biological parameter set per sampling
-    regime, using the data set conditioned on sampling regime having an effect, 
-    under the (correct) assumptions of a strict clock and Jukes-Cantor site model.
-  }
-\end{figure}
-
-\begin{figure}[!htbp]
-  \includegraphics[width=\textwidth]{fig_model_selection.png}
-  \caption{
-    Model preference on the general parameter set. 
-  }
-\end{figure}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Glossary}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Please supply, as a separate list, the definitions of field-specific terms used in your article.
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{table}
-  \centering 
-  \begin{tabular}{l p{0.65\textwidth}}
-    \hline
-    Term                  & Definition \\
-    \hline
-    \hline
-    Phylogenetics         & The inference of evolutionary relationships of groups of organisms using genetics \\
-    Model prior           & Knowledge or assumptions about the ontogeny of evolutionary histories \\
-    Posterior             & A collection of phylogenies and parameter estimates, in which more probable combinations (determined by the data and the model prior) are presented more frequently \\
-    Protracted speciation & The process in which speciation takes two events: a speciation-initiation event and a speciation-completion event  \\
-    Speciation initiation & The start of a speciation event creating an incipient species \\
-    Speciation completion & The end of a speciation event, in which an incipient species becomes or is recognized as a good species \\
-    \hline
-  \end{tabular}
-  \caption{
-    Glossary
-  }
-  \label{table:glossary}
-\end{table}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-% Bibliography
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% MEE style
-\bibliographystyle{mee}
-\bibliography{article}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\appendix
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\begin{figure}[!htbp]
-  \includegraphics[width=\textwidth]{fig_sampling_methods.png}
-  \caption{
-    Sampling a species tree from an incipient species tree. At the top left,
-    an incipient species tree is shown, of three different good species (the
-    first and second number in the taxon label) and four different subspecies (the
-    third number in the taxon tabel). The other four trees are species trees,
-    that use a different sampling method to determine which sub-species is picked to
-    represent a good species. These are: 'Youngest', 'Oldest', 'Shortest' and
-    'Longest'. With 'Youngest' the youngest sub-species is picked to represent the
-    good species. With 'Oldest' the oldest sub-species is picked to represent the
-    good species. 'Shortest' is the sampling method in which the sub-species are
-    picked to assure the shortest branch lengths. 'Longest' is the sampling method 
-    in which the sub-species are picked to assure the longest branch lengths.
-  }
-\end{figure}
-
-\begin{figure}[!htbp]
-  \includegraphics[width=\textwidth]{fig_general.png}
-  \caption{
-    nLTT statistic distribution per biological parameter set, using the
-    general data set, 
-    under the (correct) assumptions of a strict clock and Jukes-Cantor site model.
-  }
-\end{figure}
-
-\begin{figure}[!htbp]
-  \includegraphics[width=\textwidth]{fig_clock_model.png}
-  \caption{
-    nLTT statistic distribution per biological parameter set per clock model,
-    using the general data set, 
-    under the (correct) assumption of a Jukes-Cantor site model.
-  }
-\end{figure}
-
-\begin{figure}[!htbp]
-  \includegraphics[width=\textwidth]{fig_site_model.png}
-  \caption{
-    nLTT statistic distribution per biological parameter set per site model,
-    using the general data set, 
-    under the (correct) assumption of a strict clock model.
-  }
-\end{figure}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{table}
-  \centering 
-  \begin{tabular}{p{0.05\textwidth} p{0.6\textwidth} p{0.2\textwidth}}
-    \hline
-                          & Description & Values \\
-    \hline
-    \hline
-    $b_g$                 & Speciation initiation rate of a good species & derived, see \ref{table:pbd_parameters} \\
-    $b_i$                 & Speciation initiation rate of an incipient species & derived, see \ref{table:pbd_parameters} \\
-    $\lambda$             & Speciation completion rate & 0.1, 0.3, 1.0, $10^9$ \\
-    $\mu_g$               & Extinction rate of a good species & 0.0, 0.1, 0.2 \\
-    $\mu_i$               & Extinction rate of an incipient species & 0.0, 0.1, 0.2 \\
-    \hline
-    $n$                   & Number of good taxa & 50, 100, 200 \\
-    $t_c$                 & Crown age & 15 \\
-    $\sigma_c$            & Standard deviation around crown age & 0.001 \\
-    $M_s$                 & Sampling method & S, L, R \\
-    $M_c$                 & Clock model & S, RLN \\
-    $M_t$                 & Site model & JC69, GTR \\
-    $r$                   & Mutation rate & $\frac{1}{15}$ \\
-    $l_a$                 & DNA alignment length & $15K$ \\
-    $f_i$                 & MCMC sampling interval & 1K or more \\
-    $R_i$                 & RNG seed incipient tree and randomly sampled species tree & 1, 2, etc. \\
-    $R_a$                 & RNG seed alignment simulation & $R_i$ \\
-    $R_b$                 & RNG seed BEAST2 & $R_i$ \\
-    \hline
-  \end{tabular}
-  \caption{
-    Overview of the simulation parameters. Above the horizontal line is 
-    the biological parameter set. 
-    \richel{Start new}
-    The RNG seed $R_i$ is 1 for the first simulation of the general data set, 2 for the next,
-    and so on.  
-    The RNG seeds for the data set investigating the effect 
-    of sampling continue from there, but only those RNG seeds are used in which sampling has an effect.
-    \richel{End new}
-    The sampling methods are abbreviated as such: 'R' denotes random
-    sampling, 'S' is 'shortest' and 'L' is 'longest'. Sampling method $M_s$ is random for the general
-    data set. For the data set exploring the effect of sampling, we use 'shortest'
-    and 'longest' for each value of $R_i$ (which are random seeds in which sampling has an effect).
-    The clock models are abbreviated as 'S' for a strict and 'RLN' for a relaxed log-normal model.
-    The site models are abbreviated as 'JC69' for Jukes-Cantor (\cite{jc69}) and 'GTR' for the generalized 
-    time-reversible model (\cite{gtr}).
-  }
-  \label{table:parameters}
-\end{table}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{table}
-  \begin{tabular}{|l|l|l|l|l|}
-    \hline
-       & $\mu$ & $n$ & $\lambda$  & $b$ \\
-    \hline
-    1  & 0   & 50  & 0.1   & 0.30944 \\
-    2  & 0.1 & 50  & 0.1   & 0.39674 \\
-    3  & 0.2 & 50  & 0.1   & 0.48667 \\
-    4  & 0   & 100 & 0.1   & 0.36344 \\
-    5  & 0.1 & 100 & 0.1   & 0.45283 \\
-    6  & 0.2 & 100 & 0.1   & 0.54425 \\
-    7  & 0   & 200 & 0.1   & 0.41669 \\
-    8  & 0.1 & 200 & 0.1   & 0.50759 \\
-    9  & 0.2 & 200 & 0.1   & 0.6001  \\
-    10 & 0   & 50  & 0.3   & 0.25717 \\
-    11 & 0.1 & 50  & 0.3   & 0.34003 \\
-    12 & 0.2 & 50  & 0.3   & 0.42648 \\
-    13 & 0   & 100 & 0.3   & 0.30862 \\
-    14 & 0.1 & 100 & 0.3   & 0.39455 \\
-    15 & 0.2 & 100 & 0.3   & 0.48328 \\
-    16 & 0   & 200 & 0.3   & 0.35991 \\
-    17 & 0.1 & 200 & 0.3   & 0.44804 \\
-    18 & 0.2 & 200 & 0.3   & 0.53841 \\
-    19 & 0   & 50  & 1     & 0.2297  \\
-    20 & 0.1 & 50  & 1     & 0.30759 \\
-    21 & 0.2 & 50  & 1     & 0.38984 \\
-    22 & 0   & 100 & 1     & 0.2778  \\
-    23 & 0.1 & 100 & 1     & 0.35961 \\
-    24 & 0.2 & 100 & 1     & 0.44481 \\
-    25 & 0   & 200 & 1     & 0.32617 \\
-    26 & 0.1 & 200 & 1     & 0.41078 \\
-    27 & 0.2 & 200 & 1     & 0.49818 \\
-    28 & 0   & 50  & $10^9$ & 0.21589 \\
-    29 & 0.1 & 50  & $10^9$ & 0.28896 \\
-    30 & 0.2 & 50  & $10^9$ & 0.36635 \\
-    31 & 0   & 100 & $10^9$ & 0.26146 \\
-    32 & 0.1 & 100 & $10^9$ & 0.33872 \\
-    33 & 0.2 & 100 & $10^9$ & 0.41945 \\
-    34 & 0   & 200 & $10^9$ & 0.30733 \\
-    35 & 0.1 & 200 & $10^9$ & 0.38768 \\
-    36 & 0.2 & 200 & $10^9$ & 0.47099 \\
-    \hline
-  \end{tabular}
-  \caption{
-    The speciation parameters used. Starting from extinction rate $\mu$ ($\mu = \mu_g = \mu_i$), the expected
-    mean number of good species $n$, speciation completion rate $\lambda$, the speciation initation
-    rate $b$ ($b = b_g = b_i$) follows.  
-  }
-  \label{table:pbd_parameters}
-\end{table}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{table}
-  \centering 
-  \begin{tabular}{l l}
-    \hline
-    $n$ & Description \\
-    \hline
-    \hline
-    $12$   & simulation parameters, see table \ref{table:parameters} \\
-    $1000$ & nLTT statistic values \\
-    $11$ & ESSes of all parameters estimated by BEAST2 (see specs below) \\
-    \hline
-  \end{tabular}
-  \caption{
-    Specification of the data sets. Each row will contain one experiment,
-    where the columns contain parameters, measurements and diagnostics.
-    This table displays the content of the columns. 
-    $n$ denotes the number
-    of columns a certain item will occupy, resulting in a table of 
-    1023 \richel{recalc} columns and 20K rows.
-  }
-  \label{table:specs}
-\end{table}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{table}
-  \centering 
-  \begin{tabular}{l l}
-    \hline
-    \# & Description \\
-    \hline
-    \hline
-    1 & posterior \\
-    2 & likelihood \\
-    3 & prior \\
-    4 & treeLikelihood \\
-    5 & TreeHeight \\
-    6 & BirthDeath \\
-    7 & BDBirthRate \\
-    8 & BDDeathRate \\
-    9 & logP.mrca \\
-    10 & mrcatime \\
-    11 & clockRate \\
-    \hline
-  \end{tabular}
-  \caption{
-    Overview of the 11 parameters estimated by BEAST2
-  }
-  \label{table:estimated_parameters}
-\end{table}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Acknowledgements}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\richel{put this section here, as the journal does not request for this}
-We would like to thank the Center for Information Technology of the University of Groningen for their support
-and for providing access to the Peregrine high performance computing cluster.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Authors' contributions}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\richel{put this section here, as the journal does not request for this}
-RSE conceived the idea for this experiment. 
-RJCB created and tested the experiment, and wrote the first draft of the manuscript. 
-RSE contributed substantially to revisions.
+\include{content}
 

### 2019-05-06 content.tex
Move content to content.tex

\begin{abstract}

  % From 'How to construct a Nature summary paragraph'

  % A short abstract of fewer than 200 words is acceptable.

  % One or two sentences providing a basic
  % introduction to the field,
  % comprehensible to a scientist in any discipline.
  The tools for reconstructing phylogenetic relationships between taxonomic 
  units (e.g. species) have become very advanced in the last three decades. 
    % Two to three sentences of
  % more detailed background, comprehensible to
  % scientists in related disciplines.
  Among the most popular tools are Bayesian approaches, 
  such as \\ BEAST, MrBayes and RevBayes, 
  that use efficient tree sampling routines to create a posterior probability distribution 
  of the phylogenetic tree. 
  A feature of these approaches is the possibility to incorporate known or hypothesized structure of the phylogenetic tree through the tree prior. 
  It has been shown that the effect of the prior on the posterior distribution 
  of trees can be substantial.

  % One sentence clearly stating the general
  % problem being addressed by this particular
  % study.
  Currently implemented tree priors assume that speciation is instantaneous,
  where we know that speciation can be a gradual process.

  % One sentence summarising the main
  % result (with the words “here we show”
  % their equivalent).
  Here we explore the effects of ignoring 
  the protractedness of the speciation process with an extensive simulation study.

  % Two or three sentences explaining what
  % the main result reveals in direct
  % comparison to what was thought to be the case
  % previously, or how the main result adds to
  % previous knowledge.

  % One or two sentences to put the results into a
  % more general context.

  % Two or three sentences to provide a
  % broader perspective, readily comprehensible
  % to a scientist in any discipline, may be included
  % in the first paragraph if the editor considers that the accessibility of the paper is significantly enhanced
  % by their inclusion. Under these circumstances, the length of the paragraph can be up to 300 words.
  % (The above example is 190 words without the final section, and 250 words with it).

  We compare the inferred tree to the simulated tree, and find that ....

\end{abstract}

{\bf Keywords:} computational biology, evolution, phylogenetics, Bayesian analysis, tree prior

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

The computational tools that are currently available 
to the phylogeneticists go beyond the wildest 
imagination of those living four decades ago.
Advances in computational power allowed the first cladograms to be inferred 
from DNA alignments in 1981 (\cite{felsenstein1981}), and  
the first Bayesian tools emerged in 1996 (\cite{rannala1996}),
providing unprecedented flexibility in the setup of a phylogenetic model.

Currently, the most popular Bayesian phylogenetics tools are \\ 
BEAST (\cite{beast}) and its offshoot BEAST2 (\cite{beast2}), 
MrBayes (\cite{mrbayes}) and RevBayes (\cite{revbayes}). 
They allow to incorporate known or hypothesized structure of a phylogenetic 
tree-to-be-inferred through model priors. 
With these priors and an alignment of DNA, RNA or protein sequences, 
they create a sample of the posterior distribution
of phylogenies and parameter estimates (of the models used as a prior), 
in which more probable combinations are represented more often.
Each of these tools use efficient tree sampling routines to rapidly create an 
informative posterior.

The model priors in Bayesian phylogenetic reconstruction 
can be grouped into three categories: (1) site model, specifying 
nucleotide substitutions, (2) clock model, specifying
the rate of mutation per lineage in time, and (3) tree model, 
constituting the speciation model underlying branching events (speciation) 
and branch termination (extinction).
The choice of site model (\cite{posada_and_buckley_2004}), 
clock model (\cite{baele_et_al_2012}) 
or tree prior (\cite{moller2018, yang_and_ranalla_2005}) is known to affect
the posterior.

Current phylogenetic tools use tree priors 
that assume speciation is instantaneous, whilst we know that, 
speciation is often a gradual process (\cite{schluter2009}).
The (constant-rate) birth-death (BD) model is a commonly 
used tree prior, but it ignores this temporal aspect of speciation.
The protracted birth-death (PBD) model, an extension of 
the BD model, does incorporate the idea that speciation takes time.
In this model, a branching event does not give rise to a new species, but to
a new species-to-be, called an incipient species. Such an incipient
species may go extinct, finish its speciation to become a good species, or give
rise to new incipient species. Protracted speciation may explain observed 
declines in lineage accumulation (\cite{etienne_and_rosindell_2012}).

Unfortunately, a tree prior according to this model, 
providing the probability of a species tree under the PBD model, 
is unavailable in current Bayesian phylogenetic tools. 
Whilst an approximate formula for this probability has been derived (\cite{lambert_et_al_2015}) 
and the approximation is very good (\cite{simonet_et_al_2018}), 
it has not been implemented as tree prior yet. 
There are various reasons for this. 
First, the computation of this probability involves solving a set of 
non-linear differential equations, and while this computation is quite fast, 
it still takes much more time than the corresponding probability 
of the BD model which is a simple analytical formula. 
In a Bayesian MCMC chain, the tree prior probability must be calculated many times, 
and hence the total computation will take considerably longer with a PBD tree prior. 
Furthermore, the approximate probability is a probability for the species tree 
assuming an underlying incipient species tree. 
It can be safely used as tree prior when only one individual per species is sampled, 
but if one has multiple samples per species - which is currently often the case - the methods 
to account for this such as the multi-species coalescent (\cite{heled_and_drummond_2009}) 
may not be compatible with the underlying incipient species tree. 
More precisely, the phylogeny under the PBD model may contain paraphylies, 
while the multi-species coalescent was developed exactly to avoid 
these by explaining them as arising from incomplete lineage sorting. 
Because of these paraphylies there is no such thing as a true species tree in the PBD model. 
To get a species-level tree one must sample one incipient species per species. 
Which incipient species is sampled may therefore have an impact on the species tree.

Here we aim to explore the effect of using the
BD prior on PBD simulated phylogenies, taking into account possible sampling effects.
In brief, we simulate protracted phylogenies using the PBD process,
from which we sample a species tree in two very different ways. Given this species tree, 
we simulate a DNA sequence alignment. Then, we use BEAST2 on these alignments
to infer a posterior of phylogenies, using a BD prior. We quantify the difference
between the (BD) posterior phylogenies and the simulated (PBD) species tree.
Furthermore, while we evidently know the clock and site models used in the simulation, 
using a different clock and/or site model prior in inference 
may compensate or increase this difference between inferred and simulated tree. 
To study this, we also explore the effect of 
a different clock and site model prior in inference.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Methods (but we are not allowed to use this header)
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\richel{Start new}
The PBD model has five parameters, depicted in table \ref{table:parameters}. 
The per species speciation completion rates $\lambda$ 
we use are $0.1$, $0.3$, $1.0$ and $10^9$. 
This means that there is a $0.15\thinspace dt$ 
($0.35\thinspace dt$, $1.0\thinspace dt$, $10^9\thinspace dt$) 
probability of speciation completion occurring in an infinitesimal time $dt$. 
We use per species extinction rates of $\mu = \mu_g = \mu_i$ $0.0$, $0.1$ and $0.2$.
We use tree sizes of $n$ of 50, 100 and 200 good taxa.
For each combination of $\lambda$, $\mu$ and $n$, 
we use a speciation initiation rate $b = b_i = b_g$ 
so that the expected mean number of species $\mathbf{E}(\mean{n}; b, \lambda, \mu)$,
given a $b$, $\lambda$ and $\mu$,
equals the desired number of species $n$.
$b$ is calculated using the PBD R package (\cite{pbd})
for each parameter combination and shown in table \ref{table:pbd_parameters}.
We use $\lambda = 10^9 \approx \infty$ 
as our control for which the PBD model reduces to the BD model.

We simulate protracted birth-death trees, using the PBD 
package (\cite{pbd}) in the R programming language (\cite{r}).
The first tree has a random number generator seed of 1, 
which is incremented by 1 for each simulated tree.
For each combination of ${\lambda, \mu, b, n}$, 
we generate incipient species trees with a crown age of 15 million years.
Only trees with the desired number of good taxa are kept.

We create two data sets: a general one, to explore parameter space, 
and one to investigate the effect of sampling incipient species (see below).
For the general data set, all the trees with the correct number of good species are kept.
There is an additional selection criterion, for the data set to investigate sampling:
to generate that data set, only incipient species trees are kept, 
on which the two sampling methods (see below) result in different species trees.
As sampling will never have an effect for $\lambda = \infty$, 
this parameter value is absent in that data set.
\richel{End new}

From each incipient species tree, we construct a species tree,
by sampling one incipient/good species per good species. 
For example, when an
incipient species branched off from its mother lineage, 
both of these subspecies are recognized as representing the species, 
and hence both can be picked as an (equally good) representative of the species. 
Here, we use three sampling scenarios,
in which we pick the representative randomly or in such a way that this
results in either the shortest or longest branch lengths.

See the supplementary information for a visualization of these sampling methods.
Based on the sampled species tree, we simulate a DNA alignment that has the same history
as this species tree, using the \verb;phangorn; package (\cite{phangorn}). 
We set the nucleotides of the DNA alignment to follow a Jukes-Cantor (\cite{jc69})
nucleotide substitution model, in which all nucleotide-to-nucleotide transitions
are equally likely. 
The DNA sequence of the root ancestor consists of four equally sized single-nucleotide blocks of adenine, cytosine, guanine and thymine respectively. 
For example, for a DNA sequence length of 12, this would be AAACCCGGGTTT. 
The order of nucleotides does not matter in this study, 
because we do not consider several partitions of the sequence with their own parameters. 
Only the frequency of occurrence matters.
In our Bayesian inference (see below) we use the same site model as the (obviously correct) site model prior,
but we also explore the effect of assuming a more complex site model prior.
We predict with the more complex substitution model, 
that there will be more noise and hence our inference error will increase.
On the other hand, we dare not rule out that the inference error will decrease,
due to more flexibility in the more complex prior.
We set the mutation rate in such a way to maximize the information contained in the alignment.
To do so, we set the mutation rate such that we expect on average one (possibly silent) mutation per nucleotide
between crown age and present, which equates to $\frac{1}{15}$ mutations
per million years.
The DNA sequence length is chosen to provide a
resolution of $10^3$ years, 
% If sims take too long, we will decrease this resolution
that is, to have one expected nucleotide change 
per $10^3$ years per lineage on average. As one nucleotide is expected 
to have on average one (possibly silent) mutation per 15 million years, $15 \cdot 10^3$
nucleotides result in 1 mutation per alignment per $10^3$ years (which is
coincidentally the same as \cite{moller2018}). 
The simulation of these DNA aligments follows a strict clock model, 
which we will specify as one of the two clock models assumed in the Bayesian inference (see below).

From an alignment, we run a Bayesian analysis and 
create a posterior distribution of trees and parameters
using the \verb;babette; (\cite{babette}) package
that sets the input parameters similar to BEAUti 2 and then runs BEAST2. 
For our site model, we assume either a Jukes-Cantor or GTR nucleotide substitution model.
The Jukes-Cantor model is the correct one, as it is used for simulating that alignment,
where the GTR model is the site model that is picked as a default by most users.
For our clock model, we assume either a strict or relaxed log-normal 
clock model. 
Also here, the strict clock model is the correct one, as it is used for simulating the alignment,
but the relaxed log-normal clock model is the one most commonly used.
We set the BD model as a tree prior, 
as gauging the effect of this incorrect assumption is the goal of this study. 
We assume an MRCA prior with a tight normal distribution
around the crown age, by choosing the crown age as mean, and a standard deviation 
of $0.5 \cdot 10^{-3}$ time units,
resulting in 95\% of the crown ages inferred have the same resolution (of $10^{-3}$ time 
units) as the alignment. 
We ran the MCMC chain to generate 1111 states,
of which we remove the first 10\% (also called the 'burn-in'). 
Of the remaining
1000 MCMC states, the effective sample size (ESS) of the posterior 
must at least be 200
for a strong enough inference (\cite{beastbook}). An ESS can be increased by increasing
the number of samples or decreasing the autocorrelation between samples. 
If the ESS is less than 200, we decrease autocorrelation by doubling 
the MCMC sampling interval of that simulation, until the ESS exceeds 200.

We compare each posterior phylogeny to the (sampled) species tree
using the nLTT statistic (\cite{janzen2015}), from the \verb;nLTT; package (\cite{nltt}). 
The nLTT statistic equals the area between the normalized
lineages-through-time-plots of two phylogenies, which has a range 
from zero (for identical phylogenies) to one. We use inference error 
and nLTT statistic interchangeably. Comparing the simulated species tree
with each of the posterior species trees yields a distribution of nLTT statistics. 
The input trees generated with a $\lambda = 10^9$ allow us to measure the noise of the experiment.
For $\lambda = \infty$, the PBD model that generates the starting trees reduces to a BD model.
In the subsequent steps, sampling will have no effect in this case, because BEAST2 will assume the correct speciation model,
so the difference between inferred tree and true species tree are explained purely due to this experimental noise.

\richel{
  Note there is no twinning, as in the razzo package.
  Already sent an email to discuss this with RSE.
}

\richel{Start new}
As described above, per alignment, we do four different Bayian analyses,
as we use two different site models and two different clock models.
We know the generative site model (which is JC69) and clock 
model (a strict clock), but want to explore how often the correct model
is indeed preferred. Per alignment, we measure the estimated marginal 
likelihood (which is the probability of the data given the model)
for each of the four models and measure their relative propertions.
To estimate the marginal likelihood, we use the novel 
Nested Sampling approach (\cite{maturana_et_al_2018}),
which we configure to have an effective sample size (in estimating
the marginal likelihood) of at least 200, by using a single-particle
sub-chain with a length equal to the MCMC interval.
\richel{
  There are no rules described in the setup of the NS
  (for example, what should the relation between MCMC chain length
  and NS sub-chain length be?),
  but one can experiment with a sub-chain length and number of particles.
  I will need to do some pilot studies to verify if this is
  the correct way to do and describe this.  
}
\richel{End new}

We produce two data sets as a comma-separated file.
The general data set has 144 \richel{recalc} different combinations
of biological parameter combinations, site and clock models.
The data set to investigate sampling has ?552 \richel{recalc} different combinations
of biological parameter combinations, site models, clock models 
and sampling methods. The experiment is computationally intensive:
pilot experiments show that the experiment takes roughly 100 days
of CPU time and 20 days of wall clock time (which includes the queued 
waiting for computational resources) per replicate. 
Due to this, we choose to perform ten replicates, so that the complete
experiment will take an acceptable time of roughly seven months.

\richel{Start new}
In this article, we showcase the effect of sampling
and the certainty when selecting a (site and clock) model. 
We show the effect sampling has on the inference error,
for the nLTT distributions generated from assuming the (correct) Jukes-Cantor 
site model and a strict clock model. We do so per parameter combination,
displaying the distribution as a violin plot.
In the supplementary information, we display these error distributions 
of the general data set all combined, or separated per site or clock model.
The certainty in the model selection is again shown per parameter setting,
as a stacked bar. 
\richel{
  I am unsure how to depict that stacked bar and its uncertainty.
  Sure, the stacked bars (without uncertainty) will work,
  but information about uncertainty is lost. 
  See figure 'Showing a stacked bar with uncertainty'
  for three equivalent ways to display a stacked bar with uncertainty.
}

\begin{figure}[!htbp]
  \includegraphics[width=\textwidth]{fig_stacked_bar_with_uncertainty.png}
  \caption{
    Showing a stacked bar with uncertainty. These three different ways
    display the same underlying data.
  }
\end{figure}

\richel{End new}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Results}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{figure}[!htbp]
  \includegraphics[width=\textwidth]{fig_sampling.png}
  \caption{
    nLTT statistic distribution per biological parameter set per sampling
    regime, using the data set conditioned on sampling regime having an effect, 
    under the (correct) assumptions of a strict clock and Jukes-Cantor site model.
  }
\end{figure}

\begin{figure}[!htbp]
  \includegraphics[width=\textwidth]{fig_model_selection.png}
  \caption{
    Model preference on the general data set.
  }
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Glossary}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Please supply, as a separate list, the definitions of field-specific terms used in your article.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{table}
  \centering 
  \begin{tabular}{l p{0.65\textwidth}}
    \hline
    Term                  & Definition \\
    \hline
    \hline
    Phylogenetics         & The inference of evolutionary relationships of groups of organisms using genetics \\
    Model prior           & Knowledge or assumptions about the ontogeny of evolutionary histories \\
    Posterior             & A collection of phylogenies and parameter estimates, in which more probable combinations (determined by the data and the model prior) are presented more frequently \\
    Protracted speciation & The process in which speciation takes two events: a speciation-initiation event and a speciation-completion event  \\
    Speciation initiation & The start of a speciation event creating an incipient species \\
    Speciation completion & The end of a speciation event, in which an incipient species becomes or is recognized as a good species \\
    \hline
  \end{tabular}
  \caption{
    Glossary
  }
  \label{table:glossary}
\end{table}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% Bibliography
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% MEE style
\bibliographystyle{mee}
\bibliography{article}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\appendix
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{figure}[!htbp]
  \includegraphics[width=\textwidth]{fig_sampling_methods.png}
  \caption{
    Sampling a species tree from an incipient species tree. At the top left,
    an incipient species tree is shown, of three different good species (the
    first and second number in the taxon label) and four different subspecies (the
    third number in the taxon tabel). The other four trees are species trees,
    that use a different sampling method to determine which sub-species is picked to
    represent a good species. These are: 'Youngest', 'Oldest', 'Shortest' and
    'Longest'. With 'Youngest' the youngest sub-species is picked to represent the
    good species. With 'Oldest' the oldest sub-species is picked to represent the
    good species. 'Shortest' is the sampling method in which the sub-species are
    picked to assure the shortest branch lengths. 'Longest' is the sampling method 
    in which the sub-species are picked to assure the longest branch lengths.
  }
\end{figure}

\begin{figure}[!htbp]
  \includegraphics[width=\textwidth]{fig_general.png}
  \caption{
    nLTT statistic distribution per biological parameter set, using the
    general data set, 
    under the (correct) assumptions of a strict clock and Jukes-Cantor site model.
  }
\end{figure}

\begin{figure}[!htbp]
  \includegraphics[width=\textwidth]{fig_clock_model.png}
  \caption{
    nLTT statistic distribution per biological parameter set per clock model,
    using the general data set, 
    under the (correct) assumption of a Jukes-Cantor site model.
  }
\end{figure}

\begin{figure}[!htbp]
  \includegraphics[width=\textwidth]{fig_site_model.png}
  \caption{
    nLTT statistic distribution per biological parameter set per site model,
    using the general data set, 
    under the (correct) assumption of a strict clock model.
  }
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{table}
  \centering 
  \begin{tabular}{p{0.05\textwidth} p{0.6\textwidth} p{0.2\textwidth}}
    \hline
                          & Description & Values \\
    \hline
    \hline
    $b_g$                 & Speciation initiation rate of a good species & derived, see \ref{table:pbd_parameters} \\
    $b_i$                 & Speciation initiation rate of an incipient species & derived, see \ref{table:pbd_parameters} \\
    $\lambda$             & Speciation completion rate & 0.1, 0.3, 1.0, $10^9$ \\
    $\mu_g$               & Extinction rate of a good species & 0.0, 0.1, 0.2 \\
    $\mu_i$               & Extinction rate of an incipient species & 0.0, 0.1, 0.2 \\
    \hline
    $n$                   & Number of good taxa & 50, 100, 200 \\
    $t_c$                 & Crown age & 15 \\
    $\sigma_c$            & Standard deviation around crown age & 0.001 \\
    $M_s$                 & Sampling method & S, L, R \\
    $M_c$                 & Clock model & S, RLN \\
    $M_t$                 & Site model & JC69, GTR \\
    $r$                   & Mutation rate & $\frac{1}{15}$ \\
    $l_a$                 & DNA alignment length & $15K$ \\
    $f_i$                 & MCMC sampling interval & 1K or more \\
    $R_i$                 & RNG seed incipient tree and randomly sampled species tree & 1, 2, etc. \\
    $R_a$                 & RNG seed alignment simulation & $R_i$ \\
    $R_b$                 & RNG seed BEAST2 & $R_i$ \\
    \hline
  \end{tabular}
  \caption{
    Overview of the simulation parameters. Above the horizontal line is 
    the biological parameter set. 
    \richel{Start new}
    The RNG seed $R_i$ is 1 for the first simulation of the general data set, 2 for the next,
    and so on.  
    The RNG seeds for the data set investigating the effect 
    of sampling continue from there, but only those RNG seeds are used in which sampling has an effect.
    \richel{End new}
    The sampling methods are abbreviated as such: 'R' denotes random
    sampling, 'S' is 'shortest' and 'L' is 'longest'. Sampling method $M_s$ is random for the general
    data set. For the data set exploring the effect of sampling, we use 'shortest'
    and 'longest' for each value of $R_i$ (which are random seeds in which sampling has an effect).
    The clock models are abbreviated as 'S' for a strict and 'RLN' for a relaxed log-normal model.
    The site models are abbreviated as 'JC69' for Jukes-Cantor (\cite{jc69}) and 'GTR' for the generalized 
    time-reversible model (\cite{gtr}).
  }
  \label{table:parameters}
\end{table}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{table}
  \begin{tabular}{|l|l|l|l|l|}
    \hline
       & $\mu$ & $n$ & $\lambda$  & $b$ \\
    \hline
    1  & 0   & 50  & 0.1   & 0.30944 \\
    2  & 0.1 & 50  & 0.1   & 0.39674 \\
    3  & 0.2 & 50  & 0.1   & 0.48667 \\
    4  & 0   & 100 & 0.1   & 0.36344 \\
    5  & 0.1 & 100 & 0.1   & 0.45283 \\
    6  & 0.2 & 100 & 0.1   & 0.54425 \\
    7  & 0   & 200 & 0.1   & 0.41669 \\
    8  & 0.1 & 200 & 0.1   & 0.50759 \\
    9  & 0.2 & 200 & 0.1   & 0.6001  \\
    10 & 0   & 50  & 0.3   & 0.25717 \\
    11 & 0.1 & 50  & 0.3   & 0.34003 \\
    12 & 0.2 & 50  & 0.3   & 0.42648 \\
    13 & 0   & 100 & 0.3   & 0.30862 \\
    14 & 0.1 & 100 & 0.3   & 0.39455 \\
    15 & 0.2 & 100 & 0.3   & 0.48328 \\
    16 & 0   & 200 & 0.3   & 0.35991 \\
    17 & 0.1 & 200 & 0.3   & 0.44804 \\
    18 & 0.2 & 200 & 0.3   & 0.53841 \\
    19 & 0   & 50  & 1     & 0.2297  \\
    20 & 0.1 & 50  & 1     & 0.30759 \\
    21 & 0.2 & 50  & 1     & 0.38984 \\
    22 & 0   & 100 & 1     & 0.2778  \\
    23 & 0.1 & 100 & 1     & 0.35961 \\
    24 & 0.2 & 100 & 1     & 0.44481 \\
    25 & 0   & 200 & 1     & 0.32617 \\
    26 & 0.1 & 200 & 1     & 0.41078 \\
    27 & 0.2 & 200 & 1     & 0.49818 \\
    28 & 0   & 50  & $10^9$ & 0.21589 \\
    29 & 0.1 & 50  & $10^9$ & 0.28896 \\
    30 & 0.2 & 50  & $10^9$ & 0.36635 \\
    31 & 0   & 100 & $10^9$ & 0.26146 \\
    32 & 0.1 & 100 & $10^9$ & 0.33872 \\
    33 & 0.2 & 100 & $10^9$ & 0.41945 \\
    34 & 0   & 200 & $10^9$ & 0.30733 \\
    35 & 0.1 & 200 & $10^9$ & 0.38768 \\
    36 & 0.2 & 200 & $10^9$ & 0.47099 \\
    \hline
  \end{tabular}
  \caption{
    The speciation parameters used. Starting from extinction rate $\mu$ ($\mu = \mu_g = \mu_i$), the expected
    mean number of good species $n$, speciation completion rate $\lambda$, the speciation initation
    rate $b$ ($b = b_g = b_i$) follows.  
  }
  \label{table:pbd_parameters}
\end{table}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{table}
  \centering 
  \begin{tabular}{l l}
    \hline
    $n$ & Description \\
    \hline
    \hline
    $12$   & simulation parameters, see table \ref{table:parameters} \\
    $1000$ & nLTT statistic values \\
    $11$ & ESSes of all parameters estimated by BEAST2 (see specs below) \\
    $1$ & Marginal likelihood estimate \\
    $1$ & Marginal likelihood estimation uncertainty \\
    $1$ & Marginal likelihood ESS \\
    \hline
  \end{tabular}
  \caption{
    Specification of the data sets. Each row will contain one experiment,
    where the columns contain parameters, measurements and diagnostics.
    This table displays the content of the columns. 
    $n$ denotes the number
    of columns a certain item will occupy, resulting in a table of 
    1023 \richel{recalc} columns and 20K rows.
  }
  \label{table:specs}
\end{table}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{table}
  \centering 
  \begin{tabular}{l l}
    \hline
    \# & Description \\
    \hline
    \hline
    1 & posterior \\
    2 & likelihood \\
    3 & prior \\
    4 & treeLikelihood \\
    5 & TreeHeight \\
    6 & BirthDeath \\
    7 & BDBirthRate \\
    8 & BDDeathRate \\
    9 & logP.mrca \\
    10 & mrcatime \\
    11 & clockRate \\
    \hline
  \end{tabular}
  \caption{
    Overview of the 11 parameters estimated by BEAST2
  }
  \label{table:estimated_parameters}
\end{table}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Acknowledgements}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\richel{put this section here, as the journal does not request for this}
We would like to thank the Center for Information Technology of the University of Groningen for their support
and for providing access to the Peregrine high performance computing cluster.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Authors' contributions}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\richel{put this section here, as the journal does not request for this}
RSE conceived the idea for this experiment. 
RJCB created and tested the experiment, and wrote the first draft of the manuscript. 
RSE contributed substantially to revisions.

### 2019-05-07 content.tex
No appendix, use 'Supplement'
Improved title

@@
 
-\richel{Start new}
 The PBD model has five parameters, depicted in table \ref{table:parameters}. 
@@
 this parameter value is absent in that data set.
-\richel{End new}
 
@@
 
-\richel{Start new}
 As described above, per alignment, we do four different Bayian analyses,
@@
 }
-\richel{End new}
 
@@
 
-\richel{Start new}
 In this article, we showcase the effect of sampling
@@
 
-\richel{End new}
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \end{figure}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Acknowledgements}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\richel{put this section here, as the journal does not request for this}
+We would like to thank the Center for Information Technology of the University of Groningen for their support
+and for providing access to the Peregrine high performance computing cluster.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Authors' contributions}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\richel{put this section here, as the journal does not request for this}
+RSE conceived the idea for this experiment. 
+RJCB created and tested the experiment, and wrote the first draft of the manuscript. 
+RSE contributed substantially to revisions.
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\appendix
+\section{Supplement}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
     the biological parameter set. 
-    \richel{Start new}
     The RNG seed $R_i$ is 1 for the first simulation of the general data set, 2 for the next,
@@
     of sampling continue from there, but only those RNG seeds are used in which sampling has an effect.
-    \richel{End new}
     The sampling methods are abbreviated as such: 'R' denotes random
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Acknowledgements}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\richel{put this section here, as the journal does not request for this}
-We would like to thank the Center for Information Technology of the University of Groningen for their support
-and for providing access to the Peregrine high performance computing cluster.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Authors' contributions}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\richel{put this section here, as the journal does not request for this}
-RSE conceived the idea for this experiment. 
-RJCB created and tested the experiment, and wrote the first draft of the manuscript. 
-RSE contributed substantially to revisions.

### 2019-05-08 content.tex
Travis uses Xenial

@@
   Among the most popular tools are Bayesian approaches, 
-  such as \\ BEAST, MrBayes and RevBayes, 
-  that use efficient tree sampling routines to create a posterior probability distribution 
+  such as BEAST, MrBayes and RevBayes, 
+  that use efficient tree sampling routines to create a 
+  posterior probability distribution 
   of the phylogenetic tree. 
-  A feature of these approaches is the possibility to incorporate known or hypothesized structure of the phylogenetic tree through the tree prior. 
+  A feature of these approaches is the possibility to incorporate known 
+  or hypothesized structure of the phylogenetic tree through the tree prior. 
   It has been shown that the effect of the prior on the posterior distribution 
@@
 or tree prior (\cite{moller2018, yang_and_ranalla_2005}) is known to affect
-the posterior.
+the posterior. There is evidence, however, that the 
+tree prior and molecular clock do not do so substantially affect 
+the estimation of diversification rates (\cite{sarver2019choice})).
 
@@
 
-\richel{
-  Note there is no twinning, as in the razzo package.
-  Already sent an email to discuss this with RSE.
-}
+TWINNING
 
@@
 Nested Sampling approach (\cite{maturana_et_al_2018}),
-which we configure to have an effective sample size (in estimating
-the marginal likelihood) of at least 200, by using a single-particle
-sub-chain with a length equal to the MCMC interval.
-\richel{
-  There are no rules described in the setup of the NS
-  (for example, what should the relation between MCMC chain length
-  and NS sub-chain length be?),
-  but one can experiment with a sub-chain length and number of particles.
-  I will need to do some pilot studies to verify if this is
-  the correct way to do and describe this.  
-}
+which we configure to have a relative error $\epsilon$ of $1^{-12}$.
 
@@
 as a stacked bar.
-\richel{
-  I am unsure how to depict that stacked bar and its uncertainty.
-  Sure, the stacked bars (without uncertainty) will work,
-  but information about uncertainty is lost. 
-  See figure 'Showing a stacked bar with uncertainty'
-  for three equivalent ways to display a stacked bar with uncertainty.
-}
+
+HIERO: DESCRIBE PLOTTING.
 
@@
 
-\richel{put this section here, as the journal does not request for this}
 We would like to thank the Center for Information Technology of the University of Groningen for their support
@@
 
-\richel{put this section here, as the journal does not request for this}
 RSE conceived the idea for this experiment. 

### 2019-12-17 content.tex
Use new naming

@@
 It can be safely used as tree prior when only one individual per species is sampled, 
-but if one has multiple samples per species - which is currently often the case - the methods 
+but if one has multiple samples per species - which 
+is currently often the case - the methods 
 to account for this such as the multi-species coalescent (\cite{heled_and_drummond_2009}) 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Methods (but we are not allowed to use this header)
+\section{Hypotheses} \richel{but we are not allowed to use this header}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section{Hypotheses}
+
+$\mathcal{H}_1$: RJCB expects that the inference error is lowest when the true
+tree is generated under PBD parameter settings without protractedness, 
+i.e $SCR = \infty$. For such parameters settings, the
+true tree is in practice generated by a BD model, which
+matches the tree prior used in inference. Without a mismatch between
+true tree prior and assumed tree prior, this source of errors will be absent,
+where the other two sources of errors (stochasticity in the
+simulation of the alignment and the MCMC algorithm) 
+will remain the same. \richel{figure 1a and b}
+
+$\mathcal{H}_2$: RJCB expects that, on average, the inference error is highest for
+parameter settings with a lower speciation rate, 
+as for these settings, 
+the mismatch between the the speciation model that
+generated the true tree (which is profoundly-P PBD) is biggest with
+the tree prior assumed to be generative (which is BD). \richel{figure 1a and b}
+
+$\mathcal{H}_3$: RJCB expects that the inference error is higher for true trees
+with an observable/actual higher percentage of extant incipient species,
+regardless of the PBD parameters.
+These incipient species are one of the three (and the most 
+interesting) sources of error, as a BD model -as a feature- will never
+infer the branch-length distribution of protracted species. \richel{figure 2}
+
+$\mathcal{H}_4$: RJCB expects that, on average, the inference error is equal for
+parameter settings with an equal SIRI, SIRG and SCR, i.e. extinction
+has no effect on the error made, because extinction affects all species equally. 
+\richel{figure 1a and b, or a new figure}
+
+$\mathcal{H}_5$: RJCB expects that an increased number of taxa
+has a negative effect on the variance of the errors made, as there
+is more information available to base inference on.
+\richel{figure 3}
+
+$\mathcal{H}_6$: RJCB expects that an increased extinction rate
+increases the variance of the errors made,
+due to the decrease of the number of taxa, reducing the amount of information
+to base inference on.
+\richel{figure 4}
+
+$\mathcal{H}_7$: RJCB expects that the nLTT statistic between a true and twin
+tree at the start of the pipeline,
+will correlate strongly with the difference
+between the highest posterior density (HPD) of the 
+errors of the true and twin tree 
+generated at the end of the pipeline. 
+This hypothesis
+stems from the idea of assuming a setup without noise; that is, with
+a DNA alignment of infinite length. From such a DNA alignment 
+with infinite information, the MCMC
+is able to infer the close-to-correct phylogeny.
+\richel{if this is true, we can use this shortcut (instead of running
+multiple pirouette replicates) to draw stronger conclusions}
+\richel{figure 5}
+
+\begin{figure}[!htbp]
+  \includegraphics[width=\textwidth]{20191126_nltt_as_proxy.png}
+  \caption{
+    The nLTT statistic between true and twin tree correlates strongly
+    with the difference between (the highest posterior density
+    of) the errors of the true and twin tree.
+  }
+  \label{fig:nltt_as_proxy}
+\end{figure}
+
+$\mathcal{H}_8$: RJCB expects that for settings without extinctions,
+the candidate model with JC69, strict and Yule will come up as the best
+model most, as that model is the generative model, where the HKY, strict, BD
+will be seleted least, as both its site and tree model are needlessly complex.
+For settings with extinctions, RJCB predict HKY, strict and Yule will come up 
+as the worst model mostly, as it uses both an incorrect site model and incorrect
+tree model.
+
+$\mathcal{H}_9$: RJCB expects that different ways of sampling have 
+no effect.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Methods} \richel{but we are not allowed to use this header}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
